### 2021-07-29 Code/Manuscript.Rmd
updating analyses

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah J. Hart"
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = FALSE
)

ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)>0) {
             install.packages(new.pkg, dependencies = TRUE)
    }
    easypackages::libraries(pkg)
}

options(repos = c(CRAN = "http://cran.rstudio.com"))

### Import libraries
ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'foreach'))

#foreach', 'parallel', 'doParallel', 'quickPlot', 'reshape2','vegan', 'FSA', 
# Set seed
set.seed(2020)

# Set custom plotting theme
theme_new <- function(base_size = 9,base_family = "Helvetica"){
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.line.x = element_line(color="black", size = 0.25),
      axis.line.y = element_line(color="black", size = 0.25),
      axis.title = element_text(size = 9),
      axis.text = element_text(colour="black", size=8),
      legend.key=element_rect(colour=NA, fill =NA),
      panel.grid = element_blank(),   
      plot.background = element_rect(fill = NA, colour = NA),
      panel.border = element_rect(fill = NA, colour = NA),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(size = 9)

    )
}
theme_black = function(base_size = 12, base_family = "Helvetica") {

  theme_grey(base_size = base_size, base_family = base_family) %+replace%

    theme(
      # Specify axis options
      axis.line = element_blank(),  
      axis.text.x = element_text(size = base_size*0.8, color = "white", lineheight = 0.9),  
      axis.text.y = element_text(size = base_size*0.8, color = "white", lineheight = 0.9),  
      axis.ticks = element_line(color = "white", size  =  0.2),  
      axis.title.x = element_text(size = base_size, color = "white", margin = margin(10, 0, 0, 0)),  
      axis.title.y = element_text(size = base_size, color = "white", angle = 90, margin = margin(0, 10, 0, 0)),  
      axis.ticks.length = unit(0.3, "lines"),   
      # Specify legend options
      legend.background = element_rect(color = NA, fill = rgb(50, 50, 50, maxColorValue = 255)),  
      legend.key = element_rect(color = "white",  fill = rgb(50, 50, 50, maxColorValue = 255)),  
      legend.key.size = unit(1.2, "lines"),  
      legend.key.height = NULL,  
      legend.key.width = NULL,      
      legend.text = element_text(size = base_size*0.8, color = "white"),  
      legend.title = element_text(size = base_size*0.8, face = "bold", hjust = 0, color = "white"),  
      legend.position = "right",  
      legend.text.align = NULL,  
      legend.title.align = NULL,  
      legend.box = NULL, 
      # Specify panel options
      panel.background = element_rect(fill = rgb(50, 50, 50, maxColorValue = 255), color  =  NA),  
      panel.border = element_rect(fill = NA, color = "white"),  
      panel.grid.major = element_line(color = NA),  
      panel.grid.minor = element_line(color = NA),  
      panel.margin = unit(0.5, "lines"),   
      # Specify facetting options
      strip.background = element_rect(fill = "grey30", color = "grey10"),  
      strip.text.x = element_text(size = base_size*0.8, color = "white"),  
      strip.text.y = element_text(size = base_size*0.8, color = "white",angle = -90),  
      # Specify plot options
      plot.background = element_rect(color =rgb(50, 50, 50, maxColorValue = 255), fill = rgb(50, 50, 50, maxColorValue = 255)),  
      plot.title = element_text(size = base_size*1.2, color = "white"),  
      plot.margin = unit(rep(1, 4), "lines")

    )

}
theme_set(theme_new())

options(ggplot2.continuous.colour="viridis")
scale_colour_discrete <- scale_fill_npg
scale_fill_discrete <- scale_fill_npg
options(scipen=999)

p1col=c(rgb(104, 17, 10,  maxColorValue = 255), rgb(79, 130, 156, maxColorValue = 255), rgb(243, 173, 1, maxColorValue = 255), rgb(112, 146, 115,  maxColorValue = 255), rgb(112, 146, 115, maxColorValue = 255))

p1fill=c(rgb(104, 17, 10, alpha=230, maxColorValue = 255), rgb(79, 130, 156, alpha=230, maxColorValue = 255), rgb(243, 173, 1, alpha=50, maxColorValue = 255), rgb(112, 146, 115, alpha=230,  maxColorValue = 255), rgb(112, 146, 115, alpha=230, maxColorValue = 255))

p2col =c(rgb(20,20,20, maxColorValue = 255), rgb(54, 124, 111, maxColorValue = 255), rgb(235, 137, 52, maxColorValue = 255), rgb(83, 74, 90, maxColorValue = 255), rgb(232, 177, 60, maxColorValue = 255),rgb(90, 129, 154, maxColorValue = 255),rgb(113, 0, 2, maxColorValue = 255), rgb(166, 166, 166,  maxColorValue = 255))

fig.width1 <- 3.4
fig.width2 <- 7
fig.width1.5 <- 4.48

pt.width1 <-fig.width1 * 72
pt.width2 <-fig.width2 * 72
pt.width1.5 <- fig.width1.5 * 72

# Set projection - NAD83 / Conus Albers
aea.proj <- "EPSG:5070"

options(timeout=60*30) #timeout downloads that last longer than 30 minutes

# Set directory structure for project
dir.create(here("Data/"), showWarnings = FALSE)
dir.create(here("Code/"), showWarnings = FALSE)
dir.create(here("Results/"), showWarnings = FALSE)
```

# Abstract

# Keywords

# Introduction

Paragraph 1: Broad applied context

In this context there is a particular need to understand interactions between outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca. 18 million hectares more than wildfire (van Lierop et al. 2015).

- insect outbreaks are causing widespread tree mortality in temperate forest worldwide
- future changes in climate are expected to increase tree mortality
- forests provide key ecosystem services
- management

Paragraph 2: Theoretical context - diversity & natural enemies
- Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources.
- Key hypotheses:
1 - associational resistance - the abundance or damage of herbivores is lower when plant diversity is greater [@barbosa_associational_2009, @jactel_tree_2021]
hypotheses:
1 - resource concentration: the presence of heterospecific neighbors around a tree of a focal species grown in mixed stands leads to a lower probability of host tree finding by insect herbivores due to lower host abundance or frequency [@root_organization_1973]
2 - host apparency: 
3 - decoy: preference for non-host trees

However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]

suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.

Most studies have focused on individual- or population-level outcomes.

For instance, increased host community diversity may either increase or decrease the susceptibility of individuals to their natural enemies.

At the population-scale these effects may either scale-up or result in counter-intuitive effects. However, little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists. Further, the effects of different natural enemies on their focal host populations often differ greatly, due to differences in susceptibility and mortality rates and the availability and quality (as viewed by their natural enemies) of resource communities. Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and natural enemy-resource relationships are highly dynamic.

*** paragraph about irruptive species and community host diversity*** 
- stand scale
- landscape scale

To better understand how community diversity influences interactions between natural enemies and their resources, we use a natural system with inherently low resource and natural enemy diversity. Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D. rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively. We use this system to ask whether host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale? We hypothesize that
greater host richness will increase the probability of outbreak occurrence

Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year. Bark beetles bore through the bark, where they mate and oviposit their eggs. Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death. Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles. To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree. Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.For instance, in the continental western United States bark beetles have killed more than XXXX trees over the past X decades. Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses. 
For instance, a primary goal of the Western Bark Beetle is to promote resilience of forests to bark beetle outbreaks by increasing the diversity of age classes and tree species.

If greater host richness increases niche space for bark beetles, then community-level occurrence of bark beetle activity will be greater in more diverse communities. community-level severity, here the amount and proportion of basal area affected by bark beetles, will be lower in stands with more host species.

To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/). Specifically we ask: (1) Is the occurrence or severity of bark beetle

# Methods

## Study area
```{r set species and agent codes}
# SPECIES CODES

# MPBhosts
PICO <- 108
PIAL <- 101
PIAR <- 102
PIFL <- 113
PIST <- 114
PIMO <- 119
PIPO <- 122

# SBhosts
PIEN <- 93
PIPU <- 96

#SFDhosts
ABLA <-19 
ABCO <- 15
ABGR <- 17

ohostsMPB <- c( PIAL, PIAR, PIFL, PIST, PIMO, PIPO)
ohostsWBBB <- c(ABCO, ABGR)
ohostsSB <- PIPU
otherhosts <- c(ABCO, ABGR, PIPU, PIAL, PIAR, PIFL, PIST, PIMO, PIPO)

# AGENT CODES
# DCA
bb <- 11000 #bark beetle
mpb <- 11006 #mountain pine beetle
sb <- 11009 #spruce beetle
wbbb <- 11015 # western balsam bark beetle
sfm <- 80002 # subalpine fir mortality
```

The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming).

## Data
### FIA data

The FIA program is a single inventory program that includes all public and private forested land (>= 0.4 ha in size and >= 10% canopy cover) in the US. In the Western US, all plots are visited once every ten years [@gray_forest_2012]. The spatially and temporally distributed probabilistic sampling design is useful studies of the distribution of tree species [e.g., @iverson_predicting_1998; @rehfeldt_empirical_2006], forest insects [e.g., @derose_effects_2013], and tree mortality [e.g., @shaw_forest_2005]

At each FIA plot, field crews collect data for trees (>= 12.7 cm DBH) within four 7.32 m radius subplots arranged in a fixed pattern. Data on the proximate cause of death is collected for any tree (>= 12.7 cm DBH) that was alive at the previous visit and at revisit is dead using visible evidence (e.g., fire scars, bark beetle galleries).

```{r download FIA data, eval=F}
dir.create(here("Data/"), showWarnings = FALSE)
dir.create(here("Data/FIA"), showWarnings = FALSE)
# reference data
dir.create(here("Data/FIA/Reference"), showWarnings = FALSE)
temp <- here("/Data/FIA/Reference/FIADB_REFERENCE.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/FIADB_REFERENCE.zip", temp)
unzip(temp, exdir=here("/Data/FIA/Reference"))

# AZ
dir.create(here("Data/FIA/AZ"), showWarnings = FALSE)
temp <- here("/Data/FIA/AZ/AZ.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/AZ.zip", temp)
unzip(temp, exdir=here("/Data/FIA/AZ"))

# CO
dir.create(here("Data/FIA/CO"), showWarnings = FALSE)
temp <- here("/Data/FIA/CO/CO.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/CO.zip", temp)
unzip(temp, exdir=here("/Data/FIA/CO"))

# ID
dir.create(here("Data/FIA/ID"), showWarnings = FALSE)
temp <- here("/Data/FIA/ID/ID.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/ID.zip", temp)
unzip(temp, exdir=here("/Data/FIA/ID"))

# MT
dir.create(here("Data/FIA/MT"), showWarnings = FALSE)
temp <- here("/Data/FIA/MT/MT.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/MT.zip", temp)
unzip(temp, exdir=here("/Data/FIA/MT"))

# NM
dir.create(here("Data/FIA/NM"), showWarnings = FALSE)
temp <- here("/Data/FIA/NM/NM.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/NM.zip", temp)
unzip(temp, exdir=here("/Data/FIA/NM"))

# NV
dir.create(here("Data/FIA/NV"), showWarnings = FALSE)
temp <- here("/Data/FIA/NV/NV.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/NV.zip", temp)
unzip(temp, exdir=here("/Data/FIA/NV"))

# UT
dir.create(here("Data/FIA/UT"), showWarnings = FALSE)
temp <- here("/Data/FIA/UT/UT.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/UT.zip", temp)
unzip(temp, exdir=here("/Data/FIA/UT"))

# WY
dir.create(here("Data/FIA/WY"), showWarnings = FALSE)
temp <- here("/Data/FIA/WY/WY.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/WY.zip", temp)
unzip(temp, exdir=here("/Data/FIA/WY"))
```

```{r download spatial data, eval=F}
# states
dir.create(here("Data/Spatial/"), showWarnings = FALSE)
dir.create(here("Data/Spatial/States"), showWarnings = FALSE)
temp <- here("/Data/Spatial/States/cb_2018_us_state_20m.zip")
download.file("https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_state_20m.zip", temp)
unzip(temp, exdir=here("/Data/Spatial/States"))

# usfs western bark beetle strategy
dir.create(here("Data/Spatial/USFS/"), showWarnings = FALSE)
temp <- here("/Data/Spatial/USFS/WBBS_PL.gdb.zip")
download.file("https://data.fs.usda.gov/geodata/edw/edw_resources/fc/S_USA.Activity_WBBS_PL.gdb.zip", temp)
unzip(temp, exdir=here("/Data/Spatial/USFS/"))

# usfs timber harvest
dir.create(here("Data/Spatial/USFS/"), showWarnings = FALSE)
temp <- here("/Data/Spatial/USFS/Activity_TimberHarvest.gdb.zip")
download.file("https://data.fs.usda.gov/geodata/edw/edw_resources/fc/S_USA.Activity_TimberHarvest.gdb.zip", temp)
unzip(temp, exdir=here("/Data/Spatial/USFS/"))
```

```{r import raw data, eval=F}
wd <- here("Data/FIA")

# List directories containing data for each state
dirs <- list.dirs(wd)

dirs <- dirs[c(-1, -8)]

# Create empty lists to hold dataframe for each state
dats <- list() 
datp <- list()
datc <- list()
datt <- list()
datspc <- list()

for(i in 1:length(dirs)){
  state <- strsplit(dirs[i], split="FIA/")[[1]][2]
  # Survey 
  dats[[state]] <- read.csv(paste0(dirs[i], "/",state, "_SURVEY.csv"), stringsAsFactors = F)
  # Plot
  datp[[state]] <- read.csv(paste0(dirs[i], "/",state, "_PLOT.csv"), stringsAsFactors = F)
  # Condition
  datc[[state]] <- read.csv(paste0(dirs[i], "/",state, "_COND.csv"), stringsAsFactors = F)
  # Tree
  datt[[state]] <- read.csv(paste0(dirs[i], "/",state, "_TREE.csv"), stringsAsFactors = F)
  # Subplot Condition
  datspc[[state]] <- read.csv(paste0(dirs[i], "/",state, "_SUBP_COND.csv"), stringsAsFactors = F)
}

### Merge data
datsm <- do.call(rbind, dats)
datpm <- do.call(rbind, datp)
datcm <- do.call(rbind, datc)
dattm<- do.call(rbind, datt)
datspcm<- do.call(rbind, datspc)

remove(datt, datp, datc, dats,datspc)

# Merge Subplot & Plot data
df <- merge(datcm[,c("PLT_CN", "CONDID","COND_STATUS_CD","OWNCD", "STDAGE")], datspcm[,c("PLT_CN", "CONDID", "SUBP")], by=c("PLT_CN", "CONDID"))

# Merge Survey & Plot data
datz <- merge(datsm[,c("CN", "ANN_INVENTORY", "STATEAB")], datpm[,c("CN","SRV_CN","INVYR", "COUNTYCD","PLOT", "PLOT_STATUS_CD", "KINDCD", "DESIGNCD", "MANUAL", "SAMP_METHOD_CD", "REMPER", "MEASYEAR", "ELEV", "LAT", "LON", "PREV_PLT_CN", "UNITCD")],  by.y="SRV_CN", by.x="CN", all==TRUE)
colnames(datz)[which(colnames(datz) %in% c("CN", "CN.y"))] <- c("SRV_CN", "PLT_CN")

#confirm each plot record is in data frame only once 
if(max(sort(table(datz$PLT_CN)))!=1){
  print("stop there are duplicated plot records")
}
datz.sp<-datz
```

```{r merge survey, plot, condition & tree data, eval=F}
## select annual inventory
datz <-datz[datz$ANN_INVENTORY=='Y',]; nrow.annual <- nrow(datz)

## select plots where all of the subplots are forested
allspforested <- unique(df$PLT_CN[!(df$PLT_CN %in% subset(aggregate(COND_STATUS_CD~PLT_CN, FUN=function(x){length(x[x!=1])}, df), COND_STATUS_CD>0)$PLT_CN)])
allspforested <- allspforested[allspforested %in% datz$PLT_CN]
datz <- datz[datz$PLT_CN %in% allspforested, ]

#Merge Survey-Plot-Condition & Tree data
dattm.sub <-subset(dattm, PLT_CN %in% datz$PLT_CN)
datz <- merge(datz, dattm.sub[, c("PLT_CN", "CN", "STATECD", "COUNTYCD","TREE", "DIST", "STATUSCD", "DIA", "SPCD","CCLCD", "AGENTCD", "DAMAGE_AGENT_CD1","DAMAGE_AGENT_CD2", "DAMAGE_AGENT_CD3", "TPA_UNADJ")],  by="PLT_CN", all=FALSE)
datz$PLOT.STATEAB <- paste(datz$PLOT, datz$STATEAB, sep=".")

# subset data to only trees greater than 5 inches (12.7 cm) DBH
datz5 <- subset(datz, DIA>=5)

# subset data to only trees within subplot
datz5 <- subset(datz5, DIST<=24)

# remove trees with missing values for TPA count 
datz5 <- datz5[is.na(datz5$TPA_UNADJ)==FALSE, ] ## not sure why missing values occur here
datz5.bak <- datz5

spruce.plots <-unique(datz5[datz5$SPCD == PIEN,"PLT_CN"])
pine.plots <-unique(datz5[datz5$SPCD == PICO,"PLT_CN"])
fir.plots <- unique(datz5[datz5$SPCD == ABLA,"PLT_CN"])
datz5 <- datz5[datz5$PLT_CN %in% c(spruce.plots, pine.plots, fir.plots),]

## subset to recently killed trees
datz5.bak <- datz5
datz5$STATUSCD <- factor(datz5$STATUSCD , levels=c(0:3), labels=c("No Status", "Live", "Dead", "Removed"))
datz5$STATUSCD.2<- ifelse(datz5$STATUSCD=="Dead" & is.na(datz5$AGENTCD) == FALSE, "Recently killed",as.character(datz5$STATUSCD))
datz5 <- datz5[datz5$STATUSCD.2 %in% c("Live", "Recently killed"),]
```

#### Determination of presence/absence of bark beetle activity

For each live tree (>= 12.7 cm DBH), field crews record up to three damaging agents, which are defined as agents that are likely to prevent the tree from surviving >2 years, reduce the growth of the tree in the near term, or negatively affect the tree's marketable products [@burrill_forest_2017].

Cause of death codes are very broad (e.g., "insect" or "disease"). Accuracy of FIA data is commonly assessed using blind checks, where two crews perform independent inventories. Agreement between mortality agent codes recorded in the two inventories is generally >80% [@anderegg_tree_2015]. Active damage is easier to identify, thus the codes for damaging agents are much more specific.
```{r determine presence of BB agents in focal hosts, eval=F}
datz5$WBBB.l <- NA
datz5$WBBB.d <- NA
datz5$WBBB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD==ABLA |datz5$DAMAGE_AGENT_CD1 %in% c(wbbb, sfm) & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB <- rowSums(datz5[,c("WBBB.d", "WBBB.l")], na.rm=T)
datz5$WBBB[is.na(datz5$WBBB)==T]<-0

datz5$MPB.l <- NA
datz5$MPB.d <- NA
datz5$MPB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD ==PICO | datz5$DAMAGE_AGENT_CD1 ==11006 & datz5$SPCD ==PICO, 1, 0) 
datz5$MPB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD ==PICO , 1, 0) 
datz5$MPB <- rowSums(datz5[,c("MPB.d", "MPB.l")], na.rm=T)
datz5$MPB[is.na(datz5$MPB)==T]<-0

datz5$SB.l <- NA
datz5$SB.d <- NA
datz5$SB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD == PIEN | datz5$DAMAGE_AGENT_CD1==11009 & datz5$SPCD == PIEN, 1, 0) 
datz5$SB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD == PIEN, 1, 0)  
datz5$SB <- rowSums(datz5[,c("SB.d", "SB.l")], na.rm=T)
datz5$SB[is.na(datz5$SB)==T]<-0
```

#### Calculation of stand characteristics
For each plot, we then calculated
1 - basal area by host species
2 - quadratic mean diameter (QMD)
3 - basal area dominance (% total basal area) by species
4 - presence and severity of bark beetle activity (% total basal area) by bark beetle species

```{r stand characteristics, eval=F}
## BA
datz5$BA <- datz5$DIA*datz5$DIA * pi/(4*144) *datz5$TPA_UNADJ # sq ft per acre
datz5$BA <- datz5$BA * 1/10.7639 * 2.47105/1 # m2 per ha
BA <- aggregate(BA~PLT_CN+SPCD, datz5, FUN=sum, na.rm=T)
BA.dat <- dcast(BA, PLT_CN~SPCD, value.var="BA", drop=F)
BA.dat$BA.total <- rowSums(BA.dat[,-1], na.rm=T)
BA.dat$BA.Allhosts <- rowSums(BA.dat[,as.character(c(PICO, PIEN, ABLA))], na.rm=T)
BA.dat$BA.Otherhosts <- rowSums(BA.dat[,as.character(otherhosts)], na.rm=T)
BA.dat <- BA.dat[, c("PLT_CN", PICO, PIEN, ABLA, "BA.total", "BA.Allhosts", "BA.Otherhosts")]
colnames(BA.dat) <- c("PLT_CN", paste0("BA.", c("PICO", "PIEN", "ABLA", "total", "Allhosts", "Otherhosts")))
BA.dat[is.na(BA.dat)==TRUE] <- 0

## Trees per ha
datz5$TPH <- datz5$TPA_UNADJ * 2.47105 # trees per ha
TPH <- aggregate(TPH~PLT_CN+SPCD, datz5, FUN=sum, na.rm=T)
TPH.dat <- dcast(TPH, PLT_CN~SPCD, value.var="TPH", drop=F)
TPH.dat$TPH.total <- rowSums(TPH.dat[,-1], na.rm=T)
TPH.dat$TPH.Allhosts <- rowSums(TPH.dat[,as.character(c(PICO, PIEN, ABLA))], na.rm=T)
TPH.dat <- TPH.dat[, c("PLT_CN", PICO, PIEN, ABLA, "TPH.total", "TPH.Allhosts")]
colnames(TPH.dat) <- c("PLT_CN", paste0("TPH.", c("PICO", "PIEN", "ABLA", "total", "Allhosts")))
TPH.dat[is.na(TPH.dat)==TRUE] <- 0

## QMD
datz5$DIAcm <- datz5$DIA * 2.54 # convert to cm
qmd <- aggregate(DIAcm~PLT_CN+SPCD, datz5, FUN=function(x){sqrt(mean(x*x,na.rm=T))})
qmd.dat<- dcast(qmd, PLT_CN~SPCD , value.var="DIAcm", drop=F)
qmd.dat<- qmd.dat[,c("PLT_CN", PICO, PIEN, ABLA)]
colnames(qmd.dat) <- c("PLT_CN", paste0("QMD.",c("PICO", "PIEN", "ABLA")))
qmd.dat.hosts <- aggregate(DIAcm~PLT_CN, datz5[datz5$SPCD %in% c(PICO, PIEN, ABLA),], FUN=function(x){sqrt(mean(x*x,na.rm=T))})
colnames(qmd.dat.hosts) <- c("PLT_CN", "QMD.Allhosts")
qmd.dat.all <- aggregate(DIAcm~PLT_CN, datz5, FUN=function(x){sqrt(mean(x*x,na.rm=T))})
colnames(qmd.dat.all) <- c("PLT_CN", "QMD.total")
qmd.dat<-merge(qmd.dat, qmd.dat.hosts, by="PLT_CN", all=T)
qmd.dat<-merge(qmd.dat, qmd.dat.all, by="PLT_CN", all=T)
qmd.dat[is.na(qmd.dat)] <- 0

## SDI (additive)
datz5$SDIi <- (datz5$DIA/10)^1.6
sdi <- aggregate(SDIi~PLT_CN+SPCD, datz5, FUN=sum)
sdi.dat<- dcast(sdi, PLT_CN~SPCD , value.var="SDIi", drop=F)
sdi.dat<- sdi.dat[,c("PLT_CN", PICO, PIEN, ABLA)]
colnames(sdi.dat) <- c("PLT_CN", paste0("SDI.",c("PICO", "PIEN", "ABLA")))
sdi.dat[is.na(sdi.dat)] <- 0

## Trees per plot of bark beetle affected trees
notreesBB.dat <- merge(merge(aggregate(MPB~PLT_CN,datz5, FUN=sum, na.rm=T),  aggregate(SB~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T), aggregate(WBBB~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T)
colnames(notreesBB.dat) <- c("PLT_CN", paste0("stand.", c("MPB", "SB", "WBBB")))
notreesBB.dat$stand.allBBs <- rowSums(notreesBB.dat[,2:4])
notreesBB.dat[is.na(notreesBB.dat)] <- 0

## Total BA of bark beetle affected trees
BABB.dat <- merge(merge(aggregate(MPB*BA~PLT_CN,datz5, FUN=sum, na.rm=T),  aggregate(SB*BA~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T), aggregate(WBBB*BA~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T)
colnames(BABB.dat) <- c("PLT_CN", paste0("BA.", c("MPB", "SB", "WBBB")))
BABB.dat$BA.allBBs <- rowSums(BABB.dat[,2:4])
BABB.dat[is.na(BABB.dat)] <- 0

df<- unique(datz5[,c("PLT_CN", "PREV_PLT_CN", "PLOT.STATEAB", "INVYR", "ELEV", "LAT", "LON" ),])
df <-merge(df, TPH.dat, by="PLT_CN", all=T)
df <-merge(df, qmd.dat, by="PLT_CN", all=T)
df <-merge(df, BA.dat, by="PLT_CN", all=T)
df <-merge(df, sdi.dat, by="PLT_CN", all=T)
df <- merge(df, notreesBB.dat, by="PLT_CN", all=T)
df <- merge(df, BABB.dat, by="PLT_CN", all=T)
df[is.na(df$PREV_PLT_CN)==TRUE, "PREV_PLT_CN"] <- -9999
df <- na.omit(df)

df$agentSB <- ifelse(df$stand.SB>0, "SB", "absent")
df$agentMPB <- ifelse(df$stand.MPB>0, "MPB", "absent")
df$agentWBBB <- ifelse(df$stand.WBBB>0, "WBBB", "absent")

df$agents <-  do.call(paste, c(df[c("agentSB", "agentMPB", "agentWBBB")],sep="-"))
df$agents <- factor(df$agents, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))

df$nagents <- df$agents
df$nagents <- factor(df$nagents, levels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))

df$hostsPICO <- ifelse(df$TPH.PICO>0, "PICO", "absent")
df$hostsPIEN <- ifelse(df$TPH.PIEN>0, "PIEN", "absent")
df$hostsABLA <- ifelse(df$TPH.ABLA>0, "ABLA", "absent")
df$hosts <-  do.call(paste, c(df[c("hostsPICO", "hostsPIEN", "hostsABLA")],sep="-"))
df$hosts <- factor(df$hosts, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA', 'none'), ordered=TRUE)

df$nhosts <- df$hosts
df$nhosts <- factor(df$nhosts, levels=c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))

write.csv(df, here("Data/Processed/FIAdat-df.csv"),row.names=F)
```

#### Selection of plots
We selected all plots that were part of the annnual inventory and where all subplots were inventoried.

Within this plots, we characterized stand structure and composition using only live and recently killed (i.e. killed within the past ~10 years) trees greater that 12.7 cm DBH within the subplot.

```{r select plots}
df <- read.csv(here("Data/Processed/FIAdat-df.csv"))

##select most recent inventories
df.recent <- df

df.recent <-df.recent[!(df.recent$PREV_PLT_CN %in% df.recent$PLT_CN), ] # select plots that have not been subsequently sampled

##select inventories conducted within the last 10 years
df.recent <- df.recent[df.recent$INVYR %in% 2010:2019,]
df.recent$LONx <- df.recent$LON
df.recent$LATx <- df.recent$LAT
##select inventories with coordinates located in the Intermountain West
df.recent.sf <- st_as_sf(df.recent, coords = c("LONx", "LATx"), crs = 4326)
df.recent.sf <- st_transform(df.recent.sf,  crs = aea.proj)
states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
states <- st_transform(states, crs = aea.proj)
df.recent.sf.j.states <- st_join(df.recent.sf, states)
df.recent.sf.j.states<- df.recent.sf.j.states[df.recent.sf.j.states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]

df.recent.sf <- df.recent.sf.j.states
df.recent <- df.recent.sf
st_geometry(df.recent) <- NULL

write.csv(df.recent, here("Data/Processed/FIAdat-recent.csv"))
```

#### Regional bark beetle pressure
Additionally, given the effects of stand structure and composition on the susceptibility to bark beetle infestation are expected to be greatest in the initial stages of outbreak development [@raffa_cross-scale_2008, @hart_summer_2017], we calculated a proxy of regional bark beetle population density. To do so, we first identified all plots located within 25 km of the focal plot (hereafter *neighboring plots*). We then calculated the regional bark beetle pressure as:

$BP_{regional} = BA_{infested}/BA_{host_{inventoried}} * BA_{host_{total}}$

where $BP_{regional}$ is the regional beetle pressure, $BA_{infested}$ is the total basal area of infested trees within neighboring plots inventoried in the focal or preceding years, $BA_{host_{inventoried}}$ is the total basal area of host trees within neighboring plots inventoried in same year or preceding years, $BA_{host_{total}}$ is the total basal area of host trees for all neighboring plots, $n_{plots}$ is the number of neighboring plots.

```{r pressure, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-df.csv"))
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))

plots <- unique(df[,c("PLT_CN", "INVYR", "LAT", "LON")])
plotspdf <- SpatialPointsDataFrame(plots[,c("LON", "LAT")], data=plots, proj4string=CRS("+init=epsg:4269"))

plotspdf.t <- spTransform(plotspdf,
CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"))

# calculate distances between plots
plot.dists <- rgeos::gDistance(plotspdf.t, byid=T)
colnames(plot.dists) <- plotspdf$PLT_CN
rownames(plot.dists) <- plotspdf$PLT_CN
pressure <- data.frame(PLT_CN=plotspdf$PLT_CN, hostBA.MPB=NA, hostBA.SB=NA, hostBA.WBBB=NA, hostBA.MPB.samp=NA, hostBA.SB.samp=NA, hostBA.WBBB.samp=NA, hostBA.allBBs.samp=NA, nplots=NA, pressureMPB=NA, pressureSB=NA, pressureWBBB=NA, pressureallBBs=NA)

for(j in 1:ncol(plot.dists)){

  # determine plots within 25 km 
  x <- plot.dists[,j] 
  x <- x[-j] # remove plot j 
  x25 <- x[x<=25000]

  # set inventory year for plot j
  invyr <- plots[j, "INVYR"]

  # select plots that were inventoried before plot j
  pre_PLT_CNs <- plots[plots$INVYR<=invyr, "PLT_CN"]

  # select plots within 25 km that were also inventoried before plot j
  x25r <- x25[names(x25) %in% pre_PLT_CNs]

  # calculate area of host basal area within 25 km using only the most recent plot inventories 
  if(length(x25)>=0){
    pressure[j,  "hostBA.MPB"] <- sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.PICO"], na.rm=T) 
    pressure[j,  "hostBA.SB"] <-sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.PIEN"], na.rm=T) 
    pressure[j,  "hostBA.WBBB"] <- sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.ABLA"], na.rm=T) 
    pressure[j,  "hostBA.allBBs"] <- sum(rowSums(df.recent[df.recent$PLT_CN %in% names(x25), c( "BA.PICO", "BA.PIEN", "BA.ABLA")], na.rm=T), na.rm=T) 
  }

  # calculate regional pressure
  if(length(x25r)>0){
    df.x25r <- df[df$PLT_CN %in% names(x25r),]

    # calculate total basal area of PICO, PIEN, and ABLA sampled for all plots within 25 km sampled in the current or previous years 
    pressure[j,  "hostBA.MPB.samp"]  <- sum(df.x25r[, "BA.PICO"], na.rm=T) 
    pressure[j,  "hostBA.SB.samp"] <-sum(df.x25r[, "BA.PIEN"], na.rm=T) 
    pressure[j,  "hostBA.WBBB.samp"] <- sum(df.x25r[, "BA.ABLA"], na.rm=T) 
    pressure[j,  "hostBA.allBBs.samp"]   <- sum(rowSums(df.x25r[, c( "BA.PICO", "BA.PIEN", "BA.ABLA")], na.rm=T), na.rm=T)

    # calculate number of plots sampled for all plots within 25 km sampled in the current or previous years 
    pressure[j,  "nplots"]   <- nrow(df.x25r)

   # calculate total basal area of trees affected by MPB, SB, and WBBB sampled for all plots within 25 km sampled in the current or previous years 
    pressure[j,  "pressureMPB"] <- sum(df.x25r[, "BA.MPB"], na.rm=T)  
    pressure[j,  "pressureSB"] <- sum(df.x25r[, "BA.SB"], na.rm=T)   
    pressure[j,  "pressureWBBB"] <- sum(df.x25r[, "BA.WBBB"], na.rm=T) 
    pressure[j,  "pressureallBBs"] <- sum(df.x25r[, c("BA.MPB", "BA.SB", "BA.WBBB")], na.rm=T) 
  }

}

pressure$Rpressure.MPB <- (pressure$pressureMPB/pressure$hostBA.MPB.samp) * pressure$hostBA.MPB 
pressure$Rpressure.SB <- (pressure$pressureSB/pressure$hostBA.SB.samp) *pressure$hostBA.SB
pressure$Rpressure.WBBB <- (pressure$pressureWBBB/pressure$hostBA.WBBB.samp) *pressure$hostBA.WBBB 
pressure$Rpressure.allBBs <- ( pressure$pressureallBBs/pressure$hostBA.allBBs.samp) * pressure$hostBA.allBBs
pressure[is.na(pressure)==T]<-0

pressure <- pressure[pressure$PLT_CN %in% df.recent$PLT_CN,]
write.csv(pressure, paste0(here("Data/Processed/FIAdat-pressure.csv")),row.names=F)
```

#### Climate data
Finally, because bark beetle developmental rates and overwintering mortality are sensitive to summer and winter temperatures [@miller_cold-hardiness_1987, @bentz_temperature-dependent_1991, @hansen_temperature-based_2001], respectively, we acquired the average maximum temperatures for July (JulTMAX) and average minimum temperatures for December (DecTMIN). Because prior research shows associations between precipitation and outbreak [@hebertson_climate_2008; @chapman_spatiotemporal_2012], we also obtained monthly total precipitation and calculated the total water year (i.e. September to October) precipitation (WYPPT). As indices of host stress due to drought, we used the Standardized Evapotranspiration Precipitation Index (SPEI). SPEI is a multiscalar index that uses precipitation and temperature data in a simple Thornthwaite water balance model [@thornthwaite_approach_1948] to quantify drought [vicente-serrano_multiscalar_2009,]. Specifically we obtained SPEI calculated over the June–July period (JJSPEI) and a 5-yr period, which spanned September to the October five years prior (i.e., five water years; WYSPEI) [@hart_summer_2017]. SPEI data were produced from PRISM data and downloaded from the West Wide Drought Tracker [@abatzoglou_west_2017]. For each plot, we calculated the 1982-2017 climatic normals. For each plot we assigned the climate normal value of the 4 x 4 km pixel underlying each point.

```{r download and process climate data, eval=F}
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
plots <- unique(df.recent[,c("PLT_CN", "LAT", "LON", "INVYR")])
plotspdf <- SpatialPointsDataFrame(plots[,c("LON", "LAT")], data=plots, proj4string=CRS("+init=epsg:4269"))

# PRISM
dir.create(here("Data/Spatial/PRISM/"), showWarnings = FALSE)

## Normals
dir.create(here("Data/Spatial/PRISM/Normals/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/Normals/"))
get_prism_normals("ppt", resolution="4km", annual = TRUE)
get_prism_normals("tmax", resolution="4km", mon=7, annual = FALSE)
get_prism_normals("tmin", resolution="4km", mon=12, annual = FALSE)

normals <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))
normals.df <- raster::extract(normals, plotspdf, method="simple", df=T)
normals.df$PLT_CN <- plots$PLT_CN
colnames(normals.df) <- c("nPPT12", "nTMAX1.7", "nTMIN1.12", "PLT_CN")
write.csv(normals.df, here("Data/Processed/Normals.csv"), row.names = F)

## Jul MAX Temperature
dir.create(here("Data/Spatial/PRISM/TMAX1_7/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/TMAX1_7/"))
tmax <- get_prism_monthlys(type="tmax", mon=7, years=1981:2020)
TMAX1.7 <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))

options(prism.path=here("Data/Spatial/PRISM/Normals/"))
nTMAX1.7 <- raster(pd_to_file(prism_archive_ls()[2]))
TMAX1.7.anom <- (TMAX1.7+100) - (nTMAX1.7+100)

TMAX1.7.df <- raster::extract(TMAX1.7, plotspdf, method="simple", df=T)
TMAX1.7.df$PLT_CN <- plots$PLT_CN
write.csv(TMAX1.7.df, here("Data/Processed/TMAX1-7.csv"), row.names = F)

TMAX1.7.anom.df <- raster::extract(TMAX1.7.anom, plotspdf, method="simple", df=T)
TMAX1.7.anom.df$PLT_CN <- plots$PLT_CN
write.csv(TMAX1.7.anom.df, here("Data/Processed/TMAX1-7anom.csv"), row.names = F)

## Dec MIN Temperature
dir.create(here("Data/Spatial/PRISM/TMIN1_12/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/TMIN1_12/"))
TMIN12 <- get_prism_monthlys(type="tmin", mon=12, years=1981:2020)
TMIN1.12 <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))

options(prism.path=here("Data/Spatial/PRISM/Normals/"))
nTMIN1.12 <- raster(pd_to_file(prism_archive_ls()[3]))
TMIN1.12.anom <- (TMIN1.12+100) - (nTMIN1.12+100)

tmin1.12.df <- raster::extract(TMIN1.12, plotspdf, method="simple", df=T)
tmin1.12.df$PLT_CN <- plots$PLT_CN
write.csv(tmin1.12.df, here("Data/Processed/TMIN1-12.csv"), row.names = F)

tmin1.12.anom.df <- raster::extract(TMIN1.12.anom, plotspdf, method="simple", df=T)
tmin1.12.anom.df$PLT_CN <- plots$PLT_CN
write.csv(tmin1.12.anom.df, here("Data/Processed/TMIN1-12anom.csv"), row.names = F)

## Annual PPT
dir.create(here("Data/Spatial/PRISM/PPT12/"), showWarnings = FALSE)
options(prism.path=here("Data/Spatial/PRISM/PPT12/"))
get_prism_annual(type="ppt", years=1981:2020)
ppt <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))

options(prism.path=here("Data/Spatial/PRISM/Normals/"))
nppt <- raster(pd_to_file(prism_archive_ls()[1]))
pptanom <- ppt - nppt

ppt.df <- raster::extract(ppt, plotspdf, method="simple", df=T)
ppt.df$PLT_CN <- plots$PLT_CN
write.csv(ppt.df, here("Data/Processed/PPT12.csv"), row.names = F)

pptanom.df <- raster::extract(pptanom, plotspdf, method="simple", df=T)
pptanom.df$PLT_CN <- plots$PLT_CN
write.csv(pptanom.df, here("Data/Processed/PPT12anom.csv"), row.names = F)

## JUN-JUL SPEI
dir.create(here("Data/Spatial/PRISM/SPEI2_7"), showWarnings = FALSE)
temp <- here("/Data/Spatial/PRISM/spei2_7_PRISM.nc")
download.file("http://www.wrcc.dri.edu/wwdt/data/PRISM/spei2/spei2_7_PRISM.nc", temp)

spei2 <- brick(here("Data/Spatial/PRISM/SPEI2_7/spei2_7_PRISM.nc"))
spei2.df <-raster:::extract(spei2, plotspdf, method="simple", df=T)[,-1]
spei2.df$PLT_CN <- plots$PLT_CN
write.csv(spei2.df,paste0(here("Data/Processed"),"SPEI2.csv"), row.names = F)

## 5 WATER YEAR SPEI
dir.create(here("Data/Spatial/PRISM/SPEI60_9"), showWarnings = FALSE)
temp <- here("Data/Spatial/PRISM/SPEI60_9/spei60_9_PRISM.nc")
download.file("http://www.wrcc.dri.edu/wwdt/data/PRISM/spei60/spei60_9_PRISM.nc", temp)

spei60 <- brick( here("Data/Spatial/PRISM/SPEI60_9/spei60_9_PRISM.nc"))
spei60.df <-raster:::extract(spei60, plotspdf, method="simple", df=T)[,-1]
spei60.df$PLT_CN <- plots$PLT_CN
write.csv(spei60.df,paste0(here("Data/Processed/"),"SPEI60.csv"), row.names = F)

spei60.df <- read.csv(here("Data/Processed/SPEI60.csv"))
spei2.df <- read.csv(here("Data/Processed/SPEI2.csv"))

ppt.df <- read.csv( here("Data/Processed/PPT12.csv"))
pptanom.df <- read.csv( here("Data/Processed/PPT12anom.csv"))

tmin1.12.df <- read.csv(here("Data/Processed/TMIN1-12.csv"))
tmin1.12.anom.df <- read.csv(here("Data/Processed/TMIN1-12anom.csv"))

tmax1.7.df <- read.csv(here("Data/Processed/TMAX1-7.csv"))
tmax1.7.anom.df <- read.csv(here("Data/Processed/TMAX1-7anom.csv"))

climatedata <- plots  
for(j in 1:nrow(climatedata)){
  INVYR <- climatedata[j, "INVYR"]
  end.yr <- INVYR 
  start.yr <- INVYR-9

  spei2.ts <- ts(t(spei2.df[j,-1*ncol(spei2.df)]),start=1895, freq=1)
  spei2.tsw <- window(spei2.ts, end=end.yr, start=start.yr)
  climatedata[j, "JJSPEI"] <- mean(spei2.tsw , na.rm=T)

  spei60.ts <- ts(t(spei60.df[j,-1*ncol(spei60.df)]),start=1895, freq=1)
  spei60.tsw <- window(spei60.ts, end=end.yr, start=start.yr)
  climatedata[j, "WYSPEI"] <- mean(spei60.tsw , na.rm=T)

 tmin1.12.ts <- ts(t(tmin1.12.df[j,-1*ncol(tmin1.12.df )]),start=1981, freq=1)
  tmin1.12.tsw <- window(tmin1.12.ts, end=end.yr, start=start.yr)
  climatedata[j, "DecTMIN"] <- mean(tmin1.12.tsw[,1], na.rm=T)

  tmin1.12a.ts <- ts(t(tmin1.12.anom.df[j,-1*ncol(tmin1.12.anom.df )]),start=1981, freq=1)
  tmin1.12a.tsw <- window(tmin1.12a.ts, end=end.yr, start=start.yr)
  climatedata[j, "DecTMINa"] <- mean(tmin1.12a.tsw[,1], na.rm=T)

   tmax1.7.ts <- ts(t(tmax1.7.df[j,-1*ncol(tmax1.7.df )]),start=1981, freq=1)
  tmax1.7.tsw <- window(tmax1.7.ts, end=end.yr, start=start.yr)
  climatedata[j, "JulTMAX"] <- mean(tmax1.7.tsw[,1], na.rm=T)

tmax1.7a.ts <- ts(t(tmax1.7.anom.df[j,-1*ncol(tmax1.7.anom.df )]),start=1981, freq=1)

  tmax1.7a.tsw <- window(tmax1.7a.ts, end=end.yr, start=start.yr)
  climatedata[j, "JulTMAXa"] <- mean(tmax1.7a.tsw[,1], na.rm=T)

  ppt.ts <- ts(t(ppt.df [j,-1*ncol(ppt.df )]),start=1981, freq=1)
  ppt.tsw <- window( ppt.ts, end=end.yr, start=start.yr)
  climatedata[j, "PPT"] <- mean(ppt.tsw[,1], na.rm=T)

  ppta.ts <- ts(t(pptanom.df [j,-1*ncol(pptanom.df )]),start=1981, freq=1)
 ppta.tsw <- window(ppta.ts, end=end.yr, start=start.yr)
  climatedata[j, "PPTa"] <- mean(ppta.tsw[,1], na.rm=T)
}

write.csv(climatedata, here("Data/Processed/climatedata.csv"), row.names = F)
```

```{r merge data, eval=F}
df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
pressure <- read.csv(here("Data/Processed/FIAdat-pressure.csv"))
climatedata <- read.csv(here("Data/Processed/climatedata.csv"))

df <- merge(df.recent, pressure, by="PLT_CN", all=T)
df <- merge(df, climatedata[, -2:-4], by="PLT_CN", all=T)

write.csv(df, here("Data/Processed/FIAdat-df-merge.csv"), row.names=F)
```

```{r reorganize by focal host, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-df-merge.csv"))

formatp <- function(d, host="PIEN", agent="SB"){
  df.p <- d[which(d[, paste0("TPH.", host)]>0),]
  #df.p <- d
  df.p$TPH.host <- df.p[, paste0("TPH.", host)]
  df.p$QMD.host <- df.p[, paste0("QMD.", host)]
  df.p$BA.host <- df.p[, paste0("BA.", host)]

  df.p$stand.BB <- df.p[, paste0("stand.", agent)]
  df.p$BA.BB <- df.p[, paste0("BA.", agent)]
  df.p$affected <- ifelse(df.p$stand.BB>0, "affected", "unaffected")
  df.p$focalhost <- host

  df.p$BAD.host <- df.p$BA.host/df.p$BA.total
  df.p$TPHD.host <- df.p$TPH.host/df.p$TPH.total
  df.p$pressure.BB <- df.p[,paste0("pressure", agent)]

  return(df.p)
}

df <- rbind(rbind(formatp(df), formatp(df, host="PICO", agent="MPB")), formatp(df, host="ABLA", agent="WBBB"), formatp(df, host="Allhosts", agent="allBBs"))

### Proportion of total BA affected
df$propTBA <- df$BA.BB/df$BA.total

### Proportion of host BA affected
df$propHBA <- df$BA.BB/df$BA.Allhosts

### Proportion of total trees affected
df$propTTPH <- (df$stand.BB*unique(6.018046)*2.47105)/df$TPH.total

### Proportion of host trees affected
df$propHTPH <- (df$stand.BB*unique(6.018046)*2.47105)/df$TPH.host

### Affected
df$affected01 <- ifelse(df$stand.BB>0, "affected", "unaffected")

write.csv(df, here("Data/Processed/FIAdat-df-merge-re.csv"),row.names=F)
```

## Analyses

# Descriptive analyses
```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), ordered=T)
df$hostsFPlong <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB & WBBB"), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB\n& SB","MPB &\nWBBB",  "SB &\nWBBB", "MPB,\nSB, &\nWBBB"), ordered=T)
df$agentsFplong <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB & WBBB"), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB, & WBBB"), ordered=T)

df$focalhost <- factor(df$focalhost, levels=c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("PICO", "PIEN", "ABLA", "Any"), ordered=T)

df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Any"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

df$propBA.PICO <- df$BA.PICO / df$BA.total *100
df$propBA.PIEN <- df$BA.PIEN / df$BA.total *100
df$propBA.ABLA <- df$BA.ABLA / df$BA.total *100
```

```{r map study area of hosts, eval=F}
f19 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf", proxy=F)
f93 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf", proxy=F)
f108 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf", proxy=F)

# reclassify to rasters to presence (1) and absence (0)
f19[f19>0] <- 1 
f93[f93>0] <- 1
f108[f108>0] <- 1

# define areas of multiple hosts
f108f93 <- f108 + f93 # add 
f108f93[f108f93<2]<- 0 # reclassify
f108f93[f108f93==2]<- 1 # reclassify
f108f19 <- f108 + f19
f108f19[f108f19<2]<- 0 # reclassify
f108f19[f108f19==2]<- 1 # reclassify
f93f19 <- f93 + f19
f93f19[f93f19<2]<- 0  # reclassify
f93f19[f93f19==2]<- 1 # reclassify

f108f93f19 <- f108 + f93 + f19

# define areas of one host
f108f93f19re1 <-f108f93f19
f108f93f19re1 [f108f93f19re1 ==1 ]<-1
f108f93f19re1 [f108f93f19re1>1]<-0

# define areas of two hosts
f108f93f19re2 <-f108f93f19
f108f93f19re2 [f108f93f19re2 ==1 ]<-0
f108f93f19re2 [f108f93f19re2==3]<-0 
f108f93f19re2 [f108f93f19re2 ==2 ]<-1

# limit maps of presence (1) and absence (0) to only areas with one host
f108 <- f108 * f108f93f19re1 
f93 <- f93 * f108f93f19re1
f19 <- f19 * f108f93f19re1 
f93[f93==1]<- 2 # give unique labels
f19[f19==1]<- 3 # give unique labels

# create maps of two hosts limited to areas with two hosts
f108f93  <- f108f93 * f108f93f19re2 
f108f93[f108f93==1]<-4 # give unique labels

f108f19 <- f108f19 * f108f93f19re2
f108f19[f108f19==1]<-5 # give unique label

f93f19 <- f93f19 * f108f93f19re2
f93f19[f93f19==1]<-6 # give unique labels

# create maps of all three hosts
f108f93f19[f108f93f19<=2]<-0
f108f93f19[f108f93f19==3]<-7 # give unique labels

fplot <- f108 + f93 + f19 + f108f93 + f108f19 + f93f19+ f108f93f19
st_crs(fplot) <- "EPSG:5070" # assign projection

fplot[fplot==0] <-NA

states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
notIMW <- states[!states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
IMW <- st_transform(IMW, "EPSG:5070")
notIMW <- st_transform(notIMW, "EPSG:5070")

ext <- as.matrix(extent(IMW))

sa.tmap <- tm_shape(IMW, bbox=tmaptools::bb(ext)) + tm_fill(col="#f5f5f5")+tm_borders()+tm_shape(fplot) +tm_raster(style = "cat", palette = c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), colorNA=NULL, legend.show = TRUE, title="Hosts                    ", labels = c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"))+ tm_shape(notIMW) + tm_fill(col="white")+tm_compass(position=c(0.95, 0.88), just=c("center", "bottom"))+tm_layout(legend.bg.color="white")+tm_graticules(lines=FALSE)+tm_layout(legend.outside=TRUE,legend.text.size=0.55)+tm_scale_bar(position = c(0.01, 0.001), just="left", breaks=c(0,100),text.size=1.1)+tm_shape(IMW) +tm_borders()

Fig.file <- here("Results/Figures/Figure-Studyarea.jpg")
jpeg(Fig.file, width=fig.width1.5,height=fig.width1.5,units="in", res=300)
sa.tmap 
whatever <- dev.off()
```

```{r describe frequency of hosts and agents, eval=T}
## Hosts
hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))

## Affected hosts
affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]

affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]

## Agents given host presence
agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
  mutate(freq = n / sum(n))
PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))

## Agents given presence of any agent
agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
  mutate(freq = n / sum(n))

agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))

p1 <- ggplot(df[df$focalhost=="Any",], aes(x=hosts, fill=interaction(hosts, affected))) + geom_bar(width=0.75, stat="count", col="black")+ylab("number of plots")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1),legend.position = "none", legend.title=element_blank())+scale_fill_manual(values=c(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), adjust_transparency(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), alpha=0.5)))

Fig.file <- here("Results/Figures/Figure-PresencebyHostcombos.jpg")
jpeg(Fig.file, width=fig.width1,height=fig.width1/1.5,units="in", res=300)
p1
whatever <- dev.off()
```

### Model occurrence

```{r select plots to control for sampling effort, eval=T}
#df.sub <- df[df$nplots>5, ]
df.sub <- df
## remove plots with other hosts
df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
#nrow(datBB.c)
df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]
```

```{r select plots with large & abundant, eval=T}
PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.10)
PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.10)
ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.10)

PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.10)
PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.10)
ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.10)
n
# PICO, PIEN & ABLA
df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PICO.PIEN.ABLA$focalhosts2 <- "PICO, PIEN & ABLA"
df.sub.PICO.PIEN.ABLA$nhosts2 <- 3

# PICO & PIEN
df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]

df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]

df.sub.PICO.PIEN$focalhosts2 <- "PICO & PIEN"
df.sub.PICO.PIEN < - df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
df.sub.PICO.PIEN$nhosts2 <- 2

# PICO & ABLA
df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PICO.ABLA$focalhosts2 <- "PICO & ABLA"
df.sub.PICO.ABLA$nhosts2 <- 2

df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]

# PIEN & ABLA
df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]

df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]

df.sub.PIEN.ABLA$focalhosts2 <- "PIEN & ABLA"
df.sub.PIEN.ABLA$nhosts2 <- 2

df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]

###
df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)

## Affected hosts
affectedhosts2 <- df.focal2 %>% group_by(focalhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhosts2[affectedhosts2$affected=="affected",]
affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts2[affectednhosts2$affected=="affected",]
### MPB

MPBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))

### PIEN

### ABLA

```

```{r, eval=T}
datBB.c <- df.sub[df.sub$focalhost=="Any",]
splitit <- initial_split(df.sub, prop=0.5)
datz.train <- training(splitit)
datz.test <- testing(splitit)
datz.train.cv <- vfold_cv(datz.train, v=10)
```

```{r fit MPB RF model, eval=T}
response <- "agentMPB"
#vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO", "pressureMPB", "hostBA.MPB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")

form <- paste(response, "~", paste(vars, collapse = " + "))

wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
step_corr(all_predictors())%>% prep()

# set model
rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")

# set the workflow
rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)

# tune model
tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))

tuneres <- rf_tune_results %>% collect_metrics()

param_best<- rf_tune_results %>% select_best(metric = "roc_auc")

# Use backward selection to identify important variables
#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
vars.final <- vars

BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.

test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here("Results/RF-testperformance-MPB-occurrence.csv"), row.names=F)

# generate predictions from the training dataset
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
write.csv(test_predictions , here("Results/RF-pred-MPB-occurrence.csv"), row.names=F)

# generate a confusion matrix
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
write.csv(confusion$table, here("Results/RF-confusiontable-MPB-occurrence.csv"), row.names=F)

# variable importance
final_model <- fit(rf_workflow_final, data=datz.train)
ranger_obj <- pull_workflow_fit(final_model)$fit

x <- as.data.frame(ranger_obj$variable.importance)
x$var <- row.names(x)
x$prop <- x[,1]/sum(x[,1])*100
x <- x[order(x$prop),]
x$var <- factor(x$var, levels=x$var, ordered=T)
write.csv(x, here("Results/RF-vimp-MPB-occurrence.csv"), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var", which.class="MPB")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here("Results/RF-partialdependence-MPB-occurrence.csv"), row.names=F)
```

```{r fit SB RF model, eval=T}
response <- "agentSB"
#vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN", "pressureSB", "hostBA.SB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")

form <- paste(response, "~", paste(vars, collapse = " + "))

wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%step_corr(all_predictors())%>% prep()

# set model
rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")

# set the workflow
rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)

# tune model
tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))

tuneres <- rf_tune_results %>% collect_metrics()

param_best<- rf_tune_results %>% select_best(metric = "roc_auc")

# Use backward selection to identify important variables
dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
registerDoSEQ()
vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))

BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.

test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here("Results/RF-testperformance-SB-occurrence.csv"), row.names=F)

# generate predictions from the training dataset
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
write.csv(test_predictions , here("Results/RF-pred-SB-occurrence.csv"), row.names=F)

# generate a confusion matrix
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
write.csv(confusion$table, here("Results/RF-confusiontable-SB-occurrence.csv"), row.names=F)

# variable importance
final_model <- fit(rf_workflow_final, data=datz.train)
ranger_obj <- pull_workflow_fit(final_model)$fit

x <- as.data.frame(ranger_obj$variable.importance)
x$var <- row.names(x)
x$prop <- x[,1]/sum(x[,1])*100
x <- x[order(x$prop),]
x$var <- factor(x$var, levels=x$var, ordered=T)
write.csv(x, here("Results/RF-vimp-SB-occurrence.csv"), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z)
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var", which.class="SB")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here("Results/RF-partialdependence-SB-occurrence.csv"), row.names=F)
```

```{r fit WBBB RF model, eval=T}
response <- "agentWBBB"
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA", "pressureWBBB", "hostBA.WBBB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")

form <- paste(response, "~", paste(vars, collapse = " + "))

wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
wts <- wts/sum(wts)

# set formula
BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
step_corr(all_predictors())%>% prep()

# set model
rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")

# set the workflow
rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)

# tune model
tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))

tuneres <- rf_tune_results %>% collect_metrics()

param_best<- rf_tune_results %>% select_best(metric = "roc_auc")

# Use backward selection to identify important variables
dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
registerDoSEQ()
vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))

BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)

rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")

rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.

test_performance <- rf_workflow_final_fit %>% collect_metrics()
write.csv(test_performance, here("Results/RF-testperformance-WBBB-occurrence.csv"), row.names=F)

# generate predictions from the training dataset
test_predictions <- rf_workflow_final_fit %>% collect_predictions()
write.csv(test_predictions , here("Results/RF-pred-WBBB-occurrence.csv"), row.names=F)

# generate a confusion matrix
confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
write.csv(confusion$table, here("Results/RF-confusiontable-WBBB-occurrence.csv"), row.names=F)

# variable importance
final_model <- fit(rf_workflow_final, data=datz.train)
ranger_obj <- pull_workflow_fit(final_model)$fit

x <- as.data.frame(ranger_obj$variable.importance)
x$var <- row.names(x)
x$prop <- x[,1]/sum(x[,1])*100
x <- x[order(x$prop),]
x$var <- factor(x$var, levels=x$var, ordered=T)
write.csv(x, here("Results/RF-vimp-WBBB-occurrence.csv"), row.names=F)

# partial dependence
pd <- NULL
for(z in vars.final){
  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class="WBBB")
  pdz$var <- z
  colnames(pdz) <- c("value","yhat","var")
  pd <- rbind(pd, pdz)
}
write.csv(pd,  here("Results/RF-partialdependence-WBBB-occurrence.csv"), row.names=F)
```

```{r, eval=F}
mpbo <- read.csv(here("Results/RF-pred-MPB-occurrence.csv"))
sbo <- read.csv(here("Results/RF-pred-SB-occurrence.csv"))
wbbbo <- read.csv(here("Results/RF-pred-WBBB-occurrence.csv"))

bbo <- datz.test
bbo$mpbpred <- mpbo$.pred_MPB
bbo$sbpred <- sbo$.pred_SB 
bbo$wbbbpred <- wbbbo$.pred_WBBB
bbo$mpba<- mpbo$agentMPB
bbo$sba<- sbo$agentSB
bbo$wbbba<- wbbbo$agentWBBB

bbo.sub <- bbo[bbo$mpbpred>0.25 &bbo$wbbbpred>0.25 & bbo$sbpred>0.25,   ]
nrow(bbo.sub)
bbo$actualSUM <- ifelse(bbo$mpba=="absent",0, 1) + ifelse(bbo$sba=="absent",0, 1) + ifelse(bbo$wbbba=="absent",0, 1)

bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)

bbo$pres <- factor(ifelse(bbo$actualSUM >0, 1, 0), "affected", "unaffected")

ggplot(bbo, aes(x=predSUM, y=actualSUM))+geom_boxplot()

bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)
bbo

bbo$rMPB <- ifelse(bbo$agentMPB=="MPB", 1,0) - bbo$mpbpred 
bbo$rSB <- ifelse(bbo$agentSB=="SB", 1,0) - bbo$sbpred 
bbo$rWBBB <- ifelse(bbo$agentWBBB=="WBBB", 1,0) - bbo$wbbbpred
bbo$rSUM <- bbo$rMPB + bbo$rSB + bbo$rWBBB

x <- read.csv(here("Results/RF-partialdependence-SB-occurrence.csv"))
```

#### Severity
##### MPB
##### SB
##### WBBB

# Results

### Patterns

Across the Intermountain West, the most commonly occurring species was lodgepole pine, which was present in `r as.numeric(round(hosts[hosts$focalhost=="PICO", "freq"]*100,0))`% of plots, followed by subalpine fir (`r as.numeric(round(hosts[hosts$focalhost=="ABLA", "freq"]*100,0))`% of plots), and Engelmann spruce (`r as.numeric(round(hosts[hosts$focalhost=="PIEN", "freq"]*100,0))`% of plots). Given the presence of the focal host, the most commonly occurring bark beetle species was the MPB, which was found in `r PICOwMPB`% of plots with lodgepole pine. SB and WBBB, were found at much lower rates; SB was found in `r PIENwSB`% of plots with Engelmann spruce and WBBB was found in `r ABLAwWBBB`% of plots with subalpine fir.

The most commonly occurring tree community included only lodgepole pine (n=`r as.numeric(hostgroups[hostgroups$hosts=="PICO", "n"])`), followed by plots with all three hosts (n=`r as.numeric(hostgroups[hostgroups$hosts=="PICO, PIEN,\n& ABLA", "n"])`) and Engelmann spruce-sublapine fir plots (n=`r as.numeric(hostgroups[hostgroups$hosts=="PIEN \n& ABLA", "n"])`) (Figs. \@ref(fig:FigSA) -  \@ref(fig:FigPresHosts)). Logepole pine-Engelmann spruce communities were the least common (n=`r as.numeric(hostgroups[hostgroups$hosts=="PICO\n& PIEN", "n"])`), followed by plots with only subalpine fir (n=`r as.numeric(hostgroups[hostgroups$hosts=="ABLA", "n"])`) and only Engelmann spruce ((n=`r as.numeric(hostgroups[hostgroups$hosts=="PIEN", "n"])`)).

Consistent with diversity begets diversity hypothesis, the relative occurrence of bark beetle activity was greatest in stands with all three hosts (Fig \@ref(fig:FigPresHosts); `r as.numeric(round(affectednhosts[affectednhosts$nhosts==3, "freq"] *100, 0))`% of plots with all three hosts were affected by bark beetles). Stands with only one host were least likely to experience bark beetle infestation and stands with (`r as.numeric(round(affectednhosts[affectednhosts$nhosts==1, "freq"] *100, 0))`% of plots with only one host) and stands with two hosts were intermediate  (`r as.numeric(round(affectednhosts[affectednhosts$nhosts==2, "freq"] *100, 0))`% of plots with two hosts).

However host richness also varied with other factors known to influence outbreak presence and severity, notably stand structure and composition (Fig. \@ref(fig:FigStandSC)), climate (Fig. \@ref(fig:FigClimate)), and landscape scale patterns of host availability and bark beetle pressure (Fig. \@ref(fig:FigPressure)).

```{r FigSA, fig.cap = "The distribution of host species presence across the study area. Data are from the Individual Tree Species Atlas (Ellenwood et al. 2015) and represent conditions in ca. 2002.", out.width = "322.56pt"}
knitr::include_graphics( here("Results/Figures/Figure-Studyarea.jpg"))
```

```{r FigPresHosts, fig.cap = "The frequency of FIA plots inventoried by host identity and presence (dark) and absence (light) of bark beetle activity across the study area. ", out.width = "244.8pt"}
knitr::include_graphics( here("Results/Figures/Figure-PresencebyHostcombos.jpg"))
```

## Occurrence

```{r FigPropAgentxHosts, fig.cap = "A caption for figure 1", out.width = "504pt"}
knitr::include_graphics( here("Results/Figures/Figure-PropAffectedByAgentXHosts.jpg"))
```

## Co-occurrence of agents

Few stands were affected by multiple agents.

```{r FigOEACbyHI, fig.cap = "A caption for figure 2", out.width = "504pt"}
knitr::include_graphics(here("Results/Figures/Figure-ObsExpCo-ocurrenceAgentXHosts.jpg"))
```

## Ecological effects

# Discussion

# References

<div id="refs"></div>

```{r, child = "appendix1.Rmd"}
```

## Supplemental figures
### Comparison of stand structure and composition, climate, and regional bark beetle activity  
```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt"}
datBB.c <-df.sub[!df.sub$focalhost=="Any",]

QMD.melt <- melt(datBB.c[,c("hosts",paste0("QMD.", c("PIEN", "PICO", "ABLA")))], "hosts")
QMD.melt$variable <- factor(QMD.melt$variable, levels=paste0("QMD.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
QMD.melt$facet <- "QMD"

TPH.melt <- melt(datBB.c[,c("hosts",paste0("TPH.", c("PIEN", "PICO", "ABLA")))], "hosts")
TPH.melt$variable <- factor(TPH.melt$variable, levels=paste0("TPH.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
TPH.melt$facet <- "TPH"

BA.melt <- melt(datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA")))], "hosts")
BA.melt$variable <- factor(BA.melt$variable, levels=paste0("BA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
BA.melt$facet <- "BA"

propBA.melt <-datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA", "total")))]
propBA.melt$propBA.PIEN <- propBA.melt$BA.PIEN / propBA.melt$BA.total *100
propBA.melt$propBA.ABLA <- propBA.melt$BA.ABLA / propBA.melt$BA.total *100
propBA.melt$propBA.PICO <- propBA.melt$BA.PICO / propBA.melt$BA.total *100

propBA.melt <-  melt(propBA.melt[,c("hosts",paste0("propBA.", c("PIEN", "PICO", "ABLA")))], "hosts")
propBA.melt$variable <- factor(propBA.melt$variable, levels=paste0("propBA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
propBA.melt$facet <- "BA dominance (%)"

plotdat <- rbind(propBA.melt, TPH.melt, BA.melt, QMD.melt)

ggss <- ggplot(plotdat[plotdat$value>0,], aes(x=hosts, y=value,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BuPu")+ylab("value")+xlab("")+facet_wrap(.~facet, scales="free_y", ncol=1)+theme(legend.position = "bottom", legend.title=element_blank())

ggsave(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"), plot=ggss , width=fig.width1.5, height=fig.width1*2, units="in")
knitr::include_graphics(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"))
```

```{r FigPressure, fig.cap = "A caption", out.width = "322.56pt"}
## select plots where regional pressure can be estimated (> 5= plots)

RpressureSB.melt <- datBB.c[,c("hosts", "Rpressure.SB", "hostBA.SB", "nplots")]
colnames(RpressureSB.melt) <- c("hosts", "pressure", "hostBA", "nplots")
RpressureSB.melt$variable <- "SB"

RpressureMPB.melt <- datBB.c[, c("hosts", "Rpressure.MPB", "hostBA.MPB", "nplots")]
colnames(RpressureMPB.melt) <- c("hosts", "pressure", "hostBA", "nplots")
RpressureMPB.melt$variable <- "MPB"

RpressureWBBB.melt <- datBB.c[, c("hosts", "Rpressure.WBBB", "hostBA.WBBB", "nplots")]
colnames(RpressureWBBB.melt) <- c("hosts", "pressure", "hostBA", "nplots")
RpressureWBBB.melt$variable <- "WBBB"

Rpressure.melt <- rbind(RpressureMPB.melt,RpressureSB.melt)
Rpressure.melt <- rbind(Rpressure.melt,RpressureWBBB.melt)

Rpressure.melt$variable <- factor(Rpressure.melt$variable, levels=c("MPB","SB",  "WBBB"), labels=c("MPB","SB",  "WBBB"), ordered=T)
Rpressure.melt$propRegional <- Rpressure.melt$pressure /Rpressure.melt$hostBA
Rpressure.melt$host <- factor(Rpressure.melt$variable, levels=c("MPB","SB", "WBBB"), labels=c("PICO", "PIEN", "ABLA"), ordered=T)

ggpropRegional <- ggplot(Rpressure.melt, aes(x=hosts, y=propRegional,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BrBG")+ylab("Regional fraction\nof plots affected")+xlab("")+scale_fill_manual(values=rev(p2col)[2:4])

gghostBA <- ggplot(Rpressure.melt, aes(x=hosts, y=hostBA,fill=host))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BuPu")+ylab(expression(atop("Regional basal", paste("area (", m^2, ha^-1, ")"))))+xlab("")+theme(axis.text.x = element_blank(), axis.title.x = element_blank())

ggpressure <- ggplot(Rpressure.melt, aes(x=hosts, y=pressure,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_manual(values=rev(p2col)[2:4])+ylab("Regional pressure")+xlab("host composition")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1), axis.title.x = element_blank(), legend.position = "top", legend.title = element_blank())

ggcombo <- gghostBA + theme(axis.text.x=element_blank()) + ggpropRegional +  plot_layout(ncol=1)+ plot_annotation(tag_levels = 'A')+ plot_layout(guides = "collect") & theme(legend.position='bottom', legend.title=element_blank(),legend.box="vertical", legend.margin=margin(), plot.tag = element_text(face = 'bold'))
ggsave(here("Results/Figures/Figure-PressureComboByhostIdentity.jpg"), plot=ggcombo , width=fig.width1.5, height=fig.width1*1.5, units="in")

knitr::include_graphics(here("Results/Figures/Figure-PressureComboByhostIdentity.jpg"))
```

```{r FigClimate, fig.cap = "A caption", out.width = "244.8pt"}
ggDecTMINa <- ggplot(datBB.c, aes(x=hostsFPlong, y=DecTMINa))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Dec TMIN\n Anomaly")

ggJulTMAXa<- ggplot(datBB.c, aes(x=hostsFPlong, y=JulTMAXa))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Jul TMAX\n Anomaly")

ggPPTa<- ggplot(datBB.c, aes(x=hostsFPlong, y=PPTa))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Annual PPT\n Anomaly")

ggJJSPEI <- ggplot(datBB.c, aes(x=hostsFPlong, y=JJSPEI))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Jun-Jul SPEI")

ggWYSPEI <- ggplot(datBB.c, aes(x=hostsFPlong, y=WYSPEI))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("5-year SPEI")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1), axis.title.x = element_blank())

ggclimatecombo <- ggDecTMINa + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggJulTMAXa + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggPPTa + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggJJSPEI + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggWYSPEI + plot_layout(ncol=1) + plot_annotation(tag_levels = 'A') + plot_layout(guides = "collect") & theme(legend.position='bottom', legend.title=element_blank(), plot.tag = element_text(face = 'bold'))
ggsave(here("Results/Figures/Figure-ClimateByHostIdentity.jpg"), plot=ggclimatecombo , width=fig.width1, height=fig.width1*2, units="in")

knitr::include_graphics(here("Results/Figures/Figure-ClimateByHostIdentity.jpg"))
```

Stands with Engelmann spruce present occur at more southerly latitudes than stands with

```{r generate maps, eval=F}
df.all <- df.sub[df.sub$focalhost == "Any", ]

df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)

states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]

ext <- as.matrix(extent(IMW))
#ext[1,1] <- ext[1,1] -0.25
#ext[1,2] <- ext[1,2] +0.25
#ext[2,1] <- ext[2,1] -1
#ext[2,2] <- ext[2,2] +1

map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.9, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")

map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")

map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")

map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")

map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")

map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")

map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")

legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))

Fig.file <- here("Results/Figures/Figure-Map.jpg")
jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
whatever <- dev.off()

```

```{r FigMap, fig.cap="a caption", out.width = "504.8pt"}
knitr::include_graphics(here("Results/Figures/Figure-Map.jpg"))
```

### 2021-07-29 Code/Manuscript.Rmd
working on analyses

@@
 dir.create(here("Data/"), showWarnings = FALSE)
-dir.create(here("Code/"), showWarnings = FALSE)
 dir.create(here("Results/"), showWarnings = FALSE)
+dir.create(here("Results/Figures/"), showWarnings = FALSE)
 ```
@@
 
-The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming).
+The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming; Fig.
+\@ref(fig:FigSA)
 
@@
 #### Selection of plots
-We selected all plots that were part of the annnual inventory and where all subplots were inventoried.
+We selected all plots that were part of the annual inventory and where all subplots were inventoried.
 
@@
 ```{r map study area of hosts, eval=F}
-f19 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf", proxy=F)
-f93 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf", proxy=F)
-f108 <- stars::read_stars("/Volumes/GoogleDrive/My Drive/RESEARCH/GitHub/bbdiversity/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf", proxy=F)
+f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
+f93 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
+f108 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)
 
@@
 ## remove plots with other hosts
-df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
+#df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
 #nrow(datBB.c)
 df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]
-```
-
-```{r select plots with large & abundant, eval=T}
-PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.10)
-PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.10)
-ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.10)
-
-PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.10)
-PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.10)
-ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.10)
-n
-# PICO, PIEN & ABLA
-df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PICO.PIEN.ABLA$focalhosts2 <- "PICO, PIEN & ABLA"
-df.sub.PICO.PIEN.ABLA$nhosts2 <- 3
-
-# PICO & PIEN
-df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]
-
-df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]
-
-df.sub.PICO.PIEN$focalhosts2 <- "PICO & PIEN"
-df.sub.PICO.PIEN < - df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
-df.sub.PICO.PIEN$nhosts2 <- 2
-
-# PICO & ABLA
-df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PICO.ABLA$focalhosts2 <- "PICO & ABLA"
-df.sub.PICO.ABLA$nhosts2 <- 2
-
-df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
-
-# PIEN & ABLA
-df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PIEN.ABLA$focalhosts2 <- "PIEN & ABLA"
-df.sub.PIEN.ABLA$nhosts2 <- 2
-
-df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
-
-###
-df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
-df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
-df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)
-
-## Affected hosts
-affectedhosts2 <- df.focal2 %>% group_by(focalhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectedhosts2[affectedhosts2$affected=="affected",]
-affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts2[affectednhosts2$affected=="affected",]
-### MPB
-
-MPBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-
-### PIEN
-
-### ABLA
+
+# select plots where focal agent is all
+
+df.sub <- df.sub[df.sub$focalagent=="Any", ]
+```
+
+```{r select plots with large & abundant hosts, eval=T}
+threshold <- 0.10
+PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, threshold)
+PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, threshold)
+ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, threshold)
+
+PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, threshold)
+PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, threshold)
+ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, threshold)
+
+df.sub$hosts2.PICO <- "absent"
+df.sub[df.sub$QMD.PICO >PICO.QMD.threshold & df.sub$BA.PICO >PICO.BA.threshold, "hosts2.PICO"] <- "PICO"
+df.sub$hosts2.PIEN <- "absent"
+df.sub[df.sub$QMD.PIEN >PIEN.QMD.threshold & df.sub$BA.PIEN >PIEN.BA.threshold, "hosts2.PIEN"] <- "PIEN"
+df.sub$hosts2.ABLA<- "absent"
+df.sub[df.sub$QMD.ABLA >ABLA.QMD.threshold & df.sub$BA.ABLA >ABLA.BA.threshold, "hosts2.ABLA"] <- "ABLA"
+
+df.sub$hosts2<- NA
+df.sub$hosts2<- do.call(paste, c(df.sub[c("hosts2.PICO", "hosts2.PIEN", "hosts2.ABLA")],sep="-"))
+df.sub <- df.sub[!df.sub$hosts2 =='absent-absent-absent',]
+df.sub$hosts2 <- factor(df.sub$hosts2, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA'), ordered=TRUE)
+df.sub$nhosts2 <- as.numeric(factor(df.sub$hosts2, levels=c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA'), labels=c(1, 1, 1, 2,2,2, 3)))
+
+### Number of hosts
+
+#### Any hosts
+affectednhosts2 <- df.sub %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts2 <- affectednhosts2[affectednhosts2$affected=="affected",]
+colnames(affectednhosts2) <- c("nhosts", "agent", "n", "freq")
+affectednhosts2$agent <- "Any"
+
+#### MPB
+MPBnhosts2 <- df.sub %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+MPBnhosts2 <- MPBnhosts2[MPBnhosts2$agentMPB=="MPB",]
+colnames(MPBnhosts2) <- c("nhosts", "agent", "n", "freq")
+
+#### SB
+SBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentSB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+SBnhosts2 <- SBnhosts2[SBnhosts2$agentSB=="SB",]
+colnames(SBnhosts2) <- c("nhosts", "agent", "n", "freq")
+
+#### WBBB
+WBBBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentWBBB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+WBBBnhosts2 <- WBBBnhosts2[WBBBnhosts2$agentWBBB=="WBBB",]
+colnames(WBBBnhosts2) <- c("nhosts", "agent", "n", "freq")
+
+#### Merge
+affectednhosts2 <- rbind(as.data.frame(affectednhosts2),as.data.frame(MPBnhosts2))
+affectednhosts2 <- rbind(affectednhosts2, as.data.frame(SBnhosts2))
+affectednhosts2 <- rbind(affectednhosts2, as.data.frame(WBBBnhosts2))
+
+### Host identity
+#### Any hosts
+affectedhosts2 <- df.sub %>% group_by(hosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectedhosts2 <- affectedhosts2[affectedhosts2$affected=="affected",]
+colnames(affectedhosts2) <- c("hosts", "agent", "n", "freq")
+affectedhosts2$agent <- "Any"
+
+#### MPB
+MPBhosts2 <- df.sub %>% group_by(hosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+MPBhosts2 <- MPBhosts2[MPBhosts2$agentMPB=="MPB",]
+colnames(MPBhosts2) <- c("hosts", "agent", "n", "freq")
+
+#### SB
+SBhosts2 <- df.sub  %>% group_by(hosts2, agentSB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+SBhosts2 <- SBhosts2[SBhosts2$agentSB=="SB",]
+colnames(SBhosts2) <- c("hosts", "agent", "n", "freq")
+
+#### WBBB
+WBBBhosts2 <-df.sub  %>% group_by(hosts2, agentWBBB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+WBBBhosts2 <- WBBBhosts2[WBBBhosts2$agentWBBB=="WBBB",]
+colnames(WBBBhosts2) <- c("hosts", "agent", "n", "freq")
+
+#### Merge
+affectedhosts2 <- rbind(as.data.frame(affectedhosts2),as.data.frame(MPBhosts2))
+affectedhosts2 <- rbind(affectedhosts2, as.data.frame(SBhosts2))
+affectedhosts2 <- rbind(affectedhosts2, as.data.frame(WBBBhosts2))
+affectedhosts2$hosts <- factor(affectedhosts2$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+
+### test for significant differences
+affectedhosts3 <- as.data.frame(table(df.sub$hosts2))
+colnames(affectedhosts3) <- c("hosts", "N")
+
+affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="Any", c("hosts", "n")], by="hosts", all=T)
+colnames (affectedhosts3 ) <- c("hosts", "n", "Any")
+affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="MPB", c("hosts", "n")], by="hosts", all=T)
+colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB")
+affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="SB", c("hosts", "n")], by="hosts", all=T)
+colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB", "SB")
+affectedhosts3 <- merge(affectedhosts3, affectedhosts2[affectedhosts2$agent=="WBBB", c("hosts", "n")], by="hosts", all=T)
+colnames (affectedhosts3 ) <- c("hosts", "n", "Any", "MPB", "SB", "WBBB")
+
+affectedhosts3$label <- paste0(do.call(paste, c(affectedhosts3[c("hosts", "n")],sep="\n(n=")), ")")
+
+affectedhosts2$p.value <- NA
+affectedhosts2$statistic <- NA
+for(j in unique(affectedhosts2$hosts)){
+  affectedhosts2j <- affectedhosts2[affectedhosts2$hosts==j, ]
+  affectedhosts3j <- affectedhosts3[affectedhosts3$hosts==j,]
+  for(k in unique(affectedhosts2j$agent)[!unique(affectedhosts2j$agent)=="Any"]){
+
+    test <- prop.test(x=do.call(c,affectedhosts3j[, c("Any", k)]), n=rep(affectedhosts3j[, "n"],2), alternative="greater")
+  affectedhosts2[affectedhosts2$hosts==j & affectedhosts2$agent==k, "p.value"] <- round(test$p.value,3)
+    affectedhosts2[affectedhosts2$hosts==j & affectedhosts2$agent==k, "statistic"] <- test$statistic
+  }
+}
+affectedhosts2$symbol <- ifelse( affectedhosts2$p.value<0.05, "*", "")
+affectedhosts2$agent <- factor(affectedhosts2$agent, levels=c("MPB", "SB", "WBBB", "Any"), ordered=T)
+
+affectedhosts2 <- affectedhosts2[affectedhosts2$hosts %in% c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ]
+affectedhosts2$hosts <- factor(affectedhosts2$hosts, levels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=affectedhosts3$label[c(4,3,7,5)], ordered=T)
+
+Fig.file <- here("Results/Figures/Figure-OccurenceGivenSuitableStands.jpg")
+jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
+ggplot(affectedhosts2, aes(x=agent, y=freq, fill=agent))+geom_bar(stat="identity")+
+geom_text(aes(label = symbol), vjust = 0.2)+facet_wrap(.~hosts, scales="free", nrow=1)+ylab("proportion affected")+xlab("bark beetle species")+scale_fill_manual(values=c(rev(p2col)[2:4], grey(0.25)))+theme(legend.position="none")
+whatever <- dev.off()
 
@@
 
-```{r, child = "appendix1.Rmd"}
-```
-
-## Supplemental figures
+## Supplement
 ### Comparison of stand structure and composition, climate, and regional bark beetle activity  

### 2021-09-13 Code/Manuscript-SMOTE.Rmd
updating

--- 2021-07-29 Code/Manuscript.Rmd
@@
 ### Import libraries
-ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'foreach'))
+ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach'))
 
@@
 Active damage is easier to identify, thus the codes for damaging agents are much more specific.
+
 ```{r determine presence of BB agents in focal hosts, eval=F}
@@
 
-#### Regional bark beetle pressure
-Additionally, given the effects of stand structure and composition on the susceptibility to bark beetle infestation are expected to be greatest in the initial stages of outbreak development [@raffa_cross-scale_2008, @hart_summer_2017], we calculated a proxy of regional bark beetle population density.
-To do so, we first identified all plots located within 25 km of the focal plot (hereafter *neighboring plots*).
-We then calculated the regional bark beetle pressure as:
-
-$BP_{regional} = BA_{infested}/BA_{host_{inventoried}} * BA_{host_{total}}$
-
-where $BP_{regional}$ is the regional beetle pressure, $BA_{infested}$ is the total basal area of infested trees within neighboring plots inventoried in the focal or preceding years, $BA_{host_{inventoried}}$ is the total basal area of host trees within neighboring plots inventoried in same year or preceding years, $BA_{host_{total}}$ is the total basal area of host trees for all neighboring plots, $n_{plots}$ is the number of neighboring plots.
-
-```{r pressure, eval=F}
-df <- read.csv(here("Data/Processed/FIAdat-df.csv"))
-df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
-
-plots <- unique(df[,c("PLT_CN", "INVYR", "LAT", "LON")])
-plotspdf <- SpatialPointsDataFrame(plots[,c("LON", "LAT")], data=plots, proj4string=CRS("+init=epsg:4269"))
-
-plotspdf.t <- spTransform(plotspdf,
-CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"))
-
-# calculate distances between plots
-plot.dists <- rgeos::gDistance(plotspdf.t, byid=T)
-colnames(plot.dists) <- plotspdf$PLT_CN
-rownames(plot.dists) <- plotspdf$PLT_CN
-pressure <- data.frame(PLT_CN=plotspdf$PLT_CN, hostBA.MPB=NA, hostBA.SB=NA, hostBA.WBBB=NA, hostBA.MPB.samp=NA, hostBA.SB.samp=NA, hostBA.WBBB.samp=NA, hostBA.allBBs.samp=NA, nplots=NA, pressureMPB=NA, pressureSB=NA, pressureWBBB=NA, pressureallBBs=NA)
-
-for(j in 1:ncol(plot.dists)){
-
-  # determine plots within 25 km 
-  x <- plot.dists[,j] 
-  x <- x[-j] # remove plot j 
-  x25 <- x[x<=25000]
-
-  # set inventory year for plot j
-  invyr <- plots[j, "INVYR"]
-
-  # select plots that were inventoried before plot j
-  pre_PLT_CNs <- plots[plots$INVYR<=invyr, "PLT_CN"]
-
-  # select plots within 25 km that were also inventoried before plot j
-  x25r <- x25[names(x25) %in% pre_PLT_CNs]
-
-  # calculate area of host basal area within 25 km using only the most recent plot inventories 
-  if(length(x25)>=0){
-    pressure[j,  "hostBA.MPB"] <- sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.PICO"], na.rm=T) 
-    pressure[j,  "hostBA.SB"] <-sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.PIEN"], na.rm=T) 
-    pressure[j,  "hostBA.WBBB"] <- sum(df.recent[df.recent$PLT_CN %in% names(x25), "BA.ABLA"], na.rm=T) 
-    pressure[j,  "hostBA.allBBs"] <- sum(rowSums(df.recent[df.recent$PLT_CN %in% names(x25), c( "BA.PICO", "BA.PIEN", "BA.ABLA")], na.rm=T), na.rm=T) 
-  }
-
-  # calculate regional pressure
-  if(length(x25r)>0){
-    df.x25r <- df[df$PLT_CN %in% names(x25r),]
-
-    # calculate total basal area of PICO, PIEN, and ABLA sampled for all plots within 25 km sampled in the current or previous years 
-    pressure[j,  "hostBA.MPB.samp"]  <- sum(df.x25r[, "BA.PICO"], na.rm=T) 
-    pressure[j,  "hostBA.SB.samp"] <-sum(df.x25r[, "BA.PIEN"], na.rm=T) 
-    pressure[j,  "hostBA.WBBB.samp"] <- sum(df.x25r[, "BA.ABLA"], na.rm=T) 
-    pressure[j,  "hostBA.allBBs.samp"]   <- sum(rowSums(df.x25r[, c( "BA.PICO", "BA.PIEN", "BA.ABLA")], na.rm=T), na.rm=T)
-
-    # calculate number of plots sampled for all plots within 25 km sampled in the current or previous years 
-    pressure[j,  "nplots"]   <- nrow(df.x25r)
-
-   # calculate total basal area of trees affected by MPB, SB, and WBBB sampled for all plots within 25 km sampled in the current or previous years 
-    pressure[j,  "pressureMPB"] <- sum(df.x25r[, "BA.MPB"], na.rm=T)  
-    pressure[j,  "pressureSB"] <- sum(df.x25r[, "BA.SB"], na.rm=T)   
-    pressure[j,  "pressureWBBB"] <- sum(df.x25r[, "BA.WBBB"], na.rm=T) 
-    pressure[j,  "pressureallBBs"] <- sum(df.x25r[, c("BA.MPB", "BA.SB", "BA.WBBB")], na.rm=T) 
-  }
-
-}
-
-pressure$Rpressure.MPB <- (pressure$pressureMPB/pressure$hostBA.MPB.samp) * pressure$hostBA.MPB 
-pressure$Rpressure.SB <- (pressure$pressureSB/pressure$hostBA.SB.samp) *pressure$hostBA.SB
-pressure$Rpressure.WBBB <- (pressure$pressureWBBB/pressure$hostBA.WBBB.samp) *pressure$hostBA.WBBB 
-pressure$Rpressure.allBBs <- ( pressure$pressureallBBs/pressure$hostBA.allBBs.samp) * pressure$hostBA.allBBs
-pressure[is.na(pressure)==T]<-0
-
-pressure <- pressure[pressure$PLT_CN %in% df.recent$PLT_CN,]
-write.csv(pressure, paste0(here("Data/Processed/FIAdat-pressure.csv")),row.names=F)
-```
-
-#### Climate data
-Finally, because bark beetle developmental rates and overwintering mortality are sensitive to summer and winter temperatures [@miller_cold-hardiness_1987, @bentz_temperature-dependent_1991, @hansen_temperature-based_2001], respectively, we acquired the average maximum temperatures for July (JulTMAX) and average minimum temperatures for December (DecTMIN).
-Because prior research shows associations between precipitation and outbreak [@hebertson_climate_2008; @chapman_spatiotemporal_2012], we also obtained monthly total precipitation and calculated the total water year (i.e.
-September to October) precipitation (WYPPT).
-As indices of host stress due to drought, we used the Standardized Evapotranspiration Precipitation Index (SPEI).
-SPEI is a multiscalar index that uses precipitation and temperature data in a simple Thornthwaite water balance model [@thornthwaite_approach_1948] to quantify drought [vicente-serrano_multiscalar_2009,].
-Specifically we obtained SPEI calculated over the June–July period (JJSPEI) and a 5-yr period, which spanned September to the October five years prior (i.e., five water years; WYSPEI) [@hart_summer_2017].
-SPEI data were produced from PRISM data and downloaded from the West Wide Drought Tracker [@abatzoglou_west_2017].
-For each plot, we calculated the 1982-2017 climatic normals.
-For each plot we assigned the climate normal value of the 4 x 4 km pixel underlying each point.
-
-```{r download and process climate data, eval=F}
-df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
-plots <- unique(df.recent[,c("PLT_CN", "LAT", "LON", "INVYR")])
-plotspdf <- SpatialPointsDataFrame(plots[,c("LON", "LAT")], data=plots, proj4string=CRS("+init=epsg:4269"))
-
-# PRISM
-dir.create(here("Data/Spatial/PRISM/"), showWarnings = FALSE)
-
-## Normals
-dir.create(here("Data/Spatial/PRISM/Normals/"), showWarnings = FALSE)
-options(prism.path=here("Data/Spatial/PRISM/Normals/"))
-get_prism_normals("ppt", resolution="4km", annual = TRUE)
-get_prism_normals("tmax", resolution="4km", mon=7, annual = FALSE)
-get_prism_normals("tmin", resolution="4km", mon=12, annual = FALSE)
-
-normals <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))
-normals.df <- raster::extract(normals, plotspdf, method="simple", df=T)
-normals.df$PLT_CN <- plots$PLT_CN
-colnames(normals.df) <- c("nPPT12", "nTMAX1.7", "nTMIN1.12", "PLT_CN")
-write.csv(normals.df, here("Data/Processed/Normals.csv"), row.names = F)
-
-## Jul MAX Temperature
-dir.create(here("Data/Spatial/PRISM/TMAX1_7/"), showWarnings = FALSE)
-options(prism.path=here("Data/Spatial/PRISM/TMAX1_7/"))
-tmax <- get_prism_monthlys(type="tmax", mon=7, years=1981:2020)
-TMAX1.7 <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))
-
-options(prism.path=here("Data/Spatial/PRISM/Normals/"))
-nTMAX1.7 <- raster(pd_to_file(prism_archive_ls()[2]))
-TMAX1.7.anom <- (TMAX1.7+100) - (nTMAX1.7+100)
-
-TMAX1.7.df <- raster::extract(TMAX1.7, plotspdf, method="simple", df=T)
-TMAX1.7.df$PLT_CN <- plots$PLT_CN
-write.csv(TMAX1.7.df, here("Data/Processed/TMAX1-7.csv"), row.names = F)
-
-TMAX1.7.anom.df <- raster::extract(TMAX1.7.anom, plotspdf, method="simple", df=T)
-TMAX1.7.anom.df$PLT_CN <- plots$PLT_CN
-write.csv(TMAX1.7.anom.df, here("Data/Processed/TMAX1-7anom.csv"), row.names = F)
-
-## Dec MIN Temperature
-dir.create(here("Data/Spatial/PRISM/TMIN1_12/"), showWarnings = FALSE)
-options(prism.path=here("Data/Spatial/PRISM/TMIN1_12/"))
-TMIN12 <- get_prism_monthlys(type="tmin", mon=12, years=1981:2020)
-TMIN1.12 <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))
-
-options(prism.path=here("Data/Spatial/PRISM/Normals/"))
-nTMIN1.12 <- raster(pd_to_file(prism_archive_ls()[3]))
-TMIN1.12.anom <- (TMIN1.12+100) - (nTMIN1.12+100)
-
-tmin1.12.df <- raster::extract(TMIN1.12, plotspdf, method="simple", df=T)
-tmin1.12.df$PLT_CN <- plots$PLT_CN
-write.csv(tmin1.12.df, here("Data/Processed/TMIN1-12.csv"), row.names = F)
-
-tmin1.12.anom.df <- raster::extract(TMIN1.12.anom, plotspdf, method="simple", df=T)
-tmin1.12.anom.df$PLT_CN <- plots$PLT_CN
-write.csv(tmin1.12.anom.df, here("Data/Processed/TMIN1-12anom.csv"), row.names = F)
-
-## Annual PPT
-dir.create(here("Data/Spatial/PRISM/PPT12/"), showWarnings = FALSE)
-options(prism.path=here("Data/Spatial/PRISM/PPT12/"))
-get_prism_annual(type="ppt", years=1981:2020)
-ppt <- brick(apply(as.data.frame(pd_to_file(prism_archive_ls())), FUN=raster, MARGIN=1))
-
-options(prism.path=here("Data/Spatial/PRISM/Normals/"))
-nppt <- raster(pd_to_file(prism_archive_ls()[1]))
-pptanom <- ppt - nppt
-
-ppt.df <- raster::extract(ppt, plotspdf, method="simple", df=T)
-ppt.df$PLT_CN <- plots$PLT_CN
-write.csv(ppt.df, here("Data/Processed/PPT12.csv"), row.names = F)
-
-pptanom.df <- raster::extract(pptanom, plotspdf, method="simple", df=T)
-pptanom.df$PLT_CN <- plots$PLT_CN
-write.csv(pptanom.df, here("Data/Processed/PPT12anom.csv"), row.names = F)
-
-## JUN-JUL SPEI
-dir.create(here("Data/Spatial/PRISM/SPEI2_7"), showWarnings = FALSE)
-temp <- here("/Data/Spatial/PRISM/spei2_7_PRISM.nc")
-download.file("http://www.wrcc.dri.edu/wwdt/data/PRISM/spei2/spei2_7_PRISM.nc", temp)
-
-spei2 <- brick(here("Data/Spatial/PRISM/SPEI2_7/spei2_7_PRISM.nc"))
-spei2.df <-raster:::extract(spei2, plotspdf, method="simple", df=T)[,-1]
-spei2.df$PLT_CN <- plots$PLT_CN
-write.csv(spei2.df,paste0(here("Data/Processed"),"SPEI2.csv"), row.names = F)
-
-## 5 WATER YEAR SPEI
-dir.create(here("Data/Spatial/PRISM/SPEI60_9"), showWarnings = FALSE)
-temp <- here("Data/Spatial/PRISM/SPEI60_9/spei60_9_PRISM.nc")
-download.file("http://www.wrcc.dri.edu/wwdt/data/PRISM/spei60/spei60_9_PRISM.nc", temp)
-
-spei60 <- brick( here("Data/Spatial/PRISM/SPEI60_9/spei60_9_PRISM.nc"))
-spei60.df <-raster:::extract(spei60, plotspdf, method="simple", df=T)[,-1]
-spei60.df$PLT_CN <- plots$PLT_CN
-write.csv(spei60.df,paste0(here("Data/Processed/"),"SPEI60.csv"), row.names = F)
-
-spei60.df <- read.csv(here("Data/Processed/SPEI60.csv"))
-spei2.df <- read.csv(here("Data/Processed/SPEI2.csv"))
-
-ppt.df <- read.csv( here("Data/Processed/PPT12.csv"))
-pptanom.df <- read.csv( here("Data/Processed/PPT12anom.csv"))
-
-tmin1.12.df <- read.csv(here("Data/Processed/TMIN1-12.csv"))
-tmin1.12.anom.df <- read.csv(here("Data/Processed/TMIN1-12anom.csv"))
-
-tmax1.7.df <- read.csv(here("Data/Processed/TMAX1-7.csv"))
-tmax1.7.anom.df <- read.csv(here("Data/Processed/TMAX1-7anom.csv"))
-
-climatedata <- plots  
-for(j in 1:nrow(climatedata)){
-  INVYR <- climatedata[j, "INVYR"]
-  end.yr <- INVYR 
-  start.yr <- INVYR-9
-
-  spei2.ts <- ts(t(spei2.df[j,-1*ncol(spei2.df)]),start=1895, freq=1)
-  spei2.tsw <- window(spei2.ts, end=end.yr, start=start.yr)
-  climatedata[j, "JJSPEI"] <- mean(spei2.tsw , na.rm=T)
-
-  spei60.ts <- ts(t(spei60.df[j,-1*ncol(spei60.df)]),start=1895, freq=1)
-  spei60.tsw <- window(spei60.ts, end=end.yr, start=start.yr)
-  climatedata[j, "WYSPEI"] <- mean(spei60.tsw , na.rm=T)
-
- tmin1.12.ts <- ts(t(tmin1.12.df[j,-1*ncol(tmin1.12.df )]),start=1981, freq=1)
-  tmin1.12.tsw <- window(tmin1.12.ts, end=end.yr, start=start.yr)
-  climatedata[j, "DecTMIN"] <- mean(tmin1.12.tsw[,1], na.rm=T)
-
-  tmin1.12a.ts <- ts(t(tmin1.12.anom.df[j,-1*ncol(tmin1.12.anom.df )]),start=1981, freq=1)
-  tmin1.12a.tsw <- window(tmin1.12a.ts, end=end.yr, start=start.yr)
-  climatedata[j, "DecTMINa"] <- mean(tmin1.12a.tsw[,1], na.rm=T)
-
-   tmax1.7.ts <- ts(t(tmax1.7.df[j,-1*ncol(tmax1.7.df )]),start=1981, freq=1)
-  tmax1.7.tsw <- window(tmax1.7.ts, end=end.yr, start=start.yr)
-  climatedata[j, "JulTMAX"] <- mean(tmax1.7.tsw[,1], na.rm=T)
-
-tmax1.7a.ts <- ts(t(tmax1.7.anom.df[j,-1*ncol(tmax1.7.anom.df )]),start=1981, freq=1)
-
-  tmax1.7a.tsw <- window(tmax1.7a.ts, end=end.yr, start=start.yr)
-  climatedata[j, "JulTMAXa"] <- mean(tmax1.7a.tsw[,1], na.rm=T)
-
-  ppt.ts <- ts(t(ppt.df [j,-1*ncol(ppt.df )]),start=1981, freq=1)
-  ppt.tsw <- window( ppt.ts, end=end.yr, start=start.yr)
-  climatedata[j, "PPT"] <- mean(ppt.tsw[,1], na.rm=T)
-
-  ppta.ts <- ts(t(pptanom.df [j,-1*ncol(pptanom.df )]),start=1981, freq=1)
- ppta.tsw <- window(ppta.ts, end=end.yr, start=start.yr)
-  climatedata[j, "PPTa"] <- mean(ppta.tsw[,1], na.rm=T)
-}
-
-write.csv(climatedata, here("Data/Processed/climatedata.csv"), row.names = F)
-```
-
 ```{r merge data, eval=F}
 df.recent <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
-pressure <- read.csv(here("Data/Processed/FIAdat-pressure.csv"))
-climatedata <- read.csv(here("Data/Processed/climatedata.csv"))
-
-df <- merge(df.recent, pressure, by="PLT_CN", all=T)
-df <- merge(df, climatedata[, -2:-4], by="PLT_CN", all=T)
-
 write.csv(df, here("Data/Processed/FIAdat-df-merge.csv"), row.names=F)
@@
   df.p$TPHD.host <- df.p$TPH.host/df.p$TPH.total
-  df.p$pressure.BB <- df.p[,paste0("pressure", agent)]
+  #df.p$pressure.BB <- df.p[,paste0("pressure", agent)]
 
@@
 
-# Descriptive analyses
+### Descriptive analyses
 ```{r import processed data}
@@
 
-### Model occurrence
-
-```{r select plots to control for sampling effort, eval=T}
-#df.sub <- df[df$nplots>5, ]
-df.sub <- df
-## remove plots with other hosts
-df.sub  <- df.sub [df.sub$BA.Otherhosts==0,]
-#nrow(datBB.c)
-df.sub  <- df.sub [df.sub$BA.Allhosts/df.sub$BA.total>0.75,]
-```
-
-```{r select plots with large & abundant, eval=T}
-PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.10)
-PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.10)
-ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.10)
-
-PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.10)
-PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.10)
-ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.10)
-n
-# PICO, PIEN & ABLA
-df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PICO.PIEN.ABLA$focalhosts2 <- "PICO, PIEN & ABLA"
-df.sub.PICO.PIEN.ABLA$nhosts2 <- 3
-
-# PICO & PIEN
-df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]
-
-df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]
-
-df.sub.PICO.PIEN$focalhosts2 <- "PICO & PIEN"
-df.sub.PICO.PIEN < - df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
-df.sub.PICO.PIEN$nhosts2 <- 2
-
-# PICO & ABLA
-df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PICO.ABLA$focalhosts2 <- "PICO & ABLA"
-df.sub.PICO.ABLA$nhosts2 <- 2
-
-df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
-
-# PIEN & ABLA
-df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PIEN.ABLA$focalhosts2 <- "PIEN & ABLA"
-df.sub.PIEN.ABLA$nhosts2 <- 2
-
-df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
-
-###
-df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
-df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
-df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)
-
-## Affected hosts
-affectedhosts2 <- df.focal2 %>% group_by(focalhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectedhosts2[affectedhosts2$affected=="affected",]
-affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts2[affectednhosts2$affected=="affected",]
-### MPB
-
-MPBnhosts2 <- df.focal2 %>% group_by(nhosts2, agentMPB) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-
-### PIEN
-
-### ABLA
-
-```
-
-```{r, eval=T}
-datBB.c <- df.sub[df.sub$focalhost=="Any",]
-splitit <- initial_split(df.sub, prop=0.5)
+```{r}
+Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
+df$hostsFPshort<- df$hostsFPlong
+df$hostsFPshort <- factor(df$hostsFPlong, levels=levels(df$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+df$agentsFplongrev <- df$agentsFplong
+df$agentsFplongrev <- factor(df$agentsFplongrev, levels=rev(levels( df$agentsFplong)))
+jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
+ggplot(df[df$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
+whatever <- dev.off()
+```
+
+### Given suitable stand conditions, is co-occurrence common?
+
+Construct random forest models using synthetic minority oversampling technique (SMOTE)
+
+#### MPB
+
+```{r fit MPB RF model, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response)
 datz.train <- training(splitit)
 datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, v=10)
-```
-
-```{r fit MPB RF model, eval=T}
-response <- "agentMPB"
-#vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO", "pressureMPB", "hostBA.MPB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
-vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
-
-form <- paste(response, "~", paste(vars, collapse = " + "))
-
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
+
+# set sample weights
 wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
@@
 
-# set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
-step_corr(all_predictors())%>% prep()
-
 # set model
-rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
 
@@
 # tune model
-tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
 
@@
 # Use backward selection to identify important variables
-#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
 #vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
 vars.final <- vars
-
 BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
 
-rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
 
 rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
-
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here("Results/RF-testperformance-MPB-occurrence.csv"), row.names=F)
-
-# generate predictions from the training dataset
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+
+# generate a confusion matrix for test set
 test_predictions <- rf_workflow_final_fit %>% collect_predictions()
-write.csv(test_predictions , here("Results/RF-pred-MPB-occurrence.csv"), row.names=F)
-
-# generate a confusion matrix
 confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
-write.csv(confusion$table, here("Results/RF-confusiontable-MPB-occurrence.csv"), row.names=F)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
 
@@
 
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+
 x <- as.data.frame(ranger_obj$variable.importance)
-x$var <- row.names(x)
-x$prop <- x[,1]/sum(x[,1])*100
-x <- x[order(x$prop),]
-x$var <- factor(x$var, levels=x$var, ordered=T)
-write.csv(x, here("Results/RF-vimp-MPB-occurrence.csv"), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
 
@@
 for(z in vars.final){
-  pdz <- partial(ranger_obj, train=datz.train, pred.var=z)
-  pdz$var <- z
-  colnames(pdz) <- c("value","yhat","var", which.class="MPB")
-  pd <- rbind(pd, pdz)
-}
-write.csv(pd,  here("Results/RF-partialdependence-MPB-occurrence.csv"), row.names=F)
-```
-
-```{r fit SB RF model, eval=T}
-response <- "agentSB"
-#vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN", "pressureSB", "hostBA.SB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
-vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
-
-form <- paste(response, "~", paste(vars, collapse = " + "))
-
-wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
-wts <- wts/sum(wts)
-
-# set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%step_corr(all_predictors())%>% prep()
-
-# set model
-rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")
-
-# set the workflow
-rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
-
-# tune model
-tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))
-
-tuneres <- rf_tune_results %>% collect_metrics()
-
-param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
-
-# Use backward selection to identify important variables
-dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
-registerDoSEQ()
-vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
-
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
-
-rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")
-
-rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
-
-test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here("Results/RF-testperformance-SB-occurrence.csv"), row.names=F)
-
-# generate predictions from the training dataset
-test_predictions <- rf_workflow_final_fit %>% collect_predictions()
-write.csv(test_predictions , here("Results/RF-pred-SB-occurrence.csv"), row.names=F)
-
-# generate a confusion matrix
-confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
-write.csv(confusion$table, here("Results/RF-confusiontable-SB-occurrence.csv"), row.names=F)
-
-# variable importance
-final_model <- fit(rf_workflow_final, data=datz.train)
-ranger_obj <- pull_workflow_fit(final_model)$fit
-
-x <- as.data.frame(ranger_obj$variable.importance)
-x$var <- row.names(x)
-x$prop <- x[,1]/sum(x[,1])*100
-x <- x[order(x$prop),]
-x$var <- factor(x$var, levels=x$var, ordered=T)
-write.csv(x, here("Results/RF-vimp-SB-occurrence.csv"), row.names=F)
-
-# partial dependence
-pd <- NULL
-for(z in vars.final){
-  pdz <- partial(ranger_obj, train=datz.train, pred.var=z)
-  pdz$var <- z
-  colnames(pdz) <- c("value","yhat","var", which.class="SB")
-  pd <- rbind(pd, pdz)
-}
-write.csv(pd,  here("Results/RF-partialdependence-SB-occurrence.csv"), row.names=F)
-```
-
-```{r fit WBBB RF model, eval=T}
-response <- "agentWBBB"
-vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA", "pressureWBBB", "hostBA.WBBB", "PPTa", "JJSPEI", "WYSPEI", "JulTMAXa", "DecTMINa")
-vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
-
-form <- paste(response, "~", paste(vars, collapse = " + "))
-
-wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
-wts <- wts/sum(wts)
-
-# set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
-step_corr(all_predictors())%>% prep()
-
-# set model
-rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), class.weights=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")
-
-# set the workflow
-rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
-
-# tune model
-tgrid <- expand.grid(mtry = 3, trees = c(100,500), min_n=c(50,100,500))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))
-
-tuneres <- rf_tune_results %>% collect_metrics()
-
-param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
-
-# Use backward selection to identify important variables
-dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
-registerDoSEQ()
-vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
-
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
-
-rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  class.weights=wts,  importance = "permutation", probability=T) %>% set_mode("classification")
-
-rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
-
-test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here("Results/RF-testperformance-WBBB-occurrence.csv"), row.names=F)
-
-# generate predictions from the training dataset
-test_predictions <- rf_workflow_final_fit %>% collect_predictions()
-write.csv(test_predictions , here("Results/RF-pred-WBBB-occurrence.csv"), row.names=F)
-
-# generate a confusion matrix
-confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
-write.csv(confusion$table, here("Results/RF-confusiontable-WBBB-occurrence.csv"), row.names=F)
-
-# variable importance
-final_model <- fit(rf_workflow_final, data=datz.train)
-ranger_obj <- pull_workflow_fit(final_model)$fit
-
-x <- as.data.frame(ranger_obj$variable.importance)
-x$var <- row.names(x)
-x$prop <- x[,1]/sum(x[,1])*100
-x <- x[order(x$prop),]
-x$var <- factor(x$var, levels=x$var, ordered=T)
-write.csv(x, here("Results/RF-vimp-WBBB-occurrence.csv"), row.names=F)
-
-# partial dependence
-pd <- NULL
-for(z in vars.final){
-  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class="WBBB")
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
   pdz$var <- z
@@
 }
-write.csv(pd,  here("Results/RF-partialdependence-WBBB-occurrence.csv"), row.names=F)
-```
-
-```{r, eval=F}
-mpbo <- read.csv(here("Results/RF-pred-MPB-occurrence.csv"))
-sbo <- read.csv(here("Results/RF-pred-SB-occurrence.csv"))
-wbbbo <- read.csv(here("Results/RF-pred-WBBB-occurrence.csv"))
-
-bbo <- datz.test
-bbo$mpbpred <- mpbo$.pred_MPB
-bbo$sbpred <- sbo$.pred_SB 
-bbo$wbbbpred <- wbbbo$.pred_WBBB
-bbo$mpba<- mpbo$agentMPB
-bbo$sba<- sbo$agentSB
-bbo$wbbba<- wbbbo$agentWBBB
-
-bbo.sub <- bbo[bbo$mpbpred>0.25 &bbo$wbbbpred>0.25 & bbo$sbpred>0.25,   ]
-nrow(bbo.sub)
-bbo$actualSUM <- ifelse(bbo$mpba=="absent",0, 1) + ifelse(bbo$sba=="absent",0, 1) + ifelse(bbo$wbbba=="absent",0, 1)
-
-bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
-bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)
-
-bbo$pres <- factor(ifelse(bbo$actualSUM >0, 1, 0), "affected", "unaffected")
-
-ggplot(bbo, aes(x=predSUM, y=actualSUM))+geom_boxplot()
-
-bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
-bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)
-bbo
-
-bbo$rMPB <- ifelse(bbo$agentMPB=="MPB", 1,0) - bbo$mpbpred 
-bbo$rSB <- ifelse(bbo$agentSB=="SB", 1,0) - bbo$sbpred 
-bbo$rWBBB <- ifelse(bbo$agentWBBB=="WBBB", 1,0) - bbo$wbbbpred
-bbo$rSUM <- bbo$rMPB + bbo$rSB + bbo$rWBBB
-
-x <- read.csv(here("Results/RF-partialdependence-SB-occurrence.csv"))
-```
-
-#### Severity
-##### MPB
-##### SB
-##### WBBB
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+```
+
+#### SB
+```{r fit SB RF model, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response)
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics()
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+```
+
+#### WBBB
+```{r fit WBBB RF model, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response)
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics()
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+```
+
+```{r}
+MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
+MPBpp$focalagent <- "MPB"
+SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
+SBpp$focalagent <- "SB"
+WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
+WBBBpp$focalagent <- "WBBB"
+
+ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
+ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
+ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
+ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
+
+x <- df[!df$focalagent=="Any",]
+x$propBA <- x$BA.host/x$BA.total
+x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
+x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
+
+xm <- melt(x, id.vars=c("focalagent", "affected01"))
+colnames(xm) <- c("focalagent", "affected01", "var", "value")
+xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
+
+pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
+
+pBA1 + pBA2 +plot_layout(ncol=1)
+
+pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
+
+ppropBA1 + ppropBA2 +plot_layout(ncol=1)
+
+ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("propBA")
+
+Fig.file <- here("Results/Figures/PartialDependence.jpg")
+jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
+pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
+whatever <- dev.off()
+```
+
+```{r}
+MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence.csv"))
+SBpred <- read.csv(here("Results/RF-pred-agentSB-occurrence.csv"))
+WBBBpred <- read.csv(here("Results/RF-pred-agentWBBB-occurrence.csv"))
+
+BBpred <- merge(merge(MPBpred, SBpred, by="PLT_CN"), WBBBpred, by="PLT_CN")
+colnames(BBpred) <- c("PLT_CN", "MPBpred", "SBpred", "WBBBpred")
+
+threshold <- 0.25
+BBpred$MPBpred01 <- ifelse(BBpred$MPBpred>threshold, "MPB", "absent")
+BBpred$SBpred01 <- ifelse(BBpred$SBpred>threshold, "SB", "absent")
+BBpred$WBBBpred01 <- ifelse(BBpred$WBBBpred>threshold, "WBBB", "absent")
+
+BBpred$agentspred <-  do.call(paste, c(BBpred[c("SBpred01", "MPBpred01", "WBBBpred01")],sep="-"))
+BBpred$agentspred01 <- factor(BBpred$agentspred, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))
+
+dfpred <- merge(BBpred, df[df$focalagent=="Any",], id.vars="PLT_CN", all=T)
+```
+
+```{r}
+mpb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="MPB","affected"], bbo.dat[bbo.dat$focalagent=="MPB","hostspred"]))
+sb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="SB","affected"], bbo.dat[bbo.dat$focalagent=="SB","hostspred"]))
+wbbb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="WBBB","affected"], bbo.dat[bbo.dat$focalagent=="WBBB","hostspred"]))
+any<- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="Any","affected"], bbo.dat[bbo.dat$focalagent=="Any","hostspred"]))
+
+results <- NULL
+for(j in levels(bbo.dat$hostspred)[5:8]){
+  anyj <- any[any$Var2==j,]
+  mpbj <- mpb[mpb$Var2==j,]
+  sbj <- sb[sb$Var2==j,]
+  wbbbj <- wbbb[wbbb$Var2==j,]
+
+  x <- prop.test(x=c(mpbj[1,3],anyj[1,3]), n=c(mpbj[1,3]+mpbj[2,3], anyj[1,3]+anyj[2,3]))
+
+  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$hostspred <-j
+  res$barkbeetle <- "MPB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  res$prop1 <- x$estimate[1]
+  res$prop2 <- x$estimate[2]
+  res$lower <- x$conf.in[1]
+  res$upper <- x$conf.in[2]
+  x <-NULL
+  results <- rbind(results, res)
+
+    x <- prop.test(x=c(sbj[1,3],anyj[1,3]), n=c(sbj[1,3]+sbj[2,3], anyj[1,3]+anyj[2,3]))
+
+  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$hostspred <-j
+  res$barkbeetle <- "SB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  res$prop1 <- x$estimate[1]
+  res$prop2 <- x$estimate[2]
+  res$lower <- x$conf.in[1]
+  res$upper <- x$conf.in[2]
+  x <-NULL
+  results <- rbind(results, res)
+
+   x <- prop.test(x=c(wbbbj[1,3],anyj[1,3]), n=c(wbbbj[1,3]+wbbbj[2,3], anyj[1,3]+anyj[2,3]), alternative="less")
+
+  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$hostspred <-j
+  res$barkbeetle <- "WBBB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  res$prop1 <- x$estimate[1]
+  res$prop2 <- x$estimate[2]
+  res$lower <- x$conf.in[1]
+  res$upper <- x$conf.in[2]
+  x <-NULL
+  results <- rbind(results, res)
+}
+
+results$p <- round(results$p,3)
+
+x <- bbo.dat[is.na(bbo.dat$hostspred)==FALSE,]
+x <- x[x$hostspred %in% levels(x$hostspred)[5:8],]
+x[x$hostspred == "PICO\n& PIEN" & x$focalagent=="WBBB","affected"]<- 0
+x[x$hostspred == "PICO\n& ABLA" & x$focalagent=="SB","affected"]<- 0
+x[x$hostspred == "PIEN\n& ABLA" & x$focalagent=="MPB","affected"]<- 0
+
+xx <- x %>% group_by(hostspred,focalagent) %>% count(affected)
+xxx <- x %>% group_by(hostspred) %>% count()
+
+xxxx <- merge(xx, xxx, by="hostspred")
+xxxx$prop <- xxxx$n.x/xxxx$n.y
+xxxx <- xxxx[xxxx$affected=="affected",]
+xxxx <- xxxx[,c(1,2,6)]
+colnames(xxxx) <- c("hostspred", "barkbeetle", "prop")
+
+resultsx <- merge(results, xxxx, id.vars=c("hostspred", "barkbeetle"), all=T)
+resultsx$barkbeetle <- factor(resultsx$barkbeetle, levels=c("MPB", 'SB', "WBBB", "Any"), ordered=T)
+resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
+resultsx$hostspred <- factor(resultsx$hostspred, levels=levels(bbo.dat$hostspred)[5:8], ordered=T)
+resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
+resultsx$pplot[is.na(resultsx$pplot) ==TRUE] <- ""
+
+Fig.file <- here("Results/Figures/Figurey.jpg")
+jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
+ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~hostspred, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")
+whatever <- dev.off()
+```
+
+```{r}
+Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
+bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
+bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
+bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
+#jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
+ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
+#whatever <- dev.off()
+```
+
+### Is occurrence more likely in stands with multiple hosts?
+
+### Is severity greater in stands with multiple hosts?
+
+To determine if host diversity or identity influences the occurrence tree mortality attributed to bark beetles within a stand, we used simple Chi-square tests.
+Given a significant difference among groups, we then used pairwise Fisher's exact tests to determine pairwise differences, while accounting for multiple comparisons [@Benjamini_controlling_1995].
 
@@
 ## Occurrence 
-
-```{r FigPropAgentxHosts, fig.cap = "A caption for figure 1", out.width = "504pt"}
+```{r FigPropAgentxHosts, fig.cap = "A caption for figure 1", out.width = "504pt", eval=F}
 knitr::include_graphics( here("Results/Figures/Figure-PropAffectedByAgentXHosts.jpg"))
@@
 ```
-
-Stands with Engelmann spruce present occur at more southerly latitudes than stands with
 

### 2021-09-13 Code/Manuscript.Rmd
updating

--- 2021-07-29 Code/Manuscript.Rmd
@@
 }
-write.csv(pd,  here("Results/RF-partialdependence-WBBB-occurrence.csv"), row.names=F)
-```
-
-```{r, eval=F}
-mpbo <- read.csv(here("Results/RF-pred-MPB-occurrence.csv"))
-sbo <- read.csv(here("Results/RF-pred-SB-occurrence.csv"))
-wbbbo <- read.csv(here("Results/RF-pred-WBBB-occurrence.csv"))
-
-bbo <- datz.test
-bbo$mpbpred <- mpbo$.pred_MPB
-bbo$sbpred <- sbo$.pred_SB 
-bbo$wbbbpred <- wbbbo$.pred_WBBB
-bbo$mpba<- mpbo$agentMPB
-bbo$sba<- sbo$agentSB
-bbo$wbbba<- wbbbo$agentWBBB
-
-bbo.sub <- bbo[bbo$mpbpred>0.25 &bbo$wbbbpred>0.25 & bbo$sbpred>0.25,   ]
-nrow(bbo.sub)
-bbo$actualSUM <- ifelse(bbo$mpba=="absent",0, 1) + ifelse(bbo$sba=="absent",0, 1) + ifelse(bbo$wbbba=="absent",0, 1)
-
-bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
-bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)
-
-bbo$pres <- factor(ifelse(bbo$actualSUM >0, 1, 0), "affected", "unaffected")
-
-ggplot(bbo, aes(x=predSUM, y=actualSUM))+geom_boxplot()
-
-bbo$predSUM <- bbo$mpbpred + bbo$sbpred +bbo$wbbbpred
-bbo$predSUM01 <- ifelse(bbo$predSUM >0.5, 1, 0)
-bbo
-
-bbo$rMPB <- ifelse(bbo$agentMPB=="MPB", 1,0) - bbo$mpbpred 
-bbo$rSB <- ifelse(bbo$agentSB=="SB", 1,0) - bbo$sbpred 
-bbo$rWBBB <- ifelse(bbo$agentWBBB=="WBBB", 1,0) - bbo$wbbbpred
-bbo$rSUM <- bbo$rMPB + bbo$rSB + bbo$rWBBB
-
-x <- read.csv(here("Results/RF-partialdependence-SB-occurrence.csv"))
-```
-
-#### Severity
-##### MPB
-##### SB
-##### WBBB
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+```
+
+#### SB
+```{r fit SB RF model, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response)
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics()
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+```
+
+#### WBBB
+```{r fit WBBB RF model, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response)
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics()
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+```
+
+```{r}
+MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
+MPBpp$focalagent <- "MPB"
+SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
+SBpp$focalagent <- "SB"
+WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
+WBBBpp$focalagent <- "WBBB"
+
+ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
+ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
+ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
+ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
+
+x <- df[!df$focalagent=="Any",]
+x <- x[,c("QMD.host", "propTBA", "BA.host", "focalagent", "affected01")]
+x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
+
+xm <- melt(x, id.vars=c("focalagent", "affected01"))
+colnames(xm) <- c("focalagent", "yhat", "var", "value")
+xm$var <- factor(xm$var, levels=c("QMD.host", "propTBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
+xm$focalagent <- as.character(xm$focalagent)
+p <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+geom_point(data=xm[xm$var=="BA",], size=0, )+ylim(0.1,0.6)
+ggplot(xm[xm$var=="BA",],aes(x=value, y=yhat, col=focalagent)
+
+ggplot(xm, aes(x=focalagent, y=value))
+```
+
+```{r}
+
+mpb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="MPB","affected"], bbo.dat[bbo.dat$focalagent=="MPB","hostspred"]))
+sb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="SB","affected"], bbo.dat[bbo.dat$focalagent=="SB","hostspred"]))
+wbbb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="WBBB","affected"], bbo.dat[bbo.dat$focalagent=="WBBB","hostspred"]))
+any<- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="Any","affected"], bbo.dat[bbo.dat$focalagent=="Any","hostspred"]))
+
+results <- NULL
+for(j in levels(bbo.dat$hostspred)[5:8]){
+  anyj <- any[any$Var2==j,]
+  mpbj <- mpb[mpb$Var2==j,]
+  sbj <- sb[sb$Var2==j,]
+  wbbbj <- wbbb[wbbb$Var2==j,]
+
+  x <- prop.test(x=c(mpbj[1,3],anyj[1,3]), n=c(mpbj[1,3]+mpbj[2,3], anyj[1,3]+anyj[2,3]))
+
+  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$hostspred <-j
+  res$barkbeetle <- "MPB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  res$prop1 <- x$estimate[1]
+  res$prop2 <- x$estimate[2]
+  res$lower <- x$conf.in[1]
+  res$upper <- x$conf.in[2]
+  x <-NULL
+  results <- rbind(results, res)
+
+    x <- prop.test(x=c(sbj[1,3],anyj[1,3]), n=c(sbj[1,3]+sbj[2,3], anyj[1,3]+anyj[2,3]))
+
+  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$hostspred <-j
+  res$barkbeetle <- "SB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  res$prop1 <- x$estimate[1]
+  res$prop2 <- x$estimate[2]
+  res$lower <- x$conf.in[1]
+  res$upper <- x$conf.in[2]
+  x <-NULL
+  results <- rbind(results, res)
+
+   x <- prop.test(x=c(wbbbj[1,3],anyj[1,3]), n=c(wbbbj[1,3]+wbbbj[2,3], anyj[1,3]+anyj[2,3]), alternative="less")
+
+  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$hostspred <-j
+  res$barkbeetle <- "WBBB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  res$prop1 <- x$estimate[1]
+  res$prop2 <- x$estimate[2]
+  res$lower <- x$conf.in[1]
+  res$upper <- x$conf.in[2]
+  x <-NULL
+  results <- rbind(results, res)
+}
+
+results$p <- round(results$p,3)
+
+x <- bbo.dat[is.na(bbo.dat$hostspred)==FALSE,]
+x <- x[x$hostspred %in% levels(x$hostspred)[5:8],]
+x[x$hostspred == "PICO\n& PIEN" & x$focalagent=="WBBB","affected"]<- 0
+x[x$hostspred == "PICO\n& ABLA" & x$focalagent=="SB","affected"]<- 0
+x[x$hostspred == "PIEN\n& ABLA" & x$focalagent=="MPB","affected"]<- 0
+
+xx <- x %>% group_by(hostspred,focalagent) %>% count(affected)
+xxx <- x %>% group_by(hostspred) %>% count()
+
+xxxx <- merge(xx, xxx, by="hostspred")
+xxxx$prop <- xxxx$n.x/xxxx$n.y
+xxxx <- xxxx[xxxx$affected=="affected",]
+xxxx <- xxxx[,c(1,2,6)]
+colnames(xxxx) <- c("hostspred", "barkbeetle", "prop")
+
+resultsx <- merge(results, xxxx, id.vars=c("hostspred", "barkbeetle"), all=T)
+resultsx$barkbeetle <- factor(resultsx$barkbeetle, levels=c("MPB", 'SB', "WBBB", "Any"), ordered=T)
+resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
+resultsx$hostspred <- factor(resultsx$hostspred, levels=levels(bbo.dat$hostspred)[5:8], ordered=T)
+resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
+resultsx$pplot[is.na(resultsx$pplot) ==TRUE] <- ""
+
+Fig.file <- here("Results/Figures/Figurey.jpg")
+jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
+ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~hostspred, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")
+whatever <- dev.off()
+```
+
+```{r}
+Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
+bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
+bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
+bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
+#jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
+ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
+#whatever <- dev.off()
+```
+
+### Is occurrence more likely in stands with multiple hosts?
+
+### Is severity greater in stands with multiple hosts?
+
+To determine if host diversity or identity influences the occurrence tree mortality attributed to bark beetles within a stand, we used simple Chi-square tests.
+Given a significant difference among groups, we then used pairwise Fisher's exact tests to determine pairwise differences, while accounting for multiple comparisons [@Benjamini_controlling_1995].
 
@@
 
-```{r FigPropAgentxHosts, fig.cap = "A caption for figure 1", out.width = "504pt"}
+```{r FigPropAgentxHosts, fig.cap = "A caption for figure 1", out.width = "504pt", eval=F}
 knitr::include_graphics( here("Results/Figures/Figure-PropAffectedByAgentXHosts.jpg"))

### 2021-09-13 Code/Manuscript.Rmd
updating

@@
 ### Import libraries
-ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach'))
+ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'foreach'))
 
@@
 
-### Given suitable stand conditions, is co-occurrence common?
-
-Construct random forest models using synthetic minority oversampling technique (SMOTE)
+### Under what stand conditions are bark beetle outbreaks likely?
+```{r, eval=T}
+datBB.c <- df[df$focalhost=="Any",]
+splitit <- initial_split(datBB.c, prop=0.5)
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, v=10)
+```
 
 #### MPB
-
 ```{r fit MPB RF model, eval=F}
@@
 vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
-datBB.c <- df[df$focalhost=="Any",]
-
-splitit <- initial_split(datBB.c, prop=.7, strata=response)
-datz.train <- training(splitit)
-datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
-
-# set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
-
-# set sample weights
+
+form <- paste(response, "~", paste(vars, collapse = " + "))
+
 wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
@@
 
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%step_corr(all_predictors())%>% prep()
+
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), sample.fraction=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")
 
@@
 # tune model
-tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
+tgrid <- expand.grid(mtry = c(1,3), trees = c(100,500), min_n=c(50,100,500))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))
 
@@
 # Use backward selection to identify important variables
-#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
-#registerDoSEQ()
-#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
-vars.final <- vars
+dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
+registerDoSEQ()
+vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+
 BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
 
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  sample.fraction=wts,  importance = "permutation", probability=T) %>% set_mode("classification")
 
 rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(test_performance, here("Results/RF-testperformance-MPB-occurrence.csv"), row.names=F)
 
@@
 confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
-
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(confusion$table, here("Results/RF-confusiontable-MPB-occurrence.csv"), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "MPBpred")
+write.csv(x, here("Results/RF-pred-MPB-occurrence.csv"), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+x$var <- row.names(x)
+x$prop <- x[,1]/sum(x[,1])*100
+x <- x[order(x$prop),]
+x$var <- factor(x$var, levels=x$var, ordered=T)
+write.csv(x, here("Results/RF-vimp-MPB-occurrence.csv"), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class="MPB")
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here("Results/RF-partialdependence-MPB-occurrence.csv"), row.names=F)
+
+pdz <- partial(ranger_obj, train=datz.train, pred.var=c("QMD.PICO", "BA.PICO"), which.class="MPB")
+
+```
+
+#### SB
+```{r fit SB RF model, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+
+form <- paste(response, "~", paste(vars, collapse = " + "))
+
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%step_corr(all_predictors())%>% prep()
+
+# set model
+rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), sample.fraction=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(mtry = c(1,3), trees = c(100,500), min_n=c(50,100,500))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set( accuracy))
+
+tuneres <- rf_tune_results %>% collect_metrics()
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
+
+# Use backward selection to identify important variables
+dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=param_best$mtry, ntree=param_best$trees, min.node.size=param_best$min_n)
+registerDoSEQ()
+vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger",  sample.fraction=wts,  importance = "permutation", probability=T) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+write.csv(confusion$table, here("Results/RF-confusiontable-SB-occurrence.csv"), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here("Results/RF-pred-SB-occurrence.csv"), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+x$var <- row.names(x)
+x$prop <- x[,1]/sum(x[,1])*100
+x <- x[order(x$prop),]
+x$var <- factor(x$var, levels=x$var, ordered=T)
+write.csv(x, here("Results/RF-vimp-SB-occurrence.csv"), row.names=F)
 
@@
 for(z in vars.final){
-  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class="SB")
   pdz$var <- z
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
-```
-
-#### SB
-```{r fit SB RF model, eval=F}
-response <- "agentSB"
-vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
-datBB.c <- df[df$focalhost=="Any",]
-
-splitit <- initial_split(datBB.c, prop=.7, strata=response)
-datz.train <- training(splitit)
-datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
-
-# set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
-
-# set sample weights
+write.csv(pd,  here("Results/RF-partialdependence-SB-occurrence.csv"), row.names=F)
+```
+
+#### WBBB
+```{r fit WBBB RF model, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+
+form <- paste(response, "~", paste(vars, collapse = " + "))
+
 wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
@@
 
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>%
+step_corr(all_predictors())%>% prep()
+
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args(mtry = tune(), trees=tune(), min_n=tune(), sample.fraction=wts) %>% set_engine("ranger", importance = "permutation", probability=T) %>% set_mode("classification")
 
@@
 # tune model
-tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
+tgrid <- expand.grid(mtry = c(1,3), trees = c(100,500), min_n=c(30,100,500))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy))
 
@@
 # Use backward selection to identify important variables
-#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
-#registerDoSEQ()
-#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
-vars.final <- vars
+dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, mtry=max(param_best$mtry,length(vars.final)), ntree=param_best$trees, min.node.size=param_best$min_n)
+registerDoSEQ()
+vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+
 BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
 
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+#BBoccurrence_recipe_final <-  recipe(formula("agentWBBB~1"), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(mtry = param_best$mtry, trees=param_best$trees, min_n=param_best$min_n, splitrule="gini", probability=T) %>% set_engine("ranger", sample.fraction=wts,  importance = "permutation", probability=T) %>% set_mode("classification")
 
 rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
-test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
-
-# generate a confusion matrix for test set
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit)
+#emulates the process where, after determining the best model, the #final fit on the entire training set is needed and is then evaluated #on the test set.
+
+## generate a confusion matrix for test set
 test_predictions <- rf_workflow_final_fit %>% collect_predictions()
 confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
-
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
-
-# variable importance
+write.csv(confusion$table, here("Results/RF-confusiontable-WBBB-occurrence.csv"), row.names=F)
+
+## variable importance
 final_model <- fit(rf_workflow_final, data=datz.train)
@@
 
-# generate predictions for entire dataset
+## generate predictions for entire dataset
 x <- datBB.c$PLT_CN
 x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
-colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+colnames(x) <- c("PLT_CN", "WBBBpred")
+write.csv(x, here("Results/RF-pred-WBBB-occurrence.csv"), row.names=F)
+
+## variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
-
-# partial dependence
+x$var <- row.names(x)
+x$prop <- x[,1]/sum(x[,1])*100
+x <- x[order(x$prop),]
+x$var <- factor(x$var, levels=x$var, ordered=T)
+write.csv(x, here("Results/RF-vimp-WBBB-occurrence.csv"), row.names=F)
+
+## partial dependence
 pd <- NULL
 for(z in vars.final){
-  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class="WBBB")
   pdz$var <- z
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
-```
-
-#### WBBB
-```{r fit WBBB RF model, eval=F}
-response <- "agentWBBB"
-vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
-datBB.c <- df[df$focalhost=="Any",]
-
-splitit <- initial_split(datBB.c, prop=.7, strata=response)
-datz.train <- training(splitit)
-datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
-
-# set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
-
-# set sample weights
-wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
-wts <- wts/sum(wts)
-
-# set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
-
-# set the workflow
-rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
-
-# tune model
-tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
-
-tuneres <- rf_tune_results %>% collect_metrics()
-
-param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
-
-# Use backward selection to identify important variables
-#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
-#registerDoSEQ()
-#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
-vars.final <- vars
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
-
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
-
-rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
-test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
-
-# generate a confusion matrix for test set
-test_predictions <- rf_workflow_final_fit %>% collect_predictions()
-confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
-
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
-
-# variable importance
-final_model <- fit(rf_workflow_final, data=datz.train)
-ranger_obj <- pull_workflow_fit(final_model)$fit
-
-# generate predictions for entire dataset
-x <- datBB.c$PLT_CN
-x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
-colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
-
-x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
-
-# partial dependence
-pd <- NULL
-for(z in vars.final){
-  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
-  pdz$var <- z
-  colnames(pdz) <- c("value","yhat","var")
-  pd <- rbind(pd, pdz)
-}
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(pd,  here("Results/RF-partialdependence-WBBB-occurrence.csv"), row.names=F)
 ```
@@
 ```{r}
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x <- x[,c("QMD.host", "propTBA", "BA.host", "focalagent", "affected01")]
-x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "yhat", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propTBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-xm$focalagent <- as.character(xm$focalagent)
-p <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+geom_point(data=xm[xm$var=="BA",], size=0, )+ylim(0.1,0.6)
-ggplot(xm[xm$var=="BA",],aes(x=value, y=yhat, col=focalagent)
-
-ggplot(xm, aes(x=focalagent, y=value))
-```
+wbbbp <- read.csv(here("Results/RF-partialdependence-WBBB-occurrence.csv"))
+
+sbp <- read.csv(here("Results/RF-partialdependence-SB-occurrence.csv"))
+
+mpbp <- read.csv(here("Results/RF-partialdependence-MPB-occurrence.csv"))
+
+```
+
+```{r, eval=F}
+mpbo <- read.csv(here("Results/RF-pred-MPB-occurrence.csv"))
+sbo <- read.csv(here("Results/RF-pred-SB-occurrence.csv"))
+wbbbo <- read.csv(here("Results/RF-pred-WBBB-occurrence.csv"))
+
+threshold <-0.5
+
+bbo <- merge(mpbo, sbo, by="PLT_CN")
+bbo <- merge(bbo, wbbbo, by="PLT_CN")
+
+bbo$MPBpred01 <- ifelse(bbo$MPBpred>threshold,"MPB", "")
+bbo$SBpred01  <- ifelse(bbo$SBpred>threshold,"SB", "")
+bbo$WBBBpred01  <- ifelse(bbo$WBBBpred>threshold,"WBBB", "")
+
+nrow(bbo[(bbo$MPBpred>threshold | bbo$SBpred> threshold | bbo$WBBBpred > threshold),])/nrow(bbo) # 97% of stands had conditions suitable for one agent
+
+nrow(bbo[bbo$MPBpred>threshold ,])/nrow(bbo) # 55% of stands had conditions suitable for MPB
+
+nrow(bbo[bbo$SBpred>threshold ,])/nrow(bbo) # 52% of stands had conditions suitable for SB
+
+nrow(bbo[bbo$WBBBpred>threshold ,])/nrow(bbo) # 55% of stands had conditions suitable for WBBB
+
+nrow(bbo[bbo$MPBpred>threshold & bbo$SBpred>threshold  ,])/nrow(bbo) # 55% of stands had conditions suitable for WBBB
+
+nrow(bbo[bbo$SBpred>threshold & bbo$WBBBpred>threshold  ,])/nrow(bbo) # 55% of stands had conditions suitable for WBBB
+
+nrow(bbo[bbo$SBpred>threshold & bbo$WBBBpred>threshold & bbo$MPBpred>threshold  ,])/nrow(bbo) # 55% of stands had conditions suitable for WBBB
+
+bbo$agentspred <- do.call(paste, c(bbo[c("MPBpred01", "SBpred01", "WBBBpred01")],sep="-"))
+bbo$agentspred <- factor(bbo$agentspred, level=c("--","MPB--","-SB-","--WBBB","MPB-SB-","MPB--WBBB","-SB-WBBB", "MPB-SB-WBBB"), labels=c("none", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"), ordered=T)
+
+bbo$hostspred <- factor(bbo$agentspred , levels=c("none", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"),  labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), ordered=T)
+
+bbo.sub <- bbo[!bbo$agentspred == "none",   ]
+
+bbo.dat <- merge(bbo.sub,df, by="PLT_CN", all=T)
+
+```
+
+### Given suitable stand conditions, is co-occurrence common?
 
 ```{r}
-
+Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
+bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
+bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
+bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
+#jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
+ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
+#whatever <- dev.off()
+```
+
+### Is occurrence more likely in stands with multiple hosts?
+
+### Is severity greater in stands with multiple hosts?
+
+```{r}
 mpb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="MPB","affected"], bbo.dat[bbo.dat$focalagent=="MPB","hostspred"]))
@@
 
-```{r}
-Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
-bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
-bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
-bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
-#jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
-ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
-#whatever <- dev.off()
-```
-
-### Is occurrence more likely in stands with multiple hosts?
-
-### Is severity greater in stands with multiple hosts?
-
 To determine if host diversity or identity influences the occurrence tree mortality attributed to bark beetles within a stand, we used simple Chi-square tests.

### 2021-09-14 Code/Manuscript-SMOTE.Rmd

@@
 	echo = F,
-	fig.align = "center",
 	message = FALSE,
@@
 
-```{r}
+```{r determine co-occurrence of hosts}
 Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
@@
 df$agentsFplongrev <- factor(df$agentsFplongrev, levels=rev(levels( df$agentsFplong)))
-jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
+jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in",res=300)
 ggplot(df[df$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
@@
 
-### Given suitable stand conditions, is co-occurrence common?
-
+### Determine suitable hosts
 Construct random forest models using synthetic minority oversampling technique (SMOTE)
@@
 #### MPB
-
 ```{r fit MPB RF model, eval=F}
@@
 
-```{r}
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
-x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
-
-pBA1 + pBA2 +plot_layout(ncol=1)
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
-
-ppropBA1 + ppropBA2 +plot_layout(ncol=1)
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("propBA")
-
-Fig.file <- here("Results/Figures/PartialDependence.jpg")
-jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
-pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-whatever <- dev.off()
-```
-
-```{r}
+```{r determine suitable stands}
 MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence.csv"))
@@
 
-threshold <- 0.25
-BBpred$MPBpred01 <- ifelse(BBpred$MPBpred>threshold, "MPB", "absent")
-BBpred$SBpred01 <- ifelse(BBpred$SBpred>threshold, "SB", "absent")
-BBpred$WBBBpred01 <- ifelse(BBpred$WBBBpred>threshold, "WBBB", "absent")
-
-BBpred$agentspred <-  do.call(paste, c(BBpred[c("SBpred01", "MPBpred01", "WBBBpred01")],sep="-"))
-BBpred$agentspred01 <- factor(BBpred$agentspred, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))
-
-dfpred <- merge(BBpred, df[df$focalagent=="Any",], id.vars="PLT_CN", all=T)
-```
-
-```{r}
-mpb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="MPB","affected"], bbo.dat[bbo.dat$focalagent=="MPB","hostspred"]))
-sb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="SB","affected"], bbo.dat[bbo.dat$focalagent=="SB","hostspred"]))
-wbbb <- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="WBBB","affected"], bbo.dat[bbo.dat$focalagent=="WBBB","hostspred"]))
-any<- as.data.frame(table(bbo.dat[bbo.dat$focalagent=="Any","affected"], bbo.dat[bbo.dat$focalagent=="Any","hostspred"]))
+threshold <-0.5
+BBpred$MPBsuitable <- ifelse(BBpred$MPBpred > threshold, "MPB", "absent")
+BBpred$SBsuitable <- ifelse(BBpred$SBpred > threshold, "SB", "absent")
+BBpred$WBBBsuitable <- ifelse(BBpred$WBBBpred >threshold, "WBBB", "absent")
+
+BBpred$agentsuitable <-  do.call(paste, c(BBpred[c("SBsuitable", "MPBsuitable", "WBBBsuitable")],sep="-"))
+BBpred$agentsuitable <- factor(BBpred$agentsuitable, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))
+
+BBpred$suitablehosts <- factor(BBpred$agentsuitable, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB",  "MPB & WBBB", "SB & WBBB",  "MPB, SB & WBBB"), labels=c("none", "PICO", "PIEN", "ABLA","PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
+
+```
+
+## Host co-occurrence
+
+### Given suitable stand conditions, is co-occurrence of suitable hosts common?
+```{r determine co-occurrence of suitable stands}
+bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
+bbo.dat <- bbo.dat[bbo.dat$focalagent=="Any",]
+cocount <-  as.data.frame(table(bbo.dat$agentsFplongrev))
+MPBco <-sum(cocount[cocount$Var1 %in% c("MPB & SB",  "MPB & WBBB","MPB, SB & WBBB"),]$Freq)
+SBco <-sum(cocount[cocount$Var1 %in% c("MPB & SB",  "SB & WBBB","MPB, SB & WBBB"),]$Freq)
+WBBBco <-sum(cocount[cocount$Var1 %in% c("MPB & WBBB",  "SB & WBBB","MPB, SB & WBBB"),]$Freq)
+
+cocount <-  as.data.frame(table(bbo.dat$suitablehosts))
+PICOco <-sum(cocount[cocount$Var1 %in% c("PICO & PIEN",  "PICO & ABLA","PICO, PIEN & ABLA"),]$Freq)
+PIENco <-sum(cocount[cocount$Var1 %in% c("PICO & PIEN",  "PIEN & ABLA","PICO, PIEN & ABLA"),]$Freq)
+ABLAco <-sum(cocount[cocount$Var1 %in% c("PICO & ABLA",  "PIEN & ABLA","PICO, PIEN & ABLA"),]$Freq)
+
+codat <- data.frame(agent =c( "MPB", "SB", "WBBB"), agentco=c(MPBco, SBco, WBBBco), hostco=c(PICOco, PIENco, ABLAco))
+
+bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
+bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
+bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
+
+ggfreq <- ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
+
+ggsave( here("Results/Figures/FigureAgentfrequencybyhosts.jpg"), plot=ggfreq,   width=fig.width2,height=fig.width1.5,units="in")
+
+```
+
+### Is occurrence more likely or severity greater in stands with multiple hosts?
+
+```{r occurrence by suitable hosts}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+mpb <- as.data.frame(table(dfpred[dfpred$focalagent=="MPB","affected"], dfpred[dfpred$focalagent=="MPB","suitablehosts"]))
+sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
+wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","suitablehosts"]))
+any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","suitablehosts"]))
 
 results <- NULL
-for(j in levels(bbo.dat$hostspred)[5:8]){
+for(j in c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA")){
   anyj <- any[any$Var2==j,]
@@
 
-  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
-  res$hostspred <-j
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$suitablehosts <-j
   res$barkbeetle <- "MPB"
@@
 
-  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
-  res$hostspred <-j
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$suitablehosts <-j
   res$barkbeetle <- "SB"
@@
 
-  res <- data.frame(hostspred=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
-  res$hostspred <-j
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$suitablehosts <-j
   res$barkbeetle <- "WBBB"
@@
 
-x <- bbo.dat[is.na(bbo.dat$hostspred)==FALSE,]
-x <- x[x$hostspred %in% levels(x$hostspred)[5:8],]
-x[x$hostspred == "PICO\n& PIEN" & x$focalagent=="WBBB","affected"]<- 0
-x[x$hostspred == "PICO\n& ABLA" & x$focalagent=="SB","affected"]<- 0
-x[x$hostspred == "PIEN\n& ABLA" & x$focalagent=="MPB","affected"]<- 0
-
-xx <- x %>% group_by(hostspred,focalagent) %>% count(affected)
-xxx <- x %>% group_by(hostspred) %>% count()
-
-xxxx <- merge(xx, xxx, by="hostspred")
+x <- dfpred[is.na(dfpred$suitablehosts)==FALSE,]
+x <- x[x$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"),]
+x[x$suitablehosts == "PICO & PIEN" & x$focalagent=="WBBB","affected"]<- 0
+x[x$suitablehosts == "PICO & ABLA" & x$focalagent=="SB","affected"]<- 0
+x[x$suitablehosts == "PIEN & ABLA" & x$focalagent=="MPB","affected"]<- 0
+
+xx <- x %>% group_by(suitablehosts,focalagent) %>% count(affected)
+xxx <- x %>% group_by(suitablehosts) %>% count()
+
+xxxx <- merge(xx, xxx, by="suitablehosts")
 xxxx$prop <- xxxx$n.x/xxxx$n.y
@@
 xxxx <- xxxx[,c(1,2,6)]
-colnames(xxxx) <- c("hostspred", "barkbeetle", "prop")
-
-resultsx <- merge(results, xxxx, id.vars=c("hostspred", "barkbeetle"), all=T)
+colnames(xxxx) <- c("suitablehosts", "barkbeetle", "prop")
+
+resultsx <- merge(results, xxxx, id.vars=c("suitablehosts", "barkbeetle"), all=T)
 resultsx$barkbeetle <- factor(resultsx$barkbeetle, levels=c("MPB", 'SB', "WBBB", "Any"), ordered=T)
 resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
-resultsx$hostspred <- factor(resultsx$hostspred, levels=levels(bbo.dat$hostspred)[5:8], ordered=T)
+resultsx$suitablehosts <- factor(resultsx$suitablehosts, levels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
 resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
@@
 
-Fig.file <- here("Results/Figures/Figurey.jpg")
-jpeg(Fig.file, width=fig.width2,height=fig.width1,units="in", res=300)
-ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~hostspred, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")
-whatever <- dev.off()
-```
-
-```{r}
-Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
-bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
-bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
-bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
-#jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in", res=300)
-ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
-#whatever <- dev.off()
-```
-
-### Is occurrence more likely in stands with multiple hosts?
-
-### Is severity greater in stands with multiple hosts?
-
-To determine if host diversity or identity influences the occurrence tree mortality attributed to bark beetles within a stand, we used simple Chi-square tests.
-Given a significant difference among groups, we then used pairwise Fisher's exact tests to determine pairwise differences, while accounting for multiple comparisons [@Benjamini_controlling_1995].
+ggoccurrence <- ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~suitablehosts, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")
+
+ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1,units="in")
+
+```
+
+```{r severity by suitable hosts}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+mpb <- dfpred[dfpred$focalagent=="MPB" & dfpred$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA",   "PICO, PIEN, & ABLA"), ]
+sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
+wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","suitablehosts"]))
+any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","suitablehosts"]))
+
+dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ]
+
+dfpred <- dfpred[dfpred$affected01=="affected",]
+
+results <- NULL
+for(j in c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA")){
+
+  x <- kruskal.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'MPB'),])
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "MPB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  x <-NULL
+  results <- rbind(results, res)
+
+    x <- kruskal.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'SB'),])
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "SB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  x <-NULL
+  results <- rbind(results, res)
+
+      x <- kruskal.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'WBBB'),])
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "WBBB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  x <-NULL
+  results <- rbind(results, res)
+
+}
+
+results$pplot <- ifelse(results$p<0.05, "*", NA)
+
+dfpred$pplot <- NA
+for(j in c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA")){
+
+  x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB',]
+  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='MPB', "pplot"]
+
+    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB',]
+  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='SB', "pplot"]
+
+    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB',]
+  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='WBBB', "pplot"]
+
+}
+
+dfpred <- dfpred[!(dfpred$suitablehosts == "PICO & PIEN" & dfpred$focalagent=="WBBB"),]
+dfpred <- dfpred[!(dfpred$suitablehosts == "PICO & ABLA" & dfpred$focalagent=="SB"),]
+dfpred <- dfpred[!(dfpred$suitablehosts == "PIEN & ABLA" & dfpred$focalagent=="MPB"),]
+
+ggseverity <- ggplot(dfpred, aes(x=focalagent, y=propTBA*100, fill=focalagent, colour=focalagent))+geom_boxplot(notch=T)+facet_grid(.~suitablehosts, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,100)+geom_text(aes(label=pplot), vjust=-0.25,size=10, col="black")+ylab("Severity (% BA affected)")
+
+ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1,units="in")
+
+```
+
+## Agent co-occurrence
+
+### Given suitable stand conditions, is co-occurrence of multiple agents common?
+
+### What are the effects of multiple agents?
+```{r severity by suitable hosts}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ]
+
+dfpred <- dfpred[dfpred$affected01=="affected",]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PICO & PIEN" & dfpred$agentsF %in% c("WBBB", "MPB &\nWBBB", "SB &\nWBBB", "MPB,\nSB, &\nWBBB")),]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PICO & ABLA" & dfpred$agentsF %in% c("SB", "MPB\n& SB", "SB &\nWBBB", "MPB,\nSB, &\nWBBB")),]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN & ABLA" & dfpred$agentsF %in% c("MPB", "MPB\n& SB", "MPB &\nWBBB", "MPB,\nSB, &\nWBBB")),]
+
+ggmultipelseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, y=propTBA*100, fill=agentsF))+geom_boxplot()+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")
+
+ggsave(here("Results/Figures/FigureSeverityMultiple.jpg"), plot=ggmultipelseverity , width=fig.width2,height=fig.width1,units="in")
+```
 
@@
 
-```{r FigPresHosts, fig.cap = "The frequency of FIA plots inventoried by host identity and presence (dark) and absence (light) of bark beetle activity across the study area. ", out.width = "244.8pt"}
-knitr::include_graphics( here("Results/Figures/Figure-PresencebyHostcombos.jpg"))
-```
-
-## Occurrence 
-```{r FigPropAgentxHosts, fig.cap = "A caption for figure 1", out.width = "504pt", eval=F}
-knitr::include_graphics( here("Results/Figures/Figure-PropAffectedByAgentXHosts.jpg"))
-```
-
 ## Co-occurrence of agents
 
+```{r FigCoOccurrence, fig.cap = "Co-occurrence",  out.width = "504.8pt"}
+
+knitr::include_graphics( here("Results/Figures/FigureAgentfrequencybyhosts.jpg"))
+```
+
 Few stands were affected by multiple agents.
 
-```{r FigOEACbyHI, fig.cap = "A caption for figure 2", out.width = "504pt"}
-knitr::include_graphics(here("Results/Figures/Figure-ObsExpCo-ocurrenceAgentXHosts.jpg"))
-```
-
-## Ecological effects
+### Is occurrence more likely or severity greater in stands with multiple hosts?
+
+```{r FigOccurrence, fig.cap = "Severity",  out.width = "504.8pt"}
+
+knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
+```
+
+```{r FigSeverity, fig.cap = "Severity", out.width = "504.8pt"}
+
+knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
+```
 
@@
 ## Supplement
+### Randon Forest Modeling
+```{r FigPartial,  fig.cap = "A caption", out.width = "504.8pt"}
+MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
+MPBpp$focalagent <- "MPB"
+SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
+SBpp$focalagent <- "SB"
+WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
+WBBBpp$focalagent <- "WBBB"
+
+ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
+ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
+ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
+ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
+
+x <- df[!df$focalagent=="Any",]
+x$propBA <- x$BA.host/x$BA.total
+x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
+x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
+
+xm <- melt(x, id.vars=c("focalagent", "affected01"))
+colnames(xm) <- c("focalagent", "affected01", "var", "value")
+xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
+
+pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
+
+pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
+
+ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("propBA")
+
+ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
+
+ggsave(here("Results/Figures/PartialDependence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
+knitr::include_graphics(here("Results/Figures/PartialDependence.jpg"))
+
+```
+
 ### Comparison of stand structure and composition, climate, and regional bark beetle activity  
-```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt"}
+```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt", eval=F}
 datBB.c <-df.sub[!df.sub$focalhost=="Any",]
@@
 
-```{r FigPressure, fig.cap = "A caption", out.width = "322.56pt"}
-## select plots where regional pressure can be estimated (> 5= plots)
-
-RpressureSB.melt <- datBB.c[,c("hosts", "Rpressure.SB", "hostBA.SB", "nplots")]
-colnames(RpressureSB.melt) <- c("hosts", "pressure", "hostBA", "nplots")
-RpressureSB.melt$variable <- "SB"
-
-RpressureMPB.melt <- datBB.c[, c("hosts", "Rpressure.MPB", "hostBA.MPB", "nplots")]
-colnames(RpressureMPB.melt) <- c("hosts", "pressure", "hostBA", "nplots")
-RpressureMPB.melt$variable <- "MPB"
-
-RpressureWBBB.melt <- datBB.c[, c("hosts", "Rpressure.WBBB", "hostBA.WBBB", "nplots")]
-colnames(RpressureWBBB.melt) <- c("hosts", "pressure", "hostBA", "nplots")
-RpressureWBBB.melt$variable <- "WBBB"
-
-Rpressure.melt <- rbind(RpressureMPB.melt,RpressureSB.melt)
-Rpressure.melt <- rbind(Rpressure.melt,RpressureWBBB.melt)
-
-Rpressure.melt$variable <- factor(Rpressure.melt$variable, levels=c("MPB","SB",  "WBBB"), labels=c("MPB","SB",  "WBBB"), ordered=T)
-Rpressure.melt$propRegional <- Rpressure.melt$pressure /Rpressure.melt$hostBA
-Rpressure.melt$host <- factor(Rpressure.melt$variable, levels=c("MPB","SB", "WBBB"), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-
-ggpropRegional <- ggplot(Rpressure.melt, aes(x=hosts, y=propRegional,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BrBG")+ylab("Regional fraction\nof plots affected")+xlab("")+scale_fill_manual(values=rev(p2col)[2:4])
-
-gghostBA <- ggplot(Rpressure.melt, aes(x=hosts, y=hostBA,fill=host))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BuPu")+ylab(expression(atop("Regional basal", paste("area (", m^2, ha^-1, ")"))))+xlab("")+theme(axis.text.x = element_blank(), axis.title.x = element_blank())
-
-ggpressure <- ggplot(Rpressure.melt, aes(x=hosts, y=pressure,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_manual(values=rev(p2col)[2:4])+ylab("Regional pressure")+xlab("host composition")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1), axis.title.x = element_blank(), legend.position = "top", legend.title = element_blank())
-
-ggcombo <- gghostBA + theme(axis.text.x=element_blank()) + ggpropRegional +  plot_layout(ncol=1)+ plot_annotation(tag_levels = 'A')+ plot_layout(guides = "collect") & theme(legend.position='bottom', legend.title=element_blank(),legend.box="vertical", legend.margin=margin(), plot.tag = element_text(face = 'bold'))
-ggsave(here("Results/Figures/Figure-PressureComboByhostIdentity.jpg"), plot=ggcombo , width=fig.width1.5, height=fig.width1*1.5, units="in")
-
-knitr::include_graphics(here("Results/Figures/Figure-PressureComboByhostIdentity.jpg"))
-```
-
-```{r FigClimate, fig.cap = "A caption", out.width = "244.8pt"}
-ggDecTMINa <- ggplot(datBB.c, aes(x=hostsFPlong, y=DecTMINa))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Dec TMIN\n Anomaly")
-
-ggJulTMAXa<- ggplot(datBB.c, aes(x=hostsFPlong, y=JulTMAXa))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Jul TMAX\n Anomaly")
-
-ggPPTa<- ggplot(datBB.c, aes(x=hostsFPlong, y=PPTa))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Annual PPT\n Anomaly")
-
-ggJJSPEI <- ggplot(datBB.c, aes(x=hostsFPlong, y=JJSPEI))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("Jun-Jul SPEI")
-
-ggWYSPEI <- ggplot(datBB.c, aes(x=hostsFPlong, y=WYSPEI))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2)+ylab("5-year SPEI")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1), axis.title.x = element_blank())
-
-ggclimatecombo <- ggDecTMINa + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggJulTMAXa + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggPPTa + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggJJSPEI + theme(axis.text.x=element_blank(),axis.title.x = element_blank()) + ggWYSPEI + plot_layout(ncol=1) + plot_annotation(tag_levels = 'A') + plot_layout(guides = "collect") & theme(legend.position='bottom', legend.title=element_blank(), plot.tag = element_text(face = 'bold'))
-ggsave(here("Results/Figures/Figure-ClimateByHostIdentity.jpg"), plot=ggclimatecombo , width=fig.width1, height=fig.width1*2, units="in")
-
-knitr::include_graphics(here("Results/Figures/Figure-ClimateByHostIdentity.jpg"))
-```
-
 ```{r generate maps, eval=F}
-df.all <- df.sub[df.sub$focalhost == "Any", ]
+bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
+
+df.all <- bbo.dat[bbo.dat$focalhost == "Any", ]
 
@@
 Fig.file <- here("Results/Figures/Figure-Map.jpg")
-jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
+jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in")
 tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)

### 2021-10-04 Code/Manuscript-SMOTE.Rmd
analyses complete

@@
 title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
-author: "Sarah J. Hart"
+author: "Sarah J. Hart et al."
 email: "sarah.hart@colostate.edu"
@@
 citationcolor: blue
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 ### Import libraries
-ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach'))
-
-#foreach', 'parallel', 'doParallel', 'quickPlot', 'reshape2','vegan', 'FSA', 
+ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'kableExtra', 'bookdown',  'pdp', 'ranger', 'VSURF', 'themis', 'foreach',  'FSA', 'parallel'))
+
 # Set seed
 set.seed(2020)
+
+# Set number of cores for parralell processing
+cores <- parallel::detectCores()
 
@@
 Paragraph 2: Theoretical context - diversity & natural enemies
-- Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources.
-- Key hypotheses:
+
+Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources.
+In the case of plant-herbivore interactions, the greater plant diversity often reduces herbivory effects on plants at both the individual and population level [i.e.
+'associational resistance, @barbosa_associational_2009, @jactel_tree_2021].
+The resource concentration hypothesis suggests that specialists insect herbivores are less likely to find and stay in stands where their hosts are less abundant [@root_organization_1973].
+This may occur if insects are less likely to find a host tree due [i.e.
+'host dilution effect', @] or chemical masking of the focal tree by heterospecific neighbors [i.e.
+'host apparency hypothesis', @Castagneryrol_effects_2014]
+
+[i.e., 'resource concentration hypothesis', @root_organization_1973].
+
+However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
+suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.
+
 1 - associational resistance - the abundance or damage of herbivores is lower when plant diversity is greater [@barbosa_associational_2009, @jactel_tree_2021]
@@
 
-However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
-
-suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.
-
+H
 Most studies have focused on individual- or population-level outcomes.
@@
 Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.
+
 For instance, a primary goal of the Western Bark Beetle is to promote resilience of forests to bark beetle outbreaks by increasing the diversity of age classes and tree species.
 
-If greater host richness increases niche space for bark beetles, then community-level occurrence of bark beetle activity will be greater in more diverse communities.
-community-level severity, here the amount and proportion of basal area affected by bark beetles, will be lower in stands with more host species.
-
 To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).
-Specifically we ask: (1) Is the occurrence or severity of bark beetle
+
+We expect that co-occurrence of trees of different host species will be common, but co-occurrence of stand conditions suitable for multiple bark beetle species will not.
+Given stands where conditions are suitable for multiple bark beetle species, occurrence of bark beetle-driven tree mortality will be greater (i.e.
+diversity begets diversity), but severity will be lower (i.e.
+resource concentration).
+
+We also expected the severity of bark beetle infestation to vary with the number of agents present, but with two alternative hypotheses.
+If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent (i.e.
+additive effects).
+Alternatively, lower tree mortality may occur if concurrent outbreaks of bark beetles of different species cause semiochemical confusion or if competitive release increases tree defensive capacity.
 
@@
 
-The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming; Fig.
-\@ref(fig:FigSA)
+The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming)
 
@@
 df$agents <-  do.call(paste, c(df[c("agentSB", "agentMPB", "agentWBBB")],sep="-"))
-df$agents <- factor(df$agents, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))
+df$agents <- factor(df$agents, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))
 
 df$nagents <- df$agents
-df$nagents <- factor(df$nagents, levels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))
+df$nagents <- factor(df$nagents, levels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))
 
@@
 df$hosts <-  do.call(paste, c(df[c("hostsPICO", "hostsPIEN", "hostsABLA")],sep="-"))
-df$hosts <- factor(df$hosts, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA', 'none'), ordered=TRUE)
+df$hosts <- factor(df$hosts, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), ordered=TRUE)
 
 df$nhosts <- df$hosts
-df$nhosts <- factor(df$nhosts, levels=c('PICO', 'ABLA', 'PIEN', 'PICO & PIEN', 'PICO & ABLA', 'PIEN & ABLA', 'PICO, PIEN, & ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))
+df$nhosts <- factor(df$nhosts, levels=c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))
 
@@
 
-### Descriptive analyses
 ```{r import processed data}
@@
 df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
-df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), ordered=T)
-df$hostsFPlong <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO\n& PIEN", "PICO\n& ABLA", "PIEN \n& ABLA", "PICO, PIEN,\n& ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
+df$hostsFP <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
 
 df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
-df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB & WBBB"), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB\n& SB","MPB &\nWBBB",  "SB &\nWBBB", "MPB,\nSB, &\nWBBB"), ordered=T)
-df$agentsFplong <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB & WBBB"), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB","MPB & WBBB",  "SB & WBBB", "MPB, SB, & WBBB"), ordered=T)
+df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)
 
@@
 
-```{r map study area of hosts, eval=F}
-f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
-f93 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
-f108 <- stars::read_stars(here("/Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)
-
-# reclassify to rasters to presence (1) and absence (0)
-f19[f19>0] <- 1 
-f93[f93>0] <- 1
-f108[f108>0] <- 1
-
-# define areas of multiple hosts
-f108f93 <- f108 + f93 # add 
-f108f93[f108f93<2]<- 0 # reclassify
-f108f93[f108f93==2]<- 1 # reclassify
-f108f19 <- f108 + f19
-f108f19[f108f19<2]<- 0 # reclassify
-f108f19[f108f19==2]<- 1 # reclassify
-f93f19 <- f93 + f19
-f93f19[f93f19<2]<- 0  # reclassify
-f93f19[f93f19==2]<- 1 # reclassify
-
-f108f93f19 <- f108 + f93 + f19
-
-# define areas of one host
-f108f93f19re1 <-f108f93f19
-f108f93f19re1 [f108f93f19re1 ==1 ]<-1
-f108f93f19re1 [f108f93f19re1>1]<-0
-
-# define areas of two hosts
-f108f93f19re2 <-f108f93f19
-f108f93f19re2 [f108f93f19re2 ==1 ]<-0
-f108f93f19re2 [f108f93f19re2==3]<-0 
-f108f93f19re2 [f108f93f19re2 ==2 ]<-1
-
-# limit maps of presence (1) and absence (0) to only areas with one host
-f108 <- f108 * f108f93f19re1 
-f93 <- f93 * f108f93f19re1
-f19 <- f19 * f108f93f19re1 
-f93[f93==1]<- 2 # give unique labels
-f19[f19==1]<- 3 # give unique labels
-
-# create maps of two hosts limited to areas with two hosts
-f108f93  <- f108f93 * f108f93f19re2 
-f108f93[f108f93==1]<-4 # give unique labels
-
-f108f19 <- f108f19 * f108f93f19re2
-f108f19[f108f19==1]<-5 # give unique label
-
-f93f19 <- f93f19 * f108f93f19re2
-f93f19[f93f19==1]<-6 # give unique labels
-
-# create maps of all three hosts
-f108f93f19[f108f93f19<=2]<-0
-f108f93f19[f108f93f19==3]<-7 # give unique labels
-
-fplot <- f108 + f93 + f19 + f108f93 + f108f19 + f93f19+ f108f93f19
-st_crs(fplot) <- "EPSG:5070" # assign projection
-
-fplot[fplot==0] <-NA
-
-states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
-IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-notIMW <- states[!states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-IMW <- st_transform(IMW, "EPSG:5070")
-notIMW <- st_transform(notIMW, "EPSG:5070")
-
-ext <- as.matrix(extent(IMW))
-
-sa.tmap <- tm_shape(IMW, bbox=tmaptools::bb(ext)) + tm_fill(col="#f5f5f5")+tm_borders()+tm_shape(fplot) +tm_raster(style = "cat", palette = c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), colorNA=NULL, legend.show = TRUE, title="Hosts                    ", labels = c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"))+ tm_shape(notIMW) + tm_fill(col="white")+tm_compass(position=c(0.95, 0.88), just=c("center", "bottom"))+tm_layout(legend.bg.color="white")+tm_graticules(lines=FALSE)+tm_layout(legend.outside=TRUE,legend.text.size=0.55)+tm_scale_bar(position = c(0.01, 0.001), just="left", breaks=c(0,100),text.size=1.1)+tm_shape(IMW) +tm_borders()
-
-Fig.file <- here("Results/Figures/Figure-Studyarea.jpg")
-jpeg(Fig.file, width=fig.width1.5,height=fig.width1.5,units="in", res=300)
-sa.tmap 
-whatever <- dev.off()
-```
-
-```{r describe frequency of hosts and agents, eval=T}
-## Hosts
-hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-
-## Affected hosts
-affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
-affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]
-
-affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
-affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]
-
-## Agents given host presence
-agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
-  mutate(freq = n / sum(n))
-PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
-PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
-ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))
-
-## Agents given presence of any agent
-agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
-  mutate(freq = n / sum(n))
-
-agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-
-p1 <- ggplot(df[df$focalhost=="Any",], aes(x=hosts, fill=interaction(hosts, affected))) + geom_bar(width=0.75, stat="count", col="black")+ylab("number of plots")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1),legend.position = "none", legend.title=element_blank())+scale_fill_manual(values=c(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), adjust_transparency(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), alpha=0.5)))
-
-Fig.file <- here("Results/Figures/Figure-PresencebyHostcombos.jpg")
-jpeg(Fig.file, width=fig.width1,height=fig.width1/1.5,units="in", res=300)
-p1
-whatever <- dev.off()
-```
-
-```{r determine co-occurrence of hosts}
-Fig.file <- here("Results/Figures/Figure-AgentbyHostcombo.jpg")
-df$hostsFPshort<- df$hostsFPlong
-df$hostsFPshort <- factor(df$hostsFPlong, levels=levels(df$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-df$agentsFplongrev <- df$agentsFplong
-df$agentsFplongrev <- factor(df$agentsFplongrev, levels=rev(levels( df$agentsFplong)))
-jpeg(Fig.file, width=fig.width2,height=fig.width1.5,units="in",res=300)
-ggplot(df[df$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
-whatever <- dev.off()
-```
-
-### Determine suitable hosts
-Construct random forest models using synthetic minority oversampling technique (SMOTE)
-
-#### MPB
-```{r fit MPB RF model, eval=F}
+### Model the suitable stand conditions
+Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data.
+The SMOTE approach oversamples the minority class by synthesizing new cases from the minority class.
+We compared this approach with two other common approaches for dealing with imbalanced data in a random forest framework - balanced and weighted random forests and selected the approach that classified affected plots with the greatest accuracy.
+
+```{r fit MPB RF model SMOTE, eval=F}
 response <- "agentMPB"
-vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
-datBB.c <- df[df$focalhost=="Any",]
-
-splitit <- initial_split(datBB.c, prop=.7, strata=response)
-datz.train <- training(splitit)
-datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
+datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit) # create object with training data
+datz.test <- testing(splitit) # create object with testing data
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
 
 # set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train)
-
-# set sample weights
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights (not used in SMOTE approach)
 wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
 
@@
 tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
-rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, accuracy, sens, spec))
-
-tuneres <- rf_tune_results %>% collect_metrics()
-
-param_best<- rf_tune_results %>% select_best(metric = "roc_auc")
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
 
@@
 vars.final <- vars
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
-
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
 
 rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
-```
-
-#### SB
-```{r fit SB RF model, eval=F}
-response <- "agentSB"
-vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit MPB RF model Balanced, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
 datBB.c <- df[df$focalhost=="Any",]
 
-splitit <- initial_split(datBB.c, prop=.7, strata=response)
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
 datz.train <- training(splitit)
 datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
 
@@
 BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
-
 # set sample weights
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts) %>% set_mode("classification")
 
@@
 
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
 
 rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
-```
-
-#### WBBB
-```{r fit WBBB RF model, eval=F}
-response <- "agentWBBB"
-vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
+```
+
+```{r fit MPB RF model Weighted, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
 datBB.c <- df[df$focalhost=="Any",]
 
-splitit <- initial_split(datBB.c, prop=.7, strata=response)
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
 datz.train <- training(splitit)
 datz.test <- testing(splitit)
-datz.train.cv <- vfold_cv(datz.train, strata=response, v=10)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
 
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", probability=T, seed=913, sample.fraction=wts) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts) %>% set_mode("classification")
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
 
 rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
-rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, accuracy, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-2.csv")), row.names=F)
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+```
+
+```{r fit SB RF model SMOTE, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit SB RF model Balanced, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
+```
+
+```{r fit SB RF model Weighted, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model SMOTE, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model Balanced, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model Weighted, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 ```
@@
 ```{r determine suitable stands}
-MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence.csv"))
-SBpred <- read.csv(here("Results/RF-pred-agentSB-occurrence.csv"))
-WBBBpred <- read.csv(here("Results/RF-pred-agentWBBB-occurrence.csv"))
+MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence-SMOTE.csv"))
+SBpred <- read.csv(here("Results/RF-pred-agentSB-occurrence-SMOTE.csv"))
+WBBBpred <- read.csv(here("Results/RF-pred-agentWBBB-occurrence-SMOTE.csv"))
 
@@
 BBpred$agentsuitable <-  do.call(paste, c(BBpred[c("SBsuitable", "MPBsuitable", "WBBBsuitable")],sep="-"))
-BBpred$agentsuitable <- factor(BBpred$agentsuitable, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB & SB", "WBBB", "SB & WBBB", "MPB & WBBB", "MPB, SB & WBBB"))
-
-BBpred$suitablehosts <- factor(BBpred$agentsuitable, levels=c("none", "MPB", "SB", "WBBB", "MPB & SB",  "MPB & WBBB", "SB & WBBB",  "MPB, SB & WBBB"), labels=c("none", "PICO", "PIEN", "ABLA","PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
-
-```
-
-## Host co-occurrence
-
-### Given suitable stand conditions, is co-occurrence of suitable hosts common?
+BBpred$agentsuitable <- factor(BBpred$agentsuitable, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))
+
+BBpred$suitablehosts <- factor(BBpred$agentsuitable, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB",  "MPB+WBBB", "SB+WBBB",  "MPB+SB+WBBB"), ordered=T)
+
+```
+
+# Results
+## Is co-occurrence of multiple hosts common?
+```{r describe frequency of hosts and agents, eval=T}
+## Hosts
+hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+
+nhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+
+## Affected hosts
+affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
+affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]
+
+affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
+affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]
+
+## Agents given host presence
+agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
+  mutate(freq = n / sum(n))
+PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
+PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
+ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))
+
+## Agents given presence of any agent
+agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
+  mutate(freq = n / sum(n))
+
+agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+
+p1 <- ggplot(df[df$focalhost=="Any",], aes(x=hosts, fill=interaction(hosts, affected))) + geom_bar(width=0.75, stat="count", col="black")+ylab("number of plots")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1),legend.position = "none", legend.title=element_blank())+scale_fill_manual(values=c(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), adjust_transparency(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), alpha=0.5)))
+
+Fig.file <- here("Results/Figures/Figure-PresencebyHostcombos.jpg")
+jpeg(Fig.file, width=fig.width1.5,height=2.5,units="in", res=300)
+p1
+whatever <- dev.off()
+```
+
+```{r}
+gghostgroups <- ggplot(hostgroups, aes(x=hosts, y=freq, fill=hosts))+geom_bar(stat="identity")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")+xlab("")+ylab("frequency")
+
+ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")
+```
+
+Yes, 66% (n=5,850) FIA plots had at least two host species present.
+Subalpine fir was most likely to occur with other hosts (50% of stands), followed by Engelmann spruce (45% stands), and lodgepole pine (36% of stands) (Fig.
+\@ref(fig:FigHostFreq)).
+
+```{r FigHostFreq, fig.cap = ".", out.width = "322.56pt"}
+knitr::include_graphics( here("Results/Figures/FigureHostFreq.jpg"))
+```
+
+## Is co-occurrence of the stands conditions suitable for multiple bark beetle species common?
+
+20% of stands were suitable for more than one bark beetle species. Conditions suitable for WBBB were most likely to co-occur with conditions suitable for another bark beetle (17% of stands)
+
 ```{r determine co-occurrence of suitable stands}
@@
 bbo.dat <- bbo.dat[bbo.dat$focalagent=="Any",]
-cocount <-  as.data.frame(table(bbo.dat$agentsFplongrev))
-MPBco <-sum(cocount[cocount$Var1 %in% c("MPB & SB",  "MPB & WBBB","MPB, SB & WBBB"),]$Freq)
-SBco <-sum(cocount[cocount$Var1 %in% c("MPB & SB",  "SB & WBBB","MPB, SB & WBBB"),]$Freq)
-WBBBco <-sum(cocount[cocount$Var1 %in% c("MPB & WBBB",  "SB & WBBB","MPB, SB & WBBB"),]$Freq)
-
-cocount <-  as.data.frame(table(bbo.dat$suitablehosts))
-PICOco <-sum(cocount[cocount$Var1 %in% c("PICO & PIEN",  "PICO & ABLA","PICO, PIEN & ABLA"),]$Freq)
-PIENco <-sum(cocount[cocount$Var1 %in% c("PICO & PIEN",  "PIEN & ABLA","PICO, PIEN & ABLA"),]$Freq)
-ABLAco <-sum(cocount[cocount$Var1 %in% c("PICO & ABLA",  "PIEN & ABLA","PICO, PIEN & ABLA"),]$Freq)
-
-codat <- data.frame(agent =c( "MPB", "SB", "WBBB"), agentco=c(MPBco, SBco, WBBBco), hostco=c(PICOco, PIENco, ABLAco))
-
-bbo.dat$hostsFPshort<- bbo.dat$hostsFPlong
-bbo.dat$hostsFPshort <- factor(bbo.dat$hostsFPlong, levels=levels(bbo.dat$hostsFPlong), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-bbo.dat$agentsFplongrev <- bbo.dat$agentsFplong
+
+suitablehostgroups <- bbo.dat[bbo.dat$focalagent=="Any",]%>% group_by(suitablehosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+
+bbo.dat$hostsFPshort<- bbo.dat$hosts
+bbo.dat$hostsFPshort <- factor(bbo.dat$hosts, levels=levels(bbo.dat$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+bbo.dat$agentsFplongrev <- bbo.dat$agentsF
 bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
 
-ggfreq <- ggplot(bbo.dat[bbo.dat$focalhost=="Any",], aes(x=hostsFPshort,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=p2col)
-
-ggsave( here("Results/Figures/FigureAgentfrequencybyhosts.jpg"), plot=ggfreq,   width=fig.width2,height=fig.width1.5,units="in")
-
-```
-
-### Is occurrence more likely or severity greater in stands with multiple hosts?
+bbo.sub <- bbo.dat[bbo.dat$focalhost=="Any",]
+bbo.sub <- bbo.dat[!bbo.dat$suitablehosts=="none",]
+
+bbo.sub$suitablehostsP <- factor(bbo.sub$suitablehosts, levels=levels(bbo.sub$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+
+ggfreq <- ggplot(bbo.sub, aes(x=suitablehostsP,fill=suitablehostsP))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")
+
+ggsave( here("Results/Figures/FigureSuitable.jpg"), plot=ggfreq,width=fig.width1,height=fig.width1*0.75,units="in")
+```
+
+(Fig. \@ref(fig:FigureSuitable)).
+
+```{r FigureSuitable, fig.cap = ".", out.width = "322.56pt"}
+knitr::include_graphics( here("Results/Figures/FigureSuitable.jpg"))
+```
+
+## Is occurrence or severity cumulative bark beetle mortality greater in stands with multiple hosts?
 
 ```{r occurrence by suitable hosts}
+
 dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
 mpb <- as.data.frame(table(dfpred[dfpred$focalagent=="MPB","affected"], dfpred[dfpred$focalagent=="MPB","suitablehosts"]))
@@
 results <- NULL
-for(j in c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA")){
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
   anyj <- any[any$Var2==j,]
@@
 x <- dfpred[is.na(dfpred$suitablehosts)==FALSE,]
-x <- x[x$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"),]
-x[x$suitablehosts == "PICO & PIEN" & x$focalagent=="WBBB","affected"]<- 0
-x[x$suitablehosts == "PICO & ABLA" & x$focalagent=="SB","affected"]<- 0
-x[x$suitablehosts == "PIEN & ABLA" & x$focalagent=="MPB","affected"]<- 0
+x <- x[x$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"),]
+x[x$suitablehosts == "PICO+PIEN" & x$focalagent=="WBBB","affected"]<- 0
+x[x$suitablehosts == "PICO+ABLA" & x$focalagent=="SB","affected"]<- 0
+x[x$suitablehosts == "PIEN+ABLA" & x$focalagent=="MPB","affected"]<- 0
 
@@
 resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
-resultsx$suitablehosts <- factor(resultsx$suitablehosts, levels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
+resultsx$suitablehosts <- factor(resultsx$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
 resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
@@
 
-ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1,units="in")
-
-```
+ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.66,units="in")
+
+```
+
+Given suitable stand conditions for multiple agents, the probability of occurrence  of any agent was significantly greater than the probability of a single agent \@ref(fig:FigureSuitable).
+```{r FigOccurrence, fig.cap = "Severity",  out.width = "504.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
+```
+
+Given suitable stand conditions for multiple agents, is the severity of cumulative bark beetle-driven tree mortality greater in stands with multiple hosts?
+
+In stands suitable for all three hosts, cumulative mortality was significantly greater than mortality by a single agent.
+\@ref(fig:FigSeverity).
+When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most common agent (MPB > SB > WBBB) and was not significantly different than the cummualtive mortality.
 
@@
 dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-mpb <- dfpred[dfpred$focalagent=="MPB" & dfpred$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA",   "PICO, PIEN, & ABLA"), ]
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+
+mpb <- dfpred[dfpred$focalagent=="MPB" & dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA",   "PICO+PIEN+ABLA"), ]
 sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
@@
 
-dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ]
+dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
 
@@
 results <- NULL
-for(j in c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA")){
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
 
@@
 dfpred$pplot <- NA
-for(j in c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA")){
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
 
@@
 
-dfpred <- dfpred[!(dfpred$suitablehosts == "PICO & PIEN" & dfpred$focalagent=="WBBB"),]
-dfpred <- dfpred[!(dfpred$suitablehosts == "PICO & ABLA" & dfpred$focalagent=="SB"),]
-dfpred <- dfpred[!(dfpred$suitablehosts == "PIEN & ABLA" & dfpred$focalagent=="MPB"),]
-
-ggseverity <- ggplot(dfpred, aes(x=focalagent, y=propTBA*100, fill=focalagent, colour=focalagent))+geom_boxplot(notch=T)+facet_grid(.~suitablehosts, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,100)+geom_text(aes(label=pplot), vjust=-0.25,size=10, col="black")+ylab("Severity (% BA affected)")
-
-ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1,units="in")
-
-```
-
-## Agent co-occurrence
-
-### Given suitable stand conditions, is co-occurrence of multiple agents common?
-
-### What are the effects of multiple agents?
-```{r severity by suitable hosts}
+dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+PIEN" & dfpred$focalagent=="WBBB"),]
+dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+ABLA" & dfpred$focalagent=="SB"),]
+dfpred <- dfpred[!(dfpred$suitablehosts == "PIEN+ABLA" & dfpred$focalagent=="MPB"),]
+dfpred$suitablehostsP <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
+
+ggseverity <- ggplot(dfpred, aes(x=focalagent, y=propTBA*100, fill=focalagent, colour=focalagent))+geom_boxplot(notch=T)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")
+
+ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
+
+```
+
+```{r FigSeverity, fig.cap = "Severity", out.width = "504.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
+```
+
+## Given suitable stand conditions, is co-occurrence of multiple agents common?
+
+Not really, only 21.3% (n=461) of stands suitable for multiple agents were affected by multiple agents.
+When multiple agents occur, the most commonly occurring agents was MPB (n=341), followed by SB (n=317) and WBBB (n=308).
+The most commonly occurring combinations of agents was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and then SB and WBBB (26% of case; n=120).
+The combination of all three agents was rare (9.5% of cases; n=44).
+
+```{r determine co-occurrence of agents by hosts}
 dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ]
-
+dfpred <- dfpred[!dfpred$agentsuitable=="none",]
+
+dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+dfpred$suitablehostsF <-dfpred$suitablehosts
+dfpred$suitablehostsF <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+dfpred$agentsFplongrev <- factor(dfpred$agents,levels=c("MPB+SB+WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB", "WBBB", "SB", "MPB", "none"), labels=c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB", "WBBB", "SB", "MPB", "none"), ordered=T)
+
+dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())
+
+MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+sum(MPBx$n)
+sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+sum(SBx$n)
+sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+sum(WBBBx$n)
+sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+### 
+table(dfpred$suitablehostsF, dfpred$agents)
+x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+xPICO <-x[x$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA",  "PICO,\nPIEN,\n&\nABLA"),]
+#
+
+## Agents given presence of any agent
+co <- dfpred[!dfpred$focalhost=="Any" ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]
+
+ggAgentsbySuitablehosts <- ggplot(dfpred[dfpred$focalhost=="Any",], aes(x=suitablehostsF,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+scale_fill_manual(values=p2col)+theme(axis.title.x=element_blank())
+
+ggsave(here("Results/Figures/Figure-AgentsbySuitablehosts.jpg"), plot=ggAgentsbySuitablehosts , width=fig.width1.5,height=fig.width1*0.75,units="in")
+```
+
+```{r FigCoOccurrence, fig.cap = "Co-occurrence",  out.width = "504.8pt", eval=F}
+knitr::include_graphics( here("Results/Figures/Figure-AgentsbySuitablehosts.jpg"))
+```
+
+##  Is severity greater in stands with multiple agents?
+```{r multiple agents}
+library(rcompanion)
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
 dfpred <- dfpred[dfpred$affected01=="affected",]
 
-dfpred <- dfpred[!(dfpred$suitablehosts=="PICO & PIEN" & dfpred$agentsF %in% c("WBBB", "MPB &\nWBBB", "SB &\nWBBB", "MPB,\nSB, &\nWBBB")),]
-
-dfpred <- dfpred[!(dfpred$suitablehosts=="PICO & ABLA" & dfpred$agentsF %in% c("SB", "MPB\n& SB", "SB &\nWBBB", "MPB,\nSB, &\nWBBB")),]
-
-dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN & ABLA" & dfpred$agentsF %in% c("MPB", "MPB\n& SB", "MPB &\nWBBB", "MPB,\nSB, &\nWBBB")),]
-
-ggmultipelseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, y=propTBA*100, fill=agentsF))+geom_boxplot()+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")
-
-ggsave(here("Results/Figures/FigureSeverityMultiple.jpg"), plot=ggmultipelseverity , width=fig.width2,height=fig.width1,units="in")
-```
-
-# Results
-
-### Patterns
-
-Across the Intermountain West, the most commonly occurring species was lodgepole pine, which was present in `r as.numeric(round(hosts[hosts$focalhost=="PICO", "freq"]*100,0))`% of plots, followed by subalpine fir (`r as.numeric(round(hosts[hosts$focalhost=="ABLA", "freq"]*100,0))`% of plots), and Engelmann spruce (`r as.numeric(round(hosts[hosts$focalhost=="PIEN", "freq"]*100,0))`% of plots).
-Given the presence of the focal host, the most commonly occurring bark beetle species was the MPB, which was found in `r PICOwMPB`% of plots with lodgepole pine.
-SB and WBBB, were found at much lower rates; SB was found in `r PIENwSB`% of plots with Engelmann spruce and WBBB was found in `r ABLAwWBBB`% of plots with subalpine fir.
-
-The most commonly occurring tree community included only lodgepole pine (n=`r as.numeric(hostgroups[hostgroups$hosts=="PICO", "n"])`), followed by plots with all three hosts (n=`r as.numeric(hostgroups[hostgroups$hosts=="PICO, PIEN,\n& ABLA", "n"])`) and Engelmann spruce-sublapine fir plots (n=`r as.numeric(hostgroups[hostgroups$hosts=="PIEN \n& ABLA", "n"])`) (Figs.
-\@ref(fig:FigSA) -  \@ref(fig:FigPresHosts)).
-Logepole pine-Engelmann spruce communities were the least common (n=`r as.numeric(hostgroups[hostgroups$hosts=="PICO\n& PIEN", "n"])`), followed by plots with only subalpine fir (n=`r as.numeric(hostgroups[hostgroups$hosts=="ABLA", "n"])`) and only Engelmann spruce ((n=`r as.numeric(hostgroups[hostgroups$hosts=="PIEN", "n"])`)).
-
-Consistent with diversity begets diversity hypothesis, the relative occurrence of bark beetle activity was greatest in stands with all three hosts (Fig \@ref(fig:FigPresHosts); `r as.numeric(round(affectednhosts[affectednhosts$nhosts==3, "freq"] *100, 0))`% of plots with all three hosts were affected by bark beetles).
-Stands with only one host were least likely to experience bark beetle infestation and stands with (`r as.numeric(round(affectednhosts[affectednhosts$nhosts==1, "freq"] *100, 0))`% of plots with only one host) and stands with two hosts were intermediate  (`r as.numeric(round(affectednhosts[affectednhosts$nhosts==2, "freq"] *100, 0))`% of plots with two hosts).
-
-However host richness also varied with other factors known to influence outbreak presence and severity, notably stand structure and composition (Fig.
-\@ref(fig:FigStandSC)), climate (Fig.
-\@ref(fig:FigClimate)), and landscape scale patterns of host availability and bark beetle pressure (Fig.
-\@ref(fig:FigPressure)).
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+
+#dfpred <- dfpred[!(dfpred$suitablehosts=="PICO" & dfpred$agentsF %in% c("SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+#dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN" & dfpred$agentsF %in% c("MPB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+#dfpred <- dfpred[!(dfpred$suitablehosts=="ABLA" & dfpred$agentsF %in% c("MPB", "SB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+PIEN" & dfpred$agentsF %in% c("WBBB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+ABLA" & dfpred$agentsF %in% c("SB", "MPB+SB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN+ABLA" & dfpred$agentsF %in% c("MPB", "MPB+SB", "MPB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <-dfpred[dfpred$focalagent=="Any",]
+
+results <- NULL
+  dfsub <- dfpred[dfpred$suitablehosts =="PICO+PIEN"  & dfpred$agents %in% c("MPB", "SB","MPB+SB"),]
+  dfsub$agents <- as.factor(as.character(dfsub$agents))
+  x <- kruskal.test(propTBA~agents, data=dfsub)
+  res <- data.frame(agentsF= c("MPB", "SB","MPB+SB"), p=NA)
+  DT <- dunnTest(propTBA~agents, data=dfsub,method="bh")
+  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
+  res$suitablehosts <- "PICO+PIEN"
+
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res$Letter2 <- c("b", "c", "a")
+  results <- rbind(results, res)
+
+  dfsub <- dfpred[dfpred$suitablehosts == "PICO+ABLA"  & dfpred$agentsF %in% c("MPB", "WBBB","MPB+WBBB"),]
+  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
+  x <- kruskal.test(propTBA~agentsF, data=dfsub)
+  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
+  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+  res$suitablehosts <- "PICO+ABLA"
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res$Letter2 <- c("b", "c", "a")
+  results <- rbind(results, res)
+
+  dfsub <- dfpred[dfpred$suitablehosts == "PIEN+ABLA"  & dfpred$agentsF %in% c("SB", "WBBB","SB+WBBB"),]
+  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
+  x <- kruskal.test(propTBA~agentsF, data=dfsub)
+  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
+  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+  res$suitablehosts <- "PIEN+ABLA"
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res$Letter2 <- c("b", "c", "a")
+  results <- rbind(results, res)
+
+  dfsub <- dfpred[dfpred$suitablehosts == "PICO+PIEN+ABLA",]
+  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
+  x <- kruskal.test(propTBA~agentsF, data=dfsub)
+  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
+  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+  res$suitablehosts <- "PICO+PIEN+ABLA"
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res <- res[order(res$median),]
+  res$Letter2 <- c("a", "ab", "abc", "b", "c", "d", "d")
+
+  results <- rbind(results, res)
+  results$agentsF <- results$Group
+
+results$suitablehosts <- factor(results$suitablehosts,
+levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+
+dfpred$agentsF <- factor(dfpred$agentsF,
+levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
+results$agentsF <- factor(results$agentsF,
+levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
+
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.ticks = element_blank(), plot.margin = unit(c(4.5,4.5,0,4.5), "pt"))
+
+ggmultipelcount<- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, fill=agentsF))+geom_bar()+facet_grid(.~suitablehosts, scales="free", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+xlab("")+ylab("frequency")+theme(strip.background = element_blank(), strip.text.x = element_blank(), plot.margin = unit(c(3.5,4.5,0,4.5), "pt"))
+
+ggmulti <- ggmultipleseverity + ggmultipelcount +plot_layout(ncol=1, heights=c(1,0.5))
+ggsave(here("Results/Figures/FigureSeverityMultiple.jpg"), plot=ggmulti  , width=fig.width2,height=fig.width1.5,units="in")
+
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("Bark beetle species present")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")
+
+ggsave(here("Results/Figures/FigureHotSpotSeverity.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1*0.75,units="in")
+
+```
+
+Generally, yes.
+
+When plots contained two tree species susceptible to bark beetles, the presence of both bark beetle species increased tree mortality relative to presence of only one bark beetle species (Fig.
+\@ref(fig:FigHotSpotSeverity)).
+The increase was greatest for stands that contained both MPB and SB; the median severity of plots with both MPB and SB was 33.7 percentage points greater than MPB alone and 38.3 percentage points greater than SB alone.
+The median severity of plots with both MPB and WBBB was 7.0 percentage points greater than MPB alone and 13.6 percentage points greater than WBBB alone.
+The median severity of plots with both SB and WBBB was 14.7 percentage points greater than SB alone and 19.8 percentage points greater than WBBB alone.
+
+Given stand conditions that made all three tree species susceptible to bark beetles, the highest rates of mortality were in plots where both MPB and SB present; whether or not WBBB was also present did not affect mortality severity (~1 percentage point difference in median severity; Fig.
+\@ref(fig:FigHotSpotSeverity)).
+
+Plots with both SB and WBBB experience similar mortality to plots
+
+The median severity of plots with all three agents was 16.2 percentage points greater than only MPB and WBBB and 27.4% percentage points greater than plots with SB and WBBB.
+
+The median severity of plots with all three agents was most similar to plots affected by both
+
+The median severity of plots with all three agents was 27.0 percentage points greater than MPB alone, 34.2 percentage points greater than SB alone, and 32.0% percentage points greater than WBBB alone.
+
+```{r FigHotSpotSeverity, fig.cap = "The severity of bark beetle mortality in plots with multiple tree species susceptible to bark beetles (columns) by the combinatin of bark beetle species present.
+Letters above boxes indicate significant (p < 0.05) differences between groups as determined by a Dunn test, a nonparametric rank sum test.
+The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.",  out.width = "504.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureHotSpotSeverity.jpg"))
+```
+
+# Discussion
+
+# References
+
+<div id="refs"></div>
+
+## Supplement
+### Random Forest Modeling
+```{r plot confusion matrices}
+MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-SMOTE.csv"))
+SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-SMOTE.csv"))
+WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-SMOTE.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "SMOTE"
+BBct1  <- BBct
+
+MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-BALANCED.csv"))
+SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-BALANCED.csv"))
+WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-BALANCED.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "Balanced"
+BBct2  <- BBct
+
+MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-WEIGHTED.csv"))
+SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-WEIGHTED.csv"))
+WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-WEIGHTED.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "Weighted"
+BBct3  <- BBct
+
+BBct <- rbind(rbind(BBct1, BBct2), BBct3)
+
+ggconfused <- ggplot(BBct, aes(x=agent, y=sensitivity, fill=approach))+geom_bar(stat="identity", position="dodge")+theme(legend.position = "none") + ggplot(BBct, aes(x=agent, y=accuracy, fill=approach))+geom_bar(stat="identity", position="dodge")+plot_layout(nrow=2, guides="collect")
+
+ggsave(here("Results/Figures/Confusion.jpg"), plot=ggconfused, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r confusionFig}
+knitr::include_graphics(here("Results/Figures/Confusion.jpg"))
+```
+
+```{r FigPartial,  fig.cap = "A caption", out.width = "504.8pt"}
+MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
+MPBpp$focalagent <- "MPB"
+SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
+SBpp$focalagent <- "SB"
+WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
+WBBBpp$focalagent <- "WBBB"
+
+ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
+ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
+ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
+ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
+
+x <- df[!df$focalagent=="Any",]
+x$propBA <- x$BA.host/x$BA.total
+x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
+x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
+
+xm <- melt(x, id.vars=c("focalagent", "affected01"))
+colnames(xm) <- c("focalagent", "affected01", "var", "value")
+xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
+
+pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
+
+pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
+
+ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value*100, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")
+
+ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
+
+ggsave(here("Results/Figures/PartialDependence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
+```
+
+In general, the probability of each bark beetle species occurring increased with host basal area, quadratic mean diameter, and percent basal area. 
+```{r partialFig}
+knitr::include_graphics(here("Results/Figures/PartialDependence.jpg"))
+```
+
+### Stand structure and composition in stands 
+```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt", eval=T}
+datBB.c <-df[!df$focalhost=="Any",]
+
+QMD.melt <- melt(datBB.c[,c("hosts",paste0("QMD.", c("PIEN", "PICO", "ABLA")))], "hosts")
+QMD.melt$variable <- factor(QMD.melt$variable, levels=paste0("QMD.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
+QMD.melt$facet <- "QMD"
+
+TPH.melt <- melt(datBB.c[,c("hosts",paste0("TPH.", c("PIEN", "PICO", "ABLA")))], "hosts")
+TPH.melt$variable <- factor(TPH.melt$variable, levels=paste0("TPH.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
+TPH.melt$facet <- "TPH"
+
+BA.melt <- melt(datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA")))], "hosts")
+BA.melt$variable <- factor(BA.melt$variable, levels=paste0("BA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
+BA.melt$facet <- "BA"
+
+propBA.melt <-datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA", "total")))]
+propBA.melt$propBA.PIEN <- propBA.melt$BA.PIEN / propBA.melt$BA.total *100
+propBA.melt$propBA.ABLA <- propBA.melt$BA.ABLA / propBA.melt$BA.total *100
+propBA.melt$propBA.PICO <- propBA.melt$BA.PICO / propBA.melt$BA.total *100
+
+propBA.melt <-  melt(propBA.melt[,c("hosts",paste0("propBA.", c("PIEN", "PICO", "ABLA")))], "hosts")
+propBA.melt$variable <- factor(propBA.melt$variable, levels=paste0("propBA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
+propBA.melt$facet <- "BA dominance (%)"
+
+plotdat <- rbind(propBA.melt, TPH.melt, BA.melt, QMD.melt)
+
+ggss <- ggplot(plotdat[plotdat$value>0,], aes(x=hosts, y=value,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BuPu")+ylab("value")+xlab("")+facet_wrap(.~facet, scales="free_y", ncol=1)+theme(legend.position = "bottom", legend.title=element_blank())
+
+ggsave(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"), plot=ggss , width=fig.width1.5, height=fig.width1*2, units="in")
+knitr::include_graphics(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"))
+```
+
+### Maps
+```{r maps all plots and hosts, eval=F}
+df.all <- df[df$focalhost == "Any", ]
+
+df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
+
+states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
+IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+
+ext <- as.matrix(extent(IMW))
+#ext[1,1] <- ext[1,1] -0.25
+#ext[1,2] <- ext[1,2] +0.25
+#ext[2,1] <- ext[2,1] -1
+#ext[2,2] <- ext[2,2] +1
+
+map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.9, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
+
+map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
+
+map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
+
+map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
+
+map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
+
+map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
+
+map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
+
+legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
+
+Fig.file <- here("Results/Figures/Figure-Map-Allplots.jpg")
+jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
+tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
+whatever <- dev.off()
+
+```
+
+```{r generate maps, eval=F}
+bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
+
+df.all <- bbo.dat[bbo.dat$focalhost == "Any", ]
+
+df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
+
+states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
+IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+
+ext <- as.matrix(extent(IMW))
+#ext[1,1] <- ext[1,1] -0.25
+#ext[1,2] <- ext[1,2] +0.25
+#ext[2,1] <- ext[2,1] -1
+#ext[2,2] <- ext[2,2] +1
+
+map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.9, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
+
+map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
+
+map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
+
+map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
+
+map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
+
+map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
+
+map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
+
+legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
+
+Fig.file <- here("Results/Figures/Figure-Map.jpg")
+jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
+tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
+whatever <- dev.off()
+
+```
+
+```{r FigMap, fig.cap="a caption", out.width = "504.8pt"}
+knitr::include_graphics(here("Results/Figures/Figure-Map.jpg"))
+```
+
+```{r map study area of hosts, eval=F}
+f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
+f93 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
+f108 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)
+
+# reclassify to rasters to presence (1) and absence (0)
+f19[f19>0] <- 1 
+f93[f93>0] <- 1
+f108[f108>0] <- 1
+
+# define areas of multiple hosts
+f108f93 <- f108 + f93 # add 
+f108f93[f108f93<2]<- 0 # reclassify
+f108f93[f108f93==2]<- 1 # reclassify
+f108f19 <- f108 + f19
+f108f19[f108f19<2]<- 0 # reclassify
+f108f19[f108f19==2]<- 1 # reclassify
+f93f19 <- f93 + f19
+f93f19[f93f19<2]<- 0  # reclassify
+f93f19[f93f19==2]<- 1 # reclassify
+
+f108f93f19 <- f108 + f93 + f19
+
+# define areas of one host
+f108f93f19re1 <-f108f93f19
+f108f93f19re1 [f108f93f19re1 ==1 ]<-1
+f108f93f19re1 [f108f93f19re1>1]<-0
+
+# define areas of two hosts
+f108f93f19re2 <-f108f93f19
+f108f93f19re2 [f108f93f19re2 ==1 ]<-0
+f108f93f19re2 [f108f93f19re2==3]<-0 
+f108f93f19re2 [f108f93f19re2 ==2 ]<-1
+
+# limit maps of presence (1) and absence (0) to only areas with one host
+f108 <- f108 * f108f93f19re1 
+f93 <- f93 * f108f93f19re1
+f19 <- f19 * f108f93f19re1 
+f93[f93==1]<- 2 # give unique labels
+f19[f19==1]<- 3 # give unique labels
+
+# create maps of two hosts limited to areas with two hosts
+f108f93  <- f108f93 * f108f93f19re2 
+f108f93[f108f93==1]<-4 # give unique labels
+
+f108f19 <- f108f19 * f108f93f19re2
+f108f19[f108f19==1]<-5 # give unique label
+
+f93f19 <- f93f19 * f108f93f19re2
+f93f19[f93f19==1]<-6 # give unique labels
+
+# create maps of all three hosts
+f108f93f19[f108f93f19<=2]<-0
+f108f93f19[f108f93f19==3]<-7 # give unique labels
+
+fplot <- f108 + f93 + f19 + f108f93 + f108f19 + f93f19+ f108f93f19
+st_crs(fplot) <- "EPSG:5070" # assign projection
+
+fplot[fplot==0] <-NA
+
+states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
+IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+notIMW <- states[!states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+IMW <- st_transform(IMW, "EPSG:5070")
+notIMW <- st_transform(notIMW, "EPSG:5070")
+
+ext <- as.matrix(extent(IMW))
+
+sa.tmap <- tm_shape(IMW, bbox=tmaptools::bb(ext)) + tm_fill(col="#f5f5f5")+tm_borders()+tm_shape(fplot) +tm_raster(style = "cat", palette = c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), colorNA=NULL, legend.show = TRUE, title="Hosts                    ", labels = c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"))+ tm_shape(notIMW) + tm_fill(col="white")+tm_compass(position=c(0.95, 0.88), just=c("center", "bottom"))+tm_layout(legend.bg.color="white")+tm_graticules(lines=FALSE)+tm_layout(legend.outside=TRUE,legend.text.size=0.55)+tm_scale_bar(position = c(0.01, 0.001), just="left", breaks=c(0,100),text.size=1.1)+tm_shape(IMW) +tm_borders()
+
+Fig.file <- here("Results/Figures/Figure-Studyarea.jpg")
+jpeg(Fig.file, width=fig.width1.5,height=fig.width1.5,units="in", res=300)
+sa.tmap 
+whatever <- dev.off()
+```
 
@@
 
-## Co-occurrence of agents
-
-```{r FigCoOccurrence, fig.cap = "Co-occurrence",  out.width = "504.8pt"}
-
-knitr::include_graphics( here("Results/Figures/FigureAgentfrequencybyhosts.jpg"))
-```
-
-Few stands were affected by multiple agents.
-
-### Is occurrence more likely or severity greater in stands with multiple hosts?
-
-```{r FigOccurrence, fig.cap = "Severity",  out.width = "504.8pt"}
-
-knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
-```
-
-```{r FigSeverity, fig.cap = "Severity", out.width = "504.8pt"}
-
-knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
-```
-
-# Discussion
-
-# References
-
-<div id="refs"></div>
-
-## Supplement
-### Randon Forest Modeling
-```{r FigPartial,  fig.cap = "A caption", out.width = "504.8pt"}
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
-x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("propBA")
-
-ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-
-ggsave(here("Results/Figures/PartialDependence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
-knitr::include_graphics(here("Results/Figures/PartialDependence.jpg"))
-
-```
-
-### Comparison of stand structure and composition, climate, and regional bark beetle activity  
-```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt", eval=F}
-datBB.c <-df.sub[!df.sub$focalhost=="Any",]
-
-QMD.melt <- melt(datBB.c[,c("hosts",paste0("QMD.", c("PIEN", "PICO", "ABLA")))], "hosts")
-QMD.melt$variable <- factor(QMD.melt$variable, levels=paste0("QMD.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-QMD.melt$facet <- "QMD"
-
-TPH.melt <- melt(datBB.c[,c("hosts",paste0("TPH.", c("PIEN", "PICO", "ABLA")))], "hosts")
-TPH.melt$variable <- factor(TPH.melt$variable, levels=paste0("TPH.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-TPH.melt$facet <- "TPH"
-
-BA.melt <- melt(datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA")))], "hosts")
-BA.melt$variable <- factor(BA.melt$variable, levels=paste0("BA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-BA.melt$facet <- "BA"
-
-propBA.melt <-datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA", "total")))]
-propBA.melt$propBA.PIEN <- propBA.melt$BA.PIEN / propBA.melt$BA.total *100
-propBA.melt$propBA.ABLA <- propBA.melt$BA.ABLA / propBA.melt$BA.total *100
-propBA.melt$propBA.PICO <- propBA.melt$BA.PICO / propBA.melt$BA.total *100
-
-propBA.melt <-  melt(propBA.melt[,c("hosts",paste0("propBA.", c("PIEN", "PICO", "ABLA")))], "hosts")
-propBA.melt$variable <- factor(propBA.melt$variable, levels=paste0("propBA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-propBA.melt$facet <- "BA dominance (%)"
-
-plotdat <- rbind(propBA.melt, TPH.melt, BA.melt, QMD.melt)
-
-ggss <- ggplot(plotdat[plotdat$value>0,], aes(x=hosts, y=value,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BuPu")+ylab("value")+xlab("")+facet_wrap(.~facet, scales="free_y", ncol=1)+theme(legend.position = "bottom", legend.title=element_blank())
-
-ggsave(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"), plot=ggss , width=fig.width1.5, height=fig.width1*2, units="in")
-knitr::include_graphics(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"))
-```
-
-```{r generate maps, eval=F}
-bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
-
-df.all <- bbo.dat[bbo.dat$focalhost == "Any", ]
-
-df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
-
-states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
-IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-
-ext <- as.matrix(extent(IMW))
-#ext[1,1] <- ext[1,1] -0.25
-#ext[1,2] <- ext[1,2] +0.25
-#ext[2,1] <- ext[2,1] -1
-#ext[2,2] <- ext[2,2] +1
-
-map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.9, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
-
-map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
-
-map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
-
-map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
-
-map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
-
-map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
-
-map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
-
-legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
-
-Fig.file <- here("Results/Figures/Figure-Map.jpg")
-jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in")
-tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
-whatever <- dev.off()
-
-```
-
-```{r FigMap, fig.cap="a caption", out.width = "504.8pt"}
-knitr::include_graphics(here("Results/Figures/Figure-Map.jpg"))
-```
+## Quatile comparison
+```{r select plots with large & abundant hosts, eval=F}
+df.sub <- df
+
+PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.1)
+PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.1)
+ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.1)
+
+PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.1)
+PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.1)
+ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.1)
+
+# PICO, PIEN & ABLA
+df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
+
+df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
+
+df.sub.PICO.PIEN.ABLA$suitablehosts2 <- "PICO, PIEN & ABLA"
+df.sub.PICO.PIEN.ABLA$nhosts2 <- 3
+
+# PICO & PIEN
+df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]
+
+df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]
+
+df.sub.PICO.PIEN$suitablehosts2 <- "PICO & PIEN"
+df.sub.PICO.PIEN <- df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
+df.sub.PICO.PIEN$nhosts2 <- 2
+
+# PICO & ABLA
+df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
+
+df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
+
+df.sub.PICO.ABLA$suitablehosts2 <- "PICO & ABLA"
+df.sub.PICO.ABLA$nhosts2 <- 2
+
+df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
+
+# PIEN & ABLA
+df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
+
+df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
+
+df.sub.PIEN.ABLA$suitablehosts2 <- "PIEN & ABLA"
+df.sub.PIEN.ABLA$nhosts2 <- 2
+
+df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
+
+###
+df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
+df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
+df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)
+
+## Affected hosts
+affectedhosts2 <- df.focal2 %>% group_by(suitablehosts2, afp2colfected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectedhosts2[affectedhosts2$affected=="affected",]
+affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts2[affectednhosts2$affected=="affected",]
+
+```

### 2021-10-06 Code/Manuscript-SMOTE.Rmd
updated analyses

@@
 
+Key findings
+
+-
+
 # Keywords
@@
 
-Paragraph 1: Broad applied context
-
-In this context there is a particular need to understand interactions between outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca.
-18 million hectares more than wildfire (van Lierop et al.
-2015).
-
-- insect outbreaks are causing widespread tree mortality in temperate forest worldwide
-- future changes in climate are expected to increase tree mortality
-- forests provide key ecosystem services
-- management
-
-Paragraph 2: Theoretical context - diversity & natural enemies
-
-Many studies have sought to understand how community diversity influences interactions between natural enemies and their resources.
-In the case of plant-herbivore interactions, the greater plant diversity often reduces herbivory effects on plants at both the individual and population level [i.e.
-'associational resistance, @barbosa_associational_2009, @jactel_tree_2021].
-The resource concentration hypothesis suggests that specialists insect herbivores are less likely to find and stay in stands where their hosts are less abundant [@root_organization_1973].
-This may occur if insects are less likely to find a host tree due [i.e.
-'host dilution effect', @] or chemical masking of the focal tree by heterospecific neighbors [i.e.
-'host apparency hypothesis', @Castagneryrol_effects_2014]
-
-[i.e., 'resource concentration hypothesis', @root_organization_1973].
-
-However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
-suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.
-
-1 - associational resistance - the abundance or damage of herbivores is lower when plant diversity is greater [@barbosa_associational_2009, @jactel_tree_2021]
-hypotheses:
-1 - resource concentration: the presence of heterospecific neighbors around a tree of a focal species grown in mixed stands leads to a lower probability of host tree finding by insect herbivores due to lower host abundance or frequency [@root_organization_1973]
-2 - host apparency: 
-3 - decoy: preference for non-host trees
-
-H
-Most studies have focused on individual- or population-level outcomes.
-
-For instance, increased host community diversity may either increase or decrease the susceptibility of individuals to their natural enemies.
-
-At the population-scale these effects may either scale-up or result in counter-intuitive effects.
-However, little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists.
+Paragraph 1: Theoretical context - plant & insect interactions
+
+Many studies have sought to understand how biodiversity influences interactions between natural enemies and their resources.
+In the case of plant-herbivore interactions, greater plant diversity often reduces herbivory on plants at both the individual and population level [i.e., 'associational resistance', @barbosa_associational_2009].
+This is hypothesized to occur when insect herbivores, particularly specialists, are less likely to find and stay in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973].
+This may occur if insects are less likely to encounter their host due to decrease in host frequency or abundance [i.e.
+'host dilution effect', @], heterospecific neighbors chemically mask host trees [i.e.
+'host apparency hypothesis', @castagneryrol_effects_2014], and/or mixed stands favor greater abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020].
+
+#However, increased plant diversity may also result in an increase in herbivore richness ['resource specialization hypothesis', @keddy_plant_1984].
+
+Yet little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists.
 Further, the effects of different natural enemies on their focal host populations often differ greatly, due to differences in susceptibility and mortality rates and the availability and quality (as viewed by their natural enemies) of resource communities.
-Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and natural enemy-resource relationships are highly dynamic.
-
-*** paragraph about irruptive species and community host diversity*** 
-- stand scale
-- landscape scale
-
-To better understand how community diversity influences interactions between natural enemies and their resources, we use a natural system with inherently low resource and natural enemy diversity.
-Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D.
-rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively.
-We use this system to ask whether host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale?
-We hypothesize that
-greater host richness will increase the probability of outbreak occurrence
+Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and plant-herbivore relationships are highly dynamic.
+In this context there is a particular need to understand the effects of tree diversity on outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca.
+18 million hectares more than wildfire [@van_lierop_global_2015], resulting in widespread changes in carbon sequestration [e.g., @kurz_mountain_2008], timber production [e.g.
+@schowalter_insect_2012], and water quality [e.g., @Mikkelson_bark_2013], among other things.
+
+Paragraph 2: Bark beetles
 
 Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year.
+For instance in the western United States, recent outbreaks of native bark beetles have killed 3.8 billion trees [1997-2018; @hicke_characterizing_2020].
 Bark beetles bore through the bark, where they mate and oviposit their eggs.
@@
 To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree.
-Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.For instance, in the continental western United States bark beetles have killed more than XXXX trees over the past X decades.
+Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.
 Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.
@@
 
-To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).
+Paragraph 3: Bark beetles
+
+To better understand how community diversity influences interactions between bark beeetles and their resources, we use a natural system with inherently low tree and bark beetle diversity.
+Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D.
+rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively.
+We use this system to ask:
+
+(1) Is co-occurrence of multiple trees susceptible to bark beetle attack common?
+(2) How often do stand conditions susceptible to multiple bark beetles occur?
+(3) Given stand stand conditions susceptible to multiple bark beetles, does the  the occurrence or severity of cumulative bark beetle mortality
 
@@
 resource concentration).
+
+host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale?
+
+To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).
+Specifically we ask:
 
@@
 ### Model the suitable stand conditions
+
+However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
+suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.
+
 Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data.
@@
 # Results
-## Is co-occurrence of multiple hosts common?
+
+## Do stand conditions suitable for multiple outbreaks commonly co-occur?
 ```{r describe frequency of hosts and agents, eval=T}
@@
 
-```{r}
-gghostgroups <- ggplot(hostgroups, aes(x=hosts, y=freq, fill=hosts))+geom_bar(stat="identity")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")+xlab("")+ylab("frequency")
-
-ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")
-```
-
-Yes, 66% (n=5,850) FIA plots had at least two host species present.
-Subalpine fir was most likely to occur with other hosts (50% of stands), followed by Engelmann spruce (45% stands), and lodgepole pine (36% of stands) (Fig.
-\@ref(fig:FigHostFreq)).
-
-```{r FigHostFreq, fig.cap = ".", out.width = "322.56pt"}
-knitr::include_graphics( here("Results/Figures/FigureHostFreq.jpg"))
-```
-
-## Is co-occurrence of the stands conditions suitable for multiple bark beetle species common?
-
-20% of stands were suitable for more than one bark beetle species. Conditions suitable for WBBB were most likely to co-occur with conditions suitable for another bark beetle (17% of stands)
-
 ```{r determine co-occurrence of suitable stands}
@@
 
+suitablehostgroups <- bbo.sub %>% group_by(suitablehostsP) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+
+```
+
+```{r}
+gghostgroups <- ggplot(hostgroups, aes(x=hosts, y=freq, fill=hosts))+geom_bar(stat="identity")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")+xlab("")+ylab("frequency")
+
 ggfreq <- ggplot(bbo.sub, aes(x=suitablehostsP,fill=suitablehostsP))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")
@@
 ggsave( here("Results/Figures/FigureSuitable.jpg"), plot=ggfreq,width=fig.width1,height=fig.width1*0.75,units="in")
-```
-
-(Fig. \@ref(fig:FigureSuitable)).
-
-```{r FigureSuitable, fig.cap = ".", out.width = "322.56pt"}
-knitr::include_graphics( here("Results/Figures/FigureSuitable.jpg"))
+
+ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")
+
+colnames(suitablehostgroups) <- colnames(hostgroups)
+hostgroups$var <- "individual host"
+suitablehostgroups$var <- "suitable host"
+
+hosts.plot.dat <- rbind(hostgroups, suitablehostgroups)
+
+gghostgroups <- ggplot(hosts.plot.dat, aes(x=hosts, y=n, fill=var))+geom_bar(stat="identity", position="dodge")+scale_fill_manual(values=c("grey", "black"))+ylab("frequency")+theme(legend.position="bottom")+theme(legend.title=element_blank(), axis.title.x=element_blank())
+ggsave( here("Results/Figures/FigureHostCo.jpg"), plot=gghostgroups,   width=fig.width1, height=fig.width1*0.75, units="in")
+```
+
+Across the 10,439 FIA plots with one of the target host species, 66% (n=5,850) contained at least two of the target host species (Fig.
+\@ref(fig:FigHostFreq)).
+Subalpine fir was most likely to occur with other target host species (50% of plots), followed by Engelmann spruce (45% plots), and lodgepole pine (36% of stands).
+However, only 21% of all plots (n=2,163) were suitable for more than one bark beetle species.
+Conditions suitable for infestation in subalpine fir were most likely to co-occur with conditions suitable for infestation in another tree species (17% of stands).
+
+```{r FigHostFreq, fig.cap = "The frequency of plots by host identity.
+Gray bars indicate host species were present (i.e., at least one individual of the focal species was present within the plot).
+Black bars indicate  stand structure and composition conditions were conducive to bark beetle infestation in focal tree species.
+For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "244.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureHostCo.jpg"))
 ```
@@
 
-ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.66,units="in")
-
-```
-
-Given suitable stand conditions for multiple agents, the probability of occurrence  of any agent was significantly greater than the probability of a single agent \@ref(fig:FigureSuitable).
-```{r FigOccurrence, fig.cap = "Severity",  out.width = "504.8pt"}
+ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.5,units="in")
+
+```
+
+Given suitable stand conditions for multiple agents, the probability of occurrence of bark beetle attributed mortality was significantly greater than the probability of a single bark beetle species \@ref(fig:FigureOccurrence).
+
+```{r FigOccurrence, fig.cap = "The proportion of plots affected by the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or any bark beetle (i.e., MPB, SB or WBBB) by the identity of hosts species susceptible to bark beetle infestation.
+For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.
+Astericks above bars indiciate that proportion plots affected by the individual bark beelte species was significantly less than the proportion of plots affected by any bark beetle species.",  out.width = "504pt"}
 knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
@@
 
-Given suitable stand conditions for multiple agents, is the severity of cumulative bark beetle-driven tree mortality greater in stands with multiple hosts?
-
-In stands suitable for all three hosts, cumulative mortality was significantly greater than mortality by a single agent.
-\@ref(fig:FigSeverity).
-When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most common agent (MPB > SB > WBBB) and was not significantly different than the cummualtive mortality.
+## Is the severity of cumulative bark beetle-driven tree mortality greater in stands with multiple susceptible hosts?
 
@@
 
-ggseverity <- ggplot(dfpred, aes(x=focalagent, y=propTBA*100, fill=focalagent, colour=focalagent))+geom_boxplot(notch=T)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")
+dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)
+
+ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")
 
 ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
-
-```
-
-```{r FigSeverity, fig.cap = "Severity", out.width = "504.8pt"}
+```
+
+In stands susceptible for all three hosts, cumulative mortality was significantly greater than mortality by a single bark beetle species (Fig.
+\@ref(fig:FigSeverity).
+When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most commonly occurring agent (MPB > SB > WBBB) and was not significantly different than the cumulative mortality (Fig.
+\@ref(fig:FigSeverity).
+
+```{r FigSeverity, fig.cap= "Boxplots illustrating the severity of mortality (% of total basal area) attributed to the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or all bark beetles (i.e., cumulative mortality due to MPB, SB and WBBB) by the identity of hosts species susceptible to bark beetle infestation.
+For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.
+The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.
+Astericks above boxes indicate that severity bark beetle mortality attributed to the individual species was significantly less than the cumulative mortality due to all bark beetles.", out.width = "504pt"}
 knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
@@
 
-## Given suitable stand conditions, is co-occurrence of multiple agents common?
-
-Not really, only 21.3% (n=461) of stands suitable for multiple agents were affected by multiple agents.
-When multiple agents occur, the most commonly occurring agents was MPB (n=341), followed by SB (n=317) and WBBB (n=308).
-The most commonly occurring combinations of agents was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and then SB and WBBB (26% of case; n=120).
-The combination of all three agents was rare (9.5% of cases; n=44).
+## Is the  co-occurrence of multiple agents common?
+
+Across all 10,439 FIA plots, only 5.3% (n=539) experience mortality due to multiple bark beetle species.
+However, MPB, SB, and WBBB were only present in 23.4% (n=2,441), 9.9% (n=1034), and 7.8% (n=813) of plots,respectively.
+Thus the number of plots with at least two bark beetle species co-occurring is more than 2x greater than would be expected if MPB, SB, WBBB occurred independently.
+
+Given only the 2,163 stands where multiple tree species were susceptible to bark beetle infestation, 21.3% (n=461) were affected by multiple agents.
+When multiple tree species were susceptible to bark beetle infestation, the most commonly occurring agents was MPB (n=341), followed by SB (n=317) and WBBB (n=308).
+The most commonly occurring combinations of agents was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and SB and WBBB (26% of case; n=120).
+The combination of all three agents was extremely rare (9.5% of cases; n=44).
 
 ```{r determine co-occurrence of agents by hosts}
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+datBB.c <- df[df$focalhost=="Any",]
+table(datBB.c$agents)
+# nrow(datBB.c[datBB.c$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,])/nrow(datBB.c) 
+# 5.3 % of plots were affected by multiple agents
+
+table(datBB.c$agentSB)[2]/nrow(datBB.c) * table(datBB.c$agentMPB)[2]/nrow(datBB.c) * table(datBB.c$agentWBBB)[2]/nrow(datBB.c)
+
+table(datBB.c$agentSB)[2]/nrow(datBB.c) * table(datBB.c$agentMPB)[2]/nrow(datBB.c)
+
+datBB.c[datBB.c$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+table(datBB.c$agentSB)
+
+dfpred <- merge(BBpred, datBB.c, id.vars="PLT_CN", all=T)
 dfpred <- dfpred[!dfpred$agentsuitable=="none",]
 
-dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+#dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
 
@@
 
-dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())
+#dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())
 
 MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-sum(MPBx$n)
-sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+#sum(MPBx$n)
+#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
 
 SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-sum(SBx$n)
-sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+#sum(SBx$n)
+#sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
 
 WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-sum(WBBBx$n)
-sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+#sum(WBBBx$n)
+#sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
 
 ### 
-table(dfpred$suitablehostsF, dfpred$agents)
+#table(dfpred$suitablehostsF, dfpred$agents)
 x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
@@
 
-co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]
+#co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]
+
+ggAgentsbySuitablehosts <- ggplot(dfpred[dfpred$focalhost=="Any",], aes(x=suitablehostsF,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+scale_fill_manual(values=p2col)+theme(axis.title.x=element_blank())
+
+ggsave(here("Results/Figures/Figure-AgentsbySuitablehosts.jpg"), plot=ggAgentsbySuitablehosts , width=fig.width1.5,height=fig.width1*0.75,units="in")
+```
+
+```{r determine co-occurrence of agents by suitable hosts}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred <- dfpred[!dfpred$agentsuitable=="none",]
+
+#dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+dfpred$suitablehostsF <-dfpred$suitablehosts
+dfpred$suitablehostsF <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+dfpred$agentsFplongrev <- factor(dfpred$agents,levels=c("MPB+SB+WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB", "WBBB", "SB", "MPB", "none"), labels=c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB", "WBBB", "SB", "MPB", "none"), ordered=T)
+
+#dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())
+
+MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+#sum(MPBx$n)
+#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+#sum(SBx$n)
+#sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+#sum(WBBBx$n)
+#sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
+
+### 
+#table(dfpred$suitablehostsF, dfpred$agents)
+x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+xPICO <-x[x$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA",  "PICO,\nPIEN,\n&\nABLA"),]
+#
+
+## Agents given presence of any agent
+co <- dfpred[!dfpred$focalhost=="Any" ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+#co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]
 
@@
 ##  Is severity greater in stands with multiple agents?
-```{r multiple agents}
+```{r multiple agents, eval=F}
 library(rcompanion)
@@
 
-Generally, yes.
-
 When plots contained two tree species susceptible to bark beetles, the presence of both bark beetle species increased tree mortality relative to presence of only one bark beetle species (Fig.
@@
 
-pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
+
+pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
 
@@
 
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value*100, y=yhat, col=focalagent))+geom_smooth()+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value*100, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
 
@@
 
-In general, the probability of each bark beetle species occurring increased with host basal area, quadratic mean diameter, and percent basal area. 
-```{r partialFig}
+Consistent with existing stand hazard ratings, the probability of each bark beetle species occurring generally increased with host basal area, quadratic mean diameter, and percent basal area (Fig.
+\@ref(fig:partialFig)).
+However, both MPB and WBBB exhibited a slight decrease in the probability of outbreak at very high basal areas and all three bark beetle species exhibited a lower probability of outbreak in stands where host trees were very large (i.e.
+high QMD values).
+WBBB also exhibited a decline in the probability of outbreak when subalpine fir comprised a relatively high proportion of the basal area.
+
+```{r partialFig, fig.cap="The associations between the probability of occurrence of the mountain pine beetle (MPB; red), spruce beetle (SB; blue); and western balsam bark beetle (WBBB; yellow) and three common measuree of stand suceptibility to bark beetl beetle outbreak, basal area (BA), quadratic mean diameter (QMD), and host dominace (% BA).
+The top row illustrates smoothed partial dependence results from Random Forest modeling, while in the bottom row boxplots illustrate the distribution of values for stands affected (dark colors) and unaffected (light colors) by each bark beetle species."}
 knitr::include_graphics(here("Results/Figures/PartialDependence.jpg"))

### 2021-11-22 Code/Manuscript-2.Rmd
including multiple ways for determining presence

---
title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
author: "Sarah Hart"
email: "sarah.hart@colostate.edu"
date: '`r format(Sys.time(), "%B %d, %Y")`'
output: 
  bookdown::word_document2:
    reference_docx: WordTemplate.docx
    fig_caption: yes
    toc: no
    number_sections: no
    df_print: kable
csl: "`r here:::here('ecology.csl')`"
bibliography: "`r here:::here('MyLib.bib')`"
urlcolor: blue
linkcolor: blue
citationcolor: blue
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = F,
	message = FALSE,
	warning = FALSE,
	dpi = 300,
	progress = FALSE,
	cache = FALSE
)

ipak <- function(pkg){
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg)>0) {
             install.packages(new.pkg, dependencies = TRUE)
    }
    easypackages::libraries(pkg)
}

options(repos = c(CRAN = "http://cran.rstudio.com"))

### Import libraries
ipak(c('tidyverse', 'tidymodels', 'reshape2', 'here', 'rgdal', 'sp', 'raster', 'stars', 'sf', 'prism', 'tigris', "tmap", 'fasterize', 'ggplot2','RColorBrewer','ggsci', 'patchwork', 'colorspace', 'grid', 'knitr',  'flextable', 'bookdown',  'pdp', 'ranger', 'ALEPlot', 'VSURF', 'themis', 'foreach',  'FSA', 'parallel'))

# Set seed
set.seed(2020)

# Set number of cores for parralell processing
cores <- parallel::detectCores()

# Set custom plotting theme
theme_new <- function(base_size = 9,base_family = "Helvetica"){
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.line.x = element_line(color="black", size = 0.25),
      axis.line.y = element_line(color="black", size = 0.25),
      axis.title = element_text(size = 9),
      axis.text = element_text(colour="black", size=8),
      legend.key=element_rect(colour=NA, fill =NA),
      panel.grid = element_blank(),   
      plot.background = element_rect(fill = NA, colour = NA),
      panel.border = element_rect(fill = NA, colour = NA),
      panel.background = element_rect(fill = "white", colour = "black"), 
      strip.background = element_rect(fill = "white"),
      strip.text = element_text(size = 9)

    )
}
theme_set(theme_new())

options(ggplot2.continuous.colour="viridis")
scale_colour_discrete <- scale_fill_npg
scale_fill_discrete <- scale_fill_npg
options(scipen=999)

p1col=c(rgb(104, 17, 10,  maxColorValue = 255), rgb(79, 130, 156, maxColorValue = 255), rgb(243, 173, 1, maxColorValue = 255), rgb(112, 146, 115,  maxColorValue = 255), rgb(112, 146, 115, maxColorValue = 255))

p1fill=c(rgb(104, 17, 10, alpha=230, maxColorValue = 255), rgb(79, 130, 156, alpha=230, maxColorValue = 255), rgb(243, 173, 1, alpha=50, maxColorValue = 255), rgb(112, 146, 115, alpha=230,  maxColorValue = 255), rgb(112, 146, 115, alpha=230, maxColorValue = 255))

p2col =c(rgb(20,20,20, maxColorValue = 255), rgb(54, 124, 111, maxColorValue = 255), rgb(235, 137, 52, maxColorValue = 255), rgb(83, 74, 90, maxColorValue = 255), rgb(232, 177, 60, maxColorValue = 255),rgb(90, 129, 154, maxColorValue = 255),rgb(113, 0, 2, maxColorValue = 255), rgb(166, 166, 166,  maxColorValue = 255))

fig.width1 <- 3.4
fig.width2 <- 7
fig.width1.5 <- 4.48

pt.width1 <-fig.width1 * 72
pt.width2 <-fig.width2 * 72
pt.width1.5 <- fig.width1.5 * 72

# Set projection - NAD83 / Conus Albers
aea.proj <- "EPSG:5070"

options(timeout=60*30) #timeout downloads that last longer than 30 minutes

# Set directory structure for project
dir.create(here("Data/"), showWarnings = FALSE)
dir.create(here("Results/"), showWarnings = FALSE)
dir.create(here("Results/Figures/"), showWarnings = FALSE)
```
# Abstract

## Key findings

# Keywords

# Introduction

In the western United States, native bark beetles (Curculionidae: Scolytinae) have killed 3.8 billion trees over the past two decades [1997-2018; @hicke_characterizing_2020]. While across the region there are more than 25 native bark beetle species, the mountain pine beetle (MPB; *Dendroctonus ponderosae*), spruce beetle (SB;  *Dendroctonus rufipennis*), and western balsam bark beetle (WBBB; *Dryocoetes confusus*) collectively have caused over 70% of the bark beetle-attributed tree mortality, most of which has occurred in high elevation forests [@hicke_chacterizing_2020].

In the case of plant-herbivore interactions, greater plant diversity often reduces the effects of herbivory [i.e., 'associational resistance', @barbosa_associational_2009], particularly when herbivores are specialists capable of rapid population growth [@otway_resource_2005]. Associational resistance is commonly hypothesized to occur due to several non-mutually exclusive mechanisms. First, herbivory may be lower in more diverse systems if herbivores are less likely to encounter, stay, and reproduce in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973]. Second, reductions in herbivory may occur if insects are less likely to find their hosts due to chemical or visual masking by heterospecific neighbors [i.e. 'host apparency hypothesis', @feeny_plant_1976, @sciebe_semiochemical_2011, @castagneryrol_effects_2014], Third, heterospecific neighbors may increase the abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020]. However, evidence for associational resistance in natural (i.e. not plantations) forested ecosystems is limited.

While increases in plant diversity may decrease the effects of herbivore species on individuals and populations of plants, it is also hypothesized to increase herbivore richness ['resource specialization hypothesis', @keddy_plant_1984], which may result in more efficient resource use and greater depletion of the plant community, especially when herbivores are specialists [@cardinale_effects_2005; @finke_niche_2008].

Yet little research has examined the effects of plant diversity on community-level outcomes of herbivory, which critically may drive key ecosystem processes. This is particularly true in natural systems, where the number of potential interactions is high, the effects of different herbivores on their focal host populations often differ greatly, and the changes in plant diversity are often convoluted with changes in host quality [@jactel_tree_2021].

Resource specialization 
Compensatory mechanisms
- only meaningful if populations susceptible to bb infestionat co- occur with tree populatoin ssucess to other

To better understand interactions between tree and bark beetle communities in subalpine forests, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/). Specifically, we ask:

(1) How often do stand conditions suitable for multiple bark beetle species occur?
(2) Does host species richness effect the occurrence or severity of cumulative bark beetle activity increase with the host species richness?
(3) Does the co-occurrence of multiple bark beetle species result in more severe cumulative bark beetle mortality?

We expect that co-occurrence of trees of different host species will be common (i.e. resource specialization hypothesis), but co-occurrence of stand conditions suitable for multiple bark beetle species will not (i.e. associational resistance).

When stand conditions are suitable for multiple bark beetle species, we expect that the

will be greater (i.e. resource specialization hypothesis), but severity will be lower (i.e. associational resistance). Alternatively, if

We also expected the severity of bark beetle infestation to vary with the number of agents present, but with two alternative hypotheses. If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent. Alternatively, lower tree mortality may occur if concurrent outbreaks of bark beetles of different species cause semiochemical confusion or if competitive release increases tree defensive capacity.

# Methods

## Study area
```{r set species and agent codes}
# SPECIES CODES

# MPBhosts
PICO <- 108
PIAL <- 101
PIAR <- 102
PIFL <- 113
PIST <- 114
PIMO <- 119
PIPO <- 122

# SBhosts
PIEN <- 93
PIPU <- 96

#SFDhosts
ABLA <-19 
ABCO <- 15
ABGR <- 17

ohostsMPB <- c( PIAL, PIAR, PIFL, PIST, PIMO, PIPO)
ohostsWBBB <- c(ABCO, ABGR)
ohostsSB <- PIPU
otherhosts <- c(ABCO, ABGR, PIPU, PIAL, PIAR, PIFL, PIST, PIMO, PIPO)

# AGENT CODES
# DCA
bb <- 11000 #bark beetle
mpb <- 11006 #mountain pine beetle
sb <- 11009 #spruce beetle
wbbb <- 11015 # western balsam bark beetle
sfm <- 80002 # subalpine fir mortality
```

The study area consists of subalpine forests with lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) that are located in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, and Wyoming;  Fig. \@ref(fig:FigMap)). Our study area was constrained to the Intermountain West due to regional differences in data collection that exist within the FIA program [@burrill_forest_2017].

## Data

The FIA program inventories all public and private forested land using a spatially and temporally distributed sampling design, making it useful for the study of tree species distributions [e.g., @iverson_predicting_1998; @rehfeldt_empirical_2006] and tree demography [e.g., @shaw_forest_2005, @derose_effects_2013]. In the Western US, all plots are visited once every ten years [@gray_forest_2012]. At each plot, field crews collect data within four 7.32 m radius subplots arranged in a fixed pattern. For each tree within the plot, field crews record the species, size (diameter at breast height; DBH), and status (live or dead) for trees (>= 12.7 cm DBH).For each live tree, field crews additionally record up to three damaging agents (e.g. MPB), defined as any agent likely to prevent the tree from surviving >2 years, reduce the growth of the tree in the near term, or negatively affect the tree's marketable products [@burrill_forest_2017]. For dead trees that were alive at the previous visit, the proximate cause of death is also collected for any tree that was alive at the previous visit and at revisit is dead using visible evidence (e.g., fire scars, bark beetle galleries). Because identifying the mortality agent is more challenging, cause of death codes are broad (e.g., "insect" or "disease").

We acquired all FIA data for the Intermountain West, but restricted our analyses to the most recent inventory for plots that contained a focal host species and where all subplots were inventoried. We then characterized stand structure and composition using only live and recently killed (i.e. killed within the past ~10 years) trees. each plot we calculated the basal area, quadratic mean diameter (QMD), and basal area dominance (% total basal area) for each host species. Additionally, we characterized the presence (i.e., at least one tree infested) and severity of bark beetle activity (% total basal area) by bark beetle species. For live Engelmann spruce, we listed the presence of SB when the agent was listed as a general bark beetle or SB specifically. Similarly, we listed the presence of MPB in lodgepole when agent was listed as a general bark beetle or MPB specifically. For live subalpine fir, we listed the presence of WBBB when the damaging was recorded as, subalpine fir decline, a general bark beetle, or WBBB specifically. Finally, we assumed the mortality of lodgepole pine, Engelmann spruce, and subalpine fir that was attributed an insect, was caused by the MPB, SB, and WBBB, respectively.

```{r download FIA data, eval=F}
dir.create(here("Data/"), showWarnings = FALSE)
dir.create(here("Data/FIA"), showWarnings = FALSE)
# reference data
dir.create(here("Data/FIA/Reference"), showWarnings = FALSE)
temp <- here("/Data/FIA/Reference/FIADB_REFERENCE.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/FIADB_REFERENCE.zip", temp)
unzip(temp, exdir=here("/Data/FIA/Reference"))

# AZ
dir.create(here("Data/FIA/AZ"), showWarnings = FALSE)
temp <- here("/Data/FIA/AZ/AZ.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/AZ.zip", temp)
unzip(temp, exdir=here("/Data/FIA/AZ"))

# CO
dir.create(here("Data/FIA/CO"), showWarnings = FALSE)
temp <- here("/Data/FIA/CO/CO.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/CO.zip", temp)
unzip(temp, exdir=here("/Data/FIA/CO"))

# ID
dir.create(here("Data/FIA/ID"), showWarnings = FALSE)
temp <- here("/Data/FIA/ID/ID.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/ID.zip", temp)
unzip(temp, exdir=here("/Data/FIA/ID"))

# MT
dir.create(here("Data/FIA/MT"), showWarnings = FALSE)
temp <- here("/Data/FIA/MT/MT.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/MT.zip", temp)
unzip(temp, exdir=here("/Data/FIA/MT"))

# NM
dir.create(here("Data/FIA/NM"), showWarnings = FALSE)
temp <- here("/Data/FIA/NM/NM.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/NM.zip", temp)
unzip(temp, exdir=here("/Data/FIA/NM"))

# NV
dir.create(here("Data/FIA/NV"), showWarnings = FALSE)
temp <- here("/Data/FIA/NV/NV.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/NV.zip", temp)
unzip(temp, exdir=here("/Data/FIA/NV"))

# UT
dir.create(here("Data/FIA/UT"), showWarnings = FALSE)
temp <- here("/Data/FIA/UT/UT.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/UT.zip", temp)
unzip(temp, exdir=here("/Data/FIA/UT"))

# WY
dir.create(here("Data/FIA/WY"), showWarnings = FALSE)
temp <- here("/Data/FIA/WY/WY.zip")
download.file("https://apps.fs.usda.gov/fia/datamart/CSV/WY.zip", temp)
unzip(temp, exdir=here("/Data/FIA/WY"))
```

```{r download spatial data, eval=F}
# states
dir.create(here("Data/Spatial/"), showWarnings = FALSE)
dir.create(here("Data/Spatial/States"), showWarnings = FALSE)
temp <- here("/Data/Spatial/States/cb_2018_us_state_20m.zip")
download.file("https://www2.census.gov/geo/tiger/GENZ2018/shp/cb_2018_us_state_20m.zip", temp)
unzip(temp, exdir=here("/Data/Spatial/States"))

# usfs western bark beetle strategy
dir.create(here("Data/Spatial/USFS/"), showWarnings = FALSE)
temp <- here("/Data/Spatial/USFS/WBBS_PL.gdb.zip")
download.file("https://data.fs.usda.gov/geodata/edw/edw_resources/fc/S_USA.Activity_WBBS_PL.gdb.zip", temp)
unzip(temp, exdir=here("/Data/Spatial/USFS/"))

# usfs timber harvest
dir.create(here("Data/Spatial/USFS/"), showWarnings = FALSE)
temp <- here("/Data/Spatial/USFS/Activity_TimberHarvest.gdb.zip")
download.file("https://data.fs.usda.gov/geodata/edw/edw_resources/fc/S_USA.Activity_TimberHarvest.gdb.zip", temp)
unzip(temp, exdir=here("/Data/Spatial/USFS/"))
```

```{r import raw data, eval=F}
wd <- here("Data/FIA")

# List directories containing data for each state
dirs <- list.dirs(wd)

dirs <- dirs[c(-1, -8)]

# Create empty lists to hold dataframe for each state
dats <- list() 
datp <- list()
datc <- list()
datt <- list()
datspc <- list()

for(i in 1:length(dirs)){
  state <- strsplit(dirs[i], split="FIA/")[[1]][2]
  # Survey 
  dats[[state]] <- read.csv(paste0(dirs[i], "/",state, "_SURVEY.csv"), stringsAsFactors = F)
  # Plot
  datp[[state]] <- read.csv(paste0(dirs[i], "/",state, "_PLOT.csv"), stringsAsFactors = F)
  # Condition
  datc[[state]] <- read.csv(paste0(dirs[i], "/",state, "_COND.csv"), stringsAsFactors = F)
  # Tree
  datt[[state]] <- read.csv(paste0(dirs[i], "/",state, "_TREE.csv"), stringsAsFactors = F)
  # Subplot Condition
  datspc[[state]] <- read.csv(paste0(dirs[i], "/",state, "_SUBP_COND.csv"), stringsAsFactors = F)
}

### Merge data
datsm <- do.call(rbind, dats)
datpm <- do.call(rbind, datp)
datcm <- do.call(rbind, datc)
dattm<- do.call(rbind, datt)
datspcm<- do.call(rbind, datspc)

remove(datt, datp, datc, dats,datspc)

# Merge Subplot & Plot data
df <- merge(datcm[,c("PLT_CN", "CONDID","COND_STATUS_CD","OWNCD", "STDAGE")], datspcm[,c("PLT_CN", "CONDID", "SUBP")], by=c("PLT_CN", "CONDID"))

# Merge Survey & Plot data
datz <- merge(datsm[,c("CN", "ANN_INVENTORY", "STATEAB")], datpm[,c("CN","SRV_CN","INVYR", "COUNTYCD","PLOT", "PLOT_STATUS_CD", "KINDCD", "DESIGNCD", "MANUAL", "SAMP_METHOD_CD", "REMPER", "MEASYEAR", "ELEV", "LAT", "LON", "PREV_PLT_CN", "UNITCD")],  by.y="SRV_CN", by.x="CN", all==TRUE)
colnames(datz)[which(colnames(datz) %in% c("CN", "CN.y"))] <- c("SRV_CN", "PLT_CN")

#confirm each plot record is in data frame only once 
if(max(sort(table(datz$PLT_CN)))!=1){
  print("stop there are duplicated plot records")
}
datz.sp<-datz
```

```{r merge survey, plot, condition & tree data, eval=F}
## select annual inventory
datz <-datz[datz$ANN_INVENTORY=='Y',]; nrow.annual <- nrow(datz)

## select plots where all of the subplots are forested
allspforested <- unique(df$PLT_CN[!(df$PLT_CN %in% subset(aggregate(COND_STATUS_CD~PLT_CN, FUN=function(x){length(x[x!=1])}, df), COND_STATUS_CD>0)$PLT_CN)])
allspforested <- allspforested[allspforested %in% datz$PLT_CN]
datz <- datz[datz$PLT_CN %in% allspforested, ]

#Merge Survey-Plot-Condition & Tree data
dattm.sub <-subset(dattm, PLT_CN %in% datz$PLT_CN)
datz <- merge(datz, dattm.sub[, c("PLT_CN", "CN", "STATECD", "COUNTYCD","TREE", "DIST", "STATUSCD", "DIA", "SPCD","CCLCD", "AGENTCD", "DAMAGE_AGENT_CD1","DAMAGE_AGENT_CD2", "DAMAGE_AGENT_CD3", "TPA_UNADJ")],  by="PLT_CN", all=FALSE)
datz$PLOT.STATEAB <- paste(datz$PLOT, datz$STATEAB, sep=".")

# subset data to only trees greater than 5 inches (12.7 cm) DBH
datz5 <- subset(datz, DIA>=5)

# subset data to only trees within subplot
datz5 <- subset(datz5, DIST<=24)

# remove trees with missing values for TPA count 
datz5 <- datz5[is.na(datz5$TPA_UNADJ)==FALSE, ] ## not sure why missing values occur here
datz5.bak <- datz5

spruce.plots <-unique(datz5[datz5$SPCD == PIEN,"PLT_CN"])
pine.plots <-unique(datz5[datz5$SPCD == PICO,"PLT_CN"])
fir.plots <- unique(datz5[datz5$SPCD == ABLA,"PLT_CN"])
datz5 <- datz5[datz5$PLT_CN %in% c(spruce.plots, pine.plots, fir.plots),]

## subset to recently killed trees
datz5.bak <- datz5
datz5$STATUSCD <- factor(datz5$STATUSCD , levels=c(0:3), labels=c("No Status", "Live", "Dead", "Removed"))
datz5$STATUSCD.2<- ifelse(datz5$STATUSCD=="Dead" & is.na(datz5$AGENTCD) == FALSE, "Recently killed",as.character(datz5$STATUSCD))
datz5 <- datz5[datz5$STATUSCD.2 %in% c("Live", "Recently killed"),]
```

```{r determine presence of BB agents in focal hosts, eval=F}
datz5$WBBB.l <- NA
datz5$WBBB.d <- NA
datz5$WBBB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD==ABLA |datz5$DAMAGE_AGENT_CD1 %in% c(wbbb, sfm) & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD == ABLA, 1, 0) 
datz5$WBBB <- rowSums(datz5[,c("WBBB.d", "WBBB.l")], na.rm=T)
datz5$WBBB[is.na(datz5$WBBB)==T]<-0

datz5$MPB.l <- NA
datz5$MPB.d <- NA
datz5$MPB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD ==PICO | datz5$DAMAGE_AGENT_CD1 ==11006 & datz5$SPCD ==PICO, 1, 0) 
datz5$MPB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD ==PICO , 1, 0) 
datz5$MPB <- rowSums(datz5[,c("MPB.d", "MPB.l")], na.rm=T)
datz5$MPB[is.na(datz5$MPB)==T]<-0

datz5$SB.l <- NA
datz5$SB.d <- NA
datz5$SB.l <- ifelse(datz5$DAMAGE_AGENT_CD1==bb & datz5$SPCD == PIEN | datz5$DAMAGE_AGENT_CD1==11009 & datz5$SPCD == PIEN, 1, 0) 
datz5$SB.d <- ifelse(datz5$AGENTCD==10 & datz5$SPCD == PIEN, 1, 0)  
datz5$SB <- rowSums(datz5[,c("SB.d", "SB.l")], na.rm=T)
datz5$SB[is.na(datz5$SB)==T]<-0
```

```{r stand characteristics, eval=F}
## BA
datz5$BA <- datz5$DIA*datz5$DIA * pi/(4*144) *datz5$TPA_UNADJ # sq ft per acre
datz5$BA <- datz5$BA * 1/10.7639 * 2.47105/1 # m2 per ha
BA <- aggregate(BA~PLT_CN+SPCD, datz5, FUN=sum, na.rm=T)
BA.dat <- dcast(BA, PLT_CN~SPCD, value.var="BA", drop=F)
BA.dat$BA.total <- rowSums(BA.dat[,-1], na.rm=T)
BA.dat$BA.Allhosts <- rowSums(BA.dat[,as.character(c(PICO, PIEN, ABLA))], na.rm=T)
BA.dat$BA.Otherhosts <- rowSums(BA.dat[,as.character(otherhosts)], na.rm=T)
BA.dat <- BA.dat[, c("PLT_CN", PICO, PIEN, ABLA, "BA.total", "BA.Allhosts", "BA.Otherhosts")]
colnames(BA.dat) <- c("PLT_CN", paste0("BA.", c("PICO", "PIEN", "ABLA", "total", "Allhosts", "Otherhosts")))
BA.dat[is.na(BA.dat)==TRUE] <- 0

## Trees per ha
datz5$TPH <- datz5$TPA_UNADJ * 2.47105 # trees per ha
TPH <- aggregate(TPH~PLT_CN+SPCD, datz5, FUN=sum, na.rm=T)
TPH.dat <- dcast(TPH, PLT_CN~SPCD, value.var="TPH", drop=F)
TPH.dat$TPH.total <- rowSums(TPH.dat[,-1], na.rm=T)
TPH.dat$TPH.Allhosts <- rowSums(TPH.dat[,as.character(c(PICO, PIEN, ABLA))], na.rm=T)
TPH.dat <- TPH.dat[, c("PLT_CN", PICO, PIEN, ABLA, "TPH.total", "TPH.Allhosts")]
colnames(TPH.dat) <- c("PLT_CN", paste0("TPH.", c("PICO", "PIEN", "ABLA", "total", "Allhosts")))
TPH.dat[is.na(TPH.dat)==TRUE] <- 0

## QMD
datz5$DIAcm <- datz5$DIA * 2.54 # convert to cm
qmd <- aggregate(DIAcm~PLT_CN+SPCD, datz5, FUN=function(x){sqrt(mean(x*x,na.rm=T))})
qmd.dat<- dcast(qmd, PLT_CN~SPCD , value.var="DIAcm", drop=F)
qmd.dat<- qmd.dat[,c("PLT_CN", PICO, PIEN, ABLA)]
colnames(qmd.dat) <- c("PLT_CN", paste0("QMD.",c("PICO", "PIEN", "ABLA")))
qmd.dat.hosts <- aggregate(DIAcm~PLT_CN, datz5[datz5$SPCD %in% c(PICO, PIEN, ABLA),], FUN=function(x){sqrt(mean(x*x,na.rm=T))})
colnames(qmd.dat.hosts) <- c("PLT_CN", "QMD.Allhosts")
qmd.dat.all <- aggregate(DIAcm~PLT_CN, datz5, FUN=function(x){sqrt(mean(x*x,na.rm=T))})
colnames(qmd.dat.all) <- c("PLT_CN", "QMD.total")
qmd.dat<-merge(qmd.dat, qmd.dat.hosts, by="PLT_CN", all=T)
qmd.dat<-merge(qmd.dat, qmd.dat.all, by="PLT_CN", all=T)
qmd.dat[is.na(qmd.dat)] <- 0

## SDI (additive)
datz5$SDIi <- (datz5$DIA/10)^1.6
sdi <- aggregate(SDIi~PLT_CN+SPCD, datz5, FUN=sum)
sdi.dat<- dcast(sdi, PLT_CN~SPCD , value.var="SDIi", drop=F)
sdi.dat<- sdi.dat[,c("PLT_CN", PICO, PIEN, ABLA)]
colnames(sdi.dat) <- c("PLT_CN", paste0("SDI.",c("PICO", "PIEN", "ABLA")))
sdi.dat[is.na(sdi.dat)] <- 0

## Trees per plot of bark beetle affected trees
notreesBB.dat <- merge(merge(aggregate(MPB~PLT_CN,datz5, FUN=sum, na.rm=T),  aggregate(SB~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T), aggregate(WBBB~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T)
colnames(notreesBB.dat) <- c("PLT_CN", paste0("stand.", c("MPB", "SB", "WBBB")))
notreesBB.dat$stand.allBBs <- rowSums(notreesBB.dat[,2:4])
notreesBB.dat[is.na(notreesBB.dat)] <- 0

## Total BA of bark beetle affected trees
BABB.dat <- merge(merge(aggregate(MPB*BA~PLT_CN,datz5, FUN=sum, na.rm=T),  aggregate(SB*BA~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T), aggregate(WBBB*BA~PLT_CN,datz5, FUN=sum, na.rm=T), by="PLT_CN", all=T)
colnames(BABB.dat) <- c("PLT_CN", paste0("BA.", c("MPB", "SB", "WBBB")))
BABB.dat$BA.allBBs <- rowSums(BABB.dat[,2:4])
BABB.dat[is.na(BABB.dat)] <- 0

df<- unique(datz5[,c("PLT_CN", "PREV_PLT_CN", "PLOT.STATEAB", "INVYR", "ELEV", "LAT", "LON" ),])
df <-merge(df, TPH.dat, by="PLT_CN", all=T)
df <-merge(df, qmd.dat, by="PLT_CN", all=T)
df <-merge(df, BA.dat, by="PLT_CN", all=T)
df <-merge(df, sdi.dat, by="PLT_CN", all=T)
df <- merge(df, notreesBB.dat, by="PLT_CN", all=T)
df <- merge(df, BABB.dat, by="PLT_CN", all=T)
df[is.na(df$PREV_PLT_CN)==TRUE, "PREV_PLT_CN"] <- -9999
df <- na.omit(df)

df$agentSB <- ifelse(df$stand.SB>0, "SB", "absent")
df$agentMPB <- ifelse(df$stand.MPB>0, "MPB", "absent")
df$agentWBBB <- ifelse(df$stand.WBBB>0, "WBBB", "absent")

df$agentSBo <- ifelse(df$TPH.PIEN>0, ifelse(df$stand.SB/(df$TPH.PIEN/6.018046/2.47105)>0.1, "SB", "absent"), "absent")
df$agentMPBo <- ifelse(df$TPH.PICO>0, ifelse(df$stand.MPB/(df$TPH.PICO/6.018046/2.47105)>0.1, "MPB", "absent"), "absent")
df$agentWBBBo <- ifelse(df$TPH.ABLA>0, ifelse(df$stand.WBBB/(df$TPH.ABLA/6.018046/2.47105)>0.1, "WBBB", "absent"), "absent")

df$agents <-  do.call(paste, c(df[c("agentSB", "agentMPB", "agentWBBB")],sep="-"))
df$agents <- factor(df$agents, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))

df$agentso <-  do.call(paste, c(df[c("agentSBo", "agentMPBo", "agentWBBBo")],sep="-"))
df$agentso <- factor(df$agentso, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))

df$nagentso <- df$agentso
df$nagentso <- factor(df$nagentso, levels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))

df$nagents <- df$agents
df$nagents <- factor(df$nagents, levels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"), labels=c(0, 1, 1, 2, 1, 2, 2, 3))

df$hostsPICO <- ifelse(df$TPH.PICO>0, "PICO", "absent")
df$hostsPIEN <- ifelse(df$TPH.PIEN>0, "PIEN", "absent")
df$hostsABLA <- ifelse(df$TPH.ABLA>0, "ABLA", "absent")
df$hosts <-  do.call(paste, c(df[c("hostsPICO", "hostsPIEN", "hostsABLA")],sep="-"))
df$hosts <- factor(df$hosts, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), ordered=TRUE)

df$nhosts <- df$hosts
df$nhosts <- factor(df$nhosts, levels=c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))

df$hostsPICO25 <- ifelse(df$BA.PICO/df$BA.total>0.25, "PICO", "absent")
df$hostsPIEN25 <- ifelse(df$BA.PIEN/df$BA.total>0.25, "PIEN", "absent")
df$hostsABLA25 <- ifelse(df$BA.ABLA/df$BA.total>0.25, "ABLA", "absent")
df$hosts25 <-  do.call(paste, c(df[c("hostsPICO25", "hostsPIEN25", "hostsABLA25")],sep="-"))
df$hosts25 <- factor(df$hosts25, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), ordered=TRUE)

df$nhosts25 <- df$hosts25
df$nhosts25 <- factor(df$nhosts25, levels=c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))

write.csv(df, here("Data/Processed/FIAdat-df.csv"),row.names=F)
```

```{r select plots, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-df.csv"))

##select most recent inventories
df.recent <- df

df.recent <-df.recent[!(df.recent$PREV_PLT_CN %in% df.recent$PLT_CN), ] # select plots that have not been subsequently sampled

##select inventories conducted within the last 10 years
df.recent <- df.recent[df.recent$INVYR %in% 2010:2019,]
df.recent$LONx <- df.recent$LON
df.recent$LATx <- df.recent$LAT
##select inventories with coordinates located in the Intermountain West
df.recent.sf <- st_as_sf(df.recent, coords = c("LONx", "LATx"), crs = 4326)
df.recent.sf <- st_transform(df.recent.sf,  crs = aea.proj)
states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
states <- st_transform(states, crs = aea.proj)
df.recent.sf.j.states <- st_join(df.recent.sf, states)
df.recent.sf.j.states<- df.recent.sf.j.states[df.recent.sf.j.states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]

df.recent.sf <- df.recent.sf.j.states
df.recent <- df.recent.sf
st_geometry(df.recent) <- NULL

write.csv(df.recent, here("Data/Processed/FIAdat-recent.csv"))
```

```{r reorganize by focal host, eval=F}
df <- read.csv(here("Data/Processed/FIAdat-recent.csv"))

formatp <- function(d, host="PIEN", agent="SB"){
  df.p <- d[which(d[, paste0("TPH.", host)]>0),]
  df.p$TPH.host <- df.p[, paste0("TPH.", host)]
  df.p$QMD.host <- df.p[, paste0("QMD.", host)]
  df.p$BA.host <- df.p[, paste0("BA.", host)]

  df.p$stand.BB <- df.p[, paste0("stand.", agent)]
  df.p$BA.BB <- df.p[, paste0("BA.", agent)]
  df.p$affected <- ifelse(df.p$stand.BB>0, "affected", "unaffected")
  df.p$focalhost <- host

  df.p$BAD.host <- df.p$BA.host/df.p$BA.total
  df.p$TPHD.host <- df.p$TPH.host/df.p$TPH.total

  return(df.p)
}

df <- rbind(rbind(formatp(df), formatp(df, host="PICO", agent="MPB")), formatp(df, host="ABLA", agent="WBBB"), formatp(df, host="Allhosts", agent="allBBs"))

### Proportion of total BA affected
df$propTBA <- df$BA.BB/df$BA.total

### Proportion of host BA affected
df$propHBA <- df$BA.BB/df$BA.Allhosts

### Proportion of total trees affected
df$propTTPH <- (df$stand.BB*unique(6.018046)*2.47105)/df$TPH.total

### Proportion of host trees affected
df$propHTPH <- (df$stand.BB*unique(6.018046)*2.47105)/df$TPH.host

### Affected
df$affected <- ifelse(df$stand.BB>0, "affected", "unaffected")

### Affected
df$affectedo <- ifelse(df$stand.BB/df$BA.Allhosts>0.1, "affected", "unaffected")

write.csv(df, here("Data/Processed/FIAdat-recent-re.csv"),row.names=F)
```

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsFP <- factor(df$hosts,levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

df$nhosts25F <-factor(df$nhosts25, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts25 <- factor(df$hosts25, levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hosts25F <- factor(df$hosts25, levels=c("none","PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsF25P <- factor(df$hosts25,levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)

df$nagentsoF <- factor(df$nagentso, levels=0:3, ordered=T)
df$agentsoF <- factor(df$agentso, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)

df$focalhost <- factor(df$focalhost, levels=c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("PICO", "PIEN", "ABLA", "Any"), ordered=T)

df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Any"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

df$propBA.PICO <- df$BA.PICO / df$BA.total *100
df$propBA.PIEN <- df$BA.PIEN / df$BA.total *100
df$propBA.ABLA <- df$BA.ABLA / df$BA.total *100
```

```{r maps all plots and hosts, eval=F}
df.all <- df[df$focalhost == "Any", ]
df.all$agentsFlong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))

df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)

states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]

ext <- as.matrix(extent(IMW))
#ext[1,1] <- ext[1,1] -0.25
#ext[1,2] <- ext[1,2] +0.25
#ext[2,1] <- ext[2,1] -1
#ext[2,2] <- ext[2,2] +1

map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")

map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PIEN",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")

map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")

map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")

map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")

map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PIEN+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")

map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")

legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN+ABLA",]) +tm_dots(col="agentsFlong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))

Fig.file <- here("Results/Figures/Figure-Map-Anyhost-Anybeetle.jpg")
jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
whatever <- dev.off()

```

```{r FigMap, fig.cap="The distribution of FIA plots affected by outbreaks of the mountain pine beetle (MPB), spruce beetle (SB), and western balsam bark beetle (WBBB) by the identity of host tree species present within each plot. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "504.8pt"}
knitr::include_graphics(here("Results/Figures/Figure-Map-Anyhost-Anybeetle.jpg"))
```

## Analyses

```{r import processed data}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsFP <- factor(df$hosts,levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

df$nhosts25F <-factor(df$nhosts25, levels=1:3, labels=c("one", "two", "three"), ordered=T)
df$hosts25 <- factor(df$hosts25, levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
df$hosts25F <- factor(df$hosts25, levels=c("none","PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
df$hostsF25P <- factor(df$hosts25,levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)

df$nagentsoF <- factor(df$nagentso, levels=0:3, ordered=T)
df$agentsoF <- factor(df$agentso, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)

df$focalhost <- factor(df$focalhost, levels=c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("PICO", "PIEN", "ABLA", "Any"), ordered=T)

df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Any"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

df$propBA.PICO <- df$BA.PICO / df$BA.total *100
df$propBA.PIEN <- df$BA.PIEN / df$BA.total *100
df$propBA.ABLA <- df$BA.ABLA / df$BA.total *100
```

## Do stand conditions conducive for outbreaks of multiple bark beetle species commonly co-occur?

Because the effects of diversity may be convoluted with stand development processes and thus host size and availability, we limited our analyses to only the plots where host trees were susceptible to outbreak. To do so we used a random forest modeling framework [@breiman_random_2001], a distribution-free approach useful for modeling nonlinear relationships and complex interactions characteristic of ecological systems [@cutler_random_2007]. We constructed separate models for the presence of MPB, SB, and WBBB and as predictor variables included host QMD, host basal area, and host dominance, which have been commonly linked to increased risk of bark beetle activity [[e.g. REF MPB, REF SB, REF WBBB].  Because most stands in our dataset were unaffected by bark beetles, we used a synthetic minority oversampling technique (SMOTE) to deal the imbalance in the response [@chawla_smote_2002]. Briefly, the SMOTE approach oversamples the minority class by synthesizing new cases from the minority class (here affected stands). Models were fit in R [@r_core_team_r:_2020] using the **Tidymodels** package [@kuhn_tidymodels_2020]. We then assumed that each host species was susceptible to infestation when the predicted probability of occurrence of the focal bark beetle species was greater than 0.5.

```{r set up RF helper model}
rf.mod <- function(response="agentMPB", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO"), datBB.c=df[df$focalhost=="Any",]){
    splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test   and training datasets stratified by the response class
  datz.train <- training(splitit) # create object with training data
  datz.test <- testing(splitit) # create object with testing data
  datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to   allow for cross validation (again stratified by the response class) for optimization of   model hyperparameters

  # set formula
  BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))),   datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate   more presence cases

  # set sample weights (not used in SMOTE approach)
  wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
  wts <- wts/sum(wts)

  # set model
  rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger",   importance = "permutation", seed=913, probability=T, num.threads = cores) %>%   set_mode("classification")

  # set the workflow
  rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)

  # tune model
  tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
  rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid,   metrics = metric_set(roc_auc, sens, spec))

  tuneres <- rf_tune_results %>% collect_metrics() # collect metrics

  param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best   hyperparameters

  # Use backward selection to identify important variables
  #dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]),   RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
  #registerDoSEQ()
  #vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
  vars.final <- vars
  BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse   = " + "))), data=datBB.c) %>% step_smote(response)

  rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n,   splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T,   num.threads = cores) %>% set_mode("classification")

  rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>%   add_model(rf_model_final)
  rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc,   sens, spec)) #emulates the process where, after determining the best model, the final fit on   the entire training set is needed and is then evaluated on the test set. 
  test_performance <- rf_workflow_final_fit %>% collect_metrics()
  write.csv(test_performance, here(paste0("Results/RF-testperformance-", response,   "-occurrence-SMOTE.csv")), row.names=F)

  # generate a confusion matrix for test set
  test_predictions <- rf_workflow_final_fit %>% collect_predictions()
  confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)

  write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response,   "-occurrence-SMOTE.csv")), row.names=F)

  # variable importance
  final_model <- fit(rf_workflow_final, data=datz.train)
  ranger_obj <- pull_workflow_fit(final_model)$fit

  # generate predictions for entire dataset
  x <- datBB.c$PLT_CN
  x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
  colnames(x) <- c("PLT_CN", paste0("pred", response))
  write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")),   row.names=F)

  x <- as.data.frame(ranger_obj$variable.importance)
  write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")),   row.names=F)

  # partial dependence
  pdale <- NULL
  for(z in vars.final){
    pdz <- partial(ranger_obj, train=datz.train, pred.var=z,   which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T,   trim.outliers=T)
    pdz$var <- z
    colnames(pdz) <- c("value","yhat","var")
    pdz$method <- "PD"

    alez <- ALEPlot(X=datz.train[,vars.final], X.model=ranger_obj, pred.fun=function(X.model,   newdata){predict(X.model, newdata)$predictions[,2]}, J=which(vars.final==z), K=50)
    alez <- data.frame(value=alez$x.values, yhat=alez$f.values, var=z, method="ALE")

    pdale <- rbind(pdale, pdz)
    pdale <- rbind(pdale, alez)
  }
  write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
}
```

```{r fit MPB RF model occurrence, eval=F, fig.show='hide'}
rf.mod(response="agentMPB", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit SB RF model occurrence, eval=F, fig.show='hide'}
rf.mod(response="agentSB", vars=c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit WBBB RF model occurrence, eval=F, fig.show='hide'}
rf.mod(response="agentWBBB", vars=c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit MPB RF model outbreak, eval=F, fig.show='hide'}
rf.mod(response="agentMPBo", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit SB RF model outbreak, eval=F, fig.show='hide'}
rf.mod(response="agentSBo", vars=c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") , datBB.c =df[df$focalhost=="Any",])
```

```{r fit WBBB RF model outbreak, eval=F, fig.show='hide'}
rf.mod(response="agentWBBBo", vars=c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") , datBB.c =df[df$focalhost=="Any",])
```

```{r determine suitable stands}
MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence-SMOTE.csv"))
SBpred <- read.csv(here("Results/RF-pred-agentSB-occurrence-SMOTE.csv"))
WBBBpred <- read.csv(here("Results/RF-pred-agentWBBB-occurrence-SMOTE.csv"))

MPBopred <- read.csv(here("Results/RF-pred-agentMPBo-occurrence-SMOTE.csv"))
SBopred <- read.csv(here("Results/RF-pred-agentSBo-occurrence-SMOTE.csv"))
WBBBopred <- read.csv(here("Results/RF-pred-agentWBBBo-occurrence-SMOTE.csv"))

BBpred <- merge(merge(MPBpred, SBpred, by="PLT_CN"), WBBBpred, by="PLT_CN")
colnames(BBpred) <- c("PLT_CN", "MPBpred", "SBpred", "WBBBpred")

BBopred <- merge(merge(MPBopred, SBopred, by="PLT_CN"), WBBBopred, by="PLT_CN")
colnames(BBopred) <- c("PLT_CN", "MPBopred", "SBopred", "WBBBopred")

BBpred <- merge(BBpred, BBopred, by="PLT_CN")

threshold <-0.5
BBpred$MPBsuitable <- ifelse(BBpred$MPBpred > threshold, "MPB", "absent")
BBpred$SBsuitable <- ifelse(BBpred$SBpred > threshold, "SB", "absent")
BBpred$WBBBsuitable <- ifelse(BBpred$WBBBpred >threshold, "WBBB", "absent")

BBpred$MPBosuitable <- ifelse(BBpred$MPBopred > threshold, "MPB", "absent")
BBpred$SBosuitable <- ifelse(BBpred$SBopred > threshold, "SB", "absent")
BBpred$WBBBosuitable <- ifelse(BBpred$WBBBopred >threshold, "WBBB", "absent")

BBpred$agentsuitable <-  do.call(paste, c(BBpred[c("SBsuitable", "MPBsuitable", "WBBBsuitable")],sep="-"))
BBpred$agentsuitable <- factor(BBpred$agentsuitable, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))

BBpred$suitablehosts <- factor(BBpred$agentsuitable, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB",  "MPB+WBBB", "SB+WBBB",  "MPB+SB+WBBB"), ordered=T)

BBpred$agentsuitableo <-  do.call(paste, c(BBpred[c("SBosuitable", "MPBosuitable", "WBBBosuitable")],sep="-"))
BBpred$agentsuitableo <- factor(BBpred$agentsuitableo, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))

BBpred$suitablehostso <- factor(BBpred$agentsuitableo, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB",  "MPB+WBBB", "SB+WBBB",  "MPB+SB+WBBB"), ordered=T)
```

```{r determine co-occurrence hosts, eval=F}
dfpred<- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts), labels=c('none', levels(dfpred$hosts)))
dfpred$suitablehostso <- factor(dfpred$suitablehostso, levels=levels(dfpred$suitablehostso), labels=c('none', levels(dfpred$hosts)))

hostgroups <- dfpred[dfpred$focalagent=="Any",]%>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
colnames(hostgroups) <- c("hosts", "n", "freq")
hostgroups$var <- "all"

suitablehostgroups <- dfpred[dfpred$focalagent=="Any",]%>% group_by(suitablehosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
colnames(suitablehostgroups) <- c("hosts", "n", "freq")
suitablehostgroups$var <- "suitable"

suitablehostgroupso <- dfpred[dfpred$focalagent=="Any",]%>% group_by(suitablehostso) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
colnames(suitablehostgroupso) <- c("hosts", "n", "freq")
suitablehostgroupso$var <- "suitableo"

hostz <- rbind(hostgroups, suitablehostgroups)
hostz <- rbind(hostz, suitablehostgroupso)

```

## Does the occurrence or severity of cumulative bark beetle activity increase with the host species richness?

To determine if the proportion of plots affected by any bark beetle species varied with the number or identity of hosts present, we first used a proportion test to determine if groups differed followed by a pairwise proportion test to test for differences between groups. To account for multiple comparisons, we adjusted p-values to account for multiple comparisons [@benjamini_false_1995].

To determine if the severity of cumulative bark beetle-driven tree mortality was greater in stands with multiple susceptible hosts, we subsetted our data to only the plots affected by at least one bark beetle species. We then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present. Followed by a Dunn test to determine if difference existed between groups, again accounting for false discovery rate using the Benjamini and Hochberg (1995) method.

```{r BB occurrence by hosts, eval=T}
res <- dfpred %>% count(hosts, affected) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

resa <- res[res$affected=="affected",]
resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affected=="affected",]
resp <- merge(resp, xxpr, by.x="hosts", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "bc", "c", "cd", "d", "e")

resp$hosts <- factor(resp$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
resp.hosts <- resp

ggpallpres <- ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentpresence-hostpresence.jpg"), plot=ggpallpres, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak occurrence by hosts, eval=T}
res <- dfpred %>% count(hosts, affectedo) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
x <- dfpred %>% count(hosts)

resa <- res[res$affectedo=="affected",]
resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affectedo=="affected",]
resp <- merge(resp, xxpr, by.x="hosts", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "b", "c", "cd", "d", "d")

resp$hosts <- factor(resp$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
resp.hosts <- resp

ggpallout <- ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-hostspresence.jpg"), plot=ggpallout, width=fig.width1,height=fig.width1,units="in")
```

```{r BB occurrence by suitablehosts}
## Suitable hosts
res <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts, affected) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts)

resa <- res[res$affected=="affected",]

resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affected=="affected",]
resp <- merge(resp, xxpr, by.x="suitablehosts", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "c", "c", "dc", "de", "e")

resp$suitablehosts <- factor(as.character(resp$suitablehosts), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

resp.suitablehosts <- resp

ggpsuitablepres <- ggplot(resp.suitablehosts, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentpresence-suitablehosts.jpg"), plot=ggpsuitablepres, width=fig.width1,height=fig.width1,units="in")

```

```{r BB outbreak occurrence by suitable hosts}
## Suitable hosts
res <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts, affectedo) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
x <- dfpred[!dfpred$suitablehosts=="none",] %>% count(suitablehosts)

resa <- res[res$affectedo=="affected",]

resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affectedo=="affected",]
resp <- merge(resp, xxpr, by.x="suitablehosts", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "b", "c", "c", "c", "c")

resp$suitablehosts <- factor(as.character(resp$suitablehosts), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

resp.suitablehostso <- resp

ggpsuitableoutbreak<- ggplot(resp.suitablehostso, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-suitablehosts.jpg"), plot=ggpsuitableoutbreak, width=fig.width1,height=fig.width1,units="in")
```

```{r BB occurrence by hosts25}
## 
res <- df[!df$hosts25=="none",] %>% count(hosts25, affected) %>% group_by(hosts25) %>% mutate(prop = prop.table(n))
x <-  df[!df$hosts25=="none",] %>% count(hosts25)

resa <- res[res$affected=="affected",]

resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affected=="affected",]
resp <- merge(resp, xxpr, by.x="hosts25", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "b", "b", "c", "c", "c", "bc")

resp$hosts25 <- factor(as.character(resp$hosts25), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

resp.hosts25 <- resp

ggphost25pres <- ggplot(resp.hosts25, aes(x=hosts25,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentpresence-hosts25.jpg"), plot=ggphost25pres, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak occurrence by hosts25}
## Hosts 25
res <- df[!df$hosts25=="none",] %>% count(hosts25, affectedo) %>% group_by(hosts25) %>% mutate(prop = prop.table(n))
x <- df[!df$hosts25=="none",] %>% count(hosts25)

resa <- res[res$affectedo=="affected",]

resa$alln <- x$n

x <- prop.test(x=resa$n, n=resa$alln)
xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
xxp <-rstatix::cor_gather(xx$p.value)
xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)

resp <- res[res$affectedo=="affected",]
resp <- merge(resp, xxpr, by.x="hosts25", by.y="Group")
resp <- resp[order(resp$prop),]
resp$Letter2 <- c("a", "abc", "b", "b", "c", "c", "c")

resp$hosts25 <- factor(as.character(resp$hosts25), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

resp.hosts25o <- resp

ggphost25outbreak <- ggplot(resp.hosts25o, aes(x=hosts25,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-hosts25.jpg"), plot=ggphost25outbreak, width=fig.width1,height=fig.width1,units="in")
```

```{r BB severity by hosts, eval=F}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]

x <- kruskal.test(propTBA~hosts, data=dfpred)
DT <- dunnTest(propTBA~hosts, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())

res3 <- merge(res, res2, by.x="Group", by.y="hosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "ab", "bc", "c", "cd", "d", "d")
res3$Group <- factor(res3$Group, levels=levels(df$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

ggpallseverity <- ggplot(dfpred, aes(x=hostsFP, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group)) + geom_text(data=res3, aes(label = n, x = Group, y = median*100 + 4, fill=Group),size = 3) + ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentpresence-hostpresence.jpg"), plot=ggpallseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak severity by hosts, eval=F}
dfpred <- df
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affectedo=="affected",]

x <- kruskal.test(propTBA~hosts, data=dfpred)
DT <- dunnTest(propTBA~hosts, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())

res3 <- merge(res, res2, by.x="Group", by.y="hosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "bc", "bc", "cd", "cd", "d")
res3$Group <- factor(res3$Group, levels=levels(df$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

ggoallseverity <- ggplot(dfpred, aes(x=hostsFP, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey") + geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group)) + geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3) + ylab("severity (% BA)") +xlab("hosts") + theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-hostpresence.jpg"), plot=ggoallseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB severity by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

x <- kruskal.test(propTBA~suitablehosts, data=dfpred)
DT <- dunnTest(propTBA~suitablehosts, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA),n=n())

res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "b", "bc", "b", "c", "bc")

ggpsuitablehostsseverity <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +2, fill=Group),size = 3)+ ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentpresence-suitbalehosts.jpg"), plot=ggpsuitablehostsseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak severity by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affectedo=="affected",]
dfpred <- dfpred[!dfpred$suitablehosts=="none",]
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

x <- kruskal.test(propTBA~suitablehosts, data=dfpred)
DT <- dunnTest(propTBA~suitablehosts, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
res2$n2 <- paste0("n = ", res2$n)
res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "bc", "bc", "cd", "cd", "d")

ggosuitablehostsseverity <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3)+ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-suitbalehosts.jpg"), plot=ggosuitablehostsseverity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB severity by hosts25, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affected=="affected",]
dfpred <- dfpred[!dfpred$hosts25=="none",]
dfpred$hosts25F <- factor(dfpred$hosts25, levels=levels(dfpred$hosts25)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

x <- kruskal.test(propTBA~hosts25F, data=dfpred)
DT <- dunnTest(propTBA~hosts25F, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(hosts25F) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
res2$n2 <- paste0("n = ", res2$n)
res3 <- merge(res, res2, by.x="Group", by.y="hosts25F")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("abc", "a", "b", "b", "bc", "c", "bc")

ggphost25severity <- ggplot(dfpred, aes(x=hosts25F, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3)+ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentpresence-hosts25.jpg"), plot=ggphost25severity, width=fig.width1,height=fig.width1,units="in")
```

```{r BB outbreak severity by hosts25, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[dfpred$affectedo=="affected",]
dfpred <- dfpred[!dfpred$hosts25=="none",]
dfpred$hosts25F <- factor(dfpred$hosts25, levels=levels(dfpred$hosts25)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)

x <- kruskal.test(propTBA~hosts25F, data=dfpred)
DT <- dunnTest(propTBA~hosts25F, data=dfpred,method="bh")
res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)

res2 <- dfpred %>% group_by(hosts25F) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
res2$n2 <- paste0("n = ", res2$n)
res3 <- merge(res, res2, by.x="Group", by.y="hosts25F")
res3 <- res3[order(res3$median),]
res3$Letter2 <- c("a", "b", "b", "ab", "b", "b", "b")
res3[res3$Group == "PICO,\nPIEN,\n&\nABLA", "n"]<- ""

ggohost25severity <- ggplot(dfpred, aes(x=hosts25F, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3)+ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")

ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-hosts25.jpg"), plot=ggohost25severity, width=fig.width1,height=fig.width1,units="in")
```

## Does the co-occurrence of multiple bark beetle species result in more severe tree mortality?

Finally, to determine if We then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present.

```{r multiple - BB pres by hosts , eval=F}
dfpred <-  df

dfpred <- dfpred[dfpred$hosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
dfpred <-dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[!dfpred$agentsF == "none",]

results <- NULL
  dfsub <- dfpred[dfpred$hosts =="PICO+PIEN"  & dfpred$agents %in% c("MPB", "SB","MPB+SB"),]
  dfsub$agents <- as.factor(as.character(dfsub$agents))
  x <- kruskal.test(propTBA~agents, data=dfsub)
  res <- data.frame(agentsF= c("MPB", "SB","MPB+SB"), p=NA)
  DT <- dunnTest(propTBA~agents, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
  res$hosts <- "PICO+PIEN"

  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "a", "b")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$hosts == "PICO+ABLA"  & dfpred$agentsF %in% c("MPB", "WBBB","MPB+WBBB"),]
  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
  x <- kruskal.test(propTBA~agentsF, data=dfsub)
  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$hosts <- "PICO+ABLA"
  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "b", "b")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$hosts == "PIEN+ABLA"  & dfpred$agentsF %in% c("SB", "WBBB","SB+WBBB"),]
  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
  x <- kruskal.test(propTBA~agentsF, data=dfsub)
  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$hosts <- "PIEN+ABLA"
  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "b", "c")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$hosts == "PICO+PIEN+ABLA",]
  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
  x <- kruskal.test(propTBA~agentsF, data=dfsub)
  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$hosts <- "PICO+PIEN+ABLA"
  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[!res$Group =="none",]
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "abc", "b", "b", "c", "d", "d")

  results <- rbind(results, res)

results$hosts <- factor(results$hosts,
levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
dfpred$hosts <- factor(dfpred$hosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

dfpred$agentsF <- factor(dfpred$agentsF,
levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
results$agentsF <- factor(results$Group,
levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)

ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(-5, 110) + geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+ geom_text(data=results, aes(label = n, y=-3),size = 3, col="black")+ylab("severity (% BA)")

ggsave(here("Results/Figures/Figure-MultipleSeverity-agents-hosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1,units="in")

```

```{r multiple - BB outbreak by hosts , eval=F}
dfpred <-  df

dfpred <- dfpred[dfpred$hosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
dfpred <-dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[!dfpred$agentso == "none",]

results <- NULL
  dfsub <- dfpred[dfpred$hosts =="PICO+PIEN"  & dfpred$agentso %in% c("MPB", "SB","MPB+SB"),]
  dfsub$agentso <- as.factor(as.character(dfsub$agentso))
  x <- kruskal.test(propTBA~agentso, data=dfsub)
  res <- data.frame(agentsoF= c("MPB", "SB","MPB+SB"), p=NA)
  DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
  res$hosts <- "PICO+PIEN"

  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "a", "b")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$hosts == "PICO+ABLA"  & dfpred$agentso %in% c("MPB", "WBBB","MPB+WBBB"),]
  dfsub$agentso<- as.factor(as.character(dfsub$agentso))
  x <- kruskal.test(propTBA~agentso, data=dfsub)
  #DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
  #res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res <- data.frame(Group=c("MPB", "WBBB","MPB+WBBB"), Letter="", MonoLetter="")
  res$hosts <- "PICO+ABLA"

  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
  res <- res[order(res$median),]
  res$Letter2 <- c("", "", "")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$hosts == "PIEN+ABLA"  & dfpred$agentso %in% c("SB", "WBBB","SB+WBBB"),]
  dfsub$agentso <- as.factor(as.character(dfsub$agentso))
  x <- kruskal.test(propTBA~agentso, data=dfsub)
  DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$hosts <- "PIEN+ABLA"
  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "b", "b")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$hosts == "PICO+PIEN+ABLA",]
  dfsub$agentso <- as.factor(as.character(dfsub$agentso))
  x <- kruskal.test(propTBA~agentso, data=dfsub)
  DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$hosts <- "PICO+PIEN+ABLA"
  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
  res <- res[!res$Group =="none",]
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "a", "ab", "abcd", "bc", "cd", "d")

  results <- rbind(results, res)

results$hosts <- factor(results$hosts,
levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
dfpred$hosts <- factor(dfpred$hosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

dfpred$agentsoF <- factor(dfpred$agentso,
levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
results$agentsoF <- factor(results$Group,
levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)

ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsoF,y=propTBA*100, fill=agentsoF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(-5, 110) + geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+ geom_text(data=results, aes(label = n, y=-3),size = 3, col="black")+ylab("severity (% BA)")

ggsave(here("Results/Figures/Figure-MultipleSeverity-agentso-hosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1,units="in")

```

```{r multiple - BB pres by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
dfpred <- dfpred[dfpred$affected=="affected",]

dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))

#dfpred <- dfpred[!(dfpred$suitablehosts=="PICO" & dfpred$agentsF %in% c("SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]

#dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN" & dfpred$agentsF %in% c("MPB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]

#dfpred <- dfpred[!(dfpred$suitablehosts=="ABLA" & dfpred$agentsF %in% c("MPB", "SB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]

dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]

dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+PIEN" & dfpred$agentsF %in% c("WBBB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]

dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+ABLA" & dfpred$agentsF %in% c("SB", "MPB+SB", "SB+WBBB", "MPB+SB+WBBB")),]

dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN+ABLA" & dfpred$agentsF %in% c("MPB", "MPB+SB", "MPB+WBBB", "MPB+SB+WBBB")),]

dfpred <-dfpred[dfpred$focalagent=="Any",]
dfpred <- dfpred[!dfpred$agentsF == "none",]

results <- NULL
  dfsub <- dfpred[dfpred$suitablehosts =="PICO+PIEN"  & dfpred$agents %in% c("MPB", "SB","MPB+SB"),]
  dfsub$agents <- as.factor(as.character(dfsub$agents))
  x <- kruskal.test(propTBA~agents, data=dfsub)
  res <- data.frame(agentsF= c("MPB", "SB","MPB+SB"), p=NA)
  DT <- dunnTest(propTBA~agents, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
  res$suitablehosts <- "PICO+PIEN"

  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "b", "c")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$suitablehosts == "PICO+ABLA"  & dfpred$agentsF %in% c("MPB", "WBBB","MPB+WBBB"),]
  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
  x <- kruskal.test(propTBA~agentsF, data=dfsub)
  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$suitablehosts <- "PICO+ABLA"
  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "a", "b")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$suitablehosts == "PIEN+ABLA"  & dfpred$agentsF %in% c("SB", "WBBB","SB+WBBB"),]
  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
  x <- kruskal.test(propTBA~agentsF, data=dfsub)
  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$suitablehosts <- "PIEN+ABLA"
  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[order(res$median),]
  res$Letter2 <- c("a", "b", "b")
  results <- rbind(results, res)

  dfsub <- dfpred[dfpred$suitablehosts == "PICO+PIEN+ABLA",]
  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
  x <- kruskal.test(propTBA~agentsF, data=dfsub)
  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
  res$suitablehosts <- "PICO+PIEN+ABLA"
  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
  res <- res[!res$Group =="none",]
  res <- res[order(res$median),]
  res$Letter2 <- c("ab", "b", "ab", "ab", "bc", "cd", "d")

  results <- rbind(results, res)

results$suitablehosts <- factor(results$suitablehosts,
levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)

dfpred$agentsF <- factor(dfpred$agentsF,
levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
results$agentsF <- factor(results$Group,
levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)

ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(-5, 110) + geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+ geom_text(data=results, aes(label = n, y=-3),size = 3, col="black")+ylab("severity (% BA)")

ggsave(here("Results/Figures/Figure-MultipleSeverity-agent-suitablehosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1,units="in")
```

# Results

## Random forest models

Models Table \@ref(tab:TestPerformance)

## Do stand conditions suitable for multiple outbreaks commonly co-occur?

```{r describe frequency of hosts and agents}
## Hosts
hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))

nhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))

## Affected hosts
affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]

affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]

## Agents given host presence
agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
  mutate(freq = n / sum(n))
PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))

## Agents given presence of any agent
agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
  mutate(freq = n / sum(n))

agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))

```

Across the 10,439 FIA plots with at least one of the target host species, 38% (n=3,915) contained two species and 19% (n=1,935) three species (Fig. \@ref(fig:FigHostFreq)). Subalpine fir was most likely to occur with other target host species (50% of plots with multiple hosts), followed by Engelmann spruce (45% of plots with multiple hosts), and lodgepole pine (36% of plots with multiple hosts).

However, random forest models predicted that 70% of plots (n=7,292) were suitable for bark beetle infestation and of these plots only 31% (n=1,855) were suitable for more than one bark beetle species (Fig. \@ref(fig:FigHostFreq)). Again, conditions suitable for bark beetle attack in subalpine fir were most likely to co-occur with conditions suitable for bark beetle attack in another tree species (17% of plots).

```{r FigHostFreq, fig.cap = "The frequency of plots by host identity. Gray bars indicate host species were present (i.e., at least one individual of the focal species was present within the plot). Black bars indicate stand structure and composition conditions were suitbale to bark beetle outbreak in focal tree species. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "244.8pt"}
knitr::include_graphics( here("Results/Figures/FigureHostCo-outbreak.jpg"))
```

## Does the occurrence or severity of cumulative bark beetle activity increase with the host species richness?

Given stand conditions suitable to bark beetle outbreak, stands with lodgepole and spruce, lodgepole and fir, or all three hosts were most likely to experience outbreak (Fig. \@ref(fig:FigBBO)). Stands with only subalpine fir were least likely to experience outbreak, followed by stands with only spruce and their combination.

```{r FigBBO, fig.cap = "The proportion of plots affected by bark beetles by the identity of suitable hosts. Letters above bars indicate significant (p<0.05) difference between groups, as determined using pairwise proportion tests. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "244.8pt"}
knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts2-outbreak.jpg"))
```

The effect of the number of suitable hosts on the severity of bark beetle infestation depended upon the identity of host species (Fig. \@ref(fig:FigSeverity). Stands with only subalpine fir experienced the lowest severity bark beetle infestation. In contrast, stands with only lodgepole pine and Engelmann spruce and their combination, experienced the highest rates of bark beetle infestation. Stands with subalpine fir in combination with other hosts, experienced intermediate rates of bark beetle infestation.

```{r FigSeverity, fig.cap= "Boxplots illustrating the severity of bark beetle infestation by the identity of host species susceptible to bark beetle outbreak. For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir. The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range. Letters above boxes indicate significant (p<0.05) difference between groups, as determined using a nonparametric Dunn test.", out.width = "244.8pt"}
knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts-outbreak.jpg"))
```

## Does the co-occurrence of multiple bark beetle species result in more severe cumulative bark beetle mortality?

Across all 10,439 FIA plots, only 5.3% (n=539) were affected by outbreaks of multiple bark beetle species.

Given the 2,163 plots where multiple tree species were susceptible to bark beetle infestation, 56.7% (n=1,227) were affected by at least one bark beetle species and 21.3% (n=461) were affected by multiple bark beetle species. The most commonly occurring agent was the MPB (n=768, 35.5% of plots), followed by the WBBB (n=503, 23.3%) and the SB (n=461, 21.3%). The most commonly occurring combination of bark beetle species was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and SB and WBBB (26% of case; n=120). The combination of all three agents was rare (9.5% of cases; n=44).

When plots contained two tree species susceptible to bark beetles, the presence of both bark beetle species increased activity relative to presence of only one bark beetle species (Fig. \@ref(fig:FigHotSpotSeverity)). The increase was greatest for stands that contained both MPB and SB; the median severity of plots with both MPB and SB was 33.7 percentage points greater than MPB alone and 38.3 percentage points greater than SB alone. The median severity of plots with both MPB and WBBB was 7.0 percentage points greater than MPB alone and 13.6 percentage points greater than WBBB alone. The median severity of plots with both SB and WBBB was 14.7 percentage points greater than SB alone and 19.8 percentage points greater than WBBB alone.

Given stand conditions that made all three tree species susceptible to bark beetle activity, the highest rates of bark beetle activity were in plots where both MPB and SB present; whether or not WBBB was also present did not affect mortality severity (~1 percentage point difference in median severity; Fig. \@ref(fig:FigHotSpotSeverity)).

```{r FigHotSpotSeverity, fig.cap = "The severity of bark beetle mortality in plots with multiple tree species susceptible to bark beetles (columns) by the combinatin of bark beetle species present. Letters above boxes indicate significant (p < 0.05) differences between groups as determined by a Dunn test, a nonparametric rank sum test. The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.",  out.width = "504.8pt"}
knitr::include_graphics( here("Results/Figures/FigureHotSpotSeverity-outbreak.jpg"))
```

# Discussion

# Acknowledgements

# Author contributions

# References

<div id="refs"></div>

## Supplement
### Appendix I: Random Forest Modeling

```{r Performance occurrence}
MPBtp <- read.csv(here("Results/RF-testperformance-agentMPB-occurrence-SMOTE.csv"))
MPBtp$focalagent <- "MPB"
SBtp <- read.csv(here("Results/RF-testperformance-agentSB-occurrence-SMOTE.csv"))
SBtp$focalagent <- "SB"
WBBBtp <- read.csv(here("Results/RF-testperformance-agentWBBB-occurrence-SMOTE.csv"))
WBBBtp$focalagent <- "WBBB"

tpdat <- rbind(rbind(MPBtp, SBtp), WBBBtp)
tpdat$.estimate <-round(tpdat$.estimate, digits=2)
tpdat2 <- tpdat[,c(".metric", "focalagent", ".estimate")] %>% spread(key=focalagent, value=.estimate)
colnames(tpdat2) <- c("Metric","MPB", "SB", "WBBB")
tpdat2$Metric <- c("ROC AUC", "sensitivity", "specificity")
kable(tpdat2,booktabs = TRUE,caption="Accuracy statistics from Random Forest models prediciting the presence/absence of the mountain pine beetle (MPB) in lodgepole pine, spruce beetle (SB) in Engelmann spruce, and western balsam bark beetle (WBBB) in subalpine fir. ROC AUC is the area under the reciever operating characteristic curve.")
```

```{r Performance outbreak}
MPBtp <- read.csv(here("Results/RF-testperformance-agentMPBo-occurrence-SMOTE.csv"))
MPBtp$focalagent <- "MPB"
SBtp <- read.csv(here("Results/RF-testperformance-agentSBo-occurrence-SMOTE.csv"))
SBtp$focalagent <- "SB"
WBBBtp <- read.csv(here("Results/RF-testperformance-agentWBBBo-occurrence-SMOTE.csv"))
WBBBtp$focalagent <- "WBBB"

tpdat <- rbind(rbind(MPBtp, SBtp), WBBBtp)
tpdat$.estimate <-round(tpdat$.estimate, digits=2)
tpdat2 <- tpdat[,c(".metric", "focalagent", ".estimate")] %>% spread(key=focalagent, value=.estimate)
colnames(tpdat2) <- c("Metric","MPB", "SB", "WBBB")
tpdat2$Metric <- c("ROC AUC", "sensitivity", "specificity")
kable(tpdat2,booktabs = TRUE,caption="Accuracy statistics from Random Forest models prediciting the presence/absence of outbreaks of mountain pine beetle (MPB) in lodgepole pine, spruce beetle (SB) in Engelmann spruce, and western balsam bark beetle (WBBB) in subalpine fir. ROC AUC is the area under the reciever operating characteristic curve.")
```

```{r plot confusion matrices, eval=F}
MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-SMOTE.csv"))
SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-SMOTE.csv"))
WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-SMOTE.csv"))

BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
BBct$var <- "presence"
BBct1  <- BBct

MPBct <- read.csv(here("Results/RF-confusiontable-agentMPBo-occurrence-SMOTE.csv"))
SBct <- read.csv(here("Results/RF-confusiontable-agentSBo-occurrence-SMOTE.csv"))
WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBBo-occurrence-SMOTE.csv"))

BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
BBct$var <- "outbreak"

BBct <- rbind(rbind(BBct1, BBct))

ggconfused <- ggplot(BBct, aes(x=agent, y=sensitivity, fill=var))+geom_bar(stat="identity", position="dodge")+theme(legend.position = "none") + ggplot(BBct, aes(x=agent, y=accuracy, fill=var))+geom_bar(stat="identity", position="dodge")+plot_layout(nrow=2, guides="collect")

ggsave(here("Results/Figures/Confusion-outbreak.jpg"), plot=ggconfused, width=fig.width1,height=fig.width1,units="in")
```

```{r plot variable importance occurrence, eval=F}
MPBvimp <- read.csv(here("Results/RF-vimp-agentMPB-occurrence-SMOTE.csv"))
MPBvimp$focalagent <- "MPB"
MPBvimp$variable <- c("BA", "QMD", "BA dominance")
MPBvimp$scale <-MPBvimp$ranger_obj.variable.importance / sum(MPBvimp$ranger_obj.variable.importance )*100

SBvimp <- read.csv(here("Results/RF-vimp-agentSB-occurrence-SMOTE.csv"))
SBvimp$focalagent <- "SB"
SBvimp$variable <- c("BA", "QMD", "BA dominance")
SBvimp$scale <-SBvimp$ranger_obj.variable.importance / sum(SBvimp$ranger_obj.variable.importance )*100

WBBBvimp <- read.csv(here("Results/RF-vimp-agentWBBB-occurrence-SMOTE.csv"))
WBBBvimp$focalagent <- "WBBB"
WBBBvimp$variable <- c("BA", "QMD", "BA dominance")
WBBBvimp$scale <-WBBBvimp$ranger_obj.variable.importance / sum(WBBBvimp$ranger_obj.variable.importance )*100

vimpdat <- rbind(rbind(MPBvimp, SBvimp), WBBBvimp)
colnames(vimpdat) <- c("importance","focalagent", "variable", "scale")
vimpdat$focalagent <- factor(vimpdat$focalagent)  
vimpdat$variable <- factor(vimpdat$variable)  
vimpplot <- ggplot(vimpdat, aes(x=variable, y=scale, fill=focalagent))+geom_bar(stat="identity", position="dodge")+theme(legend.position="bottom", legend.title=element_blank(), axis.title.y = element_blank())+coord_flip()+ylab("variable importance\n(% contribution to overall model fit)")+scale_fill_manual(values=p1col[1:3])

ggsave(here("Results/Figures/VIMP-occurrence.jpg"), plot=vimpplot, width=fig.width1,height=fig.width1,units="in")
```

```{r plot variable importance outbreak, eval=F}
MPBvimp <- read.csv(here("Results/RF-vimp-agentMPBo-occurrence-SMOTE.csv"))
MPBvimp$focalagent <- "MPB"
MPBvimp$variable <- c("BA", "QMD", "BA dominance")
MPBvimp$scale <-MPBvimp$ranger_obj.variable.importance / sum(MPBvimp$ranger_obj.variable.importance )*100

SBvimp <- read.csv(here("Results/RF-vimp-agentSBo-occurrence-SMOTE.csv"))
SBvimp$focalagent <- "SB"
SBvimp$variable <- c("BA", "QMD", "BA dominance")
SBvimp$scale <-SBvimp$ranger_obj.variable.importance / sum(SBvimp$ranger_obj.variable.importance )*100

WBBBvimp <- read.csv(here("Results/RF-vimp-agentWBBBo-occurrence-SMOTE.csv"))
WBBBvimp$focalagent <- "WBBB"
WBBBvimp$variable <- c("BA", "QMD", "BA dominance")
WBBBvimp$scale <-WBBBvimp$ranger_obj.variable.importance / sum(WBBBvimp$ranger_obj.variable.importance )*100

vimpdat <- rbind(rbind(MPBvimp, SBvimp), WBBBvimp)
colnames(vimpdat) <- c("importance","focalagent", "variable", "scale")
vimpdat$focalagent <- factor(vimpdat$focalagent)  
vimpdat$variable <- factor(vimpdat$variable)  
vimpplot <- ggplot(vimpdat, aes(x=variable, y=scale, fill=focalagent))+geom_bar(stat="identity", position="dodge")+theme(legend.position="bottom", legend.title=element_blank(), axis.title.y = element_blank())+coord_flip()+ylab("variable importance\n(% contribution to overall model fit)")+scale_fill_manual(values=p1col[1:3])

ggsave(here("Results/Figures/VIMP-outbreak.jpg"), plot=vimpplot, width=fig.width1,height=fig.width1,units="in")
```

```{r partial dependence occurrence, eval=F}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
MPBpp$focalagent <- "MPB"
SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
SBpp$focalagent <- "SB"
WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
WBBBpp$focalagent <- "WBBB"

ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"

x <- df[!df$focalagent=="Any",]
x$propBA <- x$BA.host/x$BA.total
x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
x$affected01 <- ifelse(x$affected=="affected", 1, 0)

xm <- melt(x, id.vars=c("focalagent", "affected01"))
colnames(xm) <- c("focalagent", "affected01", "var", "value")
xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
xm$value <- as.numeric(xm$value)
xm <- na.omit(xm)

pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)

pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)

pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)

pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)

ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)

ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)

ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))

ggsave(here("Results/Figures/PartialDependence-occurrence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
```

```{r partial dependence outbreak, eval=F}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPBo-occurrence-SMOTE.csv"))
MPBpp$focalagent <- "MPB"
SBpp <- read.csv(here("Results/RF-partialdependence-agentSBo-occurrence-SMOTE.csv"))
SBpp$focalagent <- "SB"
WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBBo-occurrence-SMOTE.csv"))
WBBBpp$focalagent <- "WBBB"

ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"

x <- df[!df$focalagent=="Any",]
x$propBA <- x$BA.host/x$BA.total
x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
x$affected01 <- ifelse(x$affected=="affected", 1, 0)

xm <- melt(x, id.vars=c("focalagent", "affected01"))
colnames(xm) <- c("focalagent", "affected01", "var", "value")
xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
xm$value <- as.numeric(xm$value)
xm <- na.omit(xm)

pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)

pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)

pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)

pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)

ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)

ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)

ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))

ggsave(here("Results/Figures/PartialDependence-outbreak.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
```

```{r ale occurrence, eval=F}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
MPBpp$focalagent <- "MPB"
SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
SBpp$focalagent <- "SB"
WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
WBBBpp$focalagent <- "WBBB"

ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"

x <- df[!df$focalagent=="Any",]
x$propBA <- x$BA.host/x$BA.total
x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
x$affected01 <- ifelse(x$affected=="affected", 1, 0)

xm <- melt(x, id.vars=c("focalagent", "affected01"))
colnames(xm) <- c("focalagent", "affected01", "var", "value")
xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
xm$value <- as.numeric(xm$value)
xm <- na.omit(xm)

pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)

pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)

pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)

pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)

ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)

ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)

ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))

ggsave(here("Results/Figures/ALE-occurrence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
```

```{r ale outbreak, eval=F}
df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)

MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPBo-occurrence-SMOTE.csv"))
MPBpp$focalagent <- "MPB"
SBpp <- read.csv(here("Results/RF-partialdependence-agentSBo-occurrence-SMOTE.csv"))
SBpp$focalagent <- "SB"
WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBBo-occurrence-SMOTE.csv"))
WBBBpp$focalagent <- "WBBB"

ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"

x <- df[!df$focalagent=="Any",]
x$propBA <- x$BA.host/x$BA.total
x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
x$affected01 <- ifelse(x$affected=="affected", 1, 0)

xm <- melt(x, id.vars=c("focalagent", "affected01"))
colnames(xm) <- c("focalagent", "affected01", "var", "value")
xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
xm$value <- as.numeric(xm$value)
xm <- na.omit(xm)

pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)

pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)

pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)

pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)

ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)

ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)

ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))

ggsave(here("Results/Figures/ALE-outbreak.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
```

Consistent with existing stand hazard ratings, the probability of each bark beetle species occurring generally increased with host basal area, quadratic mean diameter, and percent basal area (Fig. \@ref(fig:partialFig)). However, both MPB and WBBB exhibited a slight decrease in the probability of outbreak at very high basal areas and all three bark beetle species exhibited a lower probability of outbreak in stands where host trees were very large (i.e. high QMD values). WBBB also exhibited a decline in the probability of outbreak when subalpine fir comprised a relatively high proportion of the basal area.

```{r partialFig, fig.cap="The associations between the probability of occurrence of the mountain pine beetle (MPB; red), spruce beetle (SB; blue); and western balsam bark beetle (WBBB; yellow) and three common measuree of stand suceptibility to bark beetl beetle outbreak, basal area (BA), quadratic mean diameter (QMD), and host dominance (% BA). The top row illustrates smoothed partial dependence results from Random Forest modeling, while in the bottom row boxplots show the distribution of values for stands affected (dark colors) and unaffected (light colors) by each bark beetle species."}
knitr::include_graphics(here("Results/Figures/PartialDependence-outbreak.jpg"))
```

### Maps

```{r generate maps for suitable hosts, eval=F}
bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)

df.all <- bbo.dat[bbo.dat$focalhost == "Any", ]
df.all<- df.all[!df.all$suitablehosts == "none", ]
df.all$hostsFPlong <- df.all$hostsFP
df.all$agentsFplong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))

df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)

states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]

ext <- as.matrix(extent(IMW))
#ext[1,1] <- ext[1,1] -0.25
#ext[1,2] <- ext[1,2] +0.25
#ext[2,1] <- ext[2,1] -1
#ext[2,2] <- ext[2,2] +1

map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")

map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="SB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")

map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")

map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB+SB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")

map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB+WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")

map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="SB+WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")

map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB+SB+WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")

legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))

Fig.file <- here("Results/Figures/Figure-Map-Suitableplots-outbreak.jpg")
jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
whatever <- dev.off()
```

```{r FigSuitableMap, fig.cap="a caption", out.width = "504.8pt", eval=F}
knitr::include_graphics(here("Results/Figures/Figure-Map-Suitableplots-outbreak.jpg"))
```

```{r map study area of hosts, eval=F}
f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
f93 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
f108 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)

# reclassify to rasters to presence (1) and absence (0)
f19[f19>0] <- 1 
f93[f93>0] <- 1
f108[f108>0] <- 1

# define areas of multiple hosts
f108f93 <- f108 + f93 # add 
f108f93[f108f93<2]<- 0 # reclassify
f108f93[f108f93==2]<- 1 # reclassify
f108f19 <- f108 + f19
f108f19[f108f19<2]<- 0 # reclassify
f108f19[f108f19==2]<- 1 # reclassify
f93f19 <- f93 + f19
f93f19[f93f19<2]<- 0  # reclassify
f93f19[f93f19==2]<- 1 # reclassify

f108f93f19 <- f108 + f93 + f19

# define areas of one host
f108f93f19re1 <-f108f93f19
f108f93f19re1 [f108f93f19re1 ==1 ]<-1
f108f93f19re1 [f108f93f19re1>1]<-0

# define areas of two hosts
f108f93f19re2 <-f108f93f19
f108f93f19re2 [f108f93f19re2 ==1 ]<-0
f108f93f19re2 [f108f93f19re2==3]<-0 
f108f93f19re2 [f108f93f19re2 ==2 ]<-1

# limit maps of presence (1) and absence (0) to only areas with one host
f108 <- f108 * f108f93f19re1 
f93 <- f93 * f108f93f19re1
f19 <- f19 * f108f93f19re1 
f93[f93==1]<- 2 # give unique labels
f19[f19==1]<- 3 # give unique labels

# create maps of two hosts limited to areas with two hosts
f108f93  <- f108f93 * f108f93f19re2 
f108f93[f108f93==1]<-4 # give unique labels

f108f19 <- f108f19 * f108f93f19re2
f108f19[f108f19==1]<-5 # give unique label

f93f19 <- f93f19 * f108f93f19re2
f93f19[f93f19==1]<-6 # give unique labels

# create maps of all three hosts
f108f93f19[f108f93f19<=2]<-0
f108f93f19[f108f93f19==3]<-7 # give unique labels

fplot <- f108 + f93 + f19 + f108f93 + f108f19 + f93f19+ f108f93f19
st_crs(fplot) <- "EPSG:5070" # assign projection

fplot[fplot==0] <-NA

states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
notIMW <- states[!states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
IMW <- st_transform(IMW, "EPSG:5070")
notIMW <- st_transform(notIMW, "EPSG:5070")

ext <- as.matrix(extent(IMW))

sa.tmap <- tm_shape(IMW, bbox=tmaptools::bb(ext)) + tm_fill(col="#f5f5f5")+tm_borders()+tm_shape(fplot) +tm_raster(style = "cat", palette = c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), colorNA=NULL, legend.show = TRUE, title="Hosts                    ", labels = c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"))+ tm_shape(notIMW) + tm_fill(col="white")+tm_compass(position=c(0.95, 0.88), just=c("center", "bottom"))+tm_layout(legend.bg.color="white")+tm_graticules(lines=FALSE)+tm_layout(legend.outside=TRUE,legend.text.size=0.55)+tm_scale_bar(position = c(0.01, 0.001), just="left", breaks=c(0,100),text.size=1.1)+tm_shape(IMW) +tm_borders()

Fig.file <- here("Results/Figures/Figure-Studyarea-outbreak.jpg")
jpeg(Fig.file, width=fig.width1.5,height=fig.width1.5,units="in", res=300)
sa.tmap 
whatever <- dev.off()
```

```{r FigSA, fig.cap = "The distribution of host species presence across the study area. Data are from the Individual Tree Species Atlas (Ellenwood et al. 2015) and represent conditions in ca. 2002.", out.width = "322.56pt", eval=F}
knitr::include_graphics( here("Results/Figures/Figure-Studyarea-outbreak.jpg"))
```

```{r severity by host presence, eval=F}
dfpred <- df

dfpred <- dfpred[dfpred$hosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]

dfpred <- dfpred[dfpred$affected01=="affected",]

results <- NULL
for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){

  res <- data.frame(hosts=NA, barkbeetle=NA, X=NA, p=NA)
  res$hosts <-j
  res$barkbeetle <- "MPB"
  if(nrow(dfpred[dfpred$hosts ==j  & dfpred$focalagent =='MPB',])>0){
      x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$hosts ==j  & dfpred$focalagent %in% c('Any', 'MPB'),], alternative="less")
      res$X <- x$statistic
      res$p <- x$p.value
  }
  x <-NULL
  results <- rbind(results, res)

  res <- data.frame(hosts=NA, barkbeetle=NA, X=NA, p=NA)
  res$hosts <-j
  res$barkbeetle <- "SB"
  if(nrow(dfpred[dfpred$hosts ==j  & dfpred$focalagent =='SB',])>0){
     x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$hosts ==j  & dfpred$focalagent %in% c('Any', 'SB'),], alternative="less")
     res$X <- x$statistic
     res$p <- x$p.value
  }
  x <-NULL
  results <- rbind(results, res)

  res <- data.frame(hosts=NA, barkbeetle=NA, X=NA, p=NA)
  res$hosts <-j
  res$barkbeetle <- "WBBB"
  if(nrow(dfpred[dfpred$hosts ==j  & dfpred$focalagent =='WBBB',])>0){
    x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$hosts ==j  & dfpred$focalagent %in% c('Any', 'WBBB'),], alternative="less")
    res$X <- x$statistic
    res$p <- x$p.value
  }
  x <-NULL
  results <- rbind(results, res)

}

results$pplot <- ifelse(results$p<0.05, "*", NA)

dfpred$pplot <- NA
for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){

  x <- dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'MPB',]
  if(nrow(x)>0){
    xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
    dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'MPB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$hosts==j & results$barkbeetle=='MPB', "pplot"]
  }

  x <- dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'SB',]
  if(nrow(x)>0){
    xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
    dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'SB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$hosts==j & results$barkbeetle=='SB', "pplot"]
  }

  x <- dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'WBBB',]
  if(nrow(x)>0){
    xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
    dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'WBBB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$hosts==j & results$barkbeetle=='WBBB', "pplot"]
  }

}

dfpred <- dfpred[!(dfpred$hosts == "PICO+PIEN" & dfpred$focalagent=="WBBB"),]
dfpred <- dfpred[!(dfpred$hosts == "PICO+ABLA" & dfpred$focalagent=="SB"),]
dfpred <- dfpred[!(dfpred$hosts == "PIEN+ABLA" & dfpred$focalagent=="MPB"),]
dfpred$hostsP <- factor(dfpred$hosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)

dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)

ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~hostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,114)+geom_text(aes(label=pplot), vjust=-0.01,size=9, col="black")+ylab("Severity (% BA affected)")

ggsave(here("Results/Figures/FigureSevertitybyPresence-outbreak.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
```

```{r severity by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))

mpb <- dfpred[dfpred$focalagent=="MPB" & dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA",   "PICO+PIEN+ABLA"), ]
sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","suitablehosts"]))
any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","suitablehosts"]))

dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]

dfpred <- dfpred[dfpred$affected01=="affected",]

results <- NULL
for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){

  x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'MPB'),], alternative="less")

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "MPB"
  res$X <- x$statistic
  res$p <- x$p.value
  x <-NULL
  results <- rbind(results, res)

    x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'SB'),], alternative="less")

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "SB"
  res$X <- x$statistic
  res$p <- x$p.value
  x <-NULL
  results <- rbind(results, res)

      x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'WBBB'),], alternative="less")

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "WBBB"
  res$X <- x$statistic
  res$p <- x$p.value
  x <-NULL
  results <- rbind(results, res)

}

results$pplot <- ifelse(results$p<0.05, "*", NA)

dfpred$pplot <- NA
for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){

  x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB',]
  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='MPB', "pplot"]

    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB',]
  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='SB', "pplot"]

    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB',]
  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='WBBB', "pplot"]

}

dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+PIEN" & dfpred$focalagent=="WBBB"),]
dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+ABLA" & dfpred$focalagent=="SB"),]
dfpred <- dfpred[!(dfpred$suitablehosts == "PIEN+ABLA" & dfpred$focalagent=="MPB"),]
dfpred$suitablehostsP <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)

dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)

ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")

ggsave(here("Results/Figures/FigureSevertitybySuitablehosts-outbreak.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
```

```{r occurrence by host presence, eval=F}
dfpred <- df
mpb <- as.data.frame(table(dfpred[dfpred$focalagent=="MPB","affected"], dfpred[dfpred$focalagent=="MPB","hosts"]))
sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","hosts"]))
wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","hosts"]))
any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","hosts"]))

results <- NULL
for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
  anyj <- any[any$Var2==j,]
  mpbj <- mpb[mpb$Var2==j,]
  sbj <- sb[sb$Var2==j,]
  wbbbj <- wbbb[wbbb$Var2==j,]

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "MPB"
  if(mpbj[1,3]+mpbj[2,3]>0){
    x <- prop.test(x=c(mpbj[1,3],anyj[1,3]), n=c(mpbj[1,3]+mpbj[2,3], anyj[1,3]+anyj[2,3]))
    res$X <- x$statistic
    res$p <- x$p.value
    res$prop1 <- x$estimate[1]
    res$prop2 <- x$estimate[2]
    res$lower <- x$conf.in[1]
    res$upper <- x$conf.in[2]
  }
  x <-NULL
  results <- rbind(results, res)

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "SB"
  if(sbj[1,3]+sbj[2,3]>0){
    x <- prop.test(x=c(sbj[1,3],anyj[1,3]), n=c(sbj[1,3]+sbj[2,3], anyj[1,3]+anyj[2,3]))
    res$X <- x$statistic
    res$p <- x$p.value
    res$prop1 <- x$estimate[1]
    res$prop2 <- x$estimate[2]
    res$lower <- x$conf.in[1]
    res$upper <- x$conf.in[2]
  }
  x <-NULL
  results <- rbind(results, res)

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "WBBB"
  if(wbbbj[1,3]+wbbbj[2,3]>0){
    x <- prop.test(x=c(wbbbj[1,3],anyj[1,3]), n=c(wbbbj[1,3]+wbbbj[2,3], anyj[1,3]+anyj[2,3]), alternative="less")
    res$X <- x$statistic
    res$p <- x$p.value
    res$prop1 <- x$estimate[1]
    res$prop2 <- x$estimate[2]
    res$lower <- x$conf.in[1]
    res$upper <- x$conf.in[2]
    x <-NULL
  }
  results <- rbind(results, res)
}

results$p <- round(results$p,3)

x <- dfpred[is.na(dfpred$suitablehosts)==FALSE,]
x <- x[x$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"),]
x[x$suitablehosts == "PICO+PIEN" & x$focalagent=="WBBB","affected"]<- 0
x[x$suitablehosts == "PICO+ABLA" & x$focalagent=="SB","affected"]<- 0
x[x$suitablehosts == "PIEN+ABLA" & x$focalagent=="MPB","affected"]<- 0

xx <- x %>% group_by(suitablehosts,focalagent) %>% count(affected)
xxx <- x %>% group_by(suitablehosts) %>% count()

xxxx <- merge(xx, xxx, by="suitablehosts")
xxxx$prop <- xxxx$n.x/xxxx$n.y
xxxx <- xxxx[xxxx$affected=="affected",]
xxxx <- xxxx[,c(1,2,6)]
colnames(xxxx) <- c("suitablehosts", "barkbeetle", "prop")

resultsx <- merge(results, xxxx, id.vars=c("suitablehosts", "barkbeetle"), all=T)
resultsx$barkbeetle <- factor(resultsx$barkbeetle, levels=c("MPB", 'SB', "WBBB", "Any"), ordered=T)
resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
resultsx$suitablehosts <- factor(resultsx$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
resultsx$pplot[is.na(resultsx$pplot) ==TRUE] <- ""

ggoccurrence <- ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~suitablehosts, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")

ggsave( here("Results/Figures/FigureOccurrencebyPresence-outbreak.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.5,units="in")

```

To determine if the proportion of plots affected by any bark beetle species was significantly greater (p <0.05) than the proportion affected by an individual species, we used a one-sided proportion test. Next, to determine if the severity of cumulative bark beetle activity was greater than the severity attributed to an individual species, we subset our data to only the plots affected by at least one bark beetle species. We then used Wilcox rank sum test to determine if cumulative mortality was greater that mortality due to an individual bark beetle species.

```{r occurrence by suitable hosts, eval=F}
dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
mpb <- as.data.frame(table(dfpred[dfpred$focalagent=="MPB","affected"], dfpred[dfpred$focalagent=="MPB","suitablehosts"]))
sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","suitablehosts"]))
any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","suitablehosts"]))

results <- NULL
for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
  anyj <- any[any$Var2==j,]
  mpbj <- mpb[mpb$Var2==j,]
  sbj <- sb[sb$Var2==j,]
  wbbbj <- wbbb[wbbb$Var2==j,]

  x <- prop.test(x=c(mpbj[1,3],anyj[1,3]), n=c(mpbj[1,3]+mpbj[2,3], anyj[1,3]+anyj[2,3]))

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "MPB"
  res$X <- x$statistic
  res$p <- x$p.value
  res$prop1 <- x$estimate[1]
  res$prop2 <- x$estimate[2]
  res$lower <- x$conf.in[1]
  res$upper <- x$conf.in[2]
  x <-NULL
  results <- rbind(results, res)

  x <- prop.test(x=c(sbj[1,3],anyj[1,3]), n=c(sbj[1,3]+sbj[2,3], anyj[1,3]+anyj[2,3]))

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "SB"
  res$X <- x$statistic
  res$p <- x$p.value
  res$prop1 <- x$estimate[1]
  res$prop2 <- x$estimate[2]
  res$lower <- x$conf.in[1]
  res$upper <- x$conf.in[2]
  x <-NULL
  results <- rbind(results, res)

   x <- prop.test(x=c(wbbbj[1,3],anyj[1,3]), n=c(wbbbj[1,3]+wbbbj[2,3], anyj[1,3]+anyj[2,3]), alternative="less")

  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
  res$suitablehosts <-j
  res$barkbeetle <- "WBBB"
  res$X <- x$statistic
  res$p <- x$p.value
  res$prop1 <- x$estimate[1]
  res$prop2 <- x$estimate[2]
  res$lower <- x$conf.in[1]
  res$upper <- x$conf.in[2]
  x <-NULL
  results <- rbind(results, res)
}

results$p <- round(results$p,3)

x <- dfpred[is.na(dfpred$suitablehosts)==FALSE,]
x <- x[x$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"),]
x[x$suitablehosts == "PICO+PIEN" & x$focalagent=="WBBB","affected"]<- 0
x[x$suitablehosts == "PICO+ABLA" & x$focalagent=="SB","affected"]<- 0
x[x$suitablehosts == "PIEN+ABLA" & x$focalagent=="MPB","affected"]<- 0

xx <- x %>% group_by(suitablehosts,focalagent) %>% count(affected)
xxx <- x %>% group_by(suitablehosts) %>% count()

xxxx <- merge(xx, xxx, by="suitablehosts")
xxxx$prop <- xxxx$n.x/xxxx$n.y
xxxx <- xxxx[xxxx$affected=="affected",]
xxxx <- xxxx[,c(1,2,6)]
colnames(xxxx) <- c("suitablehosts", "barkbeetle", "prop")

resultsx <- merge(results, xxxx, id.vars=c("suitablehosts", "barkbeetle"), all=T)
resultsx$barkbeetle <- factor(resultsx$barkbeetle, levels=c("MPB", 'SB', "WBBB", "Any"), ordered=T)
resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
resultsx$suitablehosts <- factor(resultsx$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
resultsx$pplot[is.na(resultsx$pplot) ==TRUE] <- ""

ggoccurrence <- ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~suitablehosts, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")

ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts-outbreak.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.5,units="in")

```

### 2021-11-22 Code/Manuscript-SMOTE-outbreak.Rmd
including multiple ways for determining presence

--- 2021-11-22 Code/Manuscript-2.Rmd
@@
 editor_options: 
-  chunk_output_type: console
+  chunk_output_type: inline
 ---
@@
 df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-df$hostsFP <- factor(df$hosts,levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-
-df$nhosts25F <-factor(df$nhosts25, levels=1:3, labels=c("one", "two", "three"), ordered=T)
-df$hosts25 <- factor(df$hosts25, levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
-df$hosts25F <- factor(df$hosts25, levels=c("none","PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-df$hostsF25P <- factor(df$hosts25,levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+
+df$nhosts25F <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
+df$hosts25 <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
+df$hosts25F <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
 
@@
 ## Analyses
-
-```{r import processed data}
-df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
-df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
-df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
-df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-df$hostsFP <- factor(df$hosts,levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-
-df$nhosts25F <-factor(df$nhosts25, levels=1:3, labels=c("one", "two", "three"), ordered=T)
-df$hosts25 <- factor(df$hosts25, levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
-df$hosts25F <- factor(df$hosts25, levels=c("none","PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-df$hostsF25P <- factor(df$hosts25,levels=c("none", "PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("none", "PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-
-df$nagentsF <- factor(df$nagents, levels=0:3, ordered=T)
-df$agentsF <- factor(df$agents, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)
-
-df$nagentsoF <- factor(df$nagentso, levels=0:3, ordered=T)
-df$agentsoF <- factor(df$agentso, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB","MPB+WBBB",  "SB+WBBB", "MPB+SB+WBBB"), ordered=T)
-
-df$focalhost <- factor(df$focalhost, levels=c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("PICO", "PIEN", "ABLA", "Any"), ordered=T)
-
-df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Any"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)
-
-df$propBA.PICO <- df$BA.PICO / df$BA.total *100
-df$propBA.PIEN <- df$BA.PIEN / df$BA.total *100
-df$propBA.ABLA <- df$BA.ABLA / df$BA.total *100
-```
 
@@
 
-```{r set up RF helper model}
-rf.mod <- function(response="agentMPB", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO"), datBB.c=df[df$focalhost=="Any",]){
+```{r fit MPB RF model, eval=F,fig.show='hide'}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
+datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)
+
 splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
@@
 for(z in vars.final){
-    pdz <- partial(ranger_obj, train=datz.train, pred.var=z,   which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T,   trim.outliers=T)
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
   pdz$var <- z
@@
 write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
-}
-```
-
-```{r fit MPB RF model occurrence, eval=F, fig.show='hide'}
-rf.mod(response="agentMPB", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO") , datBB.c =df[df$focalhost=="Any",])
-```
-
-```{r fit SB RF model occurrence, eval=F, fig.show='hide'}
-rf.mod(response="agentSB", vars=c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") , datBB.c =df[df$focalhost=="Any",])
-```
-
-```{r fit WBBB RF model occurrence, eval=F, fig.show='hide'}
-rf.mod(response="agentWBBB", vars=c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") , datBB.c =df[df$focalhost=="Any",])
-```
-
-```{r fit MPB RF model outbreak, eval=F, fig.show='hide'}
-rf.mod(response="agentMPBo", vars=c("BA.PICO", "QMD.PICO", "propBA.PICO") , datBB.c =df[df$focalhost=="Any",])
-```
-
-```{r fit SB RF model outbreak, eval=F, fig.show='hide'}
-rf.mod(response="agentSBo", vars=c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") , datBB.c =df[df$focalhost=="Any",])
-```
-
-```{r fit WBBB RF model outbreak, eval=F, fig.show='hide'}
-rf.mod(response="agentWBBBo", vars=c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") , datBB.c =df[df$focalhost=="Any",])
+```
+
+```{r fit SB RF model, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", paste0("pred", response))
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pdale <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pdz$method <- "PD"
+
+  alez <- ALEPlot(X=datz.train[,vars.final], X.model=ranger_obj, pred.fun=function(X.model, newdata){predict(X.model, newdata)$predictions[,2]}, J=which(vars.final==z), K=50)
+  alez <- data.frame(value=alez$x.values, yhat=alez$f.values, var=z, method="ALE")
+
+  pdale <- rbind(pdale, pdz)
+  pdale <- rbind(pdale, alez)
+}
+write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", paste0("pred", response))
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pdale <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pdz$method <- "PD"
+
+  alez <- ALEPlot(X=datz.train[,vars.final], X.model=ranger_obj, pred.fun=function(X.model, newdata){predict(X.model, newdata)$predictions[,2]}, J=which(vars.final==z), K=50)
+  alez <- data.frame(value=alez$x.values, yhat=alez$f.values, var=z, method="ALE")
+
+  pdale <- rbind(pdale, pdz)
+  pdale <- rbind(pdale, alez)
+}
+write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit MPB RF model outbreak, eval=F}
+response <- "agentMPBo"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO") # predictor variables
+datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit) # create object with training data
+datz.test <- testing(splitit) # create object with testing data
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights (not used in SMOTE approach)
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", paste0("pred", response))
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit SB RF model outbreak, eval=F}
+response <- "agentSBo"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN") # predictor variables
+datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit) # create object with training data
+datz.test <- testing(splitit) # create object with testing data
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights (not used in SMOTE approach)
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", paste0("pred", response))
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model outbreak, eval=F}
+response <- "agentWBBBo"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA") # predictor variables
+datBB.c <- df[df$focalhost=="Any",] # dataframe containing response and predictors (here i had to subset it because of how my data was arranged)
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit) # create object with training data
+datz.test <- testing(splitit) # create object with testing data
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
+# set sample weights (not used in SMOTE approach)
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", paste0("pred", response))
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(strsplit(response, split="agent")[[1]][2], split="o")[[1]][1], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
 ```
@@
 
+ggplot(hostz, aes(x=hosts, y=n, fill=var))+geom_bar(stat="identity", position = position_dodge2(width = 0.9, preserve = "single"))+theme(legend.position="bottom")+xlab("")+ylab("number of plots")
+
 ```
@@
 
-To determine if the severity of cumulative bark beetle-driven tree mortality was greater in stands with multiple susceptible hosts, we subsetted our data to only the plots affected by at least one bark beetle species.
-We then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present.
-Followed by a Dunn test to determine if difference existed between groups, again accounting for false discovery rate using the Benjamini and Hochberg (1995) method.
-
-```{r BB occurrence by hosts, eval=T}
+```{r occurrence by hosts, eval=T}
 res <- dfpred %>% count(hosts, affected) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
@@
 
-ggpallpres <- ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Presence-agentpresence-hostpresence.jpg"), plot=ggpallpres, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB outbreak occurrence by hosts, eval=T}
+ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+```
+
+```{r outbreak occurrence by hosts, eval=T}
 res <- dfpred %>% count(hosts, affectedo) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
@@
 
-ggpallout <- ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-hostspresence.jpg"), plot=ggpallout, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB occurrence by suitablehosts}
+ggplot(resp.hosts, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+```
+
+```{r occurrence by suitablehosts}
 ## Suitable hosts
@@
 
-ggpsuitablepres <- ggplot(resp.suitablehosts, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Presence-agentpresence-suitablehosts.jpg"), plot=ggpsuitablepres, width=fig.width1,height=fig.width1,units="in")
-
-```
-
-```{r BB outbreak occurrence by suitable hosts}
+ggplot(resp.suitablehosts, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+
+```
+
+```{r outbreak occurrence by suitable hosts}
 ## Suitable hosts
@@
 
-ggpsuitableoutbreak<- ggplot(resp.suitablehostso, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-suitablehosts.jpg"), plot=ggpsuitableoutbreak, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB occurrence by hosts25}
-## 
-res <- df[!df$hosts25=="none",] %>% count(hosts25, affected) %>% group_by(hosts25) %>% mutate(prop = prop.table(n))
-x <-  df[!df$hosts25=="none",] %>% count(hosts25)
-
-resa <- res[res$affected=="affected",]
-
-resa$alln <- x$n
-
-x <- prop.test(x=resa$n, n=resa$alln)
-xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
-xxp <-rstatix::cor_gather(xx$p.value)
-xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
-xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
-xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
-xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)
-
-resp <- res[res$affected=="affected",]
-resp <- merge(resp, xxpr, by.x="hosts25", by.y="Group")
-resp <- resp[order(resp$prop),]
-resp$Letter2 <- c("a", "b", "b", "c", "c", "c", "bc")
-
-resp$hosts25 <- factor(as.character(resp$hosts25), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-resp.hosts25 <- resp
-
-ggphost25pres <- ggplot(resp.hosts25, aes(x=hosts25,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Presence-agentpresence-hosts25.jpg"), plot=ggphost25pres, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB outbreak occurrence by hosts25}
-## Hosts 25
-res <- df[!df$hosts25=="none",] %>% count(hosts25, affectedo) %>% group_by(hosts25) %>% mutate(prop = prop.table(n))
-x <- df[!df$hosts25=="none",] %>% count(hosts25)
-
-resa <- res[res$affectedo=="affected",]
-
-resa$alln <- x$n
-
-x <- prop.test(x=resa$n, n=resa$alln)
-xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
-xxp <-rstatix::cor_gather(xx$p.value)
-xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
-xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
-xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
-xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)
-
-resp <- res[res$affectedo=="affected",]
-resp <- merge(resp, xxpr, by.x="hosts25", by.y="Group")
-resp <- resp[order(resp$prop),]
-resp$Letter2 <- c("a", "abc", "b", "b", "c", "c", "c")
-
-resp$hosts25 <- factor(as.character(resp$hosts25), levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-resp.hosts25o <- resp
-
-ggphost25outbreak <- ggplot(resp.hosts25o, aes(x=hosts25,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Presence-agentoutbreak-hosts25.jpg"), plot=ggphost25outbreak, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB severity by hosts, eval=F}
+ggplot(resp.suitablehostso, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+```
+
+To determine if the severity of cumulative bark beetle-driven tree mortality was greater in stands with multiple susceptible hosts, we subsetted our data to only the plots affected by at least one bark beetle species.
+We then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present.
+Followed by a Dunn test to determine if difference existed between groups, again accounting for false discovery rate using the Benjamini and Hochberg (1995) method.
+
+```{r severity by hosts, eval=F}
 dfpred <- df
@@
 dfpred <- dfpred[dfpred$affected=="affected",]
+dfpred$hosts <- factor(dfpred$hosts, levels=levels(dfpred$hosts),  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
 
@@
 
-res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
+res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA))
 
@@
 res3 <- res3[order(res3$median),]
-res3$Letter2 <- c("a", "ab", "bc", "c", "cd", "d", "d")
-res3$Group <- factor(res3$Group, levels=levels(df$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-ggpallseverity <- ggplot(dfpred, aes(x=hostsFP, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group)) + geom_text(data=res3, aes(label = n, x = Group, y = median*100 + 4, fill=Group),size = 3) + ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Severity-agentpresence-hostpresence.jpg"), plot=ggpallseverity, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB outbreak severity by hosts, eval=F}
-dfpred <- df
-dfpred <- dfpred[dfpred$focalagent=="Any",]
-dfpred <- dfpred[dfpred$affectedo=="affected",]
-
-x <- kruskal.test(propTBA~hosts, data=dfpred)
-DT <- dunnTest(propTBA~hosts, data=dfpred,method="bh")
-res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-
-res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
-
-res3 <- merge(res, res2, by.x="Group", by.y="hosts")
-res3 <- res3[order(res3$median),]
-res3$Letter2 <- c("a", "b", "bc", "bc", "cd", "cd", "d")
-res3$Group <- factor(res3$Group, levels=levels(df$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-ggoallseverity <- ggplot(dfpred, aes(x=hostsFP, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey") + geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group)) + geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3) + ylab("severity (% BA)") +xlab("hosts") + theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-hostpresence.jpg"), plot=ggoallseverity, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB severity by suitable hosts, eval=F}
+res3$Letter2 <- c("a", "a", "b", "bc", "c", "c", "c")
+
+ggsi <- ggplot(dfpred, aes(x=hosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")
+
+ggsave(here("Results/Figures/FigureSeverityby-AllHosts-outbreak.jpg"), plot=ggsi, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r severity by suitable hosts, eval=F}
 dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
@@
 
-res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA),n=n())
+res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA))
 
@@
 res3 <- res3[order(res3$median),]
-res3$Letter2 <- c("a", "b", "b", "bc", "b", "c", "bc")
-
-ggpsuitablehostsseverity <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +2, fill=Group),size = 3)+ ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Severity-agentpresence-suitbalehosts.jpg"), plot=ggpsuitablehostsseverity, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB outbreak severity by suitable hosts, eval=F}
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[dfpred$focalagent=="Any",]
-dfpred <- dfpred[dfpred$affectedo=="affected",]
-dfpred <- dfpred[!dfpred$suitablehosts=="none",]
-dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-x <- kruskal.test(propTBA~suitablehosts, data=dfpred)
-DT <- dunnTest(propTBA~suitablehosts, data=dfpred,method="bh")
-res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-
-res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
-res2$n2 <- paste0("n = ", res2$n)
-res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
-res3 <- res3[order(res3$median),]
-res3$Letter2 <- c("a", "b", "bc", "bc", "cd", "cd", "d")
-
-ggosuitablehostsseverity <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3)+ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-suitbalehosts.jpg"), plot=ggosuitablehostsseverity, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB severity by hosts25, eval=F}
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[dfpred$focalagent=="Any",]
-dfpred <- dfpred[dfpred$affected=="affected",]
-dfpred <- dfpred[!dfpred$hosts25=="none",]
-dfpred$hosts25F <- factor(dfpred$hosts25, levels=levels(dfpred$hosts25)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-x <- kruskal.test(propTBA~hosts25F, data=dfpred)
-DT <- dunnTest(propTBA~hosts25F, data=dfpred,method="bh")
-res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-
-res2 <- dfpred %>% group_by(hosts25F) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
-res2$n2 <- paste0("n = ", res2$n)
-res3 <- merge(res, res2, by.x="Group", by.y="hosts25F")
-res3 <- res3[order(res3$median),]
-res3$Letter2 <- c("abc", "a", "b", "b", "bc", "c", "bc")
-
-ggphost25severity <- ggplot(dfpred, aes(x=hosts25F, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3)+ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Severity-agentpresence-hosts25.jpg"), plot=ggphost25severity, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r BB outbreak severity by hosts25, eval=F}
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[dfpred$focalagent=="Any",]
-dfpred <- dfpred[dfpred$affectedo=="affected",]
-dfpred <- dfpred[!dfpred$hosts25=="none",]
-dfpred$hosts25F <- factor(dfpred$hosts25, levels=levels(dfpred$hosts25)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
-
-x <- kruskal.test(propTBA~hosts25F, data=dfpred)
-DT <- dunnTest(propTBA~hosts25F, data=dfpred,method="bh")
-res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-
-res2 <- dfpred %>% group_by(hosts25F) %>% summarize(median=median(propTBA), max=max(propTBA), n=n())
-res2$n2 <- paste0("n = ", res2$n)
-res3 <- merge(res, res2, by.x="Group", by.y="hosts25F")
-res3 <- res3[order(res3$median),]
-res3$Letter2 <- c("a", "b", "b", "ab", "b", "b", "b")
-res3[res3$Group == "PICO,\nPIEN,\n&\nABLA", "n"]<- ""
-
-ggohost25severity <- ggplot(dfpred, aes(x=hosts25F, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ geom_text(data=res3, aes(label = n, x = Group, y = median*100 +4, fill=Group),size = 3)+ylab("severity (% BA)") +xlab("hosts")+theme(legend.position="none")
-
-ggsave(here("Results/Figures/Figure-Severity-agentoutbreak-hosts25.jpg"), plot=ggohost25severity, width=fig.width1,height=fig.width1,units="in")
+res3$Letter2 <- c("a", "b", "b", "c", "bc", "c", "c")
+
+ggss <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100))+geom_boxplot(notch=F,fill="lightgrey")+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("suitable hosts")+theme(legend.position="none")
+
+ggsave(here("Results/Figures/FigureSeverityby-Suitablehosts-outbreak.jpg"), plot=ggss, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r severity combo, eval=F}
+ggs <- ggsi+theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + ggss + plot_layout(ncol=1)
+
+ggsave( here("Results/Figures/FigureSeverityby-Hosts-outbreak.jpg"), plot=ggs,  width=fig.width1, height=fig.width1*2,units="in")
+```
+
+```{r determine co-occurrence of agents by hosts, eval=F}
+df.all <- df[df$focalagent=="Any",]
+
+obsexp.dat <- as.data.frame(df.all  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
+obsexp.dat$n.exp <- NA
+obsexp.dat$freq.exp <- NA
+
+df.all.MPB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+PIEN", "PICO+ABLA"),] %>% group_by( agentMPB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+df.all.SB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+PIEN", "PIEN+ABLA"),] %>% group_by( agentSB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+df.all.WBBB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+ABLA", "PIEN+ABLA"),] %>% group_by( agentWBBB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp <- table(df.all$hosts)["PICO+PIEN"] * df.all.MPB[2,3]*df.all.SB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp <- table(df.all$hosts)["PICO+ABLA"] * df.all.MPB[2,3]*df.all.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp <- table(df.all$hosts)["PIEN+ABLA"] * df.all.SB[2,3]*df.all.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp <- table(df.all$hosts)["PICO+PIEN+ABLA"] * df.all.MPB[2,3]*df.all.SB[2,3]*df.all.WBBB[2,3]
+
+obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
+obsexp.dat <- as.data.frame(obsexp.dat)
+obsexp.dat2 <- data.frame(agents= c(obsexp.dat$agents,obsexp.dat$agents), n=c(obsexp.dat[,"n"], do.call(c,obsexp.dat[,"n.exp"])), variable=c(rep("observed",4),rep( "expected",4)))
+```
+
+```{r determine co-occurrence of agents by suitable hosts, eval=F}
+bbo.dat.ms <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"),]
+
+obsexp.dat <- as.data.frame(bbo.dat.ms  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
+obsexp.dat$n.exp <- NA
+obsexp.dat$freq.exp <- NA
+
+bbo.dat.ms.MPB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "MPB+WBBB", "MPB+SB"),] %>% group_by( agentMPB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+bbo.dat.ms.SB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "SB+WBBB", "MPB+SB"),] %>% group_by( agentSB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+bbo.dat.ms.WBBB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "MPB+WBBB", "SB+WBBB"),] %>% group_by( agentWBBB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+SB"] * bbo.dat.ms.MPB[2,3]*bbo.dat.ms.SB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+WBBB"] * bbo.dat.ms.MPB[2,3]*bbo.dat.ms.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["SB+WBBB"] * bbo.dat.ms.SB[2,3]*bbo.dat.ms.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+SB+WBBB"]*bbo.dat.ms.SB[2,3]*bbo.dat.ms.WBBB[2,3]*bbo.dat.ms.MPB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
+
+obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
+obsexp.dat <- as.data.frame(obsexp.dat)
+obsexp.dat2 <- data.frame(agents= c(obsexp.dat$agents,obsexp.dat$agents), n=c(obsexp.dat[,"n"], do.call(c,obsexp.dat[,"n.exp"])), variable=c(rep("observed",4),rep( "expected",4)))
 ```
@@
 
-```{r multiple - BB pres by hosts , eval=F}
+```{r multiple agents by hosts, eval=F}
 dfpred <-  df
@@
 
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[order(res$median),]
@@
   res$hosts <- "PICO+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[order(res$median),]
@@
   res$hosts <- "PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[order(res$median),]
@@
   res$hosts <- "PICO+PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[!res$Group =="none",]
@@
 
-ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(-5, 110) + geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+ geom_text(data=results, aes(label = n, y=-3),size = 3, col="black")+ylab("severity (% BA)")
-
-ggsave(here("Results/Figures/Figure-MultipleSeverity-agents-hosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1,units="in")
-
-```
-
-```{r multiple - BB outbreak by hosts , eval=F}
-dfpred <-  df
-
-dfpred <- dfpred[dfpred$hosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
-dfpred <-dfpred[dfpred$focalagent=="Any",]
-dfpred <- dfpred[!dfpred$agentso == "none",]
-
-results <- NULL
-  dfsub <- dfpred[dfpred$hosts =="PICO+PIEN"  & dfpred$agentso %in% c("MPB", "SB","MPB+SB"),]
-  dfsub$agentso <- as.factor(as.character(dfsub$agentso))
-  x <- kruskal.test(propTBA~agentso, data=dfsub)
-  res <- data.frame(agentsoF= c("MPB", "SB","MPB+SB"), p=NA)
-  DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
-  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
-  res$hosts <- "PICO+PIEN"
-
-  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
-  res <- res[order(res$median),]
-  res$Letter2 <- c("a", "a", "b")
-  results <- rbind(results, res)
-
-  dfsub <- dfpred[dfpred$hosts == "PICO+ABLA"  & dfpred$agentso %in% c("MPB", "WBBB","MPB+WBBB"),]
-  dfsub$agentso<- as.factor(as.character(dfsub$agentso))
-  x <- kruskal.test(propTBA~agentso, data=dfsub)
-  #DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
-  #res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-  res <- data.frame(Group=c("MPB", "WBBB","MPB+WBBB"), Letter="", MonoLetter="")
-  res$hosts <- "PICO+ABLA"
-
-  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
-  res <- res[order(res$median),]
-  res$Letter2 <- c("", "", "")
-  results <- rbind(results, res)
-
-  dfsub <- dfpred[dfpred$hosts == "PIEN+ABLA"  & dfpred$agentso %in% c("SB", "WBBB","SB+WBBB"),]
-  dfsub$agentso <- as.factor(as.character(dfsub$agentso))
-  x <- kruskal.test(propTBA~agentso, data=dfsub)
-  DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
-  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-  res$hosts <- "PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
-  res <- res[order(res$median),]
-  res$Letter2 <- c("a", "b", "b")
-  results <- rbind(results, res)
-
-  dfsub <- dfpred[dfpred$hosts == "PICO+PIEN+ABLA",]
-  dfsub$agentso <- as.factor(as.character(dfsub$agentso))
-  x <- kruskal.test(propTBA~agentso, data=dfsub)
-  DT <- dunnTest(propTBA~agentso, data=dfsub,method="bh")
-  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-  res$hosts <- "PICO+PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agentso) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agentso")
-  res <- res[!res$Group =="none",]
-  res <- res[order(res$median),]
-  res$Letter2 <- c("a", "a", "ab", "abcd", "bc", "cd", "d")
-
-  results <- rbind(results, res)
-
-results$hosts <- factor(results$hosts,
-levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-dfpred$hosts <- factor(dfpred$hosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-
-dfpred$agentsoF <- factor(dfpred$agentso,
-levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
-results$agentsoF <- factor(results$Group,
-levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
-
-ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsoF,y=propTBA*100, fill=agentsoF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(-5, 110) + geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+ geom_text(data=results, aes(label = n, y=-3),size = 3, col="black")+ylab("severity (% BA)")
-
-ggsave(here("Results/Figures/Figure-MultipleSeverity-agentso-hosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1,units="in")
-
-```
-
-```{r multiple - BB pres by suitable hosts, eval=F}
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.ticks = element_blank(), plot.margin = unit(c(4.5,4.5,0,4.5), "pt"))
+
+ggmultipelcount<- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, fill=agentsF))+geom_bar()+facet_grid(.~hosts, scales="free", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+xlab("")+ylab("frequency")+theme(strip.background = element_blank(), strip.text.x = element_blank(), plot.margin = unit(c(3.5,4.5,0,4.5), "pt"))
+
+ggmulti <- ggmultipleseverity + ggmultipelcount +plot_layout(ncol=1, heights=c(1,0.5))
+ggsave(here("Results/Figures/FigureSeverityMultiple-outbreak-allhosts.jpg"), plot=ggmulti  , width=fig.width2,height=fig.width1.5,units="in")
+
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("Bark beetle species present")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")
+
+ggsave(here("Results/Figures/FigureHotSpotSeverity-outbreak-allhosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1*0.75,units="in")
+
+```
+
+```{r multiple agents by suitable hosts, eval=F}
 dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
 dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
-dfpred <- dfpred[dfpred$affected=="affected",]
+dfpred <- dfpred[dfpred$affected01=="affected",]
 
@@
 
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[order(res$median),]
-  res$Letter2 <- c("a", "b", "c")
+  res$Letter2 <- c("a", "a", "b")
   results <- rbind(results, res)
@@
   res$suitablehosts <- "PICO+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[order(res$median),]
-  res$Letter2 <- c("a", "a", "b")
+  res$Letter2 <- c("a", "b", "c")
   results <- rbind(results, res)
@@
   res$suitablehosts <- "PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[order(res$median),]
@@
   res$suitablehosts <- "PICO+PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA), n=n()), by.x="Group", by.y="agents")
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
   res <- res[!res$Group =="none",]
   res <- res[order(res$median),]
-  res$Letter2 <- c("ab", "b", "ab", "ab", "bc", "cd", "d")
+  res$Letter2 <- c("a", "abc", "ab", "ab", "b", "c", "c")
 
@@
 
-ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~hosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(-5, 110) + geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+ geom_text(data=results, aes(label = n, y=-3),size = 3, col="black")+ylab("severity (% BA)")
-
-ggsave(here("Results/Figures/Figure-MultipleSeverity-agent-suitablehosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1,units="in")
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.ticks = element_blank(), plot.margin = unit(c(4.5,4.5,0,4.5), "pt"))
+
+ggmultipelcount<- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, fill=agentsF))+geom_bar()+facet_grid(.~suitablehosts, scales="free", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+xlab("")+ylab("frequency")+theme(strip.background = element_blank(), strip.text.x = element_blank(), plot.margin = unit(c(3.5,4.5,0,4.5), "pt"))
+
+ggmulti <- ggmultipleseverity + ggmultipelcount +plot_layout(ncol=1, heights=c(1,0.5))
+ggsave(here("Results/Figures/FigureSeverityMultiple-outbreak-suitablehosts.jpg"), plot=ggmulti  , width=fig.width2,height=fig.width1.5,units="in")
+
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("Bark beetle species present")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")
+
+ggsave(here("Results/Figures/FigureHotSpotSeverity-outbreak-suitablehosts.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1*0.75,units="in")
+
 ```
@@
 ### Appendix I: Random Forest Modeling
-
-```{r Performance occurrence}
-MPBtp <- read.csv(here("Results/RF-testperformance-agentMPB-occurrence-SMOTE.csv"))
+```{r fit MPB RF model Balanced, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write-outbreak.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write-outbreak.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write-outbreak.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write-outbreak.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write-outbreak.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+```
+
+```{r fit MPB RF model Weighted, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write-outbreak.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write-outbreak.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write-outbreak.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write-outbreak.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write-outbreak.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+```
+
+```{r fit SB RF model Balanced, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write-outbreak.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write-outbreak.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write-outbreak.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write-outbreak.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write-outbreak.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+```
+
+```{r fit SB RF model Weighted, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write-outbreak.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write-outbreak.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write-outbreak.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write-outbreak.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write-outbreak.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model Balanced, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write-outbreak.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write-outbreak.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write-outbreak.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write-outbreak.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write-outbreak.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED-outbreak.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model Weighted, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+datBB.c <- df[df$focalhost=="Any",]
+
+splitit <- initial_split(datBB.c, prop=.7, strata=response) # partition data to create test and training datasets stratified by the response class
+datz.train <- training(splitit)
+datz.test <- testing(splitit)
+datz.train.cv <- vfold_cv(datz.train, strata=response, v=10) # partition training data to allow for cross validation (again stratified by the response class) for optimization of model hyperparameters
+
+# set formula
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+# set sample weights
+wts <- 1/(table(datBB.c[,response])/ max(table(datBB.c[,response])))
+wts <- wts/sum(wts)
+
+# set model
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+# set the workflow
+rf_workflow <- workflow() %>% add_recipe(BBoccurrence_recipe) %>% add_model(rf_model)
+
+# tune model
+tgrid <- expand.grid(trees = c(100,500), min_n=c(10, 50, 100))
+rf_tune_results <- rf_workflow %>% tune_grid(resamples = datz.train.cv,  grid = tgrid, metrics = metric_set(roc_auc, sens, spec))
+
+tuneres <- rf_tune_results %>% collect_metrics() # collect metrics
+
+param_best<- rf_tune_results %>% select_best(metric = "roc_auc") # select best hyperparameters
+
+# Use backward selection to identify important variables
+#dfo.vsurf <- VSURF::VSURF(datz.train[,vars], factor(datz.train[,response]), RFimplem="ranger", parallel=T, ntree=param_best$trees, min.node.size=param_best$min_n)
+#registerDoSEQ()
+#vars.final <- unique(c(vars[dfo.vsurf$varselect.pred]))
+vars.final <- vars
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+
+rf_workflow_final <- workflow() %>% add_recipe(BBoccurrence_recipe_final) %>% add_model(rf_model_final)
+rf_workflow_final_fit<- rf_workflow_final %>% last_fit(splitit, metrics=metric_set(roc_auc, sens, spec)) #emulates the process where, after determining the best model, the final fit on the entire training set is needed and is then evaluated on the test set.
+test_performance <- rf_workflow_final_fit %>% collect_metrics()
+write-outbreak.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# generate a confusion matrix for test set
+test_predictions <- rf_workflow_final_fit %>% collect_predictions()
+confusion <- test_predictions %>% conf_mat(response,estimate = .pred_class)
+
+write-outbreak.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# variable importance
+final_model <- fit(rf_workflow_final, data=datz.train)
+ranger_obj <- pull_workflow_fit(final_model)$fit
+
+# generate predictions for entire dataset
+x <- datBB.c$PLT_CN
+x <- cbind(x, predict(ranger_obj, datBB.c)$predictions[,2])
+colnames(x) <- c("PLT_CN", "SBpred")
+write-outbreak.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+x <- as.data.frame(ranger_obj$variable.importance)
+write-outbreak.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+
+# partial dependence
+pd <- NULL
+for(z in vars.final){
+  pdz <- partial(ranger_obj, train=datz.train, pred.var=z, which.class=strsplit(response, split="agent")[[1]][2], prob=T)
+  pdz$var <- z
+  colnames(pdz) <- c("value","yhat","var")
+  pd <- rbind(pd, pdz)
+}
+write-outbreak.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED-outbreak.csv")), row.names=F)
+```
+
+```{r plot confusion matrices, eval=F}
+MPBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentMPB-occurrence-SMOTE-outbreak.csv"))
+SBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentSB-occurrence-SMOTE-outbreak.csv"))
+WBBBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-SMOTE-outbreak.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "SMOTE"
+BBct1  <- BBct
+
+MPBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentMPB-occurrence-BALANCED-outbreak.csv"))
+SBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentSB-occurrence-BALANCED-outbreak.csv"))
+WBBBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-BALANCED-outbreak.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "Balanced"
+BBct2  <- BBct
+
+MPBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentMPB-occurrence-WEIGHTED-outbreak.csv"))
+SBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentSB-occurrence-WEIGHTED-outbreak.csv"))
+WBBBct <- read-outbreak.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-WEIGHTED-outbreak.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "Weighted"
+BBct3  <- BBct
+
+BBct <- rbind(rbind(BBct1, BBct2), BBct3)
+
+ggconfused <- ggplot(BBct, aes(x=agent, y=sensitivity, fill=approach))+geom_bar(stat="identity", position="dodge")+theme(legend.position = "none") + ggplot(BBct, aes(x=agent, y=accuracy, fill=approach))+geom_bar(stat="identity", position="dodge")+plot_layout(nrow=2, guides="collect")
+
+ggsave(here("Results/Figures/Confusion-outbreak.jpg"), plot=ggconfused, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r partial dependence, eval=F}
+MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-outbreak-SMOTE.csv"))
+MPBpp$focalagent <- "MPB"
+SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-outbreak-SMOTE.csv"))
+SBpp$focalagent <- "SB"
+WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-outbreak-SMOTE.csv"))
+WBBBpp$focalagent <- "WBBB"
+
+ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
+ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
+ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
+ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
+
+x <- df[!df$focalagent=="Any",]
+x$propBA <- x$BA.host/x$BA.total
+x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
+x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
+
+xm <- melt(x, id.vars=c("focalagent", "affected01"))
+colnames(xm) <- c("focalagent", "affected01", "var", "value")
+xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
+
+pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
+
+pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
+
+ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value*100, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")
+
+ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
+
+ggsave(here("Results/Figures/PartialDependence-outbreak.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
+```
+
+```{r TestPerformance}
+MPBtp <- read.csv(here("Results/RF-testperformance-agentMPB-occurrence-outbreak-SMOTE.csv"))
 MPBtp$focalagent <- "MPB"
-SBtp <- read.csv(here("Results/RF-testperformance-agentSB-occurrence-SMOTE.csv"))
+SBtp <- read.csv(here("Results/RF-testperformance-agentSB-occurrence-outbreak-SMOTE.csv"))
 SBtp$focalagent <- "SB"
-WBBBtp <- read.csv(here("Results/RF-testperformance-agentWBBB-occurrence-SMOTE.csv"))
+WBBBtp <- read.csv(here("Results/RF-testperformance-agentWBBB-occurrence-outbreak-SMOTE.csv"))
 WBBBtp$focalagent <- "WBBB"
@@
 tpdat2$Metric <- c("ROC AUC", "sensitivity", "specificity")
-kable(tpdat2,booktabs = TRUE,caption="Accuracy statistics from Random Forest models prediciting the presence/absence of the mountain pine beetle (MPB) in lodgepole pine, spruce beetle (SB) in Engelmann spruce, and western balsam bark beetle (WBBB) in subalpine fir.
-ROC AUC is the area under the reciever operating characteristic curve.")
-```
-
-```{r Performance outbreak}
-MPBtp <- read.csv(here("Results/RF-testperformance-agentMPBo-occurrence-SMOTE.csv"))
-MPBtp$focalagent <- "MPB"
-SBtp <- read.csv(here("Results/RF-testperformance-agentSBo-occurrence-SMOTE.csv"))
-SBtp$focalagent <- "SB"
-WBBBtp <- read.csv(here("Results/RF-testperformance-agentWBBBo-occurrence-SMOTE.csv"))
-WBBBtp$focalagent <- "WBBB"
-
-tpdat <- rbind(rbind(MPBtp, SBtp), WBBBtp)
-tpdat$.estimate <-round(tpdat$.estimate, digits=2)
-tpdat2 <- tpdat[,c(".metric", "focalagent", ".estimate")] %>% spread(key=focalagent, value=.estimate)
-colnames(tpdat2) <- c("Metric","MPB", "SB", "WBBB")
-tpdat2$Metric <- c("ROC AUC", "sensitivity", "specificity")
-kable(tpdat2,booktabs = TRUE,caption="Accuracy statistics from Random Forest models prediciting the presence/absence of outbreaks of mountain pine beetle (MPB) in lodgepole pine, spruce beetle (SB) in Engelmann spruce, and western balsam bark beetle (WBBB) in subalpine fir.
-ROC AUC is the area under the reciever operating characteristic curve.")
-```
-
-```{r plot confusion matrices, eval=F}
-MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-SMOTE.csv"))
-SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-SMOTE.csv"))
-WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-SMOTE.csv"))
-
-BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
-BBct$var <- "presence"
-BBct1  <- BBct
-
-MPBct <- read.csv(here("Results/RF-confusiontable-agentMPBo-occurrence-SMOTE.csv"))
-SBct <- read.csv(here("Results/RF-confusiontable-agentSBo-occurrence-SMOTE.csv"))
-WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBBo-occurrence-SMOTE.csv"))
-
-BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
-BBct$var <- "outbreak"
-
-BBct <- rbind(rbind(BBct1, BBct))
-
-ggconfused <- ggplot(BBct, aes(x=agent, y=sensitivity, fill=var))+geom_bar(stat="identity", position="dodge")+theme(legend.position = "none") + ggplot(BBct, aes(x=agent, y=accuracy, fill=var))+geom_bar(stat="identity", position="dodge")+plot_layout(nrow=2, guides="collect")
-
-ggsave(here("Results/Figures/Confusion-outbreak.jpg"), plot=ggconfused, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r plot variable importance occurrence, eval=F}
-MPBvimp <- read.csv(here("Results/RF-vimp-agentMPB-occurrence-SMOTE.csv"))
-MPBvimp$focalagent <- "MPB"
-MPBvimp$variable <- c("BA", "QMD", "BA dominance")
-MPBvimp$scale <-MPBvimp$ranger_obj.variable.importance / sum(MPBvimp$ranger_obj.variable.importance )*100
-
-SBvimp <- read.csv(here("Results/RF-vimp-agentSB-occurrence-SMOTE.csv"))
-SBvimp$focalagent <- "SB"
-SBvimp$variable <- c("BA", "QMD", "BA dominance")
-SBvimp$scale <-SBvimp$ranger_obj.variable.importance / sum(SBvimp$ranger_obj.variable.importance )*100
-
-WBBBvimp <- read.csv(here("Results/RF-vimp-agentWBBB-occurrence-SMOTE.csv"))
-WBBBvimp$focalagent <- "WBBB"
-WBBBvimp$variable <- c("BA", "QMD", "BA dominance")
-WBBBvimp$scale <-WBBBvimp$ranger_obj.variable.importance / sum(WBBBvimp$ranger_obj.variable.importance )*100
-
-vimpdat <- rbind(rbind(MPBvimp, SBvimp), WBBBvimp)
-colnames(vimpdat) <- c("importance","focalagent", "variable", "scale")
-vimpdat$focalagent <- factor(vimpdat$focalagent)  
-vimpdat$variable <- factor(vimpdat$variable)  
-vimpplot <- ggplot(vimpdat, aes(x=variable, y=scale, fill=focalagent))+geom_bar(stat="identity", position="dodge")+theme(legend.position="bottom", legend.title=element_blank(), axis.title.y = element_blank())+coord_flip()+ylab("variable importance\n(% contribution to overall model fit)")+scale_fill_manual(values=p1col[1:3])
-
-ggsave(here("Results/Figures/VIMP-occurrence.jpg"), plot=vimpplot, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r plot variable importance outbreak, eval=F}
-MPBvimp <- read.csv(here("Results/RF-vimp-agentMPBo-occurrence-SMOTE.csv"))
-MPBvimp$focalagent <- "MPB"
-MPBvimp$variable <- c("BA", "QMD", "BA dominance")
-MPBvimp$scale <-MPBvimp$ranger_obj.variable.importance / sum(MPBvimp$ranger_obj.variable.importance )*100
-
-SBvimp <- read.csv(here("Results/RF-vimp-agentSBo-occurrence-SMOTE.csv"))
-SBvimp$focalagent <- "SB"
-SBvimp$variable <- c("BA", "QMD", "BA dominance")
-SBvimp$scale <-SBvimp$ranger_obj.variable.importance / sum(SBvimp$ranger_obj.variable.importance )*100
-
-WBBBvimp <- read.csv(here("Results/RF-vimp-agentWBBBo-occurrence-SMOTE.csv"))
-WBBBvimp$focalagent <- "WBBB"
-WBBBvimp$variable <- c("BA", "QMD", "BA dominance")
-WBBBvimp$scale <-WBBBvimp$ranger_obj.variable.importance / sum(WBBBvimp$ranger_obj.variable.importance )*100
-
-vimpdat <- rbind(rbind(MPBvimp, SBvimp), WBBBvimp)
-colnames(vimpdat) <- c("importance","focalagent", "variable", "scale")
-vimpdat$focalagent <- factor(vimpdat$focalagent)  
-vimpdat$variable <- factor(vimpdat$variable)  
-vimpplot <- ggplot(vimpdat, aes(x=variable, y=scale, fill=focalagent))+geom_bar(stat="identity", position="dodge")+theme(legend.position="bottom", legend.title=element_blank(), axis.title.y = element_blank())+coord_flip()+ylab("variable importance\n(% contribution to overall model fit)")+scale_fill_manual(values=p1col[1:3])
-
-ggsave(here("Results/Figures/VIMP-outbreak.jpg"), plot=vimpplot, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r partial dependence occurrence, eval=F}
-df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
-df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)
-
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
-x$affected01 <- ifelse(x$affected=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-xm$value <- as.numeric(xm$value)
-xm <- na.omit(xm)
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)
-
-ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-
-ggsave(here("Results/Figures/PartialDependence-occurrence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
-```
-
-```{r partial dependence outbreak, eval=F}
-df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
-df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)
-
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPBo-occurrence-SMOTE.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSBo-occurrence-SMOTE.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBBo-occurrence-SMOTE.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
-x$affected01 <- ifelse(x$affected=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-xm$value <- as.numeric(xm$value)
-xm <- na.omit(xm)
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="PD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)
-
-ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-
-ggsave(here("Results/Figures/PartialDependence-outbreak.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
-```
-
-```{r ale occurrence, eval=F}
-df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
-df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)
-
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
-x$affected01 <- ifelse(x$affected=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-xm$value <- as.numeric(xm$value)
-xm <- na.omit(xm)
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)
-
-ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-
-ggsave(here("Results/Figures/ALE-occurrence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
-```
-
-```{r ale outbreak, eval=F}
-df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
-df$focalagent <- factor(df$focalhost, levels= c("PICO", "PIEN", "ABLA", "Allhosts"), labels=c("MPB", "SB", "WBBB", "Any"), ordered=T)
-
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPBo-occurrence-SMOTE.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSBo-occurrence-SMOTE.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBBo-occurrence-SMOTE.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected")]
-x$affected01 <- ifelse(x$affected=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-xm$value <- as.numeric(xm$value)
-xm <- na.omit(xm)
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 60)
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")+xlim(0, 60)
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD"& ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(10, 100)
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")+xlim(10, 100)
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA" & ppdat$method=="ALE",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )+xlim(0, 100)
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")+xlim(0, 100)
-
-ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-
-ggsave(here("Results/Figures/ALE-outbreak.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
-```
-
+kable(tpdat2,booktabs = TRUE,caption="A caption")
+```
 Consistent with existing stand hazard ratings, the probability of each bark beetle species occurring generally increased with host basal area, quadratic mean diameter, and percent basal area (Fig.

### 2021-11-22 Code/Manuscript-SMOTE.Rmd
including multiple ways for determining presence

@@
 title: "Host richness increases the occurrence but not the severity of bark beetle-induced tree mortality"
-author: "Sarah J. Hart et al."
+author: "Sarah Hart"
 email: "sarah.hart@colostate.edu"
@@
 ---
-
 ```{r setup, include=FALSE}
@@
 
-Key findings
-
--
+## Key findings
 
@@
 
-Paragraph 1: Theoretical context - plant & insect interactions
-
 Many studies have sought to understand how biodiversity influences interactions between natural enemies and their resources.
-In the case of plant-herbivore interactions, greater plant diversity often reduces herbivory on plants at both the individual and population level [i.e., 'associational resistance', @barbosa_associational_2009].
-This is hypothesized to occur when insect herbivores, particularly specialists, are less likely to find and stay in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973].
-This may occur if insects are less likely to encounter their host due to decrease in host frequency or abundance [i.e.
-'host dilution effect', @], heterospecific neighbors chemically mask host trees [i.e.
-'host apparency hypothesis', @castagneryrol_effects_2014], and/or mixed stands favor greater abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020].
-
-#However, increased plant diversity may also result in an increase in herbivore richness ['resource specialization hypothesis', @keddy_plant_1984].
-
-Yet little research has examined community-level outcomes, which are often hard to quantify because of the number of potential interactions increases dramatically with increasing community diversity, particularly when natural enemies are generalists.
-Further, the effects of different natural enemies on their focal host populations often differ greatly, due to differences in susceptibility and mortality rates and the availability and quality (as viewed by their natural enemies) of resource communities.
-Critically, community-level outcomes may drive ecosystem processes, particularly when community diversity is low and plant-herbivore relationships are highly dynamic.
-In this context there is a particular need to understand the effects of tree diversity on outbreaks of irruptive insects, which in recent years (2003-2012) have affected 85 million hectares of forest globally, or ca.
-18 million hectares more than wildfire [@van_lierop_global_2015], resulting in widespread changes in carbon sequestration [e.g., @kurz_mountain_2008], timber production [e.g.
-@schowalter_insect_2012], and water quality [e.g., @Mikkelson_bark_2013], among other things.
-
-Paragraph 2: Bark beetles
-
-Bark beetles (Curculionidae: Scolytinae) are among the few native insect species that can kill large numbers of trees in a single year.
-For instance in the western United States, recent outbreaks of native bark beetles have killed 3.8 billion trees [1997-2018; @hicke_characterizing_2020].
-Bark beetles bore through the bark, where they mate and oviposit their eggs.
-Concurrently, bark beetles introduce pathogenic fungi Larvae feeding upon the phloem and fungal spread stop the translocation of water and nutrients and cause tree death.
-Conifer defense against bark beetles consists primarily of resin exudation that physically expels the beetle and allelochemicals, which repel or kill beetles.
-To overcome these defenses and colonize live trees, bark beetles rely on a mass-attack strategy, where pioneering beetles emit aggregation pheromones that call conspecifics to the tree.
-Typically bark beetles exist at low population levels and attack weakened trees, but as populations increase bark beetles attack increasingly better defended trees.
-Such severe mortality occurs only during outbreaks when pheromone-mediated mass-attack allows bark beetles to overcome tree defenses.
-
-For instance, a primary goal of the Western Bark Beetle is to promote resilience of forests to bark beetle outbreaks by increasing the diversity of age classes and tree species.
-
-Paragraph 3: Bark beetles
-
-To better understand how community diversity influences interactions between bark beeetles and their resources, we use a natural system with inherently low tree and bark beetle diversity.
-Specifically our research focuses on tree mortality due to three bark beetle species, the mountain pine beetle (*Dendroctonus ponderosae*), spruce beetle (*D.
-rufipennis*), and western balsam bark beetle (*Dryocoetes confusus*), which in subalpine forests of Western North America predominantly attack lodgepole pine (*Pinus contorta*), Engelmann spruce (*Picea engelmannii*), and subalpine fir (*Abies lasiocarpa*), respectively.
+In the case of plant-herbivore interactions, greater plant diversity often reduces the effects of herbivory [i.e., 'associational resistance', @barbosa_associational_2009], particularly when herbivores are specialists capable of rapid population growth [@otway_resource_2005].
+While increases in plant diversity may decrease the effects of herbivore species on individuals and populations of plants, it is also hypothesized to increase herbivore richness ['resource specialization hypothesis', @keddy_plant_1984], which may result in more efficient resource use and greater depletion of the plant community, especially when herbivores are specialists [@cardinale_effects_2005; @finke_niche_2008].
+Thus plant community-level outcomes of herbivory may depend upon both direct and indirect effects of plant diversity.
+Yet little research has examined the effects of plant diversity on community-level outcomes of herbivory, which critically may drive key ecosystem processes.
+This is particularly true in natural systems, where the number of potential interactions is high, the effects of different herbivores on their focal host populations often differ greatly, and the changes in plant diversity are often convoluted with changes in host quality [@jactel_tree_2021].
+
+In the western United States, native bark beetles (Curculionidae: Scolytinae) have killed 3.8 billion trees over the past two decades [1997-2018; @hicke_characterizing_2020].
+While there are more than 25 native bark beetle species, the mountain pine beetle (MPB; *Dendroctonus ponderosae*), spruce beetle (SB;  *Dendroctonus rufipennis*), and western balsam bark beetle (WBBB; *Dryocoetes confusus*) collectively have caused over 70% of the bark beetle-attributed tree mortality, most of which has occurred in high elevation forests [@hicke_chacterizing_2020].
+
+Associational resistance is commonly hypothesized to occur due to several non-mutually exclusive mechanisms.
+First, herbivory may be lower in more diverse systems if herbivores are less likely to encounter, stay, and reproduce in areas where their hosts are less abundant [i.e., 'resource concentration hypothesis', @root_organization_1973].
+Second, reductions in herbivory may occur if insects are less likely to find their hosts due to chemical or visual masking by heterospecific neighbors [i.e.
+'host apparency hypothesis', @feeny_plant_1976, @sciebe_semiochemical_2011, @castagneryrol_effects_2014], Third, heterospecific neighbors may increase the abundance of herbivore natural enemies [i.e., 'enemies hypothesis', @elton_ecology_1985, @staab_influence_2020].
+However, evidence for associational resistance in natural (i.e.
+not plantations) forested ecosystems.
+
+Resource specialization 
+Compensatory mechanisms
+- only meaningful if populations susceptible to bb infestionat co- occur with tree populatoin ssucess to other
+
+To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).
+
 We use this system to ask:
 
-(1) Is co-occurrence of multiple trees susceptible to bark beetle attack common?
-(2) How often do stand conditions susceptible to multiple bark beetles occur?
-(3) Given stand stand conditions susceptible to multiple bark beetles, does the  the occurrence or severity of cumulative bark beetle mortality
-
-We expect that co-occurrence of trees of different host species will be common, but co-occurrence of stand conditions suitable for multiple bark beetle species will not.
-Given stands where conditions are suitable for multiple bark beetle species, occurrence of bark beetle-driven tree mortality will be greater (i.e.
-diversity begets diversity), but severity will be lower (i.e.
-resource concentration).
-
-host tree richness or identity influence the occurrence and severity of tree mortality due to bark beetles at the community-scale?
-
-To test these hypotheses, we used a large dataset consisting of XXX,XXX plots established by the United States Forest Service Forest Inventory and Analysis Program (FIA; https://www.fia.fs.fed.us/).
-Specifically we ask:
+(1) How often do stand conditions suitable for multiple bark beetle species occur?
+(2) Does the occurrence or severity of cumulative bark beetle activity increase with the host species richness?
+(3) Does the co-occurrence of multiple bark beetle species result in more severe cumulative bark beetle mortality?
+
+We expect that co-occurrence of trees of different host species will be common (i.e.
+resource specialization hypothesis), but co-occurrence of stand conditions suitable for multiple bark beetle species will not (i.e.
+associational resistance).
+
+When stand conditions are suitable for multiple bark beetle species, we expect that at the population-scale occurrence of bark beetle-driven tree mortality wi
+
+will be greater (i.e. resource specialization hypothesis), but severity will be lower (i.e. associational resistance). Alternatively, if
 
 We also expected the severity of bark beetle infestation to vary with the number of agents present, but with two alternative hypotheses.
-If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent (i.e.
-additive effects).
+If the population dynamics of each bark beetle species are independent, then stands with multiple agents will experience higher tree mortality than stands with only one agent.
 Alternatively, lower tree mortality may occur if concurrent outbreaks of bark beetles of different species cause semiochemical confusion or if competitive release increases tree defensive capacity.
@@
 
-The study area consists of subalpine lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) forest in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, Wyoming)
+The study area consists of subalpine forests with lodgepole pine (*Pinus contorta*), subalpine fir (*Abies lasiocarpa*), and Engelmann spruce (*Picea engelmannii*) that are located in the Intermountain West (i.e., Arizona, New Mexico, Colorado, Utah, Nevada, Idaho, Montana, and Wyoming;  Fig.
+\@ref(fig:FigMap)).
+Our study area was constrained to the Intermountain West due to regional differences in data collection that exist within the FIA program [@burrill_forest_2017].
 
 ## Data
-### FIA data
-
-The FIA program is a single inventory program that includes all public and private forested land (>= 0.4 ha in size and >= 10% canopy cover) in the US.
+
+The FIA program inventories all public and private forested land using a spatially and temporally distributed sampling design, making it useful for the study of tree species distributions [e.g., @iverson_predicting_1998; @rehfeldt_empirical_2006] and tree demography [e.g., @shaw_forest_2005, @derose_effects_2013].
 In the Western US, all plots are visited once every ten years [@gray_forest_2012].
-The spatially and temporally distributed probabilistic sampling design is useful studies of the distribution of tree species [e.g., @iverson_predicting_1998; @rehfeldt_empirical_2006], forest insects [e.g., @derose_effects_2013], and tree mortality [e.g., @shaw_forest_2005]
-
-At each FIA plot, field crews collect data for trees (>= 12.7 cm DBH) within four 7.32 m radius subplots arranged in a fixed pattern.
-Data on the proximate cause of death is collected for any tree (>= 12.7 cm DBH) that was alive at the previous visit and at revisit is dead using visible evidence (e.g., fire scars, bark beetle galleries).
+At each plot, field crews collect data within four 7.32 m radius subplots arranged in a fixed pattern.
+For each tree within the plot, field crews record the species, size (diameter at breast height; DBH), and status (live or dead) for trees (>= 12.7 cm DBH).For each live tree, field crews additionally record up to three damaging agents (e.g.
+MPB), defined as any agent likely to prevent the tree from surviving >2 years, reduce the growth of the tree in the near term, or negatively affect the tree's marketable products [@burrill_forest_2017].
+For dead trees that were alive at the previous visit, the proximate cause of death is also collected for any tree that was alive at the previous visit and at revisit is dead using visible evidence (e.g., fire scars, bark beetle galleries).
+Because identifying the mortality agent is more challenging cause of death codes are broad (e.g., "insect" or "disease").
+
+We acquired all FIA data for the Intermountain West, but restricted our analyses to the most recent inventory for plots that contained a focal host species and where all subplots were inventoried.
+We then characterized stand structure and composition using only live and recently killed (i.e.
+killed within the past ~10 years) trees.
+each plot we calculated the basal area, quadratic mean diameter (QMD), and basal area dominance (% total basal area) for each host species.
+Additionally, we characterized the presence (i.e., at least one tree infested) and severity of bark beetle activity (% total basal area) by bark beetle species.
+For live Engelmann spruce, we listed the presence of SB when the agent was listed as a general bark beetle or SB specifically.
+Similarly, we listed the presence of MPB in lodgepole when agent was listed as a general bark beetle or MPB specifically.
+For live subalpine fir, we listed the presence of WBBB when the damaging was recorded as, subalpine fir decline, a general bark beetle, or WBBB specifically.
+Finally, we assumed the mortality of lodgepole pine, Engelmann spruce, and subalpine fir that was attributed an insect, was caused by the MPB, SB, and WBBB, respectively.
 
@@
 
-#### Determination of presence/absence of bark beetle activity
-
-For each live tree (>= 12.7 cm DBH), field crews record up to three damaging agents, which are defined as agents that are likely to prevent the tree from surviving >2 years, reduce the growth of the tree in the near term, or negatively affect the tree's marketable products [@burrill_forest_2017].
-
-Cause of death codes are very broad (e.g., "insect" or "disease").
-Accuracy of FIA data is commonly assessed using blind checks, where two crews perform independent inventories.
-Agreement between mortality agent codes recorded in the two inventories is generally >80% [@anderegg_tree_2015].
-Active damage is easier to identify, thus the codes for damaging agents are much more specific.
-
 ```{r determine presence of BB agents in focal hosts, eval=F}
@@
 ```
-
-#### Calculation of stand characteristics
-For each plot, we then calculated
-1 - basal area by host species
-2 - quadratic mean diameter (QMD)
-3 - basal area dominance (% total basal area) by species
-4 - presence and severity of bark beetle activity (% total basal area) by bark beetle species
 
@@
 
+#df$agentSB <- ifelse(df$stand.SB/(df$TPH.PIEN/6.018046/2.47105)>0.1, "SB", "absent")
+#df$agentMPB <- ifelse(df$stand.MPB/(df$TPH.PICO/6.018046/2.47105)>0.1, "MPB", "absent")
+#df$agentWBBB <- ifelse(df$stand.WBBB/(df$TPH.ABLA/6.018046/2.47105)>0.1, "WBBB", "absent")
+
 df$agentSB <- ifelse(df$stand.SB>0, "SB", "absent")
@@
 
-#### Selection of plots
-We selected all plots that were part of the annual inventory and where all subplots were inventoried.
-
-Within this plots, we characterized stand structure and composition using only live and recently killed (i.e. killed within the past ~10 years) trees greater that 12.7 cm DBH within the subplot.
-
-```{r select plots}
+```{r select plots, eval=F}
 df <- read.csv(here("Data/Processed/FIAdat-df.csv"))
@@
 
-## Analyses
-
 ```{r import processed data}
@@
 
-### Model the suitable stand conditions
-
-However, benefits provided by mixtures are less evident for larger-scale disturbances [@jactel_tree_2021]
-suggesting that changes in the structure of host communities, rather than biodiversity per se, can explain when a dilution effect should be observed.
-
-Construct random forest models using synthetic minority oversampling technique (SMOTE), an approach for dealing with imbalanced data.
-The SMOTE approach oversamples the minority class by synthesizing new cases from the minority class.
-We compared this approach with two other common approaches for dealing with imbalanced data in a random forest framework - balanced and weighted random forests and selected the approach that classified affected plots with the greatest accuracy.
+```{r maps all plots and hosts, eval=F}
+df.all <- df[df$focalhost == "Any", ]
+df.all$hostsFPlong <- df.all$hostsFP
+df.all$agentsFplong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))
+
+df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
+
+states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
+IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+
+ext <- as.matrix(extent(IMW))
+#ext[1,1] <- ext[1,1] -0.25
+#ext[1,2] <- ext[1,2] +0.25
+#ext[2,1] <- ext[2,1] -1
+#ext[2,2] <- ext[2,2] +1
+
+map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
+
+map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
+
+map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
+
+map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
+
+map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
+
+map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
+
+map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
+
+legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
+
+Fig.file <- here("Results/Figures/Figure-Map-Allplots.jpg")
+jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
+tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
+whatever <- dev.off()
+
+```
+
+```{r FigMap, fig.cap="a caption", out.width = "504.8pt"}
+knitr::include_graphics(here("Results/Figures/Figure-Map-Allplots.jpg"))
+```
+
+## Analyses
+
+To determine how frequently co-occurrence of multiple bark beetle species was likely to occur, we first calculated the proportion of plots where multiple host species were present.
+
+Because the effects of diversity may be convoluted with stand development processes and thus host size and availability, we limited our analyses to only the plots where host trees were susceptible to outbreak.
+To do so we used a random forest modeling framework [@breiman_random_2001], a distribution-free approach useful for modeling nonlinear relationships and complex interactions characteristic of ecological systems [@cutler_random_2007].
+We constructed separate models for the presence of MPB, SB, and WBBB and as predictor variables included host QMD, host basal area, and host dominance, which have been commonly linked to increased risk of bark beetle activity [[e.g.
+REF MPB, REF SB, REF WBBB].
+Because most stands in our dataset were unaffected by bark beetles, we used a synthetic minority oversampling technique (SMOTE) to deal the imbalance in the response [@chawla_smote_2002].
+Briefly, the SMOTE approach oversamples the minority class by synthesizing new cases from the minority class (here affected stands).
+Models were fit in R [@r_core_team_r:_2020] using the **Tidymodels** package [@kuhn_tidymodels_2020].
+We then assumed that each host species was susceptible to infestation when the predicted probability of occurrence of the focal bark beetle species was greater than 0.5.
+
+```{r describe frequency of hosts and agents, eval=T}
+## Hosts
+hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+
+nhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+
+## Affected hosts
+affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
+affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]
+
+affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
+affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]
+
+## Agents given host presence
+agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
+  mutate(freq = n / sum(n))
+PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
+PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
+ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))
+
+## Agents given presence of any agent
+agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
+  mutate(freq = n / sum(n))
+
+agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+
+p1 <- ggplot(df[df$focalhost=="Any",], aes(x=hosts, fill=interaction(hosts, affected))) + geom_bar(width=0.75, stat="count", col="black")+ylab("number of plots")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1),legend.position = "none", legend.title=element_blank())+scale_fill_manual(values=c(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), adjust_transparency(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), alpha=0.5)))
+
+Fig.file <- here("Results/Figures/Figure-PresencebyHostcombos.jpg")
+jpeg(Fig.file, width=fig.width1.5,height=2.5,units="in", res=300)
+p1
+whatever <- dev.off()
+```
 
@@
 
-```{r fit MPB RF model Balanced, eval=F}
-response <- "agentMPB"
-vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
+```{r fit SB RF model SMOTE, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
 datBB.c <- df[df$focalhost=="Any",]
@@
 # set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
+
 # set sample weights
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
 
@@
 vars.final <- vars
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
-
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
 
@@
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
-```
-
-```{r fit MPB RF model Weighted, eval=F}
-response <- "agentMPB"
-vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r fit WBBB RF model SMOTE, eval=F}
+response <- "agentWBBB"
+vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
 datBB.c <- df[df$focalhost=="Any",]
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
 
@@
 vars.final <- vars
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
-
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
 
@@
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
-```
-
-```{r fit SB RF model SMOTE, eval=F}
-response <- "agentSB"
-vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+```
+
+```{r determine suitable stands}
+MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence-SMOTE.csv"))
+SBpred <- read.csv(here("Results/RF-pred-agentSB-occurrence-SMOTE.csv"))
+WBBBpred <- read.csv(here("Results/RF-pred-agentWBBB-occurrence-SMOTE.csv"))
+
+BBpred <- merge(merge(MPBpred, SBpred, by="PLT_CN"), WBBBpred, by="PLT_CN")
+colnames(BBpred) <- c("PLT_CN", "MPBpred", "SBpred", "WBBBpred")
+
+threshold <-0.5
+BBpred$MPBsuitable <- ifelse(BBpred$MPBpred > threshold, "MPB", "absent")
+BBpred$SBsuitable <- ifelse(BBpred$SBpred > threshold, "SB", "absent")
+BBpred$WBBBsuitable <- ifelse(BBpred$WBBBpred >threshold, "WBBB", "absent")
+
+BBpred$agentsuitable <-  do.call(paste, c(BBpred[c("SBsuitable", "MPBsuitable", "WBBBsuitable")],sep="-"))
+BBpred$agentsuitable <- factor(BBpred$agentsuitable, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))
+
+BBpred$suitablehosts <- factor(BBpred$agentsuitable, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB",  "MPB+WBBB", "SB+WBBB",  "MPB+SB+WBBB"), ordered=T)
+
+```
+
+```{r determine co-occurrence of suitable stands}
+bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
+bbo.dat <- bbo.dat[bbo.dat$focalagent=="Any",]
+
+suitablehostgroups <- bbo.dat[bbo.dat$focalagent=="Any",]%>% group_by(suitablehosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+
+bbo.dat$hostsFPshort<- bbo.dat$hosts
+bbo.dat$hostsFPshort <- factor(bbo.dat$hosts, levels=levels(bbo.dat$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+bbo.dat$agentsFplongrev <- bbo.dat$agentsF
+bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
+
+bbo.sub <- bbo.dat[bbo.dat$focalhost=="Any",]
+bbo.sub <- bbo.dat[!bbo.dat$suitablehosts=="none",]
+
+bbo.sub$suitablehostsP <- factor(bbo.sub$suitablehosts, levels=levels(bbo.sub$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+
+suitablehostgroups <- bbo.sub %>% group_by(suitablehostsP) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+```
+
+```{r create co-occurrence figure}
+gghostgroups <- ggplot(hostgroups, aes(x=hosts, y=freq, fill=hosts))+geom_bar(stat="identity")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")+xlab("")+ylab("frequency")
+
+ggfreq <- ggplot(bbo.sub, aes(x=suitablehostsP,fill=suitablehostsP))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")
+
+ggsave( here("Results/Figures/FigureSuitable.jpg"), plot=ggfreq,width=fig.width1,height=fig.width1*0.75,units="in")
+
+ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")
+
+colnames(suitablehostgroups) <- colnames(hostgroups)
+hostgroups$var <- "individual host"
+suitablehostgroups$var <- "suitable host"
+
+hosts.plot.dat <- rbind(hostgroups, suitablehostgroups)
+
+gghostgroups <- ggplot(hosts.plot.dat, aes(x=hosts, y=n, fill=var))+geom_bar(stat="identity", position="dodge")+scale_fill_manual(values=c("grey", "black"))+ylab("frequency")+theme(legend.position="bottom")+theme(legend.title=element_blank(), axis.title.x=element_blank())
+ggsave( here("Results/Figures/FigureHostCo.jpg"), plot=gghostgroups,   width=fig.width1, height=fig.width1*0.75, units="in")
+```
+
+To determine if the proportion of plots affected by any bark beetle species varied with the number or identity of hosts present, we first used a proportion test to determine if groups differed followed by a pairwise proportion test to test for differences between groups.
+To account for multiple comparisons, we adjusted p-values using methods outlined by [@benjamini_false_1995].
+
+```{r occurrence by hosts 2}
+dfpred <- df
+dfpred <- dfpred[dfpred$focalagent=="Any",]
+
+res <- dfpred %>% count(hosts, affected) %>% group_by(hosts) %>% mutate(prop = prop.table(n))
+x <- dfpred %>% count(hosts)
+
+resa <- res[res$affected=="affected",]
+resa$alln <- x$n
+
+x <- prop.test(x=resa$n, n=resa$alln)
+xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
+xxp <-rstatix::cor_gather(xx$p.value)
+xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
+xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)
+
+resp <- res[res$affected=="affected",]
+resp <- merge(resp, xxpr, by.x="hosts", by.y="Group")
+resp <- resp[order(resp$prop),]
+resp$Letter2 <- c("a", "b", "b", "c", "c", "d", "e")
+resp$hosts <- factor(resp$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
+
+ggoi <- ggplot(resp, aes(x=hosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+
+ggsave( here("Results/Figures/FigureOccurrencebHosts2.jpg"), plot=ggoi,  width=fig.width1, height=fig.width1*0.75,units="in")
+
+```
+
+```{r occurrence by suitable hosts 2}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred <- dfpred[dfpred$focalagent=="Any",]
+dfpred <- dfpred[!dfpred$suitablehosts=="none",]
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+
+res <- dfpred %>% count(suitablehosts, affected) %>% group_by(suitablehosts) %>% mutate(prop = prop.table(n))
+x <- dfpred %>% count(suitablehosts)
+
+resa <- res[res$affected=="affected",]
+resa$alln <- x$n
+
+x <- prop.test(x=resa$n, n=resa$alln)
+xx <- pairwise.prop.test(x=resa$n, n=resa$alln, p.adjust.method="BH")
+xxp <-rstatix::cor_gather(xx$p.value)
+xxp$var1 <- factor(xxp$var1, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+xxp$var2 <- factor(xxp$var2, levels=1:7, labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+xxp$Comparison <- apply(xxp[,c("var1", "var2")], 1, paste, collapse=" - ")
+xxpr <- rcompanion::cldList(cor~ Comparison, data = xxp, threshold = 0.05)
+
+resp <- res[res$affected=="affected",]
+resp <- merge(resp, xxpr, by.x="suitablehosts", by.y="Group")
+resp <- resp[order(resp$prop),]
+resp$Letter2 <- c("a", "b", "c", "d", "d", "d", "e")
+resp$suitablehosts <- factor(resp$suitablehosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
+
+ggos <- ggplot(resp, aes(x=suitablehosts,y=prop))+geom_bar(stat="identity", width=0.75)+geom_text(aes(label=Letter2), vjust=-1,size=3)+ylim(0,1)+xlab("suitable hosts")+ylab("proportion affected")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+
+ggsave( here("Results/Figures/FigureOccurrencebySuitablehosts2.jpg"), plot=ggos,  width=fig.width1, height=fig.width1*0.75,units="in")
+```
+
+```{r occurence combo}
+ggo <- ggoi+theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + ggos + plot_layout(ncol=1)
+
+ggsave( here("Results/Figures/FigureOccurrencebyHosts.jpg"), plot=ggo,  width=fig.width1, height=fig.width1*1.3,units="in")
+
+```
+
+To determine if the severity of cumulative bark beetle-driven tree mortality was greater in stands with multiple susceptible hosts, we subsetted our data to only the plots affected by at least one bark beetle species.
+We then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present.
+Followed by a Dunn test to determine if difference existed between groups, again accounting for false discovery rate using the Benjamini and Hochberg (1995) method.
+
+```{r severity by hosts}
+dfpred <- df
+dfpred <- dfpred[dfpred$focalagent=="Any",]
+dfpred <- dfpred[dfpred$affected=="affected",]
+dfpred$hosts <- factor(dfpred$hosts, levels=levels(dfpred$hosts),  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
+
+x <- kruskal.test(propTBA~hosts, data=dfpred)
+DT <- dunnTest(propTBA~hosts, data=dfpred,method="bh")
+res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+
+res2 <- dfpred %>% group_by(hosts) %>% summarize(median=median(propTBA), max=max(propTBA))
+
+res3 <- merge(res, res2, by.x="Group", by.y="hosts")
+res3 <- res3[order(res3$median),]
+res3$Letter2 <- c("a", "a", "b", "bc", "c", "c", "c")
+
+ggsi <- ggplot(dfpred, aes(x=hosts, y=propTBA*100, fill=hosts))+geom_boxplot(notch=F)+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("hosts")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+
+ggsave(here("Results/Figures/FigureSevertitybyhosts.jpg"), plot=ggsi, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r severity by suitable hosts 2}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred <- dfpred[dfpred$focalagent=="Any",]
+dfpred <- dfpred[dfpred$affected=="affected",]
+dfpred <- dfpred[!dfpred$suitablehosts=="none",]
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1],  labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"), ordered=T)
+
+x <- kruskal.test(propTBA~suitablehosts, data=dfpred)
+DT <- dunnTest(propTBA~suitablehosts, data=dfpred,method="bh")
+res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+
+res2 <- dfpred %>% group_by(suitablehosts) %>% summarize(median=median(propTBA), max=max(propTBA))
+
+res3 <- merge(res, res2, by.x="Group", by.y="suitablehosts")
+res3 <- res3[order(res3$median),]
+res3$Letter2 <- c("a", "b", "c", "d", "cd", "d", "d")
+
+ggss <- ggplot(dfpred, aes(x=suitablehosts, y=propTBA*100, fill=suitablehosts))+geom_boxplot(notch=F)+geom_text(data=res3, aes(label = Letter2, x = Group, y = 5 + max*100, fill=Group))+ ylab("severity (% BA)") +xlab("suitable hosts")+scale_fill_manual(values=colorspace::adjust_transparency(p2col, alpha=0.6))+theme(legend.position="none")
+
+ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggss, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r severity combo}
+ggs <- ggsi+theme(axis.title.x = element_blank(), axis.text.x = element_blank()) + ggss + plot_layout(ncol=1)
+
+ggsave( here("Results/Figures/FigureSeveritybyHosts.jpg"), plot=ggs,  width=fig.width1, height=fig.width1*1.3,units="in")
+```
+
+```{r determine co-occurrence of agents by hosts}
+df.all <- df[df$focalagent=="Any",]
+
+obsexp.dat <- as.data.frame(df.all  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
+obsexp.dat$n.exp <- NA
+obsexp.dat$freq.exp <- NA
+
+df.all.MPB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+PIEN", "PICO+ABLA"),] %>% group_by( agentMPB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+df.all.SB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+PIEN", "PIEN+ABLA"),] %>% group_by( agentSB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+df.all.WBBB <- df.all[df.all$hosts %in% c("PICO+PIEN+ABLA", "PICO+ABLA", "PIEN+ABLA"),] %>% group_by( agentWBBB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp <- table(df.all$hosts)["PICO+PIEN"] * df.all.MPB[2,3]*df.all.SB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp <- table(df.all$hosts)["PICO+ABLA"] * df.all.MPB[2,3]*df.all.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp <- table(df.all$hosts)["PIEN+ABLA"] * df.all.SB[2,3]*df.all.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp <- table(df.all$hosts)["PICO+PIEN+ABLA"] * df.all.MPB[2,3]*df.all.SB[2,3]*df.all.WBBB[2,3]
+
+obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
+obsexp.dat <- as.data.frame(obsexp.dat)
+obsexp.dat2 <- data.frame(agents= c(obsexp.dat$agents,obsexp.dat$agents), n=c(obsexp.dat[,"n"], do.call(c,obsexp.dat[,"n.exp"])), variable=c(rep("observed",4),rep( "expected",4)))
+```
+
+```{r determine co-occurrence of agents by suitable hosts}
+bbo.dat.ms <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"),]
+
+obsexp.dat <- as.data.frame(bbo.dat.ms  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n)))
+obsexp.dat$n.exp <- NA
+obsexp.dat$freq.exp <- NA
+
+bbo.dat.ms.MPB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "MPB+WBBB", "MPB+SB"),] %>% group_by( agentMPB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+bbo.dat.ms.SB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "SB+WBBB", "MPB+SB"),] %>% group_by( agentSB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+bbo.dat.ms.WBBB <- bbo.dat[bbo.dat$focalagent=="Any" & bbo.dat$suitablehosts %in% c("MPB+SB+WBBB", "MPB+WBBB", "SB+WBBB"),] %>% group_by( agentWBBB) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+SB"] * bbo.dat.ms.MPB[2,3]*bbo.dat.ms.SB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+SB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+WBBB"] * bbo.dat.ms.MPB[2,3]*bbo.dat.ms.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["SB+WBBB"] * bbo.dat.ms.SB[2,3]*bbo.dat.ms.WBBB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="SB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp <- table(bbo.dat.ms$suitablehosts)["MPB+SB+WBBB"]*bbo.dat.ms.SB[2,3]*bbo.dat.ms.WBBB[2,3]*bbo.dat.ms.MPB[2,3]
+
+obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$freq.exp <- obsexp.dat[obsexp.dat$agents=="MPB+SB+WBBB", ]$n.exp[[1]][1] /sum(obsexp.dat$n)
+
+obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
+
+obsexp.dat <- obsexp.dat[obsexp.dat$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB"), ]
+obsexp.dat <- as.data.frame(obsexp.dat)
+obsexp.dat2 <- data.frame(agents= c(obsexp.dat$agents,obsexp.dat$agents), n=c(obsexp.dat[,"n"], do.call(c,obsexp.dat[,"n.exp"])), variable=c(rep("observed",4),rep( "expected",4)))
+```
+
+Finally, to determine ifWe then used Kruskall-Wallis test to determine if cumulative mortality differed with the identify of bark beetle species present.
+
+```{r multiple agents, eval=F}
+library(rcompanion)
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
+dfpred <- dfpred[dfpred$affected01=="affected",]
+
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+
+#dfpred <- dfpred[!(dfpred$suitablehosts=="PICO" & dfpred$agentsF %in% c("SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+#dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN" & dfpred$agentsF %in% c("MPB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+#dfpred <- dfpred[!(dfpred$suitablehosts=="ABLA" & dfpred$agentsF %in% c("MPB", "SB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+PIEN" & dfpred$agentsF %in% c("WBBB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+ABLA" & dfpred$agentsF %in% c("SB", "MPB+SB", "SB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN+ABLA" & dfpred$agentsF %in% c("MPB", "MPB+SB", "MPB+WBBB", "MPB+SB+WBBB")),]
+
+dfpred <-dfpred[dfpred$focalagent=="Any",]
+
+results <- NULL
+  dfsub <- dfpred[dfpred$suitablehosts =="PICO+PIEN"  & dfpred$agents %in% c("MPB", "SB","MPB+SB"),]
+  dfsub$agents <- as.factor(as.character(dfsub$agents))
+  x <- kruskal.test(propTBA~agents, data=dfsub)
+  res <- data.frame(agentsF= c("MPB", "SB","MPB+SB"), p=NA)
+  DT <- dunnTest(propTBA~agents, data=dfsub,method="bh")
+  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
+  res$suitablehosts <- "PICO+PIEN"
+
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res$Letter2 <- c("b", "c", "a")
+  results <- rbind(results, res)
+
+  dfsub <- dfpred[dfpred$suitablehosts == "PICO+ABLA"  & dfpred$agentsF %in% c("MPB", "WBBB","MPB+WBBB"),]
+  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
+  x <- kruskal.test(propTBA~agentsF, data=dfsub)
+  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
+  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+  res$suitablehosts <- "PICO+ABLA"
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res$Letter2 <- c("b", "c", "a")
+  results <- rbind(results, res)
+
+  dfsub <- dfpred[dfpred$suitablehosts == "PIEN+ABLA"  & dfpred$agentsF %in% c("SB", "WBBB","SB+WBBB"),]
+  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
+  x <- kruskal.test(propTBA~agentsF, data=dfsub)
+  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
+  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+  res$suitablehosts <- "PIEN+ABLA"
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res$Letter2 <- c("b", "c", "a")
+  results <- rbind(results, res)
+
+  dfsub <- dfpred[dfpred$suitablehosts == "PICO+PIEN+ABLA",]
+  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
+  x <- kruskal.test(propTBA~agentsF, data=dfsub)
+  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
+  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
+  res$suitablehosts <- "PICO+PIEN+ABLA"
+  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
+  res <- res[order(res$median),]
+  res$Letter2 <- c("a", "ab", "abc", "b", "c", "d", "d")
+
+  results <- rbind(results, res)
+  results$agentsF <- results$Group
+
+results$suitablehosts <- factor(results$suitablehosts,
+levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+
+dfpred$agentsF <- factor(dfpred$agentsF,
+levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
+results$agentsF <- factor(results$agentsF,
+levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
+
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.ticks = element_blank(), plot.margin = unit(c(4.5,4.5,0,4.5), "pt"))
+
+ggmultipelcount<- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, fill=agentsF))+geom_bar()+facet_grid(.~suitablehosts, scales="free", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+xlab("")+ylab("frequency")+theme(strip.background = element_blank(), strip.text.x = element_blank(), plot.margin = unit(c(3.5,4.5,0,4.5), "pt"))
+
+ggmulti <- ggmultipleseverity + ggmultipelcount +plot_layout(ncol=1, heights=c(1,0.5))
+ggsave(here("Results/Figures/FigureSeverityMultiple.jpg"), plot=ggmulti  , width=fig.width2,height=fig.width1.5,units="in")
+
+ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("Bark beetle species present")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")
+
+ggsave(here("Results/Figures/FigureHotSpotSeverity.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1*0.75,units="in")
+
+```
+
+# Results
+
+## Do stand conditions suitable for multiple outbreaks commonly co-occur?
+
+Across the 10,439 FIA plots with one of the target host species, 66% (n=5,850) contained at least two species (Fig.
+\@ref(fig:FigHostFreq)).
+Subalpine fir was most likely to occur with other target host species (50% of plots), followed by Engelmann spruce (45% plots), and lodgepole pine (36% of stands).
+However, random forest models predicted that only 21% of all plots (n=2,163) were suitable for more than one bark beetle species (Fig.
+\@ref(fig:FigHostFreq)).
+Again, conditions suitable for bark beetle attack in subalpine fir were most likely to co-occur with conditions suitable for bark beetle attack in another tree species (17% of stands).
+
+```{r FigHostFreq, fig.cap = "The frequency of plots by host identity.
+Gray bars indicate host species were present (i.e., at least one individual of the focal species was present within the plot).
+Black bars indicate  stand structure and composition conditions were conducive to bark beetle infestation in focal tree species.
+For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "244.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureHostCo.jpg"))
+```
+
+### Does the occurrence or severity of cumulative bark beetle activity increase with the host species richness?
+
+Stands with all three hosts present were most likely to experience bark beetle outbreaks (Fig.
+\@ref(fig:FigBBO)).
+Similarly, plots with spruce and fir present, exhibited more bark beetle activity then plots with just spruce or fir.
+However, when lodgepole pine was present there was no effect of increased host richness on the occurrence of bark beetle activity.
+
+```{r FigBBO, fig.cap = "The proportion of plots affected by bark beetles by the identity of suitable hosts.
+Letters above bars indicate significant (p<0.05) difference between groups, as determined using a proportion test.", out.width = "244.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts2.jpg"))
+```
+
+Given suitable stand conditions for multiple bark beetle species, the probability of occurrence of bark beetle activity was significantly greater than the probability of a single bark beetle species (Fig.
+\@ref(fig:FigOccurrence)).
+This effect did not depend upon the identity of susceptible hosts present or the identity of the individual bark beetle species.
+
+```{r FigOccurrence, fig.cap = "The proportion of plots affected by the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or any bark beetle (i.e., MPB, SB or WBBB) by the identity of hosts species susceptible to bark beetle infestation.
+For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.
+Astericks above bars indiciate that proportion plots affected by the individual bark beelte species was significantly less than the proportion of plots affected by any bark beetle species.",  out.width = "504pt"}
+knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
+```
+
+However, the effect of the number of susceptible hosts on the severity of bark beetle activity depended upon the identity of host species.
+In stands susceptible to all three hosts, the proportion of trees affected by bark beetles was greater than the activity due to a single bark beetle species (Fig.
+\@ref(fig:FigSeverity).
+When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most commonly occurring agent (MPB > SB > WBBB).
+Specifically significant differences did not exist between severity attributed to all bark beetles and the most commonly occurring bark beetle, but the proportion of the basal area affected by the other bark beetle was lower than the cumulative severity (Fig.
+\@ref(fig:FigSeverity).
+
+```{r FigSeverity, fig.cap= "Boxplots illustrating the severity of mortality (% of total basal area) attributed to the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or all bark beetles (i.e., cumulative mortality due to MPB, SB and WBBB) by the identity of hosts species susceptible to bark beetle infestation.
+For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.
+The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.
+Astericks above boxes indicate that severity bark beetle mortality attributed to the individual species was significantly less than the cumulative mortality due to all bark beetles.", out.width = "504pt"}
+knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
+```
+
+## Does the co-occurrence of multiple bark beetle species result in more severe cumulative bark beetle mortality?
+
+Across all 10,439 FIA plots, only 5.3% (n=539) were affected by multiple bark beetle species.
+
+Given the 2,163 plots where multiple tree species were susceptible to bark beetle infestation, 56.7% (n=1,227) were affected by at least one bark beetle species and 21.3% (n=461) were affected by multiple bark beetle species.
+The most commonly occurring agent was the MPB (n=768, 35.5% of plots), followed by the WBBB (n=503, 23.3%) and the SB (n=461, 21.3%).
+The most commonly occurring combination of bark beetle species was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and SB and WBBB (26% of case; n=120).
+The combination of all three agents was rare (9.5% of cases; n=44).
+
+When plots contained two tree species susceptible to bark beetles, the presence of both bark beetle species increased activity relative to presence of only one bark beetle species (Fig.
+\@ref(fig:FigHotSpotSeverity)).
+The increase was greatest for stands that contained both MPB and SB; the median severity of plots with both MPB and SB was 33.7 percentage points greater than MPB alone and 38.3 percentage points greater than SB alone.
+The median severity of plots with both MPB and WBBB was 7.0 percentage points greater than MPB alone and 13.6 percentage points greater than WBBB alone.
+The median severity of plots with both SB and WBBB was 14.7 percentage points greater than SB alone and 19.8 percentage points greater than WBBB alone.
+
+Given stand conditions that made all three tree species susceptible to bark beetle activity, the highest rates of bark beetle activity were in plots where both MPB and SB present; whether or not WBBB was also present did not affect mortality severity (~1 percentage point difference in median severity; Fig.
+\@ref(fig:FigHotSpotSeverity)).
+
+```{r FigHotSpotSeverity, fig.cap = "The severity of bark beetle mortality in plots with multiple tree species susceptible to bark beetles (columns) by the combinatin of bark beetle species present.
+Letters above boxes indicate significant (p < 0.05) differences between groups as determined by a Dunn test, a nonparametric rank sum test.
+The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.",  out.width = "504.8pt"}
+knitr::include_graphics( here("Results/Figures/FigureHotSpotSeverity.jpg"))
+```
+
+# Discussion
+
+# Acknowledgements
+
+# Author contributions
+
+# References
+
+<div id="refs"></div>
+
+## Supplement
+### Random Forest Modeling
+```{r fit MPB RF model Balanced, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
 datBB.c <- df[df$focalhost=="Any",]
@@
 # set formula
-BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) %>% step_smote(response) # set model preprocessing to include SMOTE to simulate more presence cases
-
+BBoccurrence_recipe <- recipe(formula(paste(response, "~", paste(vars, collapse = " + "))), datz.train) 
 # set sample weights
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts) %>% set_mode("classification")
 
@@
 vars.final <- vars
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
-
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
-```
-
-```{r fit SB RF model Balanced, eval=F}
-response <- "agentSB"
-vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
+```
+
+```{r fit MPB RF model Weighted, eval=F}
+response <- "agentMPB"
+vars <- c("BA.PICO", "QMD.PICO", "propBA.PICO")
 datBB.c <- df[df$focalhost=="Any",]
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
-```
-
-```{r fit SB RF model Weighted, eval=F}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+```
+
+```{r fit SB RF model Balanced, eval=F}
 response <- "agentSB"
 vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
-
 datBB.c <- df[df$focalhost=="Any",]
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, sample.fraction=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-BALANCED.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
-```
-
-```{r fit WBBB RF model SMOTE, eval=F}
-response <- "agentWBBB"
-vars <- c("BA.ABLA", "QMD.ABLA", "propBA.ABLA")
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-BALANCED.csv")), row.names=F)
+```
+
+```{r fit SB RF model Weighted, eval=F}
+response <- "agentSB"
+vars <- c("BA.PIEN", "QMD.PIEN", "propBA.PIEN")
+
 datBB.c <- df[df$focalhost=="Any",]
@@
 # set model
-rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, num.threads = cores) %>% set_mode("classification")
+rf_model <- rand_forest() %>% set_args( trees=tune(), min_n=tune()) %>% set_engine("ranger", importance = "permutation", seed=913, probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 vars.final <- vars
-BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c) %>% step_smote(response)
-
-rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, num.threads = cores) %>% set_mode("classification")
+BBoccurrence_recipe_final <- recipe(formula(paste(response, "~", paste(vars.final, collapse = " + "))), data=datBB.c)
+
+rf_model_final <- rand_forest() %>% set_args(trees=param_best$trees, min_n=param_best$min_n, splitrule="gini") %>% set_engine("ranger", importance = "permutation", probability=T, class.weights=wts, num.threads = cores) %>% set_mode("classification")
 
@@
 test_performance <- rf_workflow_final_fit %>% collect_metrics()
-write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(test_performance, here(paste0("Results/RF-testperformance-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 
-write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(confusion$table, here(paste0("Results/RF-confusiontable-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 colnames(x) <- c("PLT_CN", "SBpred")
-write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-pred-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
 x <- as.data.frame(ranger_obj$variable.importance)
-write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(x, here(paste0("Results/RF-vimp-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 
@@
 }
-write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-WEIGHTED.csv")), row.names=F)
 ```
@@
 
-```{r determine suitable stands}
-MPBpred <- read.csv(here("Results/RF-pred-agentMPB-occurrence-SMOTE.csv"))
-SBpred <- read.csv(here("Results/RF-pred-agentSB-occurrence-SMOTE.csv"))
-WBBBpred <- read.csv(here("Results/RF-pred-agentWBBB-occurrence-SMOTE.csv"))
-
-BBpred <- merge(merge(MPBpred, SBpred, by="PLT_CN"), WBBBpred, by="PLT_CN")
-colnames(BBpred) <- c("PLT_CN", "MPBpred", "SBpred", "WBBBpred")
-
-threshold <-0.5
-BBpred$MPBsuitable <- ifelse(BBpred$MPBpred > threshold, "MPB", "absent")
-BBpred$SBsuitable <- ifelse(BBpred$SBpred > threshold, "SB", "absent")
-BBpred$WBBBsuitable <- ifelse(BBpred$WBBBpred >threshold, "WBBB", "absent")
-
-BBpred$agentsuitable <-  do.call(paste, c(BBpred[c("SBsuitable", "MPBsuitable", "WBBBsuitable")],sep="-"))
-BBpred$agentsuitable <- factor(BBpred$agentsuitable, levels=c("absent-absent-absent","absent-MPB-absent","SB-absent-absent", "SB-MPB-absent", "absent-absent-WBBB", "SB-absent-WBBB", "absent-MPB-WBBB", "SB-MPB-WBBB"), labels=c("none", "MPB", "SB", "MPB+SB", "WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB+WBBB"))
-
-BBpred$suitablehosts <- factor(BBpred$agentsuitable, levels=c("none", "MPB", "SB", "WBBB", "MPB+SB",  "MPB+WBBB", "SB+WBBB",  "MPB+SB+WBBB"), ordered=T)
-
-```
-
-# Results
-
-## Do stand conditions suitable for multiple outbreaks commonly co-occur?
-```{r describe frequency of hosts and agents, eval=T}
-## Hosts
-hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-
-nhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-
-## Affected hosts
-affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
-affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]
-
-affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
-affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]
-
-## Agents given host presence
-agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
-  mutate(freq = n / sum(n))
-PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
-PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
-ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))
-
-## Agents given presence of any agent
-agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
-  mutate(freq = n / sum(n))
-
-agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-
-p1 <- ggplot(df[df$focalhost=="Any",], aes(x=hosts, fill=interaction(hosts, affected))) + geom_bar(width=0.75, stat="count", col="black")+ylab("number of plots")+theme(axis.text.x = element_text(angle = 45, hjust=1,vjust=1),legend.position = "none", legend.title=element_blank())+scale_fill_manual(values=c(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), adjust_transparency(c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), alpha=0.5)))
-
-Fig.file <- here("Results/Figures/Figure-PresencebyHostcombos.jpg")
-jpeg(Fig.file, width=fig.width1.5,height=2.5,units="in", res=300)
-p1
+```{r plot confusion matrices}
+MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-SMOTE.csv"))
+SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-SMOTE.csv"))
+WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-SMOTE.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "SMOTE"
+BBct1  <- BBct
+
+MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-BALANCED.csv"))
+SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-BALANCED.csv"))
+WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-BALANCED.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "Balanced"
+BBct2  <- BBct
+
+MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-WEIGHTED.csv"))
+SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-WEIGHTED.csv"))
+WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-WEIGHTED.csv"))
+
+BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
+BBct$approach <- "Weighted"
+BBct3  <- BBct
+
+BBct <- rbind(rbind(BBct1, BBct2), BBct3)
+
+ggconfused <- ggplot(BBct, aes(x=agent, y=sensitivity, fill=approach))+geom_bar(stat="identity", position="dodge")+theme(legend.position = "none") + ggplot(BBct, aes(x=agent, y=accuracy, fill=approach))+geom_bar(stat="identity", position="dodge")+plot_layout(nrow=2, guides="collect")
+
+ggsave(here("Results/Figures/Confusion.jpg"), plot=ggconfused, width=fig.width1,height=fig.width1,units="in")
+```
+
+```{r partial dependence}
+MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
+MPBpp$focalagent <- "MPB"
+SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
+SBpp$focalagent <- "SB"
+WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
+WBBBpp$focalagent <- "WBBB"
+
+ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
+ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
+ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
+ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
+
+x <- df[!df$focalagent=="Any",]
+x$propBA <- x$BA.host/x$BA.total
+x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
+x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
+
+xm <- melt(x, id.vars=c("focalagent", "affected01"))
+colnames(xm) <- c("focalagent", "affected01", "var", "value")
+xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
+
+pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
+
+pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
+
+ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value*100, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
+
+ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")
+
+ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
+
+ggsave(here("Results/Figures/PartialDependence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
+```
+
+Consistent with existing stand hazard ratings, the probability of each bark beetle species occurring generally increased with host basal area, quadratic mean diameter, and percent basal area (Fig.
+\@ref(fig:partialFig)).
+However, both MPB and WBBB exhibited a slight decrease in the probability of outbreak at very high basal areas and all three bark beetle species exhibited a lower probability of outbreak in stands where host trees were very large (i.e.
+high QMD values).
+WBBB also exhibited a decline in the probability of outbreak when subalpine fir comprised a relatively high proportion of the basal area.
+
+```{r partialFig, fig.cap="The associations between the probability of occurrence of the mountain pine beetle (MPB; red), spruce beetle (SB; blue); and western balsam bark beetle (WBBB; yellow) and three common measuree of stand suceptibility to bark beetl beetle outbreak, basal area (BA), quadratic mean diameter (QMD), and host dominace (% BA).
+The top row illustrates smoothed partial dependence results from Random Forest modeling, while in the bottom row boxplots illustrate the distribution of values for stands affected (dark colors) and unaffected (light colors) by each bark beetle species."}
+knitr::include_graphics(here("Results/Figures/PartialDependence.jpg"))
+```
+
+### Maps
+
+```{r generate maps for suitable hosts, eval=F}
+bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
+
+df.all <- bbo.dat[bbo.dat$focalhost == "Any", ]
+df.all<- df.all[!df.all$suitablehosts == "none", ]
+df.all$hostsFPlong <- df.all$hostsFP
+df.all$agentsFplong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))
+
+df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
+
+states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
+IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+
+ext <- as.matrix(extent(IMW))
+#ext[1,1] <- ext[1,1] -0.25
+#ext[1,2] <- ext[1,2] +0.25
+#ext[2,1] <- ext[2,1] -1
+#ext[2,2] <- ext[2,2] +1
+
+map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
+
+map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="SB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
+
+map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
+
+map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB+SB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
+
+map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB+WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
+
+map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="SB+WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
+
+map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$suitablehosts=="MPB+SB+WBBB",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
+
+legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
+
+Fig.file <- here("Results/Figures/Figure-Map-Suitableplots.jpg")
+jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
+tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
 whatever <- dev.off()
@@
 
-```{r determine co-occurrence of suitable stands}
-bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
-bbo.dat <- bbo.dat[bbo.dat$focalagent=="Any",]
-
-suitablehostgroups <- bbo.dat[bbo.dat$focalagent=="Any",]%>% group_by(suitablehosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-
-bbo.dat$hostsFPshort<- bbo.dat$hosts
-bbo.dat$hostsFPshort <- factor(bbo.dat$hosts, levels=levels(bbo.dat$hosts), labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-bbo.dat$agentsFplongrev <- bbo.dat$agentsF
-bbo.dat$agentsFplongrev <- factor(bbo.dat$agentsFplongrev, levels=rev(levels( bbo.dat$agentsFplong)))
-
-bbo.sub <- bbo.dat[bbo.dat$focalhost=="Any",]
-bbo.sub <- bbo.dat[!bbo.dat$suitablehosts=="none",]
-
-bbo.sub$suitablehostsP <- factor(bbo.sub$suitablehosts, levels=levels(bbo.sub$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-
-suitablehostgroups <- bbo.sub %>% group_by(suitablehostsP) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-
-```
-
-```{r}
-gghostgroups <- ggplot(hostgroups, aes(x=hosts, y=freq, fill=hosts))+geom_bar(stat="identity")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")+xlab("")+ylab("frequency")
-
-ggfreq <- ggplot(bbo.sub, aes(x=suitablehostsP,fill=suitablehostsP))+geom_bar()+theme(legend.title = element_blank())+xlab("hosts")+scale_fill_manual(values=c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"))+theme(legend.position="none")
-
-ggsave( here("Results/Figures/FigureSuitable.jpg"), plot=ggfreq,width=fig.width1,height=fig.width1*0.75,units="in")
-
-ggsave( here("Results/Figures/FigureHostFreq.jpg"), plot=gghostgroups,   width=fig.width1,height=fig.width1*0.75,units="in")
-
-colnames(suitablehostgroups) <- colnames(hostgroups)
-hostgroups$var <- "individual host"
-suitablehostgroups$var <- "suitable host"
-
-hosts.plot.dat <- rbind(hostgroups, suitablehostgroups)
-
-gghostgroups <- ggplot(hosts.plot.dat, aes(x=hosts, y=n, fill=var))+geom_bar(stat="identity", position="dodge")+scale_fill_manual(values=c("grey", "black"))+ylab("frequency")+theme(legend.position="bottom")+theme(legend.title=element_blank(), axis.title.x=element_blank())
-ggsave( here("Results/Figures/FigureHostCo.jpg"), plot=gghostgroups,   width=fig.width1, height=fig.width1*0.75, units="in")
-```
-
-Across the 10,439 FIA plots with one of the target host species, 66% (n=5,850) contained at least two of the target host species (Fig.
-\@ref(fig:FigHostFreq)).
-Subalpine fir was most likely to occur with other target host species (50% of plots), followed by Engelmann spruce (45% plots), and lodgepole pine (36% of stands).
-However, only 21% of all plots (n=2,163) were suitable for more than one bark beetle species.
-Conditions suitable for infestation in subalpine fir were most likely to co-occur with conditions suitable for infestation in another tree species (17% of stands).
-
-```{r FigHostFreq, fig.cap = "The frequency of plots by host identity.
-Gray bars indicate host species were present (i.e., at least one individual of the focal species was present within the plot).
-Black bars indicate  stand structure and composition conditions were conducive to bark beetle infestation in focal tree species.
-For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.", out.width = "244.8pt"}
-knitr::include_graphics( here("Results/Figures/FigureHostCo.jpg"))
-```
-
-## Is occurrence or severity cumulative bark beetle mortality greater in stands with multiple hosts?
+```{r FigSuitableMap, fig.cap="a caption", out.width = "504.8pt"}
+knitr::include_graphics(here("Results/Figures/Figure-Map-Suitableplots.jpg"))
+```
+
+```{r map study area of hosts, eval=F}
+f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
+f93 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
+f108 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)
+
+# reclassify to rasters to presence (1) and absence (0)
+f19[f19>0] <- 1 
+f93[f93>0] <- 1
+f108[f108>0] <- 1
+
+# define areas of multiple hosts
+f108f93 <- f108 + f93 # add 
+f108f93[f108f93<2]<- 0 # reclassify
+f108f93[f108f93==2]<- 1 # reclassify
+f108f19 <- f108 + f19
+f108f19[f108f19<2]<- 0 # reclassify
+f108f19[f108f19==2]<- 1 # reclassify
+f93f19 <- f93 + f19
+f93f19[f93f19<2]<- 0  # reclassify
+f93f19[f93f19==2]<- 1 # reclassify
+
+f108f93f19 <- f108 + f93 + f19
+
+# define areas of one host
+f108f93f19re1 <-f108f93f19
+f108f93f19re1 [f108f93f19re1 ==1 ]<-1
+f108f93f19re1 [f108f93f19re1>1]<-0
+
+# define areas of two hosts
+f108f93f19re2 <-f108f93f19
+f108f93f19re2 [f108f93f19re2 ==1 ]<-0
+f108f93f19re2 [f108f93f19re2==3]<-0 
+f108f93f19re2 [f108f93f19re2 ==2 ]<-1
+
+# limit maps of presence (1) and absence (0) to only areas with one host
+f108 <- f108 * f108f93f19re1 
+f93 <- f93 * f108f93f19re1
+f19 <- f19 * f108f93f19re1 
+f93[f93==1]<- 2 # give unique labels
+f19[f19==1]<- 3 # give unique labels
+
+# create maps of two hosts limited to areas with two hosts
+f108f93  <- f108f93 * f108f93f19re2 
+f108f93[f108f93==1]<-4 # give unique labels
+
+f108f19 <- f108f19 * f108f93f19re2
+f108f19[f108f19==1]<-5 # give unique label
+
+f93f19 <- f93f19 * f108f93f19re2
+f93f19[f93f19==1]<-6 # give unique labels
+
+# create maps of all three hosts
+f108f93f19[f108f93f19<=2]<-0
+f108f93f19[f108f93f19==3]<-7 # give unique labels
+
+fplot <- f108 + f93 + f19 + f108f93 + f108f19 + f93f19+ f108f93f19
+st_crs(fplot) <- "EPSG:5070" # assign projection
+
+fplot[fplot==0] <-NA
+
+states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
+IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+notIMW <- states[!states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
+IMW <- st_transform(IMW, "EPSG:5070")
+notIMW <- st_transform(notIMW, "EPSG:5070")
+
+ext <- as.matrix(extent(IMW))
+
+sa.tmap <- tm_shape(IMW, bbox=tmaptools::bb(ext)) + tm_fill(col="#f5f5f5")+tm_borders()+tm_shape(fplot) +tm_raster(style = "cat", palette = c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), colorNA=NULL, legend.show = TRUE, title="Hosts                    ", labels = c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"))+ tm_shape(notIMW) + tm_fill(col="white")+tm_compass(position=c(0.95, 0.88), just=c("center", "bottom"))+tm_layout(legend.bg.color="white")+tm_graticules(lines=FALSE)+tm_layout(legend.outside=TRUE,legend.text.size=0.55)+tm_scale_bar(position = c(0.01, 0.001), just="left", breaks=c(0,100),text.size=1.1)+tm_shape(IMW) +tm_borders()
+
+Fig.file <- here("Results/Figures/Figure-Studyarea.jpg")
+jpeg(Fig.file, width=fig.width1.5,height=fig.width1.5,units="in", res=300)
+sa.tmap 
+whatever <- dev.off()
+```
+
+```{r FigSA, fig.cap = "The distribution of host species presence across the study area.
+Data are from the Individual Tree Species Atlas (Ellenwood et al.
+2015) and represent conditions in ca.
+2002.", out.width = "322.56pt"}
+knitr::include_graphics( here("Results/Figures/Figure-Studyarea.jpg"))
+```
+
+```{r severity by host presence}
+dfpred <- df
+
+dfpred <- dfpred[dfpred$hosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
+
+dfpred <- dfpred[dfpred$affected01=="affected",]
+
+results <- NULL
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
+
+  res <- data.frame(hosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$hosts <-j
+  res$barkbeetle <- "MPB"
+  if(nrow(dfpred[dfpred$hosts ==j  & dfpred$focalagent =='MPB',])>0){
+      x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$hosts ==j  & dfpred$focalagent %in% c('Any', 'MPB'),], alternative="less")
+      res$X <- x$statistic
+      res$p <- x$p.value
+  }
+  x <-NULL
+  results <- rbind(results, res)
+
+  res <- data.frame(hosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$hosts <-j
+  res$barkbeetle <- "SB"
+  if(nrow(dfpred[dfpred$hosts ==j  & dfpred$focalagent =='SB',])>0){
+     x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$hosts ==j  & dfpred$focalagent %in% c('Any', 'SB'),], alternative="less")
+     res$X <- x$statistic
+     res$p <- x$p.value
+  }
+  x <-NULL
+  results <- rbind(results, res)
+
+  res <- data.frame(hosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$hosts <-j
+  res$barkbeetle <- "WBBB"
+  if(nrow(dfpred[dfpred$hosts ==j  & dfpred$focalagent =='WBBB',])>0){
+    x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$hosts ==j  & dfpred$focalagent %in% c('Any', 'WBBB'),], alternative="less")
+    res$X <- x$statistic
+    res$p <- x$p.value
+  }
+  x <-NULL
+  results <- rbind(results, res)
+
+}
+
+results$pplot <- ifelse(results$p<0.05, "*", NA)
+
+dfpred$pplot <- NA
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
+
+  x <- dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'MPB',]
+  if(nrow(x)>0){
+    xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+    dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'MPB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$hosts==j & results$barkbeetle=='MPB', "pplot"]
+  }
+
+  x <- dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'SB',]
+  if(nrow(x)>0){
+    xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+    dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'SB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$hosts==j & results$barkbeetle=='SB', "pplot"]
+  }
+
+  x <- dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'WBBB',]
+  if(nrow(x)>0){
+    xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+    dfpred[dfpred$hosts ==j  & dfpred$focalagent == 'WBBB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$hosts==j & results$barkbeetle=='WBBB', "pplot"]
+  }
+
+}
+
+dfpred <- dfpred[!(dfpred$hosts == "PICO+PIEN" & dfpred$focalagent=="WBBB"),]
+dfpred <- dfpred[!(dfpred$hosts == "PICO+ABLA" & dfpred$focalagent=="SB"),]
+dfpred <- dfpred[!(dfpred$hosts == "PIEN+ABLA" & dfpred$focalagent=="MPB"),]
+dfpred$hostsP <- factor(dfpred$hosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
+
+dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)
+
+ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~hostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,114)+geom_text(aes(label=pplot), vjust=-0.01,size=9, col="black")+ylab("Severity (% BA affected)")
+
+ggsave(here("Results/Figures/FigureSevertitybyPresence.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
+```
+
+```{r severity by suitable hosts}
+dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
+dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
+
+mpb <- dfpred[dfpred$focalagent=="MPB" & dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA",   "PICO+PIEN+ABLA"), ]
+sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
+wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","suitablehosts"]))
+any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","suitablehosts"]))
+
+dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
+
+dfpred <- dfpred[dfpred$affected01=="affected",]
+
+results <- NULL
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
+
+  x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'MPB'),], alternative="less")
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "MPB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  x <-NULL
+  results <- rbind(results, res)
+
+    x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'SB'),], alternative="less")
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "SB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  x <-NULL
+  results <- rbind(results, res)
+
+      x <- wilcox.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'WBBB'),], alternative="less")
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "WBBB"
+  res$X <- x$statistic
+  res$p <- x$p.value
+  x <-NULL
+  results <- rbind(results, res)
+
+}
+
+results$pplot <- ifelse(results$p<0.05, "*", NA)
+
+dfpred$pplot <- NA
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
+
+  x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB',]
+  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='MPB', "pplot"]
+
+    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB',]
+  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='SB', "pplot"]
+
+    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB',]
+  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
+  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='WBBB', "pplot"]
+
+}
+
+dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+PIEN" & dfpred$focalagent=="WBBB"),]
+dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+ABLA" & dfpred$focalagent=="SB"),]
+dfpred <- dfpred[!(dfpred$suitablehosts == "PIEN+ABLA" & dfpred$focalagent=="MPB"),]
+dfpred$suitablehostsP <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
+
+dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)
+
+ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")
+
+ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
+```
+
+```{r occurrence by host presence}
+dfpred <- df
+mpb <- as.data.frame(table(dfpred[dfpred$focalagent=="MPB","affected"], dfpred[dfpred$focalagent=="MPB","hosts"]))
+sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","hosts"]))
+wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","hosts"]))
+any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","hosts"]))
+
+results <- NULL
+for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
+  anyj <- any[any$Var2==j,]
+  mpbj <- mpb[mpb$Var2==j,]
+  sbj <- sb[sb$Var2==j,]
+  wbbbj <- wbbb[wbbb$Var2==j,]
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "MPB"
+  if(mpbj[1,3]+mpbj[2,3]>0){
+    x <- prop.test(x=c(mpbj[1,3],anyj[1,3]), n=c(mpbj[1,3]+mpbj[2,3], anyj[1,3]+anyj[2,3]))
+    res$X <- x$statistic
+    res$p <- x$p.value
+    res$prop1 <- x$estimate[1]
+    res$prop2 <- x$estimate[2]
+    res$lower <- x$conf.in[1]
+    res$upper <- x$conf.in[2]
+  }
+  x <-NULL
+  results <- rbind(results, res)
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "SB"
+  if(sbj[1,3]+sbj[2,3]>0){
+    x <- prop.test(x=c(sbj[1,3],anyj[1,3]), n=c(sbj[1,3]+sbj[2,3], anyj[1,3]+anyj[2,3]))
+    res$X <- x$statistic
+    res$p <- x$p.value
+    res$prop1 <- x$estimate[1]
+    res$prop2 <- x$estimate[2]
+    res$lower <- x$conf.in[1]
+    res$upper <- x$conf.in[2]
+  }
+  x <-NULL
+  results <- rbind(results, res)
+
+  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA, prop1=NA, prop2 =NA, lower=NA, upper=NA)
+  res$suitablehosts <-j
+  res$barkbeetle <- "WBBB"
+  if(wbbbj[1,3]+wbbbj[2,3]>0){
+    x <- prop.test(x=c(wbbbj[1,3],anyj[1,3]), n=c(wbbbj[1,3]+wbbbj[2,3], anyj[1,3]+anyj[2,3]), alternative="less")
+    res$X <- x$statistic
+    res$p <- x$p.value
+    res$prop1 <- x$estimate[1]
+    res$prop2 <- x$estimate[2]
+    res$lower <- x$conf.in[1]
+    res$upper <- x$conf.in[2]
+    x <-NULL
+  }
+  results <- rbind(results, res)
+}
+
+results$p <- round(results$p,3)
+
+x <- dfpred[is.na(dfpred$suitablehosts)==FALSE,]
+x <- x[x$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"),]
+x[x$suitablehosts == "PICO+PIEN" & x$focalagent=="WBBB","affected"]<- 0
+x[x$suitablehosts == "PICO+ABLA" & x$focalagent=="SB","affected"]<- 0
+x[x$suitablehosts == "PIEN+ABLA" & x$focalagent=="MPB","affected"]<- 0
+
+xx <- x %>% group_by(suitablehosts,focalagent) %>% count(affected)
+xxx <- x %>% group_by(suitablehosts) %>% count()
+
+xxxx <- merge(xx, xxx, by="suitablehosts")
+xxxx$prop <- xxxx$n.x/xxxx$n.y
+xxxx <- xxxx[xxxx$affected=="affected",]
+xxxx <- xxxx[,c(1,2,6)]
+colnames(xxxx) <- c("suitablehosts", "barkbeetle", "prop")
+
+resultsx <- merge(results, xxxx, id.vars=c("suitablehosts", "barkbeetle"), all=T)
+resultsx$barkbeetle <- factor(resultsx$barkbeetle, levels=c("MPB", 'SB', "WBBB", "Any"), ordered=T)
+resultsx <- resultsx[is.na(resultsx$prop)==FALSE,]
+resultsx$suitablehosts <- factor(resultsx$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
+resultsx$pplot <- ifelse(resultsx$p<0.05, "*", "")
+resultsx$pplot[is.na(resultsx$pplot) ==TRUE] <- ""
+
+ggoccurrence <- ggplot(resultsx, aes(x=barkbeetle,y=prop, fill=barkbeetle))+geom_bar(stat="identity")+theme(legend.title = element_blank())+geom_text(aes(label=pplot), vjust=-0.25,size=6)+facet_grid(.~suitablehosts, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylab("proportion of plots")
+
+ggsave( here("Results/Figures/FigureOccurrencebyPresence.jpg"), plot=ggoccurrence,  width=fig.width2, height=fig.width1*0.5,units="in")
+
+```
+
+To determine if the proportion of plots affected by any bark beetle species was significantly greater (p <0.05) than the proportion affected by an individual species, we used a one-sided proportion test.
+Next, to determine if the severity of cumulative bark beetle activity was greater than the severity attributed to an individual species, we subset our data to only the plots affected by at least one bark beetle species.
+We then used Wilcox rank sum test to determine if cumulative mortality was greater that mortality due to an individual bark beetle species.
 
 ```{r occurrence by suitable hosts}
-
 dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
@@
 ```
-
-Given suitable stand conditions for multiple agents, the probability of occurrence of bark beetle attributed mortality was significantly greater than the probability of a single bark beetle species \@ref(fig:FigureOccurrence).
-
-```{r FigOccurrence, fig.cap = "The proportion of plots affected by the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or any bark beetle (i.e., MPB, SB or WBBB) by the identity of hosts species susceptible to bark beetle infestation.
-For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.
-Astericks above bars indiciate that proportion plots affected by the individual bark beelte species was significantly less than the proportion of plots affected by any bark beetle species.",  out.width = "504pt"}
-knitr::include_graphics( here("Results/Figures/FigureOccurrencebySuitablehosts.jpg"))
-```
-
-## Is the severity of cumulative bark beetle-driven tree mortality greater in stands with multiple susceptible hosts?
-
-```{r severity by suitable hosts}
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
-
-mpb <- dfpred[dfpred$focalagent=="MPB" & dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA",   "PICO+PIEN+ABLA"), ]
-sb <- as.data.frame(table(dfpred[dfpred$focalagent=="SB","affected"], dfpred[dfpred$focalagent=="SB","suitablehosts"]))
-wbbb <- as.data.frame(table(dfpred[dfpred$focalagent=="WBBB","affected"], dfpred[dfpred$focalagent=="WBBB","suitablehosts"]))
-any<- as.data.frame(table(dfpred[dfpred$focalagent=="Any","affected"], dfpred[dfpred$focalagent=="Any","suitablehosts"]))
-
-dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
-
-dfpred <- dfpred[dfpred$affected01=="affected",]
-
-results <- NULL
-for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
-
-  x <- kruskal.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'MPB'),])
-
-  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
-  res$suitablehosts <-j
-  res$barkbeetle <- "MPB"
-  res$X <- x$statistic
-  res$p <- x$p.value
-  x <-NULL
-  results <- rbind(results, res)
-
-    x <- kruskal.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'SB'),])
-
-  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
-  res$suitablehosts <-j
-  res$barkbeetle <- "SB"
-  res$X <- x$statistic
-  res$p <- x$p.value
-  x <-NULL
-  results <- rbind(results, res)
-
-      x <- kruskal.test(propTBA~focalagent, data=dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent %in% c('Any', 'WBBB'),])
-
-  res <- data.frame(suitablehosts=NA, barkbeetle=NA, X=NA, p=NA)
-  res$suitablehosts <-j
-  res$barkbeetle <- "WBBB"
-  res$X <- x$statistic
-  res$p <- x$p.value
-  x <-NULL
-  results <- rbind(results, res)
-
-}
-
-results$pplot <- ifelse(results$p<0.05, "*", NA)
-
-dfpred$pplot <- NA
-for(j in c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA")){
-
-  x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB',]
-  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
-  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'MPB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='MPB', "pplot"]
-
-    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB',]
-  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
-  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'SB' &dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='SB', "pplot"]
-
-    x <- dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB',]
-  xplt <- x[x$propTBA==max(x$propTBA), "PLT_CN"][1]
-  dfpred[dfpred$suitablehosts ==j  & dfpred$focalagent == 'WBBB' & dfpred$PLT_CN==xplt, ]$pplot <- results[results$suitablehosts==j & results$barkbeetle=='WBBB', "pplot"]
-
-}
-
-dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+PIEN" & dfpred$focalagent=="WBBB"),]
-dfpred <- dfpred[!(dfpred$suitablehosts == "PICO+ABLA" & dfpred$focalagent=="SB"),]
-dfpred <- dfpred[!(dfpred$suitablehosts == "PIEN+ABLA" & dfpred$focalagent=="MPB"),]
-dfpred$suitablehostsP <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), labels=c("PICO & PIEN",  "PICO & ABLA", "PIEN & ABLA",  "PICO, PIEN, & ABLA"), ordered=T)
-
-dfpred$focalagent2 <- factor(dfpred$focalagent, levels=levels(dfpred$focalagent), labels=c("MPB", "SB", "WBBB", "All"), ordered=T)
-
-ggseverity <- ggplot(dfpred, aes(x=focalagent2, y=propTBA*100, fill=focalagent2, colour=focalagent))+geom_boxplot(notch=F)+facet_grid(.~suitablehostsP, scales="free_x", space="free_x")+xlab("bark beetle species")+scale_fill_manual(values=adjust_transparency(c(p1col[1:3], gray(0.3)), 0.5))+scale_colour_manual(values=c(p1col[1:3], gray(0.3)))+theme(legend.position = "none")+ylim(1,110)+geom_text(aes(label=pplot), vjust=-0.1,size=9, col="black")+ylab("Severity (% BA affected)")
-
-ggsave(here("Results/Figures/FigureSevertitybySuitablehosts.jpg"), plot=ggseverity, width=fig.width2,height=fig.width1*0.66,units="in")
-```
-
-In stands susceptible for all three hosts, cumulative mortality was significantly greater than mortality by a single bark beetle species (Fig.
-\@ref(fig:FigSeverity).
-When stand conditions were suitable for only two bark beetles, the severity of cumulative tree mortality was driven by the most commonly occurring agent (MPB > SB > WBBB) and was not significantly different than the cumulative mortality (Fig.
-\@ref(fig:FigSeverity).
-
-```{r FigSeverity, fig.cap= "Boxplots illustrating the severity of mortality (% of total basal area) attributed to the mountain pine beetle (MPB), spruce beetle (SB), western balsam bark beetle (WBBB), or all bark beetles (i.e., cumulative mortality due to MPB, SB and WBBB) by the identity of hosts species susceptible to bark beetle infestation.
-For host identities, PICO is lodgepole pine, PIEN is Engelmann spruce, and ABLA is subalpine fir.
-The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.
-Astericks above boxes indicate that severity bark beetle mortality attributed to the individual species was significantly less than the cumulative mortality due to all bark beetles.", out.width = "504pt"}
-knitr::include_graphics( here("Results/Figures/FigureSevertitybySuitablehosts.jpg"))
-```
-
-## Is the  co-occurrence of multiple agents common?
-
-Across all 10,439 FIA plots, only 5.3% (n=539) experience mortality due to multiple bark beetle species.
-However, MPB, SB, and WBBB were only present in 23.4% (n=2,441), 9.9% (n=1034), and 7.8% (n=813) of plots,respectively.
-Thus the number of plots with at least two bark beetle species co-occurring is more than 2x greater than would be expected if MPB, SB, WBBB occurred independently.
-
-Given only the 2,163 stands where multiple tree species were susceptible to bark beetle infestation, 21.3% (n=461) were affected by multiple agents.
-When multiple tree species were susceptible to bark beetle infestation, the most commonly occurring agents was MPB (n=341), followed by SB (n=317) and WBBB (n=308).
-The most commonly occurring combinations of agents was MPB and SB (33.2% of cases; n=153), followed by MPB and WBBB (31.2% of cases; n=144) and SB and WBBB (26% of case; n=120).
-The combination of all three agents was extremely rare (9.5% of cases; n=44).
-
-```{r determine co-occurrence of agents by hosts}
-datBB.c <- df[df$focalhost=="Any",]
-table(datBB.c$agents)
-# nrow(datBB.c[datBB.c$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,])/nrow(datBB.c) 
-# 5.3 % of plots were affected by multiple agents
-
-table(datBB.c$agentSB)[2]/nrow(datBB.c) * table(datBB.c$agentMPB)[2]/nrow(datBB.c) * table(datBB.c$agentWBBB)[2]/nrow(datBB.c)
-
-table(datBB.c$agentSB)[2]/nrow(datBB.c) * table(datBB.c$agentMPB)[2]/nrow(datBB.c)
-
-datBB.c[datBB.c$agents %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agents) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-table(datBB.c$agentSB)
-
-dfpred <- merge(BBpred, datBB.c, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[!dfpred$agentsuitable=="none",]
-
-#dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-dfpred$suitablehostsF <-dfpred$suitablehosts
-dfpred$suitablehostsF <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-dfpred$agentsFplongrev <- factor(dfpred$agents,levels=c("MPB+SB+WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB", "WBBB", "SB", "MPB", "none"), labels=c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB", "WBBB", "SB", "MPB", "none"), ordered=T)
-
-#dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())
-
-MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-#sum(MPBx$n)
-#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-#sum(SBx$n)
-#sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-#sum(WBBBx$n)
-#sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-### 
-#table(dfpred$suitablehostsF, dfpred$agents)
-x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-xPICO <-x[x$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA",  "PICO,\nPIEN,\n&\nABLA"),]
-#
-
-## Agents given presence of any agent
-co <- dfpred[!dfpred$focalhost=="Any" ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-#co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]
-
-ggAgentsbySuitablehosts <- ggplot(dfpred[dfpred$focalhost=="Any",], aes(x=suitablehostsF,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+scale_fill_manual(values=p2col)+theme(axis.title.x=element_blank())
-
-ggsave(here("Results/Figures/Figure-AgentsbySuitablehosts.jpg"), plot=ggAgentsbySuitablehosts , width=fig.width1.5,height=fig.width1*0.75,units="in")
-```
-
-```{r determine co-occurrence of agents by suitable hosts}
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[!dfpred$agentsuitable=="none",]
-
-#dfpred[dfpred$focalagent=="Any" &dfpred$suitablehosts %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB")& dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB", "MPB+SB") ,]  %>% group_by( agentsF) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-dfpred$suitablehostsF <-dfpred$suitablehosts
-dfpred$suitablehostsF <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-dfpred$agentsFplongrev <- factor(dfpred$agents,levels=c("MPB+SB+WBBB", "SB+WBBB", "MPB+WBBB", "MPB+SB", "WBBB", "SB", "MPB", "none"), labels=c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB", "WBBB", "SB", "MPB", "none"), ordered=T)
-
-#dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]  %>% summarise(n = n())
-
-MPBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "MPB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-#sum(MPBx$n)
-#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-#sum(MPBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-SBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+SB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-#sum(SBx$n)
-#sum(SBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PICO\n&\nPIEN", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-WBBBx <- dfpred[dfpred$focalagent=="Any" & dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") & dfpred$agentsF %in% c("MPB+SB+WBBB",  "SB+WBBB", "MPB+WBBB") ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-#sum(WBBBx$n)
-#sum(WBBBx$n)/nrow(dfpred[dfpred$suitablehostsF %in% c( "PIEN\n&\nABLA", "PICO\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),])
-
-### 
-#table(dfpred$suitablehostsF, dfpred$agents)
-x <- dfpred[dfpred$focalhost=="Any" & dfpred$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA") &  !dfpred$agentsFplongrev =="none",]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-xPICO <-x[x$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA",  "PICO,\nPIEN,\n&\nABLA"),]
-#
-
-## Agents given presence of any agent
-co <- dfpred[!dfpred$focalhost=="Any" ,]  %>% group_by(suitablehostsF, agentsFplongrev) %>% summarise(n = n())%>% mutate(freq = n / sum(n))
-
-#co[co$agentsFplongrev %in% c("MPB, SB, & WBBB", "SB & WBBB", "MPB & WBBB", "MPB & SB") & co$suitablehostsF %in% c("PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"),]
-
-ggAgentsbySuitablehosts <- ggplot(dfpred[dfpred$focalhost=="Any",], aes(x=suitablehostsF,fill=agentsFplongrev))+geom_bar()+theme(legend.title = element_blank())+scale_fill_manual(values=p2col)+theme(axis.title.x=element_blank())
-
-ggsave(here("Results/Figures/Figure-AgentsbySuitablehosts.jpg"), plot=ggAgentsbySuitablehosts , width=fig.width1.5,height=fig.width1*0.75,units="in")
-```
-
-```{r FigCoOccurrence, fig.cap = "Co-occurrence",  out.width = "504.8pt", eval=F}
-knitr::include_graphics( here("Results/Figures/Figure-AgentsbySuitablehosts.jpg"))
-```
-
-##  Is severity greater in stands with multiple agents?
-```{r multiple agents, eval=F}
-library(rcompanion)
-dfpred <- merge(BBpred, df, id.vars="PLT_CN", all=T)
-dfpred <- dfpred[!dfpred$suitablehosts == "none", ]
-dfpred <- dfpred[dfpred$affected01=="affected",]
-
-dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=levels(dfpred$suitablehosts)[-1], labels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"))
-
-#dfpred <- dfpred[!(dfpred$suitablehosts=="PICO" & dfpred$agentsF %in% c("SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
-
-#dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN" & dfpred$agentsF %in% c("MPB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
-
-#dfpred <- dfpred[!(dfpred$suitablehosts=="ABLA" & dfpred$agentsF %in% c("MPB", "SB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
-
-dfpred <- dfpred[dfpred$suitablehosts %in% c("PICO+PIEN",  "PICO+ABLA", "PIEN+ABLA",  "PICO+PIEN+ABLA"), ]
-
-dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+PIEN" & dfpred$agentsF %in% c("WBBB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB")),]
-
-dfpred <- dfpred[!(dfpred$suitablehosts=="PICO+ABLA" & dfpred$agentsF %in% c("SB", "MPB+SB", "SB+WBBB", "MPB+SB+WBBB")),]
-
-dfpred <- dfpred[!(dfpred$suitablehosts=="PIEN+ABLA" & dfpred$agentsF %in% c("MPB", "MPB+SB", "MPB+WBBB", "MPB+SB+WBBB")),]
-
-dfpred <-dfpred[dfpred$focalagent=="Any",]
-
-results <- NULL
-  dfsub <- dfpred[dfpred$suitablehosts =="PICO+PIEN"  & dfpred$agents %in% c("MPB", "SB","MPB+SB"),]
-  dfsub$agents <- as.factor(as.character(dfsub$agents))
-  x <- kruskal.test(propTBA~agents, data=dfsub)
-  res <- data.frame(agentsF= c("MPB", "SB","MPB+SB"), p=NA)
-  DT <- dunnTest(propTBA~agents, data=dfsub,method="bh")
-  res <- rcompanion::cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.1)
-  res$suitablehosts <- "PICO+PIEN"
-
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
-  res$Letter2 <- c("b", "c", "a")
-  results <- rbind(results, res)
-
-  dfsub <- dfpred[dfpred$suitablehosts == "PICO+ABLA"  & dfpred$agentsF %in% c("MPB", "WBBB","MPB+WBBB"),]
-  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
-  x <- kruskal.test(propTBA~agentsF, data=dfsub)
-  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
-  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-  res$suitablehosts <- "PICO+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
-  res$Letter2 <- c("b", "c", "a")
-  results <- rbind(results, res)
-
-  dfsub <- dfpred[dfpred$suitablehosts == "PIEN+ABLA"  & dfpred$agentsF %in% c("SB", "WBBB","SB+WBBB"),]
-  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
-  x <- kruskal.test(propTBA~agentsF, data=dfsub)
-  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
-  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-  res$suitablehosts <- "PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
-  res$Letter2 <- c("b", "c", "a")
-  results <- rbind(results, res)
-
-  dfsub <- dfpred[dfpred$suitablehosts == "PICO+PIEN+ABLA",]
-  dfsub$agentsF <- as.factor(as.character(dfsub$agentsF))
-  x <- kruskal.test(propTBA~agentsF, data=dfsub)
-  DT <- dunnTest(propTBA~agentsF, data=dfsub,method="bh")
-  res <- cldList(P.unadj ~ Comparison, data = DT$res, threshold = 0.05)
-  res$suitablehosts <- "PICO+PIEN+ABLA"
-  res <- merge(res, dfsub %>% group_by(agents) %>% summarise(median =median(propTBA), propTBA=max(propTBA)), by.x="Group", by.y="agents")
-  res <- res[order(res$median),]
-  res$Letter2 <- c("a", "ab", "abc", "b", "c", "d", "d")
-
-  results <- rbind(results, res)
-  results$agentsF <- results$Group
-
-results$suitablehosts <- factor(results$suitablehosts,
-levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-dfpred$suitablehosts <- factor(dfpred$suitablehosts, levels=c("PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-
-dfpred$agentsF <- factor(dfpred$agentsF,
-levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
-results$agentsF <- factor(results$agentsF,
-levels=c("MPB", "SB", "WBBB", "MPB+SB", "MPB+WBBB", "SB+WBBB", "MPB+SB+WBBB"), labels=c("MPB", "SB", "WBBB", "MPB\n&\nSB", "MPB\n&\nWBBB", "SB\n&\nWBBB", "MPB,\nSB,\n&\nWBBB"), ordered=T)
-
-ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")+theme(axis.text.x=element_blank(), axis.title.x=element_blank(), axis.ticks = element_blank(), plot.margin = unit(c(4.5,4.5,0,4.5), "pt"))
-
-ggmultipelcount<- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF, fill=agentsF))+geom_bar()+facet_grid(.~suitablehosts, scales="free", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+xlab("")+ylab("frequency")+theme(strip.background = element_blank(), strip.text.x = element_blank(), plot.margin = unit(c(3.5,4.5,0,4.5), "pt"))
-
-ggmulti <- ggmultipleseverity + ggmultipelcount +plot_layout(ncol=1, heights=c(1,0.5))
-ggsave(here("Results/Figures/FigureSeverityMultiple.jpg"), plot=ggmulti  , width=fig.width2,height=fig.width1.5,units="in")
-
-ggmultipleseverity <- ggplot(dfpred[dfpred$focalagent=="Any",], aes(x=agentsF,y=propTBA*100, fill=agentsF))+geom_boxplot(notch=F)+facet_grid(.~suitablehosts, scales="free_x", space = "free_x")+theme(legend.position="none")+scale_fill_manual(values=rev(p2col)[-1])+ylab("Severity (% basal area)")+xlab("Bark beetle species present")+ylim(0,110)+geom_text(data=results, aes(label=Letter2), vjust=-1,size=4.5, col="black")
-
-ggsave(here("Results/Figures/FigureHotSpotSeverity.jpg"), plot=ggmultipleseverity, width=fig.width2,height=fig.width1*0.75,units="in")
-
-```
-
-When plots contained two tree species susceptible to bark beetles, the presence of both bark beetle species increased tree mortality relative to presence of only one bark beetle species (Fig.
-\@ref(fig:FigHotSpotSeverity)).
-The increase was greatest for stands that contained both MPB and SB; the median severity of plots with both MPB and SB was 33.7 percentage points greater than MPB alone and 38.3 percentage points greater than SB alone.
-The median severity of plots with both MPB and WBBB was 7.0 percentage points greater than MPB alone and 13.6 percentage points greater than WBBB alone.
-The median severity of plots with both SB and WBBB was 14.7 percentage points greater than SB alone and 19.8 percentage points greater than WBBB alone.
-
-Given stand conditions that made all three tree species susceptible to bark beetles, the highest rates of mortality were in plots where both MPB and SB present; whether or not WBBB was also present did not affect mortality severity (~1 percentage point difference in median severity; Fig.
-\@ref(fig:FigHotSpotSeverity)).
-
-Plots with both SB and WBBB experience similar mortality to plots
-
-The median severity of plots with all three agents was 16.2 percentage points greater than only MPB and WBBB and 27.4% percentage points greater than plots with SB and WBBB.
-
-The median severity of plots with all three agents was most similar to plots affected by both
-
-The median severity of plots with all three agents was 27.0 percentage points greater than MPB alone, 34.2 percentage points greater than SB alone, and 32.0% percentage points greater than WBBB alone.
-
-```{r FigHotSpotSeverity, fig.cap = "The severity of bark beetle mortality in plots with multiple tree species susceptible to bark beetles (columns) by the combinatin of bark beetle species present.
-Letters above boxes indicate significant (p < 0.05) differences between groups as determined by a Dunn test, a nonparametric rank sum test.
-The bottom and top limits of each box are the lower and upper quartiles, respectively; the thick black line within the box is the median; error bars equal ±1.5 times the interquartile range; and points denote outliers, values outside ±1.5 times the interquartile range.",  out.width = "504.8pt"}
-knitr::include_graphics( here("Results/Figures/FigureHotSpotSeverity.jpg"))
-```
-
-# Discussion
-
-# References
-
-<div id="refs"></div>
-
-## Supplement
-### Random Forest Modeling
-```{r plot confusion matrices}
-MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-SMOTE.csv"))
-SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-SMOTE.csv"))
-WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-SMOTE.csv"))
-
-BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
-BBct$approach <- "SMOTE"
-BBct1  <- BBct
-
-MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-BALANCED.csv"))
-SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-BALANCED.csv"))
-WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-BALANCED.csv"))
-
-BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
-BBct$approach <- "Balanced"
-BBct2  <- BBct
-
-MPBct <- read.csv(here("Results/RF-confusiontable-agentMPB-occurrence-WEIGHTED.csv"))
-SBct <- read.csv(here("Results/RF-confusiontable-agentSB-occurrence-WEIGHTED.csv"))
-WBBBct <- read.csv(here("Results/RF-confusiontable-agentWBBB-occurrence-WEIGHTED.csv"))
-
-BBct <- data.frame(agent=c("MPB", "SB", "WBBB"), sensitivity=c(MPBct[2,2]/sum(MPBct[,2]), SBct[2,2]/sum(SBct[,2]), WBBBct[2,2]/sum(WBBBct[,2])), accuracy= c((MPBct[2,2]+MPBct[1,1])/sum(MPBct), (SBct[2,2]+SBct[1,1])/sum(SBct), (WBBBct[2,2]+WBBBct[1,1])/sum(WBBBct)))
-BBct$approach <- "Weighted"
-BBct3  <- BBct
-
-BBct <- rbind(rbind(BBct1, BBct2), BBct3)
-
-ggconfused <- ggplot(BBct, aes(x=agent, y=sensitivity, fill=approach))+geom_bar(stat="identity", position="dodge")+theme(legend.position = "none") + ggplot(BBct, aes(x=agent, y=accuracy, fill=approach))+geom_bar(stat="identity", position="dodge")+plot_layout(nrow=2, guides="collect")
-
-ggsave(here("Results/Figures/Confusion.jpg"), plot=ggconfused, width=fig.width1,height=fig.width1,units="in")
-```
-
-```{r confusionFig}
-knitr::include_graphics(here("Results/Figures/Confusion.jpg"))
-```
-
-```{r FigPartial,  fig.cap = "A caption", out.width = "504.8pt"}
-MPBpp <- read.csv(here("Results/RF-partialdependence-agentMPB-occurrence-SMOTE.csv"))
-MPBpp$focalagent <- "MPB"
-SBpp <- read.csv(here("Results/RF-partialdependence-agentSB-occurrence-SMOTE.csv"))
-SBpp$focalagent <- "SB"
-WBBBpp <- read.csv(here("Results/RF-partialdependence-agentWBBB-occurrence-SMOTE.csv"))
-WBBBpp$focalagent <- "WBBB"
-
-ppdat <- rbind(rbind(MPBpp, SBpp), WBBBpp)
-ppdat$var[ppdat$var %in% paste0("BA.", c("PICO", "PIEN", "ABLA"))] <- "BA"
-ppdat$var[ppdat$var %in% paste0("QMD.", c("PICO", "PIEN", "ABLA"))] <- "QMD"
-ppdat$var[ppdat$var %in% paste0("propBA.", c("PICO", "PIEN", "ABLA"))] <- "propBA"
-
-x <- df[!df$focalagent=="Any",]
-x$propBA <- x$BA.host/x$BA.total
-x <- x[,c("QMD.host", "propBA", "BA.host", "focalagent", "affected01")]
-x$affected01 <- ifelse(x$affected01=="affected", 1, 0)
-
-xm <- melt(x, id.vars=c("focalagent", "affected01"))
-colnames(xm) <- c("focalagent", "affected01", "var", "value")
-xm$var <- factor(xm$var, levels=c("QMD.host", "propBA", "BA.host"), labels=c("QMD", "propBA", "BA"))
-
-pBA1 <- ggplot(ppdat[ppdat$var=="BA",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pBA2 <- ggplot(xm[xm$var=="BA",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot(se=F)+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("bark beetle")+xlab("basal area")
-
-pQMD1 <- ggplot(ppdat[ppdat$var=="QMD",], aes(x=value, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-pQMD2 <- ggplot(xm[xm$var=="QMD",],aes(x=value, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("QMD")
-
-ppropBA1 <- ggplot(ppdat[ppdat$var=="propBA",], aes(x=value*100, y=yhat, col=focalagent))+geom_smooth(se=F)+scale_color_manual(values=p1col)+theme(legend.position="none")+ylab("probability of occurrence")+theme(axis.title.x=element_blank(),axis.text.x=element_blank() )
-
-ppropBA2 <- ggplot(xm[xm$var=="propBA",],aes(x=value*100, y=focalagent, fill=interaction(focalagent, affected01)))+geom_boxplot()+scale_fill_manual(values=c(adjust_transparency(p1col[1:3], alpha=0.25), p1col[1:3]))+theme(legend.position="none")+ylab("")+xlab("% BA")
-
-ppplot <- pBA1  + pQMD1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank()) +ppropBA1 +theme(axis.title.y=element_blank(),axis.text.y=element_blank())+ pBA2 + pQMD2 +theme(axis.text.y=element_blank())  + ppropBA2 +theme(axis.text.y=element_blank()) + plot_layout(ncol=3, heights=c(2,1))
-
-ggsave(here("Results/Figures/PartialDependence.jpg"), plot=ppplot, width=fig.width2,height=fig.width1,units="in")
-```
-
-Consistent with existing stand hazard ratings, the probability of each bark beetle species occurring generally increased with host basal area, quadratic mean diameter, and percent basal area (Fig.
-\@ref(fig:partialFig)).
-However, both MPB and WBBB exhibited a slight decrease in the probability of outbreak at very high basal areas and all three bark beetle species exhibited a lower probability of outbreak in stands where host trees were very large (i.e.
-high QMD values).
-WBBB also exhibited a decline in the probability of outbreak when subalpine fir comprised a relatively high proportion of the basal area.
-
-```{r partialFig, fig.cap="The associations between the probability of occurrence of the mountain pine beetle (MPB; red), spruce beetle (SB; blue); and western balsam bark beetle (WBBB; yellow) and three common measuree of stand suceptibility to bark beetl beetle outbreak, basal area (BA), quadratic mean diameter (QMD), and host dominace (% BA).
-The top row illustrates smoothed partial dependence results from Random Forest modeling, while in the bottom row boxplots illustrate the distribution of values for stands affected (dark colors) and unaffected (light colors) by each bark beetle species."}
-knitr::include_graphics(here("Results/Figures/PartialDependence.jpg"))
-```
-
-### Stand structure and composition in stands 
-```{r FigStandSC,  fig.cap = "A caption", out.width = "322.56pt", eval=T}
-datBB.c <-df[!df$focalhost=="Any",]
-
-QMD.melt <- melt(datBB.c[,c("hosts",paste0("QMD.", c("PIEN", "PICO", "ABLA")))], "hosts")
-QMD.melt$variable <- factor(QMD.melt$variable, levels=paste0("QMD.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-QMD.melt$facet <- "QMD"
-
-TPH.melt <- melt(datBB.c[,c("hosts",paste0("TPH.", c("PIEN", "PICO", "ABLA")))], "hosts")
-TPH.melt$variable <- factor(TPH.melt$variable, levels=paste0("TPH.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-TPH.melt$facet <- "TPH"
-
-BA.melt <- melt(datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA")))], "hosts")
-BA.melt$variable <- factor(BA.melt$variable, levels=paste0("BA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-BA.melt$facet <- "BA"
-
-propBA.melt <-datBB.c[,c("hosts",paste0("BA.", c("PIEN", "PICO", "ABLA", "total")))]
-propBA.melt$propBA.PIEN <- propBA.melt$BA.PIEN / propBA.melt$BA.total *100
-propBA.melt$propBA.ABLA <- propBA.melt$BA.ABLA / propBA.melt$BA.total *100
-propBA.melt$propBA.PICO <- propBA.melt$BA.PICO / propBA.melt$BA.total *100
-
-propBA.melt <-  melt(propBA.melt[,c("hosts",paste0("propBA.", c("PIEN", "PICO", "ABLA")))], "hosts")
-propBA.melt$variable <- factor(propBA.melt$variable, levels=paste0("propBA.", c("PICO","PIEN",  "ABLA")), labels=c("PICO", "PIEN", "ABLA"), ordered=T)
-propBA.melt$facet <- "BA dominance (%)"
-
-plotdat <- rbind(propBA.melt, TPH.melt, BA.melt, QMD.melt)
-
-ggss <- ggplot(plotdat[plotdat$value>0,], aes(x=hosts, y=value,fill=variable))+geom_boxplot(outlier.size = 0.2,color="black",lwd=0.2,position = position_dodge(preserve = "single"))+scale_fill_brewer(palette ="BuPu")+ylab("value")+xlab("")+facet_wrap(.~facet, scales="free_y", ncol=1)+theme(legend.position = "bottom", legend.title=element_blank())
-
-ggsave(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"), plot=ggss , width=fig.width1.5, height=fig.width1*2, units="in")
-knitr::include_graphics(here("Results/Figures/Figure-HostStructureByhostIdentity.jpg"))
-```
-
-### Maps
-```{r maps all plots and hosts, eval=F}
-df.all <- df[df$focalhost == "Any", ]
-
-df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
-
-states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
-IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-
-ext <- as.matrix(extent(IMW))
-#ext[1,1] <- ext[1,1] -0.25
-#ext[1,2] <- ext[1,2] +0.25
-#ext[2,1] <- ext[2,1] -1
-#ext[2,2] <- ext[2,2] +1
-
-map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.9, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
-
-map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
-
-map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
-
-map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
-
-map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
-
-map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
-
-map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
-
-legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
-
-Fig.file <- here("Results/Figures/Figure-Map-Allplots.jpg")
-jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
-tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
-whatever <- dev.off()
-
-```
-
-```{r generate maps, eval=F}
-bbo.dat<- merge(BBpred, df, id.vars="PLT_CN", all=T)
-
-df.all <- bbo.dat[bbo.dat$focalhost == "Any", ]
-
-df.all.sf <- st_as_sf(df.all, coords = c("LON", "LAT"), crs = 4326)
-
-states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
-IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-
-ext <- as.matrix(extent(IMW))
-#ext[1,1] <- ext[1,1] -0.25
-#ext[1,2] <- ext[1,2] +0.25
-#ext[2,1] <- ext[2,1] -1
-#ext[2,2] <- ext[2,2] +1
-
-map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_compass(position=c(0.9, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
-
-map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
-
-map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
-
-map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
-
-map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
-
-map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
-
-map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
-
-legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
-
-Fig.file <- here("Results/Figures/Figure-Map.jpg")
-jpeg(Fig.file, width=fig.width1.5, height=fig.width2/2, units="in", res=300)
-tmap_arrange(map1, map2,map3, map4, map5, map6, map7,legend.map, nrow=2)
-whatever <- dev.off()
-
-```
-
-```{r FigMap, fig.cap="a caption", out.width = "504.8pt"}
-knitr::include_graphics(here("Results/Figures/Figure-Map.jpg"))
-```
-
-```{r map study area of hosts, eval=F}
-f19 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f19/w001001.adf"), proxy=F)
-f93 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f93/w001001.adf"), proxy=F)
-f108 <- stars::read_stars(here("Data/Spatial/ITSPM2002-subalpine/f108/w001001.adf"), proxy=F)
-
-# reclassify to rasters to presence (1) and absence (0)
-f19[f19>0] <- 1 
-f93[f93>0] <- 1
-f108[f108>0] <- 1
-
-# define areas of multiple hosts
-f108f93 <- f108 + f93 # add 
-f108f93[f108f93<2]<- 0 # reclassify
-f108f93[f108f93==2]<- 1 # reclassify
-f108f19 <- f108 + f19
-f108f19[f108f19<2]<- 0 # reclassify
-f108f19[f108f19==2]<- 1 # reclassify
-f93f19 <- f93 + f19
-f93f19[f93f19<2]<- 0  # reclassify
-f93f19[f93f19==2]<- 1 # reclassify
-
-f108f93f19 <- f108 + f93 + f19
-
-# define areas of one host
-f108f93f19re1 <-f108f93f19
-f108f93f19re1 [f108f93f19re1 ==1 ]<-1
-f108f93f19re1 [f108f93f19re1>1]<-0
-
-# define areas of two hosts
-f108f93f19re2 <-f108f93f19
-f108f93f19re2 [f108f93f19re2 ==1 ]<-0
-f108f93f19re2 [f108f93f19re2==3]<-0 
-f108f93f19re2 [f108f93f19re2 ==2 ]<-1
-
-# limit maps of presence (1) and absence (0) to only areas with one host
-f108 <- f108 * f108f93f19re1 
-f93 <- f93 * f108f93f19re1
-f19 <- f19 * f108f93f19re1 
-f93[f93==1]<- 2 # give unique labels
-f19[f19==1]<- 3 # give unique labels
-
-# create maps of two hosts limited to areas with two hosts
-f108f93  <- f108f93 * f108f93f19re2 
-f108f93[f108f93==1]<-4 # give unique labels
-
-f108f19 <- f108f19 * f108f93f19re2
-f108f19[f108f19==1]<-5 # give unique label
-
-f93f19 <- f93f19 * f108f93f19re2
-f93f19[f93f19==1]<-6 # give unique labels
-
-# create maps of all three hosts
-f108f93f19[f108f93f19<=2]<-0
-f108f93f19[f108f93f19==3]<-7 # give unique labels
-
-fplot <- f108 + f93 + f19 + f108f93 + f108f19 + f93f19+ f108f93f19
-st_crs(fplot) <- "EPSG:5070" # assign projection
-
-fplot[fplot==0] <-NA
-
-states <-st_read(here("Data/Spatial/States/cb_2018_us_state_20m.shp"),quiet=T)
-IMW <- states[states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-notIMW <- states[!states$STUSPS %in% c("AZ", "CO", "ID","MT", "NM", "UT", "WY"),]
-IMW <- st_transform(IMW, "EPSG:5070")
-notIMW <- st_transform(notIMW, "EPSG:5070")
-
-ext <- as.matrix(extent(IMW))
-
-sa.tmap <- tm_shape(IMW, bbox=tmaptools::bb(ext)) + tm_fill(col="#f5f5f5")+tm_borders()+tm_shape(fplot) +tm_raster(style = "cat", palette = c("#8c510a", "#bf812d", "#dfc27d", "#80cdc1", "#35978f", "#01665e", "#003c30"), colorNA=NULL, legend.show = TRUE, title="Hosts                    ", labels = c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"))+ tm_shape(notIMW) + tm_fill(col="white")+tm_compass(position=c(0.95, 0.88), just=c("center", "bottom"))+tm_layout(legend.bg.color="white")+tm_graticules(lines=FALSE)+tm_layout(legend.outside=TRUE,legend.text.size=0.55)+tm_scale_bar(position = c(0.01, 0.001), just="left", breaks=c(0,100),text.size=1.1)+tm_shape(IMW) +tm_borders()
-
-Fig.file <- here("Results/Figures/Figure-Studyarea.jpg")
-jpeg(Fig.file, width=fig.width1.5,height=fig.width1.5,units="in", res=300)
-sa.tmap 
-whatever <- dev.off()
-```
-
-```{r FigSA, fig.cap = "The distribution of host species presence across the study area.
-Data are from the Individual Tree Species Atlas (Ellenwood et al.
-2015) and represent conditions in ca.
-2002.", out.width = "322.56pt"}
-knitr::include_graphics( here("Results/Figures/Figure-Studyarea.jpg"))
-```
-
-## Quatile comparison
-```{r select plots with large & abundant hosts, eval=F}
-df.sub <- df
-
-PICO.BA.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$BA.PICO, 0.1)
-PIEN.BA.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$BA.PIEN, 0.1)
-ABLA.BA.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$BA.ABLA, 0.1)
-
-PICO.QMD.threshold <- quantile(df.sub[df.sub$agentMPB=="MPB", ]$QMD.PICO, 0.1)
-PIEN.QMD.threshold <- quantile(df.sub[df.sub$agentSB=="SB", ]$QMD.PIEN, 0.1)
-ABLA.QMD.threshold <- quantile(df.sub[df.sub$agentWBBB=="WBBB", ]$QMD.ABLA, 0.1)
-
-# PICO, PIEN & ABLA
-df.sub.PICO.PIEN.ABLA <- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PICO.PIEN.ABLA <- df.sub.PICO.PIEN.ABLA[df.sub.PICO.PIEN.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PICO.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PICO.PIEN.ABLA$suitablehosts2 <- "PICO, PIEN & ABLA"
-df.sub.PICO.PIEN.ABLA$nhosts2 <- 3
-
-# PICO & PIEN
-df.sub.PICO.PIEN<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.PIEN >= PIEN.BA.threshold, ]
-
-df.sub.PICO.PIEN <- df.sub.PICO.PIEN[df.sub.PICO.PIEN$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.PIEN$QMD.PIEN >= PIEN.QMD.threshold, ]
-
-df.sub.PICO.PIEN$suitablehosts2 <- "PICO & PIEN"
-df.sub.PICO.PIEN <- df.sub.PICO.PIEN[!(df.sub.PICO.PIEN$PLT_CN %in% df.sub.PICO.PIEN$PLT_CN), ]
-df.sub.PICO.PIEN$nhosts2 <- 2
-
-# PICO & ABLA
-df.sub.PICO.ABLA<- df.sub[df.sub$BA.PICO >= PICO.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PICO.ABLA <- df.sub.PICO.ABLA[df.sub.PICO.ABLA$QMD.PICO >= PICO.QMD.threshold & df.sub.PICO.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PICO.ABLA$suitablehosts2 <- "PICO & ABLA"
-df.sub.PICO.ABLA$nhosts2 <- 2
-
-df.sub.PICO.ABLA<- df.sub.PICO.ABLA[!(df.sub.PICO.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
-
-# PIEN & ABLA
-df.sub.PIEN.ABLA<- df.sub[df.sub$BA.PIEN >= PIEN.BA.threshold & df.sub$BA.ABLA >= ABLA.BA.threshold, ]
-
-df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[df.sub.PIEN.ABLA$QMD.PIEN >= PIEN.QMD.threshold & df.sub.PIEN.ABLA$QMD.ABLA >= ABLA.QMD.threshold, ]
-
-df.sub.PIEN.ABLA$suitablehosts2 <- "PIEN & ABLA"
-df.sub.PIEN.ABLA$nhosts2 <- 2
-
-df.sub.PIEN.ABLA <- df.sub.PIEN.ABLA[!(df.sub.PIEN.ABLA$PLT_CN %in% df.sub.PICO.PIEN.ABLA$PLT_CN), ]
-
-###
-df.focal2 <- rbind(df.sub.PICO.ABLA, df.sub.PICO.PIEN)
-df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN)
-df.focal2 <- rbind(df.focal2, df.sub.PICO.PIEN.ABLA)
-
-## Affected hosts
-affectedhosts2 <- df.focal2 %>% group_by(suitablehosts2, afp2colfected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectedhosts2[affectedhosts2$affected=="affected",]
-affectednhosts2 <- df.focal2 %>% group_by(nhosts2, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts2[affectednhosts2$affected=="affected",]
-
-```

### 2021-11-22 Code/Manuscript-many.Rmd
including multiple ways for determining presence

--- 2021-11-22 Code/Manuscript-SMOTE-outbreak.Rmd
@@
 
-df$hostsPICO25 <- ifelse(df$BA.PICO/df$BA.total>0.25, "PICO", "absent")
-df$hostsPIEN25 <- ifelse(df$BA.PIEN/df$BA.total>0.25, "PIEN", "absent")
-df$hostsABLA25 <- ifelse(df$BA.ABLA/df$BA.total>0.25, "ABLA", "absent")
-df$hosts25 <-  do.call(paste, c(df[c("hostsPICO25", "hostsPIEN25", "hostsABLA25")],sep="-"))
-df$hosts25 <- factor(df$hosts25, levels=c('PICO-absent-absent','absent-absent-ABLA', 'absent-PIEN-absent', 'PICO-PIEN-absent',   'PICO-absent-ABLA', 'absent-PIEN-ABLA', 'PICO-PIEN-ABLA', 'absent-absent-absent'), labels= c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), ordered=TRUE)
-
-df$nhosts25 <- df$hosts25
-df$nhosts25 <- factor(df$nhosts25, levels=c('PICO', 'ABLA', 'PIEN', 'PICO+PIEN', 'PICO+ABLA', 'PIEN+ABLA', 'PICO+PIEN+ABLA', 'none'), labels=c(1, 1,1,2,2,2,3,0))
-
-write.csv(df, here("Data/Processed/FIAdat-df.csv"),row.names=F)
+write.csv(df, here("Data/Processed/FIAdat-df-outbreak.csv"),row.names=F)
 ```
@@
 
-write.csv(df, here("Data/Processed/FIAdat-recent-re.csv"),row.names=F)
+write.csv(df, here("Data/Processed/FIAdat-df-merge-re.csv"),row.names=F)
+```
+
+```{r reorganize by focal host - 25%, eval=F}
+df <- read.csv(here("Data/Processed/FIAdat-recent.csv"))
+
+formatp <- function(d, host="PIEN", agent="SB"){
+  df.p <- d[which(d[, paste0("TPH.", host)]>0),]
+  df.p <- df.p[df.p$BA.host/df.p$BA.total>0.25, ]
+  df.p$TPH.host <- df.p[, paste0("TPH.", host)]
+  df.p$QMD.host <- df.p[, paste0("QMD.", host)]
+  df.p$BA.host <- df.p[, paste0("BA.", host)]
+
+  df.p$stand.BB <- df.p[, paste0("stand.", agent)]
+  df.p$BA.BB <- df.p[, paste0("BA.", agent)]
+  df.p$affected <- ifelse(df.p$stand.BB>0, "affected", "unaffected")
+  df.p$focalhost <- host
+
+  df.p$BAD.host <- df.p$BA.host/df.p$BA.total
+  df.p$TPHD.host <- df.p$TPH.host/df.p$TPH.total
+
+  return(df.p)
+}
+
+df <- rbind(rbind(formatp(df), formatp(df, host="PICO", agent="MPB")), formatp(df, host="ABLA", agent="WBBB"), formatp(df, host="Allhosts", agent="allBBs"))
+
+### Proportion of total BA affected
+df$propTBA <- df$BA.BB/df$BA.total
+
+### Proportion of host BA affected
+df$propHBA <- df$BA.BB/df$BA.Allhosts
+
+### Proportion of total trees affected
+df$propTTPH <- (df$stand.BB*unique(6.018046)*2.47105)/df$TPH.total
+
+### Proportion of host trees affected
+df$propHTPH <- (df$stand.BB*unique(6.018046)*2.47105)/df$TPH.host
+
+### Affected
+df$affected <- ifelse(df$stand.BB>0, "affected", "unaffected")
+
+### Affected
+df$affectedo <- ifelse(df$stand.BB/df$BA.Allhosts>0.1, "affected", "unaffected")
+
+write.csv(df, here("Data/Processed/FIAdat-df-merge-re.csv"),row.names=F)
 ```
@@
 ```{r import processed data}
-df <- read.csv( here("Data/Processed/FIAdat-recent-re.csv"))
+df <- read.csv( here("Data/Processed/FIAdat-df-merge-re.csv"))
 df$nhostsF <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
 df$hosts <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
-df$hostsF <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
-
-df$nhosts25F <-factor(df$nhosts, levels=1:3, labels=c("one", "two", "three"), ordered=T)
-df$hosts25 <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), ordered=T)
-df$hosts25F <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
+df$hostsFP <- factor(df$hosts, levels=c("PICO", "PIEN", "ABLA", "PICO+PIEN", "PICO+ABLA", "PIEN+ABLA", "PICO+PIEN+ABLA"), labels=c("PICO", "PIEN", "ABLA", "PICO & PIEN", "PICO & ABLA", "PIEN & ABLA", "PICO, PIEN, & ABLA"), ordered=T)
 
@@
 df.all <- df[df$focalhost == "Any", ]
-df.all$agentsFlong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))
+df.all$hostsFPlong <- df.all$hostsFP
+df.all$agentsFplong <-factor(df.all$agentsF, levels=levels(df.all$agentsF), labels=c("unaffected", "MPB", "SB", "WBBB", "MPB & SB", "MPB & WBBB", "SB & WBBB", "MPB, SB, & WBBB"))
 
@@
 
-map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
-
-map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PIEN",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
-
-map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
-
-map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
-
-map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
-
-map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PIEN+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
-
-map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN+ABLA",]) +tm_dots(col="agentsFlong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
-
-legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hosts=="PICO+PIEN+ABLA",]) +tm_dots(col="agentsFlong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
+map1 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=c(unaffected="#A6A6A6", MPB="#710002")) +tm_compass(position=c(0.1, 0.02), just=c("center", "bottom"), text.size=0.5, size=2)+tm_layout(main.title = "PICO",main.title.size=0.5,main.title.position="center")
+
+map2 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN",main.title.size=0.5,main.title.position="center")
+
+map3 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "ABLA",main.title.size=0.5, main.title.position="center")
+
+map4 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & PIEN",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & PIEN",main.title.size=0.5,main.title.position="center")
+
+map5 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PICO & ABLA", main.title.size=0.5, main.title.position="center")
+
+map6 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PIEN & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col)) +tm_layout(main.title = "PIEN & ABLA", main.title.size=0.5, main.title.position="center")
+
+map7 <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", legend.show = FALSE,size=.0025, palette=rev(p2col))+tm_layout(main.title = "PICO, PIEN, & ABLA", main.title.size=0.5, main.title.position="center")
+
+legend.map <- tm_shape(IMW, bbox=tmaptools::bb(ext)) +  tm_fill(col=grey(level=0.95)) +  tm_borders()  + tm_shape(df.all.sf[df.all.sf$hostsFPlong=="PICO, PIEN, & ABLA",]) +tm_dots(col="agentsFplong", size=4, title="bark beetle species",  palette=rev(p2col))+tm_layout(legend.only = TRUE,legend.text.size = 0.5,legend.title.size = 0.65)+tm_scale_bar(position = c(0.5,0.16), text.size=0.5, just="center", breaks=c(0,250,500))
 
@@
 
-```{r fit MPB RF model, eval=F,fig.show='hide'}
+```{r describe frequency of hosts and agents}
+## Hosts
+hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
+
+nhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
+
+## Affected hosts
+affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
+affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]
+
+affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
+affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]
+
+## Agents given host presence
+agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
+  mutate(freq = n / sum(n))
+PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
+PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
+ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))
+
+## Agents given presence of any agent
+agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
+  mutate(freq = n / sum(n))
+
+agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
+
+```
+
+```{r fit MPB RF model, eval=F}
 response <- "agentMPB"
@@
 # partial dependence
-pdale <- NULL
+pd <- NULL
 for(z in vars.final){
@@
   colnames(pdz) <- c("value","yhat","var")
-  pdz$method <- "PD"
-
-  alez <- ALEPlot(X=datz.train[,vars.final], X.model=ranger_obj, pred.fun=function(X.model, newdata){predict(X.model, newdata)$predictions[,2]}, J=which(vars.final==z), K=50)
-  alez <- data.frame(value=alez$x.values, yhat=alez$f.values, var=z, method="ALE")
-
-  pdale <- rbind(pdale, pdz)
-  pdale <- rbind(pdale, alez)
-}
-write.csv(pdale,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
+  pd <- rbind(pd, pdz)
+}
+write.csv(pd,  here(paste0("Results/RF-partialdependence-", response, "-occurrence-SMOTE.csv")), row.names=F)
 ```
@@
 ## Do stand conditions suitable for multiple outbreaks commonly co-occur?
-
-```{r describe frequency of hosts and agents}
-## Hosts
-hosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-hostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-hostgroups$hosts <- factor(hostgroups$hosts, levels=hostgroups$hosts, labels=c("PICO", "PIEN", "ABLA", "PICO\n&\nPIEN", "PICO\n&\nABLA", "PIEN\n&\nABLA", "PICO,\nPIEN,\n&\nABLA"))
-
-nhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts) %>% summarise(n = n())%>%mutate(freq = n / sum(n))
-
-## Affected hosts
-affectedhosts <- df[!df$focalhost=="Any",] %>% group_by(focalhost, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts <- df[df$focalhost=="Any",] %>% group_by(nhosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectednhosts<- affectednhosts[affectednhosts$affected=="affected", ]
-affectednhosts<- affectednhosts[order(affectednhosts$freq, decreasing=T),]
-
-affectedhostgroups <- df[df$focalhost=="Any",] %>% group_by(hosts, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-affectedhostgroups<- affectedhostgroups[affectedhostgroups$affected=="affected", ]
-affectedhostgroups<- affectedhostgroups[order(affectedhostgroups$freq, decreasing=T),]
-
-## Agents given host presence
-agents <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>%
-  mutate(freq = n / sum(n))
-PICOwMPB <- as.numeric(round(agents[agents$focalagent=="MPB" & agents$affected=="affected", "freq"]*100, 0))
-PIENwSB <- as.numeric(round(agents[agents$focalagent=="SB" & agents$affected=="affected", "freq"]*100, 0))
-ABLAwWBBB<- as.numeric(round(agents[agents$focalagent=="WBBB" & agents$affected=="affected", "freq"]*100, 0))
-
-## Agents given presence of any agent
-agentso <- df[!df$focalhost=="Any" & df$affected=="affected",] %>% group_by(focalagent) %>% summarise(n = n())%>%
-  mutate(freq = n / sum(n))
-
-agentsbyhosts <- df[!df$focalhost=="Any",] %>% group_by(focalagent, affected) %>% summarise(n = n())%>% mutate(freq = prop.table(n))
-
-```
 
